### 2023-10-02 manuscript/paper-latin-FUAs.qmd
manuscript-folder

---
title: "The influence of socioeconomic deprivation on the recovery of urban mobility after COVID-19 in Latin America"
author: "Carmen Cabrera-Arnau, Miguel González-Leonardo, Andrea Nasuto, Ruth Neville, Francisco Rowe"
format:
  pdf:
    template: template.tex
editor: visual
bibliography: refs.bib
editor_options:
  markdown:
    wrap: sentence
---

\section{Introduction}

\subsection{General context}

\subsection{New forms of digital data to track human mobility}

\subsection{Contribution}

\section{Data}

The aim of this study is to uncover the role of socioeconomic deprivation in determining the pace of post-pandemic recovery of urban mobility in Latin America.
To do this, we leverage high-resolution data from different sources.

\subsection{Meta-Facebook data}

The multinational technology conglomerate Meta offers a range of data products aimed for social good through their Data for Good programme, which is open to trusted partners including universities, non-profit organisations, and international institutions.
The data is available at small spatial and temporal scales and has the potential to improve how we respond to real-world crises or unusual events, such as earthquakes, hurricanes, floods or pandemics [@Maas19].
In particular, the Data for Good programme offers location data, gathered from Facebook app users who have the Location Services setting turned on on their smartphone.

We analyse human mobility during the COVID-19 pandemic using the datasets Coronavirus Disease Prevention Map of Facebook Population During Crisis (Tile Level) and Coronavirus Disease Prevention Map of Facebook Movements During Crisis (Tile Level), hereinafter Facebook Population and Facebook Movements datasets.
Specifically, we analyse data from four Latin American countries: Argentina, Chile, Colombia and Mexico. The datasets include data corresponding to a two-year period, starting on the 10th March 2020 and ending in mid-March 2022.
Data are temporally aggregated into three 8-hour daily time windows (00:00-08:00, 08:00-16:00 and 16:00-00:00).The data is spatially aggregated into units called Tiles, according to the Bing Maps Tile System developed by Microsoft [@microsoftBingMaps].
This widely-used system offers a variety of world partitions, where the spatial units are square cells at various levels of resolution.
The Data for Good datasets are typically generated using Bing tile levels 13 through 16, where level 13 results in tiles that are about 4.9 x 4.9 km at the Equator [@Maas19]. The datasets also include data for baseline levels before COVID-19 based on a 45-day period ending on the 10th of March of 2020.
The baseline data is computed using an average for the same time of the day and day of the week in the period preceding the crisis (e.g., average over all data collected on a Monday from 00:00 to 8:00 or average over all data collected on a Wednesday from 16:00 to 00:00).
For more details on how the baseline values are computed, see [@Maas19].

The Facebook Population data provides the number of mobile app users who have the Location Services setting turned on, aggregated by tile.
The location of each user in a given 8-hour time window is determined by the tile where they spent most of the time within that window.
The Facebook Movement data captures the number of mobile app users who have the the Location Services setting turned on moving between a pairs of tile.
The origin and destination of a movement are defined as the locations where a user spent most time between two subsequent time windows (e.g., 00:00-8:00 and 8:00-16:00).
In addition to the count data, both the Facebook Population and the Facebook Movement datasets include the percentage difference between the number of counts during the crisis period and the corresponding baseline level for each entry.

Prior to releasing the datasets, information on personal characteristics of users is removed and several techniques are applied to ensure privacy and anonymity.
Small-count dropping is one of these techniques, whereby a data entry is removed from the data set if population or movement counts are lower than 10 during the crisis period, the baseline period or both.
While this technique makes it harder to identify individual users based on their movement patterns, the removal of data entries containing locations with small counts may lead to an underrepresentation of the population in these places.
Another of these techniques consists in adding a small undisclosed amount of random noise to ensure that it is not possible to ascertain precise, true counts for sparsely populated locations.
In order to produce a smooth population count surface, spatial smoothing using inverse distance-weighted averaging is also applied (see [@Maas19] for details).

\subsection{Geographic delimitations}

Our analysis focuses on intraurban mobility.
The analysis is done at two spatial scales: for functional urban areas (FUAs) and for smaller spatial units, communes.

The rationale for this is: FUAs give us an indication of commuter regions, so they are the natural spatial unit to track intra-urban mobility.
However, there is a lot of socioeconomic inequalities in Latin America, so we want to know what happens at smaller spatial scales, such as the commune level.
For this reason, we also use data from W, X, Y and Z, which give us the geographical delimitations for FUAs in the four countries analysed here respectively, as well as delimitations for the communes forming the different FUAs.

\subsection{COVID-19 stringency data}

To understand the patterns of mobility in the context of the COVID-19 pandemic, we use the stringency index as a measure of the level of nonpharmaceutical interventions to COVID-19, such as social distancing and lockdowns.
The stringency index ranges from 0 to 100, with 100 being the value corresponding to the most strict scenario.
The values for the stringency index were retrieved from the COVID-19 government response tracker (https://www.bsg.ox.ac.uk/research/research-projects/covid-19-government-response-tracker).
For more information, see Hale et al. [@Hale2021].

\subsection{Relative deprivation data}

\subsection{Worldpop population data}

<!-- might not need this one !-->
An additional data set from WorldPop was used to capture the spatial distribution of actual population in the different countries analysed here.
The WorlPop poulation dataset is in raster format and contains gridded population data at 1 sqkm resolution.
We processed the WorlPop population data so it is spatially aggregated according to the Bing Maps Tile System.
The level of detail was chosen so it coincides with that used in the Facebook Population and the Facebook Movements data for each of the countries analysed in this study.

\section{Methods}

\subsection{Determining recovery}

Explain here the baseline level of mobility and how just after the pandemic, there was a generalised drop in mobility across the urban-rural continuum.
As the pandemic evolved and some of the stringency measures were relaxed, the levels of mobility bounced back to levels closer to those observed during the baseline period.
Show Figure illustrating this evolution.

However, for the subsequent parts of the analysis, we need a formal definition of "recovery" since we want to quantitatively assess the association between relative deprivation and the pace of return to normal levels of urban mobility.

For each spatial unit (FUA or commune) the recovery time $t_R^*$ is defined as the time at which we can reject the hypothesis ($p<0.05$) that the level of mobility $m(t)$ is below the baseline level.
The level of mobility $m$ could be the percentage change relative to the baseline of the number of trips per user in a given spatial unit.

To do this, we need to apply the following method:

```{=tex}
\begin{itemize}

\item Model the time evolution of $m$ taking into account exogenous regressors such as the specific month of the year (to account for seasonality?), the average population density and, importantly, the stringency index. The model needs to be predictive and needs to give the probability distribution of the outcome variable.
\item For the model, each sample will be formed by a spatial unit. Each sample has as many time steps as months in the movement dataset. Each sample also has information that should be incorporated as exogenous regressors.
\item Compute $t_R^*$ such that $ t_R = min\{t_R: Pr(m(t_R)<0) < 0.05\}$.
\end{itemize}
```

\subsection{Analysis}

For each spatial unit, we will be left with a value of $t_R*$.
We can also compute the average relative deprivation of each spatial unit, denoted by $d$.
Then, we can check whether there is an association between these two variables.
Do we observe a trend wherer cities that reach recovery first are those where deprivaton is lower?
If so, which are those cities?

\section{Results}

\section{Discussion}

### 2023-10-02 manuscript-old/manuscript-latin-countries.qmd
manuscript-old-folder

---
title: "The mid-term impact of COVID-19 on human mobility patterns in Latin American countries"
author: "Carmen Cabrera-Arnau, Miguel González-Leonardo, Andrea Nasuto, Ruth Neville, Francisco Rowe"
format:
  pdf:
    template: template.tex
editor: visual
bibliography: refs.bib
editor_options:
  markdown:
    wrap: sentence
---

```{=html}
<!--# The abstract has to be set up in the tex file, but I copy it below.

Recent empirical studies, predominantly from countries in the Global North, have shown that the COVID-19 pandemic disrupted human mobility at different spatial scales, but particularly in big cities, initially identified as epicentres of infections. Lockdowns, remote work, and online education decreased the demand for commuting and urban living, resulting in an "urban exodus". Despite the existing evidence, little is known about whether counterurbanisation movements have unfolded similarly in the Global South. In this study, we use anonymised, high-resolution location data from Facebook users to answer the following research questions: i) to what extent are people leaving cities? ii) are there variations across countries? iii) are there changes in mobility patterns over time? iv) how long do these changes last? The data is aggregated into spatial units according to the Bing Maps Tile System and temporally aggregated into 8-hour time windows for a period of approximately 2 years. The methodology involves three steps. Firstly, we classify the Bing tiles into different population density categories according to population data from WorldPop. Secondly, we quantify and analyse the number and characteristics of movements between different locations in each country. Finally, we use statistical modelling and spatial interaction models to assess how the population density category at the origin and destination affects the strength of the flows at different points in time. We observe that the strength of flows between Bing tiles belonging to different population density categories was systematically reduced in the early days of the pandemic and, even though by 2022 it seemed to have increased, it never returned to pre-pandemic levels. In the early days of COVID-19, we also find a systematically increase above pre-pandemic levels of the strength of the flows between Bing tiles belonding to the the same population density category. As of 2022, the frequency of these movements within population density classes have decreased with respect to 2020.

-->
```
\section{Introduction}

<!--# Add comments using this -->

The COVID-19 pandemic globally resulted in disruptive shocks to the human mobility system.
Governments stringency measures and the economic downturn caused by the pandemic constrained international migration [@internat2021; @gonzález-leonardo2022; @gonzález-leonardo2023] and population movements within countries [@nouvellet2021; @gonzález-leonardo2022a; @wang2022; @rowe2023].
During early stages of the pandemic, big cities became the main epicentres of COVID-19 infections and deaths [@Pomeroy2021] due to their higher connectivity via air travel, higher population density and concentration of public-facing jobs [@brandén2020; @bhadra2020].
As a result, governments implemented non-pharmaceutical interventions, such as lockdowns and business closures, leading to increases in unemployment and removing the urban vibrancy [@blustein2020; @BBCnews2023].
In addition, remote work and online teaching reduced the need for commuting, living close to work and education centers, and added pressure for individuals living in small and crowded spaces to move out of cities and look for more affordable and comfortable housing [@nathan2020; @USBureauLabor2021].
These changes are believed to have reduced the attractiveness of cities during the pandemic, while suburbs and rural areas have gained popularity [@florida2021].
As a result, a number of headlines have claimed the emergence of an 'urban exodus' to areas with lower population densities (e.g. [@TheGuardian2020]).

Despite their potential rise in popularity, remote locations offer limited availability and diversity of job opportunities, and tend to lack infrastructure and services, such as schooling and health, to effectively accommodate a large population influx [@Pinilla2021; @oecdreg].
They generally lack high-speed Internet connection which is vital for teleworking [@Chen2004].
Additionally, most jobs cannot be done remotely [@florida2021, @oecdsoc], and some companies have returned to in-person office work or have implemented hybrid forms of work, following the relaxation of stringency measures [@McKinsey21].
Schools and universities have also returned to face-to-face teaching.
Therefore, living close to work or study places remains relevant.
Businesses and leisure activities in cities have gradually returned to normal, reinvigorating the economic functioning and urban life.
Collectively, these arguments suggest that a potential increase in counterurbanisation movements during the pandemic was concentrated during periods of stringency measures and is likely to be short term.
Big cities have successfully weathered previous pandemics [@huremovic2019; @Glaeser20] and are likely to remain attractive places to live.
Thanks to agglomeration economies, cities are normally regarded as centres for economic growth, clusters of talent, consumer bases and spaces for face‐to‐face interaction and diversity [@storper2004; @florida2021].

Recent empirical evidence has shown a decline on human mobility during periods of severe stringency measures, coupled with a rise of movements from cities to less populated areas and a slowdown of urbanisation movements across some countries of the Global North: the United States [@ramani2021], United Kingdom [@rowe2023; @wang2022], Spain [@gonzález-leonardo2022b; @gonzález-leonardo2022a], Germany [\@; @stawarz2022], Sweden [@vogiazides2022], Norway [@tønnessen2021], Australia [@perales2022] and Japan [@fielding2021, @kotsubo2022].
The majority of movements away from cities headed to their suburbs [@vogiazides2022; @stawarz2022] or rural areas in their close proximity [@gonzález-leonardo2022b; @gonzález-leonardo2022a] showing in some instances a "doghnut effect", with a decline in population movements directed to cities and a rise in the inflow of people to less dense surrounding areas [@ramani2021].
Studies demonstrated that professionals [@tønnessen2021] and individuals with a high income [@haslag2021], those who are potentially able to practice remote working, underpinned movements away from large cities.
In addition, research has shown that a large share of counterurbanisation movements is headed to touristic locations with a high concentration of second homes, suggesting that middle and high class individuals may have been key actors in movements from cities to less populated areas during the pandemic [@gonzález-leonardo2022a].
However, most studies suggested that variations in internal population movements during the pandemic are likely to be temporary and have not altered previous macro-structures of human mobility across the rural-urban hierarchy [@rowe2023].

In Latin America and the Caribbean, a reduction in internal population movements was documented during periods of high stringency [@aromí2023], but there is no evidence on how human mobility has changed across the rural-urban continuum.
Anecdotal reports have shown that movements away from cities seem to have accelerated in two countries of the Global South, India [@irudayarajan2020] and South Africa [@ginsburg2022], while urbanisation trends have decelerated.
However, lack of suitable up-to-date data has prevented us from properly assessing the "urban exodus" hypothesis in Global South countries, as well as the extent, spatial patterns, and durability of potential changes on internal population movements.
The evidence for these two countries suggested that increasing counterurbanisation movements during the pandemic in India and South Africa could be underpinned by internal migrants residing in cities who returned to their hometowns due to a rise in unemployment [@irudayarajan2020; @ginsburg2022].
However, recent studies have shown that urban residents from wealthy neighborhoods were more likely to move to rural areas than those from low income neighborhoods during the first wave of the pandemic in Brazil, Colombia, Mexico, Indonesia, Philippines and South Africa [@lucchini2023].

It is known that human mobility within countries tends to vary in systematic ways with development levels [@bell2015].
In Latin America and the Caribbean, 28.6% of the population was living on less than \$6.8 per day in 2020 [@WorldBank2023] and the informality rate was about 53% in 2016 [@OIT2018].
Thus, we could expect different pandemic impacts on the patterns of internal population movements between countries in the Global North and the Global South.
The way in which COVID-19 unfolded in each country, with varying levels of stringency measures, may have also produced different outcomes on human mobility.

Traditionally, censuses have been used to analyse human mobility patterns within Global South countries, which lack register data [@bell2015; @bernard2017].
Censuses, however, are normally updated every 10 years, lacking the temporal granularity to explore population movements over yearly, monthly or short-time intervals.

Digital traces data from mobile phone applications, on the other hand, provide a unique opportunity to capture internal population movements with small spatial and temporal granularity in multiple countries [@green2021; @rowe2021].
Drawing on geographically granular Meta-Facebook data from March 2020 to March 2022, we aim to analyse how the pandemic has changed the patterns of human mobility across the rural-urban hierarchy in four Latin American countries: Argentina, Chile, Colombia and Mexico.
We aim to address the following research questions: 1) To what extent did people leave cities and move to less populated areas in these countries?
2) Which are the spatial patterns of potential changes on internal population movements?
3) How have the patterns of human mobility changed over the course of the pandemic?
4) Are potential changes likely to endure the pandemic?

The rest of the paper is structured as follows: ...

\section{Background}

\subsection{What do we know about internal population movements during the pandemic?}

The COVID-19 pandemic led to an overall decline in human mobility within countries [@nouvellet2021].
Research has shown that internal population movements across the rural-urban continuum decreased in the United States [@ramani2021], several European countries, Japan and Austria during 2020 [@rowe2023].
Declines ranged from 2.5% in Spain to 8.5% in Australia.
The largest drops occurred during periods of national lockdowns and the implementation of other stringency measures, such as mobility restrictions, work and school closing, while levels of internal population movements recovered or exceeded pre-pandemic figures when lockdowns were lifted.
Declines were partially attributed to a reduction in involuntary migration during lockdowns, a loss of labour market dynamism due to the pandemic, fewer people changing jobs and increasing remote work [@perales2022].

COVID-19 has also altered patterns of internal population movements across the rural-urban hierarchy [@rowe2022].
In large metro areas across the United States, population movements, as well as households and businesses, shifted from central cities to suburbs and exurbs during 2020 [@ramani2021].
These authors label this trend as a "doughnut effect", reflecting a decrease in population inflows and activities in city centers and a growth in suburban rings.
They found a sizeable "doghnut effect" in large cities, a smaller effect in medium-sized cities and almost no effect for small cities.
A similar trend was observed in Norway [@tønnessen2021], Germany [@stawarz2022], Sweden [@vogiazides2022] and Japan [@fielding2021; @kotsubo2022], where net migration rates in big cities declined, while population movement to their suburbs increased.
In addition, this trend was accompanied by a deceleration of urbanisation trends and unusual population gains in rural areas due to an acceleration in counterurbanisation movements, mostly in rural locations which are known to be holiday destinations with a high concentration of second homes and are generally located in the close proximity to big cities.
An increase of counterurbanisation was the main pandemic outcome on human mobility in Spain [@gonzález-leonardo2022b; @gonzález-leonardo2022a], the United Kingdom [@wang2022; @rowe2022] and Australia [@perales2022], where there is no evidence of a "doghnut effect", since substantial variations in suburbs were not found across these countries.

However, studies suggest that variations are likely to be temporary [@rowe2023].
In Spain, the urbanisation movement returned to pre-pandemic levels when the lockdown ended in mid-June 2020 [@gonzález-leonardo2022b] and unusually high levels of counterurbanisation persisted over 2021, although the majority of movements continued to occur between and within urban areas [@gonzález-leonardo2022a].
In Australia, the pandemic has caused minor changes in spatial patterns of internal migration, and its effects were minimal by the end of 2020 [@perales2022].
In the United Kingdom, mobility patterns returned to those registered prior to the pandemic after the easing of non-pharmaceutical interventions [@rowe2022; @wang2022].
These findings suggest that COVID-19 generated shock waves leading to temporary changes in the patterns of internal population movement across the rural-urban continuum, but it has not significantly altered the prevalent macro-structures of population movement within countries.

Despite some evidence in the Global North, lees is known about the impact of COVID-19 on human mobility in the Global South countries.
In Latin America and the Caribbean, internal population movements declined by 10% during periods of severe stringent measures, from 16-19% in Bolivia, Ecuador and Argentina to less than 3% in Paraguay and Venezuela [@aromí2023].
However, there is no evidence about pandemic outcomes on human mobility across the rural-urban continuum.
To date, surveys carried out in India [@irudayarajan2020] and South Africa [@ginsburg2022] have shown anecdotal evidence suggesting that urbanisation trends declined during the pandemic, while movements from cities to rural areas increased.
Both studies pointed out that a number of rural residents initiating a migration decreased during 2020 and some labourers in major cities seem to have returned to their hometowns due to job losses, since stringency measures have extensively impacted the functioning of the economy [@ghosh2020].
Many internal migrants residing in Indian cities lost their jobs and were unable to afford basic necessities [@ILO2020].
Collectively, these findings suggest that vulnerable populations seem to have played a role in increasing counterurbanisation movements in Global South countries.
However, recent work has shown that high-wealth groups residing in cities across Mexico, Colombia, Brazil, Indonesia, Philippines and South Africa were on average 159% more likely to have moved to rural area compared to those of low-wealth groups during early stages of the pandemic [@lucchini2023], in line with the evidencee found for Global North countries.

Despite anecdotal evidence suggesting pandemic impacts on the patterns of human mobility across the rural-urban hierarchy in the Global South, lack of suitable data has not allowed us to identify and quantify the magnitude of these impacts.
In this paper, we use Meta-Facebook data to provide evidence on the effect of the pandemic on the patterns of internal population movements across four Latin American countries.

\subsection{Contemporary trends of human mobility across the rural-urban hierarchy in Latin America}

Until 1980, rural to urban migration dominated internal population movements in Latin America with significant levels of population redistribution, especially during the rapid industrialisation process between the 1950s and 1970s when a critical mass of individuals moved from villages and towns to urban areas [@firebaugh1979; @lattes95; @sobrino12].
Urban growth due to internal mobility was primarily concentrated in large urban centres resulting in a significant population imbalance between the chief cities and other settlements [@PintoDaCunha02; @lattes2017].
As a consequence, Latin America shows high rates of urbanisation with more than 81% of the population living in cities, the highest figure after North America, and about 46% of urban residents settled in cities with more than 1 million inhabitants [@UNpopulation19].

Currently, internal population movements amongst cities dominate the migratory system in Latin America, while rural to urban flows are of less importance [@bernard2017; @rodríguez-vignoli2018; @UNpopulation19].
According to the 2010 census round, 80% of internal migrants moved between cities [@rodríguez-vignoli2018].
Contrary to the period preceding the 1980s, medium-sizes cities (those between 500K and 1 million inhabitants) experienced the highest net migration rates, while large cities (\>1 million) presented balanced rates and small cities (\>500K) lost individuals by internal population movements [@rodríguez-vignoli2018].
Changing patterns in human mobility within Latin American countries were driven by a decline in internal migration since the 1980s crisis [@chávezgalindo2016] and deconcentration dynamics in large urban centres, such as Mexico City [@sobrino2006] or Santiago de Chile [@gonzálezollino2006], where long distance inflows have declined over time.
At the same time, there has been an increasing domestic and foreign investment in export-oriented activities or tourism industries in certain middle-sized cities, fostering the geographic dispersal of employment opportunities and increasing the attractiveness of these cities [@Brea03; @Pérez-Campuzano13; @chávezgalindo2016].

Suburbanisation also represents an important type of internal population movement and has increased over time due to the growing expansion of Latin American cities [@graizbord2007; @chávezgalindo2016].
These movements are underpinned by flows of middle- and upper-class families moving away from central cities to surrounding auto-segregated areas of wealthy individuals [@borsdorf2003; @rodríguezvignoli2017].
Low-income populations also settle in specific areas within suburbs where the cost of living is usually cheaper than in cities [@janoschka2002; @rodríguezvignoli2017].
A large proportion of both wealthy individuals and low-income residents commute daily to central cities, mainly for work reasons [@chávezgalindo2016].
Recently, reurbanisation dynamics have also been observed due to gentrification processes in specific central areas [@sobrino12; @chávezgalindo2016].
In this paper, we analyse how the COVID-19 pandemic has changed patterns of internal population movements across the rural-urban hierarchy in Mexico, Colombia, Perú, Chile and Argentina.

\section{Data}

The analysis proposed in this paper is based on several data sets.
Firstly, Facebook Population and Facebook Movements are data sets based on the Facebook users' location history.
The data sets were created by Facebook's owner company Meta and can be accessed through their Data for Good programme (\url{https://dataforgood.facebook.com/}).
Prior to releasing the data sets, Meta applies three techniques to ensure privacy and anonymisation.
First, a small undisclosed amount of random noise is added to ensure that precise location cannot be identified for small population counts in sparsely populated areas.
While removing small counts may lead to an underrepresentation of the population in these places, the geographic distribution of population is still reflected in the data.
Second, spatial smoothing is applied to produce a smooth population count surface using inverse distance-weighted averaging.
Third, any remaining population counts of less than 10 are removed from the final data set (see Maas et al. @Maas19 for details).

Both data sets Facebook Population and Facebook Movements contain data corresponding to a time period comprising approximately two years, starting from March 2020, and to four Latin American countries, Argentina, Chile, Colombia and Mexico.
The data in both data sets is temporally aggregated into three 8-hour windows (00:00--08:00, 08:00--16:00 and 16:00--00:00) for every day in the aforementioned two-year period.
It is spatially aggregated into tiles according to the Bing Maps Tile System.
This geospatial indexing system was developed by Microsoft and it partitions the world into square cells at various levels of resolution.
The Facebook Movements and Facebook Population datasets are aggregated into various levels of resolution depending on the country, and the length of the Bing tile sides vary also by country according to the country's distance to the Equator, even when two countries use tiles at the same level of resolution.

The Facebook Population data set provides information on the number of active Facebook users in each tile.
The data set Facebook Movements captures the total number of Facebook users moving between pairs of origin and destination Bing tiles.
We note here that due to the nature of the Facebook Movement data, we cannot distinguish between different types of movements, for example, daily commutes to work or permanent changes of address.
However, we are still able to detect the evolution of movements between origin-destination pairs of Bing tiles and hence, we are able to capture the impact that COVID-19 has on mobility patterns.

On top of the data for the two-year period, each entry in the Facebook Population and Facebook Movements datasets include data for baseline levels before COVID-19.
The baseline values are computed based on a 45-day period ending on the 10th of March of 2020.
The data sets also include a 'quality' score indicating the number of standard deviations by which the observed data at specific locations and time windows differ from the baseline values, hence highlighting statistically significant changes.

An additional data set from WorldPop was used to capture the spatial distribution of population density in the different countries analysed here.
The WorlPop dataset is in raster format and contains gridded population data at 1 sqkm resolution.

\section{Method}

\subsection{Data pre-processing}

Ultimately, we work with a dataset where each entry represents a population flow between an origin and a destination pair.
For each entry, we consider when the movement took place, the population density at the origin and destination, the number of active Facebook users at the origin and destination and?

\subsubsection{Spatial aggregation}

The working dataset is aggregated at the same level as the Bing tiles used to track the movements in a given country.
To each Bing tile, we aggregate WorldPop population data and use it to estimate the population density of each Bing tile.

\subsubsection{Temporal aggregation}

The raw data is aggregated into 8-hour windows for every day in the two-year period.
However, we are interested in longer-term trends and not in how population flows vary throughout the day.
For this reason, we obtain estimates for the parameters in the models described in subsection \ref{subsec-models} by splitting the total data set by month.
For the visualisations in section \ref{sec-results}, we also present the data aggregated by month.

\subsection{Classifying Bing tiles according to their population density}

Here we aim to understand how the population density at the origin and the destination might influence mobility behaviours.
To help characterise the population density between orign and destination, we classify the Bing tiles into 10 discrete categories of population density according to the Jenks natural breaks classification method, hence obtaining a categorisation of Bing tiles a lot more detailed than the traditional binary rural/urban classification.

![Maps of population density categories by country with histograms, showing that the distribution of Bing tiles in each density category.](figures/Map_density_histogram.png){fig-align="center" width="625" height="640"}

\subsection{Estimating the strength of the flows between tiles of different population density categories}

\ref{subsec-models}

In this section, we propose a statistical model to capture the extent to which the strength of the population flows is systematically impacted by the density classes of the origin and destination tiles, and how this effect varies over time.
Our approach here is very similar to that proposed by \cite{Rowe}.

While we could obtain the population flows between different population classes and their evolution over time, the results would fail to capture specifically the effect of the density classes of the origin and destination tiles.
This is due to the fact that movement patterns are influenced not only by the population density of the starting and ending locations, but also by factors such as the distance, the population size, the time of the week and day at which movements take place, etc.
These additional variables can act as confounding factors, obscuring the actual impact of population density.
In order to disentangle and obtain a more accurate estimation of the importance of population density, we chose to utilise a regression modeling approach.
I AM HERE

We adopted a spatial interaction framework (Rowe, Lovelace, et al., 2022).
We used population flows between tiles by time window and day (as described in Section 3), and modelled them as a function of characteristics of the flow itself (Fij), the origin (Tilei) and destination (Tilej) tiles and the temporal nature of the flow (Tw).
Crucially, we included indicator variables that capture the pair of population classes (Figure 1) of the origin and destination tile.
In mathematical form:

(2) where µijw is the expectation of the flow of people from tile i to tile j in the time window w; ⍺ is an intercept; is a series of indicator variables that reflect the pair of population density classes of a given origin i (I) and destination tile j (J), resulting in 99 pairs (10 classes × 10 classes minus one so it is not collinear with ⍺); dij is the geographic distance between tiles i and j; qijw is a measure of the quality of the flow estimate provided by Meta-Facebook and related to the uncertainty behind the user count of the flow as described in Section 3; Popi,j represents the population at the origin (i) and destination (j) tiles; are parameters to estimate in the model linking their respective covariates to µijw; D is a trend tracking the day to which the flow relates to during the period in analysis; while Wk and H are indicator variables capturing day of the week (i.e., weekday or weekend) and hour window (i.e., 00:00--08:00, 08:00--16:00 and 16:00--00:00). Our focus in Equation (2) is centred on . Controlling for all other variables, these parameters capture the extent to which, the expected flow between a given origin-destination population density class pair of tiles (e.g., a high-density origin to a low-density destination) is systematically higher or lower than if it occurred between a baseline origin-destination population density class pair of tiles (e.g., a low-density origin to a low-density destination). Additionally, we standardised continuous variables (dij, qijw, Popi j), α so that they can be interpreted as the expected flow on the first day (D = 0), during the first temporal window (H = 00:00--08:00), on a weekday (Wk = 0), for the baseline origin-population population density class pair, when all the other variables are at their mean value. In this context, each can also be seen as the 'modulation factor' around that expectation associated with each pair of origin-destination classes. The baseline origin-destination population density class pair is the lowest population density class as origin and destination.

We used a count data regression model.
Specifically, we fitted a generalised linear model (GLM) where the error term is assumed to be distributed following a Poisson distribution, with a flow expectation of (µijw) linked to the flow count (Fijw) through a log link: (3) (4) The Poisson regression model (PRM) assumes that equidispersion, that is, equality of mean and variance in the response variable (Cameron & Trivedi, 2013).
In practice, the equidispersion property is commonly violated because of overdispersion, that is, the variance exceeds the mean.
When this occurs, the PRM may produce biased parameter estimates, causing the standard errors of the estimates to be underestimated, and compromising the statistical inference process (Hilbe, 2011).
To test for overdispersion, we used a regression-based test based on an auxiliary regression of the conditional variance as described in Cameron and Trivedi (2013).

Following Gelman and Hill (2006), we used a quasi-PRM to address overdispersion in our response variable.
This is one of the most common strategies to deal with overdispersion in count data models (Hilbe, 2011).
Intuitively, this model adjusts the standard errors of the estimates to account for the extra dispersion in the data.
To implement this, we estimated Equation (2) by using robust variance estimators.
The number of active Facebook users were used as a weight to account for the variability of the observed count of population movement over time.
This strategy is also used to mitigate for any potential biases regarding the variation in the observed number of active Facebook users changes over time across Britain.

We fitted Equation (2) using iteratively reweighted least squares (IWLS).
We separately estimated models for individual months in our data, resulting in 18 sets of estimates.
Our key aim was to generate estimates for ⍺ and , so that we focused on discussing the evolution of these estimates in a grid of line plots with 10 rows and 10 columns, each of them representing one of our population density classes.
The plot corresponding to the Ith row and Jth column displays the evolution of the parameter that tracks the intensity of population flows from tiles in population density class Ith to those in population density class Jth.

\section{Results}\label{sec-results}

\section{Discussion}

\section{Conclusion}

\section{References}

### 2023-12-27 manuscript-latin-countries/paper-latin-FUAs.tex
start-evo-to-recovery

--- 2023-10-02 manuscript-old/manuscript-latin-countries.qmd
@@
----
-title: "The mid-term impact of COVID-19 on human mobility patterns in Latin American countries"
-author: "Carmen Cabrera-Arnau, Miguel González-Leonardo, Andrea Nasuto, Ruth Neville, Francisco Rowe"
-format:
-  pdf:
-    template: template.tex
-editor: visual
-bibliography: refs.bib
-editor_options:
-  markdown:
-    wrap: sentence
----
+\RequirePackage[l2tabu,orthodox]{nag}
+\documentclass[11pt,letterpaper]{article}
+\usepackage[T1]{fontenc}
+\usepackage[utf8]{inputenc}
+\usepackage{crimson}
+\usepackage{helvet}
+\usepackage[strict,autostyle]{csquotes}
+\usepackage[USenglish]{babel}
+\usepackage{microtype}
+\usepackage{authblk}
+\usepackage{booktabs}
+\usepackage{caption}
+\usepackage{endnotes}
+\usepackage{geometry}
+\usepackage{graphicx}
+\usepackage{hyperref}
+\usepackage{natbib}
+\usepackage{rotating}
+\usepackage{setspace}
+\usepackage{titlesec}
+\usepackage{url}
+\usepackage{hyperref}
+\usepackage{soul}
+\usepackage[dvipsnames]{xcolor}
+\usepackage[many]{tcolorbox}
+\newtcolorbox{mybox}{colback = black!5!gray!50, colframe = black!75!black, segmentation style={solid}} % create text box
+\usepackage{hanging}
 
-```{=html}
-<!--# The abstract has to be set up in the tex file, but I copy it below.
+% location of figure files, via graphicx package
+\graphicspath{{./figures/}}
+
+% configure the page layout, via geometry package
+\geometry{
+	paper=letterpaper,
+	top=2.5cm,
+	bottom=2.5cm,
+	left=2.5cm,
+	right=2.5cm}
+\setstretch{1.02}
+\clubpenalty=10000
+\widowpenalty=10000
+
+% set section/subsection headings as the sans serif font
+\titleformat{\section}{\normalfont\sffamily\large\bfseries}{\thesection.}{0.3em}{}
+\titleformat{\subsection}{\normalfont\sffamily\small\bfseries}{\thesubsection.}{0.3em}{}
+
+% make figure/table captions sans-serif small font
+\captionsetup{font={footnotesize,sf},labelfont=bf,labelsep=period}
+
+% configure pdf metadata and link handling
+\hypersetup{
+	pdfauthor={Carmen Cabrera-Arnau},
+	pdftitle={The mid-term impact of COVID-19 on human mobility patterns in Latin American countries},
+	pdfsubject={Title},
+	pdfkeywords={Keywords},
+	pdffitwindow=true,
+	breaklinks=true,
+	colorlinks=false,
+	pdfborder={0 0 0}}
+
+\title{The mid-term impact of COVID-19 on human mobility patterns in Latin American countries\footnote{\textbf{Citation}: Cabrera-Arnau, C., González-Leonardo, M., Nasuto, A., Neville, R., Rowe, F.
+(2023).
+The mid-term impact of COVID-19 on human mobility patterns in Latin American countries}}
+\author[1]{Carmen Cabrera-Arnau \thanks{\textit{Corresponding author}: c.cabrera-arnau@liverpool.ac.uk}}
+\author[2]{Miguel González-Leonardo}
+\author[1]{Andrea Nasuto}
+\author[1]{Ruth Neville}
+\author[1]{Francisco Rowe}
+\affil[1]{Geographic Data Science Lab, Department of Geography and Planning, University of Liverpool, Liverpool, United Kingdom}
+\affil[2]{Center for Demographic, Urban and Environmental Studies, El Colegio de México (COLMEX), Mexico City, Mexico}
+
+\date{}
+
+% From pandoc:
+% https://github.com/jgm/pandoc-templates/blob/master/default.latex
+\setlength{\emergencystretch}{3em} % prevent overfull lines
+\providecommand{\tightlist}{%
+  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
+
+% https://stackoverflow.com/questions/41052687/rstudio-pdf-knit-fails-with-environment-shaded-undefined-error
+
+\begin{document}
+
+\maketitle
+
+\begin{abstract}
 
@@
 
--->
-```
-\section{Introduction}
+%\vspace{1cm}
+\end{abstract}
 
-<!--# Add comments using this -->
+\pagebreak
 
-The COVID-19 pandemic globally resulted in disruptive shocks to the human mobility system.
-Governments stringency measures and the economic downturn caused by the pandemic constrained international migration [@internat2021; @gonzález-leonardo2022; @gonzález-leonardo2023] and population movements within countries [@nouvellet2021; @gonzález-leonardo2022a; @wang2022; @rowe2023].
-During early stages of the pandemic, big cities became the main epicentres of COVID-19 infections and deaths [@Pomeroy2021] due to their higher connectivity via air travel, higher population density and concentration of public-facing jobs [@brandén2020; @bhadra2020].
-As a result, governments implemented non-pharmaceutical interventions, such as lockdowns and business closures, leading to increases in unemployment and removing the urban vibrancy [@blustein2020; @BBCnews2023].
-In addition, remote work and online teaching reduced the need for commuting, living close to work and education centers, and added pressure for individuals living in small and crowded spaces to move out of cities and look for more affordable and comfortable housing [@nathan2020; @USBureauLabor2021].
-These changes are believed to have reduced the attractiveness of cities during the pandemic, while suburbs and rural areas have gained popularity [@florida2021].
-As a result, a number of headlines have claimed the emergence of an 'urban exodus' to areas with lower population densities (e.g. [@TheGuardian2020]).
+% print the bibliography
+\setlength{\bibsep}{0.00cm plus 0.05cm} % no space between items
+\bibliographystyle{apalike}
+\bibliography{sim_refs}
 
-Despite their potential rise in popularity, remote locations offer limited availability and diversity of job opportunities, and tend to lack infrastructure and services, such as schooling and health, to effectively accommodate a large population influx [@Pinilla2021; @oecdreg].
-They generally lack high-speed Internet connection which is vital for teleworking [@Chen2004].
-Additionally, most jobs cannot be done remotely [@florida2021, @oecdsoc], and some companies have returned to in-person office work or have implemented hybrid forms of work, following the relaxation of stringency measures [@McKinsey21].
-Schools and universities have also returned to face-to-face teaching.
-Therefore, living close to work or study places remains relevant.
-Businesses and leisure activities in cities have gradually returned to normal, reinvigorating the economic functioning and urban life.
-Collectively, these arguments suggest that a potential increase in counterurbanisation movements during the pandemic was concentrated during periods of stringency measures and is likely to be short term.
-Big cities have successfully weathered previous pandemics [@huremovic2019; @Glaeser20] and are likely to remain attractive places to live.
-Thanks to agglomeration economies, cities are normally regarded as centres for economic growth, clusters of talent, consumer bases and spaces for face‐to‐face interaction and diversity [@storper2004; @florida2021].
-
-Recent empirical evidence has shown a decline on human mobility during periods of severe stringency measures, coupled with a rise of movements from cities to less populated areas and a slowdown of urbanisation movements across some countries of the Global North: the United States [@ramani2021], United Kingdom [@rowe2023; @wang2022], Spain [@gonzález-leonardo2022b; @gonzález-leonardo2022a], Germany [\@; @stawarz2022], Sweden [@vogiazides2022], Norway [@tønnessen2021], Australia [@perales2022] and Japan [@fielding2021, @kotsubo2022].
-The majority of movements away from cities headed to their suburbs [@vogiazides2022; @stawarz2022] or rural areas in their close proximity [@gonzález-leonardo2022b; @gonzález-leonardo2022a] showing in some instances a "doghnut effect", with a decline in population movements directed to cities and a rise in the inflow of people to less dense surrounding areas [@ramani2021].
-Studies demonstrated that professionals [@tønnessen2021] and individuals with a high income [@haslag2021], those who are potentially able to practice remote working, underpinned movements away from large cities.
-In addition, research has shown that a large share of counterurbanisation movements is headed to touristic locations with a high concentration of second homes, suggesting that middle and high class individuals may have been key actors in movements from cities to less populated areas during the pandemic [@gonzález-leonardo2022a].
-However, most studies suggested that variations in internal population movements during the pandemic are likely to be temporary and have not altered previous macro-structures of human mobility across the rural-urban hierarchy [@rowe2023].
-
-In Latin America and the Caribbean, a reduction in internal population movements was documented during periods of high stringency [@aromí2023], but there is no evidence on how human mobility has changed across the rural-urban continuum.
-Anecdotal reports have shown that movements away from cities seem to have accelerated in two countries of the Global South, India [@irudayarajan2020] and South Africa [@ginsburg2022], while urbanisation trends have decelerated.
-However, lack of suitable up-to-date data has prevented us from properly assessing the "urban exodus" hypothesis in Global South countries, as well as the extent, spatial patterns, and durability of potential changes on internal population movements.
-The evidence for these two countries suggested that increasing counterurbanisation movements during the pandemic in India and South Africa could be underpinned by internal migrants residing in cities who returned to their hometowns due to a rise in unemployment [@irudayarajan2020; @ginsburg2022].
-However, recent studies have shown that urban residents from wealthy neighborhoods were more likely to move to rural areas than those from low income neighborhoods during the first wave of the pandemic in Brazil, Colombia, Mexico, Indonesia, Philippines and South Africa [@lucchini2023].
-
-It is known that human mobility within countries tends to vary in systematic ways with development levels [@bell2015].
-In Latin America and the Caribbean, 28.6% of the population was living on less than \$6.8 per day in 2020 [@WorldBank2023] and the informality rate was about 53% in 2016 [@OIT2018].
-Thus, we could expect different pandemic impacts on the patterns of internal population movements between countries in the Global North and the Global South.
-The way in which COVID-19 unfolded in each country, with varying levels of stringency measures, may have also produced different outcomes on human mobility.
-
-Traditionally, censuses have been used to analyse human mobility patterns within Global South countries, which lack register data [@bell2015; @bernard2017].
-Censuses, however, are normally updated every 10 years, lacking the temporal granularity to explore population movements over yearly, monthly or short-time intervals.
-
-Digital traces data from mobile phone applications, on the other hand, provide a unique opportunity to capture internal population movements with small spatial and temporal granularity in multiple countries [@green2021; @rowe2021].
-Drawing on geographically granular Meta-Facebook data from March 2020 to March 2022, we aim to analyse how the pandemic has changed the patterns of human mobility across the rural-urban hierarchy in four Latin American countries: Argentina, Chile, Colombia and Mexico.
-We aim to address the following research questions: 1) To what extent did people leave cities and move to less populated areas in these countries?
-2) Which are the spatial patterns of potential changes on internal population movements?
-3) How have the patterns of human mobility changed over the course of the pandemic?
-4) Are potential changes likely to endure the pandemic?
-
-The rest of the paper is structured as follows: ...
-
-\section{Background}
-
-\subsection{What do we know about internal population movements during the pandemic?}
-
-The COVID-19 pandemic led to an overall decline in human mobility within countries [@nouvellet2021].
-Research has shown that internal population movements across the rural-urban continuum decreased in the United States [@ramani2021], several European countries, Japan and Austria during 2020 [@rowe2023].
-Declines ranged from 2.5% in Spain to 8.5% in Australia.
-The largest drops occurred during periods of national lockdowns and the implementation of other stringency measures, such as mobility restrictions, work and school closing, while levels of internal population movements recovered or exceeded pre-pandemic figures when lockdowns were lifted.
-Declines were partially attributed to a reduction in involuntary migration during lockdowns, a loss of labour market dynamism due to the pandemic, fewer people changing jobs and increasing remote work [@perales2022].
-
-COVID-19 has also altered patterns of internal population movements across the rural-urban hierarchy [@rowe2022].
-In large metro areas across the United States, population movements, as well as households and businesses, shifted from central cities to suburbs and exurbs during 2020 [@ramani2021].
-These authors label this trend as a "doughnut effect", reflecting a decrease in population inflows and activities in city centers and a growth in suburban rings.
-They found a sizeable "doghnut effect" in large cities, a smaller effect in medium-sized cities and almost no effect for small cities.
-A similar trend was observed in Norway [@tønnessen2021], Germany [@stawarz2022], Sweden [@vogiazides2022] and Japan [@fielding2021; @kotsubo2022], where net migration rates in big cities declined, while population movement to their suburbs increased.
-In addition, this trend was accompanied by a deceleration of urbanisation trends and unusual population gains in rural areas due to an acceleration in counterurbanisation movements, mostly in rural locations which are known to be holiday destinations with a high concentration of second homes and are generally located in the close proximity to big cities.
-An increase of counterurbanisation was the main pandemic outcome on human mobility in Spain [@gonzález-leonardo2022b; @gonzález-leonardo2022a], the United Kingdom [@wang2022; @rowe2022] and Australia [@perales2022], where there is no evidence of a "doghnut effect", since substantial variations in suburbs were not found across these countries.
-
-However, studies suggest that variations are likely to be temporary [@rowe2023].
-In Spain, the urbanisation movement returned to pre-pandemic levels when the lockdown ended in mid-June 2020 [@gonzález-leonardo2022b] and unusually high levels of counterurbanisation persisted over 2021, although the majority of movements continued to occur between and within urban areas [@gonzález-leonardo2022a].
-In Australia, the pandemic has caused minor changes in spatial patterns of internal migration, and its effects were minimal by the end of 2020 [@perales2022].
-In the United Kingdom, mobility patterns returned to those registered prior to the pandemic after the easing of non-pharmaceutical interventions [@rowe2022; @wang2022].
-These findings suggest that COVID-19 generated shock waves leading to temporary changes in the patterns of internal population movement across the rural-urban continuum, but it has not significantly altered the prevalent macro-structures of population movement within countries.
-
-Despite some evidence in the Global North, lees is known about the impact of COVID-19 on human mobility in the Global South countries.
-In Latin America and the Caribbean, internal population movements declined by 10% during periods of severe stringent measures, from 16-19% in Bolivia, Ecuador and Argentina to less than 3% in Paraguay and Venezuela [@aromí2023].
-However, there is no evidence about pandemic outcomes on human mobility across the rural-urban continuum.
-To date, surveys carried out in India [@irudayarajan2020] and South Africa [@ginsburg2022] have shown anecdotal evidence suggesting that urbanisation trends declined during the pandemic, while movements from cities to rural areas increased.
-Both studies pointed out that a number of rural residents initiating a migration decreased during 2020 and some labourers in major cities seem to have returned to their hometowns due to job losses, since stringency measures have extensively impacted the functioning of the economy [@ghosh2020].
-Many internal migrants residing in Indian cities lost their jobs and were unable to afford basic necessities [@ILO2020].
-Collectively, these findings suggest that vulnerable populations seem to have played a role in increasing counterurbanisation movements in Global South countries.
-However, recent work has shown that high-wealth groups residing in cities across Mexico, Colombia, Brazil, Indonesia, Philippines and South Africa were on average 159% more likely to have moved to rural area compared to those of low-wealth groups during early stages of the pandemic [@lucchini2023], in line with the evidencee found for Global North countries.
-
-Despite anecdotal evidence suggesting pandemic impacts on the patterns of human mobility across the rural-urban hierarchy in the Global South, lack of suitable data has not allowed us to identify and quantify the magnitude of these impacts.
-In this paper, we use Meta-Facebook data to provide evidence on the effect of the pandemic on the patterns of internal population movements across four Latin American countries.
-
-\subsection{Contemporary trends of human mobility across the rural-urban hierarchy in Latin America}
-
-Until 1980, rural to urban migration dominated internal population movements in Latin America with significant levels of population redistribution, especially during the rapid industrialisation process between the 1950s and 1970s when a critical mass of individuals moved from villages and towns to urban areas [@firebaugh1979; @lattes95; @sobrino12].
-Urban growth due to internal mobility was primarily concentrated in large urban centres resulting in a significant population imbalance between the chief cities and other settlements [@PintoDaCunha02; @lattes2017].
-As a consequence, Latin America shows high rates of urbanisation with more than 81% of the population living in cities, the highest figure after North America, and about 46% of urban residents settled in cities with more than 1 million inhabitants [@UNpopulation19].
-
-Currently, internal population movements amongst cities dominate the migratory system in Latin America, while rural to urban flows are of less importance [@bernard2017; @rodríguez-vignoli2018; @UNpopulation19].
-According to the 2010 census round, 80% of internal migrants moved between cities [@rodríguez-vignoli2018].
-Contrary to the period preceding the 1980s, medium-sizes cities (those between 500K and 1 million inhabitants) experienced the highest net migration rates, while large cities (\>1 million) presented balanced rates and small cities (\>500K) lost individuals by internal population movements [@rodríguez-vignoli2018].
-Changing patterns in human mobility within Latin American countries were driven by a decline in internal migration since the 1980s crisis [@chávezgalindo2016] and deconcentration dynamics in large urban centres, such as Mexico City [@sobrino2006] or Santiago de Chile [@gonzálezollino2006], where long distance inflows have declined over time.
-At the same time, there has been an increasing domestic and foreign investment in export-oriented activities or tourism industries in certain middle-sized cities, fostering the geographic dispersal of employment opportunities and increasing the attractiveness of these cities [@Brea03; @Pérez-Campuzano13; @chávezgalindo2016].
-
-Suburbanisation also represents an important type of internal population movement and has increased over time due to the growing expansion of Latin American cities [@graizbord2007; @chávezgalindo2016].
-These movements are underpinned by flows of middle- and upper-class families moving away from central cities to surrounding auto-segregated areas of wealthy individuals [@borsdorf2003; @rodríguezvignoli2017].
-Low-income populations also settle in specific areas within suburbs where the cost of living is usually cheaper than in cities [@janoschka2002; @rodríguezvignoli2017].
-A large proportion of both wealthy individuals and low-income residents commute daily to central cities, mainly for work reasons [@chávezgalindo2016].
-Recently, reurbanisation dynamics have also been observed due to gentrification processes in specific central areas [@sobrino12; @chávezgalindo2016].
-In this paper, we analyse how the COVID-19 pandemic has changed patterns of internal population movements across the rural-urban hierarchy in Mexico, Colombia, Perú, Chile and Argentina.
-
-\section{Data}
-
-The analysis proposed in this paper is based on several data sets.
-Firstly, Facebook Population and Facebook Movements are data sets based on the Facebook users' location history.
-The data sets were created by Facebook's owner company Meta and can be accessed through their Data for Good programme (\url{https://dataforgood.facebook.com/}).
-Prior to releasing the data sets, Meta applies three techniques to ensure privacy and anonymisation.
-First, a small undisclosed amount of random noise is added to ensure that precise location cannot be identified for small population counts in sparsely populated areas.
-While removing small counts may lead to an underrepresentation of the population in these places, the geographic distribution of population is still reflected in the data.
-Second, spatial smoothing is applied to produce a smooth population count surface using inverse distance-weighted averaging.
-Third, any remaining population counts of less than 10 are removed from the final data set (see Maas et al. @Maas19 for details).
-
-Both data sets Facebook Population and Facebook Movements contain data corresponding to a time period comprising approximately two years, starting from March 2020, and to four Latin American countries, Argentina, Chile, Colombia and Mexico.
-The data in both data sets is temporally aggregated into three 8-hour windows (00:00--08:00, 08:00--16:00 and 16:00--00:00) for every day in the aforementioned two-year period.
-It is spatially aggregated into tiles according to the Bing Maps Tile System.
-This geospatial indexing system was developed by Microsoft and it partitions the world into square cells at various levels of resolution.
-The Facebook Movements and Facebook Population datasets are aggregated into various levels of resolution depending on the country, and the length of the Bing tile sides vary also by country according to the country's distance to the Equator, even when two countries use tiles at the same level of resolution.
-
-The Facebook Population data set provides information on the number of active Facebook users in each tile.
-The data set Facebook Movements captures the total number of Facebook users moving between pairs of origin and destination Bing tiles.
-We note here that due to the nature of the Facebook Movement data, we cannot distinguish between different types of movements, for example, daily commutes to work or permanent changes of address.
-However, we are still able to detect the evolution of movements between origin-destination pairs of Bing tiles and hence, we are able to capture the impact that COVID-19 has on mobility patterns.
-
-On top of the data for the two-year period, each entry in the Facebook Population and Facebook Movements datasets include data for baseline levels before COVID-19.
-The baseline values are computed based on a 45-day period ending on the 10th of March of 2020.
-The data sets also include a 'quality' score indicating the number of standard deviations by which the observed data at specific locations and time windows differ from the baseline values, hence highlighting statistically significant changes.
-
-An additional data set from WorldPop was used to capture the spatial distribution of population density in the different countries analysed here.
-The WorlPop dataset is in raster format and contains gridded population data at 1 sqkm resolution.
-
-\section{Method}
-
-\subsection{Data pre-processing}
-
-Ultimately, we work with a dataset where each entry represents a population flow between an origin and a destination pair.
-For each entry, we consider when the movement took place, the population density at the origin and destination, the number of active Facebook users at the origin and destination and?
-
-\subsubsection{Spatial aggregation}
-
-The working dataset is aggregated at the same level as the Bing tiles used to track the movements in a given country.
-To each Bing tile, we aggregate WorldPop population data and use it to estimate the population density of each Bing tile.
-
-\subsubsection{Temporal aggregation}
-
-The raw data is aggregated into 8-hour windows for every day in the two-year period.
-However, we are interested in longer-term trends and not in how population flows vary throughout the day.
-For this reason, we obtain estimates for the parameters in the models described in subsection \ref{subsec-models} by splitting the total data set by month.
-For the visualisations in section \ref{sec-results}, we also present the data aggregated by month.
-
-\subsection{Classifying Bing tiles according to their population density}
-
-Here we aim to understand how the population density at the origin and the destination might influence mobility behaviours.
-To help characterise the population density between orign and destination, we classify the Bing tiles into 10 discrete categories of population density according to the Jenks natural breaks classification method, hence obtaining a categorisation of Bing tiles a lot more detailed than the traditional binary rural/urban classification.
-
-![Maps of population density categories by country with histograms, showing that the distribution of Bing tiles in each density category.](figures/Map_density_histogram.png){fig-align="center" width="625" height="640"}
-
-\subsection{Estimating the strength of the flows between tiles of different population density categories}
-
-\ref{subsec-models}
-
-In this section, we propose a statistical model to capture the extent to which the strength of the population flows is systematically impacted by the density classes of the origin and destination tiles, and how this effect varies over time.
-Our approach here is very similar to that proposed by \cite{Rowe}.
-
-While we could obtain the population flows between different population classes and their evolution over time, the results would fail to capture specifically the effect of the density classes of the origin and destination tiles.
-This is due to the fact that movement patterns are influenced not only by the population density of the starting and ending locations, but also by factors such as the distance, the population size, the time of the week and day at which movements take place, etc.
-These additional variables can act as confounding factors, obscuring the actual impact of population density.
-In order to disentangle and obtain a more accurate estimation of the importance of population density, we chose to utilise a regression modeling approach.
-I AM HERE
-
-We adopted a spatial interaction framework (Rowe, Lovelace, et al., 2022).
-We used population flows between tiles by time window and day (as described in Section 3), and modelled them as a function of characteristics of the flow itself (Fij), the origin (Tilei) and destination (Tilej) tiles and the temporal nature of the flow (Tw).
-Crucially, we included indicator variables that capture the pair of population classes (Figure 1) of the origin and destination tile.
-In mathematical form:
-
-(2) where µijw is the expectation of the flow of people from tile i to tile j in the time window w; ⍺ is an intercept; is a series of indicator variables that reflect the pair of population density classes of a given origin i (I) and destination tile j (J), resulting in 99 pairs (10 classes × 10 classes minus one so it is not collinear with ⍺); dij is the geographic distance between tiles i and j; qijw is a measure of the quality of the flow estimate provided by Meta-Facebook and related to the uncertainty behind the user count of the flow as described in Section 3; Popi,j represents the population at the origin (i) and destination (j) tiles; are parameters to estimate in the model linking their respective covariates to µijw; D is a trend tracking the day to which the flow relates to during the period in analysis; while Wk and H are indicator variables capturing day of the week (i.e., weekday or weekend) and hour window (i.e., 00:00--08:00, 08:00--16:00 and 16:00--00:00).
-Our focus in Equation (2) is centred on .
-Controlling for all other variables, these parameters capture the extent to which, the expected flow between a given origin-destination population density class pair of tiles (e.g., a high-density origin to a low-density destination) is systematically higher or lower than if it occurred between a baseline origin-destination population density class pair of tiles (e.g., a low-density origin to a low-density destination).
-Additionally, we standardised continuous variables (dij, qijw, Popi j), α so that they can be interpreted as the expected flow on the first day (D = 0), during the first temporal window (H = 00:00--08:00), on a weekday (Wk = 0), for the baseline origin-population population density class pair, when all the other variables are at their mean value.
-In this context, each can also be seen as the 'modulation factor' around that expectation associated with each pair of origin-destination classes.
-The baseline origin-destination population density class pair is the lowest population density class as origin and destination.
-
-We used a count data regression model.
-Specifically, we fitted a generalised linear model (GLM) where the error term is assumed to be distributed following a Poisson distribution, with a flow expectation of (µijw) linked to the flow count (Fijw) through a log link: (3) (4) The Poisson regression model (PRM) assumes that equidispersion, that is, equality of mean and variance in the response variable (Cameron & Trivedi, 2013).
-In practice, the equidispersion property is commonly violated because of overdispersion, that is, the variance exceeds the mean.
-When this occurs, the PRM may produce biased parameter estimates, causing the standard errors of the estimates to be underestimated, and compromising the statistical inference process (Hilbe, 2011).
-To test for overdispersion, we used a regression-based test based on an auxiliary regression of the conditional variance as described in Cameron and Trivedi (2013).
-
-Following Gelman and Hill (2006), we used a quasi-PRM to address overdispersion in our response variable.
-This is one of the most common strategies to deal with overdispersion in count data models (Hilbe, 2011).
-Intuitively, this model adjusts the standard errors of the estimates to account for the extra dispersion in the data.
-To implement this, we estimated Equation (2) by using robust variance estimators.
-The number of active Facebook users were used as a weight to account for the variability of the observed count of population movement over time.
-This strategy is also used to mitigate for any potential biases regarding the variation in the observed number of active Facebook users changes over time across Britain.
-
-We fitted Equation (2) using iteratively reweighted least squares (IWLS).
-We separately estimated models for individual months in our data, resulting in 18 sets of estimates.
-Our key aim was to generate estimates for ⍺ and , so that we focused on discussing the evolution of these estimates in a grid of line plots with 10 rows and 10 columns, each of them representing one of our population density classes.
-The plot corresponding to the Ith row and Jth column displays the evolution of the parameter that tracks the intensity of population flows from tiles in population density class Ith to those in population density class Jth.
-
-\section{Results}\label{sec-results}
-
-\section{Discussion}
-
-\section{Conclusion}
-
-\section{References}
+\end{document}

### 2023-12-27 manuscript-latin-countries/template.tex
start-evo-to-recovery

--- 2023-12-27 manuscript-latin-countries/paper-latin-FUAs.tex
@@
 \usepackage{url}
-\usepackage{hyperref}
 \usepackage{soul}
@@
 % https://github.com/jgm/pandoc-templates/blob/master/default.latex
+$if(pagestyle)$
+\pagestyle{$pagestyle$}
+$endif$
+$if(csl-refs)$
+\newlength{\cslhangindent}
+\setlength{\cslhangindent}{1.5em}
+\newlength{\csllabelwidth}
+\setlength{\csllabelwidth}{3em}
+\newlength{\cslentryspacingunit} % times entry-spacing
+\setlength{\cslentryspacingunit}{\parskip}
+\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
+ {% don't indent paragraphs
+  \setlength{\parindent}{0pt}
+  % turn on hanging indent if param 1 is 1
+  \ifodd #1
+  \let\oldpar\par
+  \def\par{\hangindent=\cslhangindent\oldpar}
+  \fi
+  % set entry spacing
+  \setlength{\parskip}{#2\cslentryspacingunit}
+ }%
+ {}
+\usepackage{calc}
+\newcommand{\CSLBlock}[1]{#1\hfill\break}
+\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
+\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
+\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
+$endif$
 \setlength{\emergencystretch}{3em} % prevent overfull lines
@@
 % https://stackoverflow.com/questions/41052687/rstudio-pdf-knit-fails-with-environment-shaded-undefined-error
+$if(highlighting-macros)$
+$highlighting-macros$
+$endif$
 
@@
 iv) how long do these changes last?
-The data is aggregated into spatial units according to the Bing Maps Tile System and temporally aggregated into 8-hour time windows for a period of approximately 2 years.
 The methodology involves three steps.
-Firstly, we classify the Bing tiles into different population density categories according to population data from WorldPop.
-Secondly, we quantify and analyse the number and characteristics of movements between different locations in each country.
-Finally, we use statistical modelling and spatial interaction models to assess how the population density category at the origin and destination affects the strength of the flows at different points in time.
-We observe that the strength of flows between Bing tiles belonging to different population density categories was systematically reduced in the early days of the pandemic and, even though by 2022 it seemed to have increased, it never returned to pre-pandemic levels.
-In the early days of COVID-19, we also find a systematically increase above pre-pandemic levels of the strength of the flows between Bing tiles belonding to the the same population density category.
-As of 2022, the frequency of these movements within population density classes have decreased with respect to 2020.
+Firstly, we quantify the number of movements between different locations in each country.
+Second, we use statistical modelling and spatial interaction models to assess the strength of the flows between specific origin-destination pairs at different points in time.
+Thirdly, for each country, we use cluster analysis to characterise different mobility behaviours according to the population density at the origin and the destination.
+We obtain population density data from Worldpop at a resolution of 1 sqkm, and we use it as a proxy of the level of urbanisation at these locations.
 
@@
 \pagebreak
+
+$body$
 

### 2023-12-27 manuscript/paper-latin-FUAs.tex
start-evo-to-recovery

\RequirePackage[l2tabu,orthodox]{nag}
\documentclass[11pt,letterpaper]{article}
\usepackage[T1]{fontenc}
\usepackage[utf8]{inputenc}
\usepackage{crimson}
\usepackage{helvet}
\usepackage[strict,autostyle]{csquotes}
\usepackage[USenglish]{babel}
\usepackage{microtype}
\usepackage{authblk}
\usepackage{booktabs}
\usepackage{caption}
\usepackage{endnotes}
\usepackage{geometry}
\usepackage{graphicx}
\usepackage{hyperref}
\usepackage{natbib}
\usepackage{rotating}
\usepackage{setspace}
\usepackage{titlesec}
\usepackage{url}
\usepackage{hyperref}
\usepackage{soul}
\usepackage[dvipsnames]{xcolor}
\usepackage[many]{tcolorbox}
\newtcolorbox{mybox}{colback = black!5!gray!50, colframe = black!75!black, segmentation style={solid}} % create text box
\usepackage{hanging}

% location of figure files, via graphicx package
\graphicspath{{./figures/}}

% configure the page layout, via geometry package
\geometry{
	paper=letterpaper,
	top=2.5cm,
	bottom=2.5cm,
	left=2.5cm,
	right=2.5cm}
\setstretch{1.02}
\clubpenalty=10000
\widowpenalty=10000

% set section/subsection headings as the sans serif font
\titleformat{\section}{\normalfont\sffamily\large\bfseries}{\thesection.}{0.3em}{}
\titleformat{\subsection}{\normalfont\sffamily\small\bfseries}{\thesubsection.}{0.3em}{}

% make figure/table captions sans-serif small font
\captionsetup{font={footnotesize,sf},labelfont=bf,labelsep=period}

% configure pdf metadata and link handling
\hypersetup{
	pdfauthor={Carmen Cabrera-Arnau},
	pdftitle={The influence of socioeconomic deprivation on the recovery of urban mobility after COVID-19 in Latin America},
	pdfsubject={Title},
	pdfkeywords={Keywords},
	pdffitwindow=true,
	breaklinks=true,
	colorlinks=false,
	pdfborder={0 0 0}}

\title{The influence of socioeconomic deprivation on the recovery of urban mobility after COVID-19 in Latin America}
\author[1]{Carmen Cabrera-Arnau \thanks{\textit{Corresponding author}: c.cabrera-arnau@liverpool.ac.uk}}
\author[2]{Miguel González-Leonardo}
\author[1]{Andrea Nasuto}
\author[1]{Ruth Neville}
\author[1]{Francisco Rowe}
\affil[1]{Geographic Data Science Lab, Department of Geography and Planning, University of Liverpool, Liverpool, United Kingdom}
\affil[2]{Center for Demographic, Urban and Environmental Studies, El Colegio de México (COLMEX), Mexico City, Mexico}

\date{}

% From pandoc:
% https://github.com/jgm/pandoc-templates/blob/master/default.latex
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}

% https://stackoverflow.com/questions/41052687/rstudio-pdf-knit-fails-with-environment-shaded-undefined-error

\begin{document}

\maketitle

\begin{abstract}

%\vspace{1cm}
\end{abstract}

\pagebreak

\section{Introduction}

\subsection{General context}

\subsection{New forms of digital data to track human mobility}

\subsection{Contribution}

\section{Data}

The aim of this study is to uncover the role of socioeconomic
deprivation in determining the pace of post-pandemic recovery of urban
mobility in Latin America. To do this, we leverage high-resolution data
on human mobility collected by Facebook and on relative deprivation
collected by .

\subsection{Facebook-Meta data}

\subsection{Relative deprivation data}

\subsection{Other data}

Our analysis focuses on intraurban mobility. The analysis is done at two
spatial scales: for functional urban areas (FUAs) and for smaller
spatial units: communes.

The rationale for this is: FUAs give us an indication of commuter
regions, so they are the natural spatial unit to track intra-urban
mobility. However, there is a lot of socioeconomic inequalities in Latin
America, so we want to know what happens at smaller spatial scales, such
as the commune level. For this reason, we also use data from W, X, Y and
Z, which give us the geographical delimitations for FUAs in the four
countries analysed here respectively, as well as delimitations for the
communes forming the different FUAs.

\section{Methods}

\subsection{Determining recovery}

Explain here the baseline level of mobility and how just after the
pandemic, there was a generalised drop in mobility across the
urban-rural continuum. As the pandemic evolved and some of the
stringency measures were relaxed, the levels of mobility bounced back to
levels closer to those observed during the baseline period. Show Figure
illustrating this evolution.

However, for the subsequent parts of the analysis, we need a formal
definition of ``recovery'' since we want to quantitatively assess the
association between relative deprivation and the pace of return to
normal levels of urban mobility.

For each spatial unit (FUA or commune) the recovery time \(t_R^*\) is
defined as the time at which we can reject the hypothesis (\(p<0.05\))
that the level of mobility \(m(t)\) is below the baseline level. The
level of mobility \(m\) could be the percentage change relative to the
baseline of the number of trips per user in a given spatial unit.

To do this, we need to apply the following method:

\begin{itemize}

\item Model the time evolution of $m$ taking into account exogenous regressors such as the specific month of the year (to account for seasonality?), the average population density and, importantly, the stringency index. The model needs to be predictive and needs to give the probability distribution of the outcome variable.
\item For the model, each sample will be formed by a spatial unit. Each sample has as many time steps as months in the movement dataset. Each sample also has information that should be incorporated as exogenous regressors.
\item Compute $t_R^*$ such that $ t_R = min\{t_R: Pr(m(t_R)<0) < 0.05\}$.
\end{itemize}

\subsection{Analysis}

For each spatial unit, we will be left with a value of \(t_R*\). We can
also compute the average relative deprivation of each spatial unit,
denoted by \(d\). Then, we can check whether there is an association
between these two variables. Do we observe a trend wherer cities that
reach recovery first are those where deprivaton is lower? If so, which
are those cities?

\section{Results}

\section{Discussion}

% print the bibliography
\setlength{\bibsep}{0.00cm plus 0.05cm} % no space between items
\bibliographystyle{apalike}
\bibliography{refs}

\end{document}

### 2024-01-15 manuscript/paper-latin-FUAs.qmd
add journal guidelines

@@
 ---
+
+[Journal guideline](https://www.nature.com/natcities/content)
+
+Article
+An Article is a substantial novel research study of high quality and general interest to urban-focused researchers.
+
+Format:
+
+Title – Up to 90 characters including spaces. Avoid punctuation and active verbs.
+Abstract – Up to 150 words, unreferenced. 
+Main text – Up to 4,000 words, divided as follows: 
+Introduction (without heading)
+Results
+Discussion
+Methods – Up to 3,000 words, following the main text as a separate section. It can include further conceptual framing to justify research design.
+References – As a guideline, we typically recommend up to 50. Please follow Nature's reference style.
+Display items – Up to 6 figures and/or tables (additional items should go in a separate Supplementary Information file).
 

### 2024-01-17 manuscript-latin-countries-old/manuscript-latin-countries.tex
file-restructure

\RequirePackage[l2tabu,orthodox]{nag}
\documentclass[11pt,letterpaper]{article}
\usepackage[T1]{fontenc}
\usepackage[utf8]{inputenc}
\usepackage{crimson}
\usepackage{helvet}
\usepackage[strict,autostyle]{csquotes}
\usepackage[USenglish]{babel}
\usepackage{microtype}
\usepackage{authblk}
\usepackage{booktabs}
\usepackage{caption}
\usepackage{endnotes}
\usepackage{geometry}
\usepackage{graphicx}
\usepackage{hyperref}
\usepackage{natbib}
\usepackage{rotating}
\usepackage{setspace}
\usepackage{titlesec}
\usepackage{url}
\usepackage{hyperref}
\usepackage{soul}
\usepackage[dvipsnames]{xcolor}
\usepackage[many]{tcolorbox}
\newtcolorbox{mybox}{colback = black!5!gray!50, colframe = black!75!black, segmentation style={solid}} % create text box
\usepackage{hanging}

% location of figure files, via graphicx package
\graphicspath{{./figures/}}

% configure the page layout, via geometry package
\geometry{
	paper=letterpaper,
	top=2.5cm,
	bottom=2.5cm,
	left=2.5cm,
	right=2.5cm}
\setstretch{1.02}
\clubpenalty=10000
\widowpenalty=10000

% set section/subsection headings as the sans serif font
\titleformat{\section}{\normalfont\sffamily\large\bfseries}{\thesection.}{0.3em}{}
\titleformat{\subsection}{\normalfont\sffamily\small\bfseries}{\thesubsection.}{0.3em}{}

% make figure/table captions sans-serif small font
\captionsetup{font={footnotesize,sf},labelfont=bf,labelsep=period}

% configure pdf metadata and link handling
\hypersetup{
	pdfauthor={Carmen Cabrera-Arnau},
	pdftitle={The mid-term impact of COVID-19 on human mobility patterns in Latin American countries},
	pdfsubject={Title},
	pdfkeywords={Keywords},
	pdffitwindow=true,
	breaklinks=true,
	colorlinks=false,
	pdfborder={0 0 0}}

\title{The mid-term impact of COVID-19 on human mobility patterns in Latin American countries\footnote{\textbf{Citation}: Cabrera-Arnau, C., González-Leonardo, M., Nasuto, A., Neville, R., Rowe, F. (2023). The mid-term impact of COVID-19 on human mobility patterns in Latin American countries}}
\author[1]{Carmen Cabrera-Arnau \thanks{\textit{Corresponding author}: c.cabrera-arnau@liverpool.ac.uk}}
\author[2]{Miguel González-Leonardo}
\author[1]{Andrea Nasuto}
\author[1]{Ruth Neville}
\author[1]{Francisco Rowe}
\affil[1]{Geographic Data Science Lab, Department of Geography and Planning, University of Liverpool, Liverpool, United Kingdom}
\affil[2]{Center for Demographic, Urban and Environmental Studies, El Colegio de México (COLMEX), Mexico City, Mexico}

\date{}

% From pandoc:
% https://github.com/jgm/pandoc-templates/blob/master/default.latex
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newlength{\cslentryspacingunit} % times entry-spacing
\setlength{\cslentryspacingunit}{\parskip}
\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
 {% don't indent paragraphs
  \setlength{\parindent}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1
  \let\oldpar\par
  \def\par{\hangindent=\cslhangindent\oldpar}
  \fi
  % set entry spacing
  \setlength{\parskip}{#2\cslentryspacingunit}
 }%
 {}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{#1\hfill\break}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}

% https://stackoverflow.com/questions/41052687/rstudio-pdf-knit-fails-with-environment-shaded-undefined-error

\begin{document}

\maketitle

\begin{abstract}

Recent empirical studies, predominantly from countries in the Global North, have shown that the COVID-19 pandemic disrupted human mobility at different spatial scales, but particularly in big cities, initially identified as epicentres of infections. Lockdowns, remote work, and online education decreased the demand for commuting and urban living, resulting in an "urban exodus". Despite the existing evidence, little is known about whether counterurbanisation movements have unfolded similarly in the Global South. In this study, we use anonymised, high-resolution location data from Facebook users to answer the following research questions: i) to what extent are people leaving cities? ii) are there variations across countries? iii) are there changes in mobility patterns over time? iv) how long do these changes last? The data is aggregated into spatial units according to the Bing Maps Tile System and temporally aggregated into 8-hour time windows for a period of approximately 2 years. The methodology involves three steps. Firstly, we classify the Bing tiles into different population density categories according to population data from WorldPop. Secondly, we quantify and analyse the number and characteristics of movements between different locations in each country. Finally, we use statistical modelling and spatial interaction models to assess how the population density category at the origin and destination affects the strength of the flows at different points in time. We observe that the strength of flows between Bing tiles belonging to different population density categories was systematically reduced in the early days of the pandemic and, even though by 2022 it seemed to have increased, it never returned to pre-pandemic levels. In the early days of COVID-19, we also find a systematically increase above pre-pandemic levels of the strength of the flows between Bing tiles belonding to the the same population density category. As of 2022, the frequency of these movements within population density classes have decreased with respect to 2020.

%\vspace{1cm}
\end{abstract}

\pagebreak

\section{Introduction}

The COVID-19 pandemic globally resulted in disruptive shocks to the
human mobility system. Governments stringency measures and the economic
downturn caused by the pandemic constrained international migration
({``International Migration 2020: Highlights''} 2021; González-Leonardo
and Spijker 2022; González-Leonardo et al. 2023) and population
movements within countries (Nouvellet et al. 2021; González-Leonardo,
Rowe, and Fresolone-Caparrós 2022; Wang et al. 2022; Rowe,
González-Leonardo, and Champion 2023). During early stages of the
pandemic, big cities became the main epicentres of COVID-19 infections
and deaths (Pomeroy and Chiney 2021) due to their higher connectivity
via air travel, higher population density and concentration of
public-facing jobs (Brandén et al. 2020; Bhadra, Mukherjee, and Sarkar
2020). As a result, governments implemented non-pharmaceutical
interventions, such as lockdowns and business closures, leading to
increases in unemployment and removing the urban vibrancy (Blustein et
al. 2020; BBC news 2023). In addition, remote work and online teaching
reduced the need for commuting, living close to work and education
centers, and added pressure for individuals living in small and crowded
spaces to move out of cities and look for more affordable and
comfortable housing (Nathan and Overman 2020; US Bureau of Labor
Statistics 2021). These changes are believed to have reduced the
attractiveness of cities during the pandemic, while suburbs and rural
areas have gained popularity (Florida, Rodríguez-Pose, and Storper
2021). As a result, a number of headlines have claimed the emergence of
an `urban exodus' to areas with lower population densities (e.g. (The
Guardian 2020)).

Despite their potential rise in popularity, remote locations offer
limited availability and diversity of job opportunities, and tend to
lack infrastructure and services, such as schooling and health, to
effectively accommodate a large population influx (Pinilla and Sáez
2021; \emph{OECD Regions and Cities at a Glance}, n.d.). They generally
lack high-speed Internet connection which is vital for teleworking (Chen
and Wellman 2004). Additionally, most jobs cannot be done remotely
{``OECD Social, Employment and Migration Working Papers''} (n.d.), and
some companies have returned to in-person office work or have
implemented hybrid forms of work, following the relaxation of stringency
measures (McKinsey \& Company 2021). Schools and universities have also
returned to face-to-face teaching. Therefore, living close to work or
study places remains relevant. Businesses and leisure activities in
cities have gradually returned to normal, reinvigorating the economic
functioning and urban life. Collectively, these arguments suggest that a
potential increase in counterurbanisation movements during the pandemic
was concentrated during periods of stringency measures and is likely to
be short term. Big cities have successfully weathered previous pandemics
(Huremović 2019; Glaeser 2020) and are likely to remain attractive
places to live. Thanks to agglomeration economies, cities are normally
regarded as centres for economic growth, clusters of talent, consumer
bases and spaces for face‐to‐face interaction and diversity (Storper and
Venables 2004; Florida, Rodríguez-Pose, and Storper 2021).

Recent empirical evidence has shown a decline on human mobility during
periods of severe stringency measures, coupled with a rise of movements
from cities to less populated areas and a slowdown of urbanisation
movements across some countries of the Global North: the United States
(Ramani and Bloom 2021), United Kingdom (Rowe, González-Leonardo, and
Champion 2023; Wang et al. 2022), Spain (González-Leonardo et al. 2022;
González-Leonardo, Rowe, and Fresolone-Caparrós 2022), Germany {[}@;
Stawarz et al. (2022){]}, Sweden (Vogiazides and Kawalerowicz 2022),
Norway (Tønnessen 2021), Australia (Perales and Bernard 2022) and Japan
Kotsubo and Nakaya (2022). The majority of movements away from cities
headed to their suburbs (Vogiazides and Kawalerowicz 2022; Stawarz et
al. 2022) or rural areas in their close proximity (González-Leonardo et
al. 2022; González-Leonardo, Rowe, and Fresolone-Caparrós 2022) showing
in some instances a ``doghnut effect'', with a decline in population
movements directed to cities and a rise in the inflow of people to less
dense surrounding areas (Ramani and Bloom 2021). Studies demonstrated
that professionals (Tønnessen 2021) and individuals with a high income
(Haslag and Weagley 2021), those who are potentially able to practice
remote working, underpinned movements away from large cities. In
addition, research has shown that a large share of counterurbanisation
movements is headed to touristic locations with a high concentration of
second homes, suggesting that middle and high class individuals may have
been key actors in movements from cities to less populated areas during
the pandemic (González-Leonardo, Rowe, and Fresolone-Caparrós 2022).
However, most studies suggested that variations in internal population
movements during the pandemic are likely to be temporary and have not
altered previous macro-structures of human mobility across the
rural-urban hierarchy (Rowe, González-Leonardo, and Champion 2023).

In Latin America and the Caribbean, a reduction in internal population
movements was documented during periods of high stringency (Aromí et al.
2023), but there is no evidence on how human mobility has changed across
the rural-urban continuum. Anecdotal reports have shown that movements
away from cities seem to have accelerated in two countries of the Global
South, India (Irudaya Rajan, Sivakumar, and Srinivasan 2020) and South
Africa (Ginsburg et al. 2022), while urbanisation trends have
decelerated. However, lack of suitable up-to-date data has prevented us
from properly assessing the ``urban exodus'' hypothesis in Global South
countries, as well as the extent, spatial patterns, and durability of
potential changes on internal population movements. The evidence for
these two countries suggested that increasing counterurbanisation
movements during the pandemic in India and South Africa could be
underpinned by internal migrants residing in cities who returned to
their hometowns due to a rise in unemployment (Irudaya Rajan, Sivakumar,
and Srinivasan 2020; Ginsburg et al. 2022). However, recent studies have
shown that urban residents from wealthy neighborhoods were more likely
to move to rural areas than those from low income neighborhoods during
the first wave of the pandemic in Brazil, Colombia, Mexico, Indonesia,
Philippines and South Africa (Lucchini et al. 2023).

It is known that human mobility within countries tends to vary in
systematic ways with development levels (Bell et al. 2015). In Latin
America and the Caribbean, 28.6\% of the population was living on less
than \$6.8 per day in 2020 (World Bank" 2023) and the informality rate
was about 53\% in 2016 (Salazar-Xirinachs 2018). Thus, we could expect
different pandemic impacts on the patterns of internal population
movements between countries in the Global North and the Global South.
The way in which COVID-19 unfolded in each country, with varying levels
of stringency measures, may have also produced different outcomes on
human mobility.

Traditionally, censuses have been used to analyse human mobility
patterns within Global South countries, which lack register data (Bell
et al. 2015; Bernard et al. 2017). Censuses, however, are normally
updated every 10 years, lacking the temporal granularity to explore
population movements over yearly, monthly or short-time intervals.

Digital traces data from mobile phone applications, on the other hand,
provide a unique opportunity to capture internal population movements
with small spatial and temporal granularity in multiple countries
(Green, Pollock, and Rowe 2021; Rowe 2021). Drawing on geographically
granular Meta-Facebook data from March 2020 to March 2022, we aim to
analyse how the pandemic has changed the patterns of human mobility
across the rural-urban hierarchy in four Latin American countries:
Argentina, Chile, Colombia and Mexico. We aim to address the following
research questions: 1) To what extent did people leave cities and move
to less populated areas in these countries? 2) Which are the spatial
patterns of potential changes on internal population movements? 3) How
have the patterns of human mobility changed over the course of the
pandemic? 4) Are potential changes likely to endure the pandemic?

The rest of the paper is structured as follows: \ldots{}

\section{Background}

\subsection{What do we know about internal population movements during the pandemic?}

The COVID-19 pandemic led to an overall decline in human mobility within
countries (Nouvellet et al. 2021). Research has shown that internal
population movements across the rural-urban continuum decreased in the
United States (Ramani and Bloom 2021), several European countries, Japan
and Austria during 2020 (Rowe, González-Leonardo, and Champion 2023).
Declines ranged from 2.5\% in Spain to 8.5\% in Australia. The largest
drops occurred during periods of national lockdowns and the
implementation of other stringency measures, such as mobility
restrictions, work and school closing, while levels of internal
population movements recovered or exceeded pre-pandemic figures when
lockdowns were lifted. Declines were partially attributed to a reduction
in involuntary migration during lockdowns, a loss of labour market
dynamism due to the pandemic, fewer people changing jobs and increasing
remote work (Perales and Bernard 2022).

COVID-19 has also altered patterns of internal population movements
across the rural-urban hierarchy (Rowe et al. 2022). In large metro
areas across the United States, population movements, as well as
households and businesses, shifted from central cities to suburbs and
exurbs during 2020 (Ramani and Bloom 2021). These authors label this
trend as a ``doughnut effect'', reflecting a decrease in population
inflows and activities in city centers and a growth in suburban rings.
They found a sizeable ``doghnut effect'' in large cities, a smaller
effect in medium-sized cities and almost no effect for small cities. A
similar trend was observed in Norway (Tønnessen 2021), Germany (Stawarz
et al. 2022), Sweden (Vogiazides and Kawalerowicz 2022) and Japan
(Fielding and Ishikawa 2021; Kotsubo and Nakaya 2022), where net
migration rates in big cities declined, while population movement to
their suburbs increased. In addition, this trend was accompanied by a
deceleration of urbanisation trends and unusual population gains in
rural areas due to an acceleration in counterurbanisation movements,
mostly in rural locations which are known to be holiday destinations
with a high concentration of second homes and are generally located in
the close proximity to big cities. An increase of counterurbanisation
was the main pandemic outcome on human mobility in Spain
(González-Leonardo et al. 2022; González-Leonardo, Rowe, and
Fresolone-Caparrós 2022), the United Kingdom (Wang et al. 2022; Rowe et
al. 2022) and Australia (Perales and Bernard 2022), where there is no
evidence of a ``doghnut effect'', since substantial variations in
suburbs were not found across these countries.

However, studies suggest that variations are likely to be temporary
(Rowe, González-Leonardo, and Champion 2023). In Spain, the urbanisation
movement returned to pre-pandemic levels when the lockdown ended in
mid-June 2020 (González-Leonardo et al. 2022) and unusually high levels
of counterurbanisation persisted over 2021, although the majority of
movements continued to occur between and within urban areas
(González-Leonardo, Rowe, and Fresolone-Caparrós 2022). In Australia,
the pandemic has caused minor changes in spatial patterns of internal
migration, and its effects were minimal by the end of 2020 (Perales and
Bernard 2022). In the United Kingdom, mobility patterns returned to
those registered prior to the pandemic after the easing of
non-pharmaceutical interventions (Rowe et al. 2022; Wang et al. 2022).
These findings suggest that COVID-19 generated shock waves leading to
temporary changes in the patterns of internal population movement across
the rural-urban continuum, but it has not significantly altered the
prevalent macro-structures of population movement within countries.

Despite some evidence in the Global North, lees is known about the
impact of COVID-19 on human mobility in the Global South countries. In
Latin America and the Caribbean, internal population movements declined
by 10\% during periods of severe stringent measures, from 16-19\% in
Bolivia, Ecuador and Argentina to less than 3\% in Paraguay and
Venezuela (Aromí et al. 2023). However, there is no evidence about
pandemic outcomes on human mobility across the rural-urban continuum. To
date, surveys carried out in India (Irudaya Rajan, Sivakumar, and
Srinivasan 2020) and South Africa (Ginsburg et al. 2022) have shown
anecdotal evidence suggesting that urbanisation trends declined during
the pandemic, while movements from cities to rural areas increased. Both
studies pointed out that a number of rural residents initiating a
migration decreased during 2020 and some labourers in major cities seem
to have returned to their hometowns due to job losses, since stringency
measures have extensively impacted the functioning of the economy
(Ghosh, Seth, and Tiwary 2020). Many internal migrants residing in
Indian cities lost their jobs and were unable to afford basic
necessities (Nair and Verma 2020). Collectively, these findings suggest
that vulnerable populations seem to have played a role in increasing
counterurbanisation movements in Global South countries. However, recent
work has shown that high-wealth groups residing in cities across Mexico,
Colombia, Brazil, Indonesia, Philippines and South Africa were on
average 159\% more likely to have moved to rural area compared to those
of low-wealth groups during early stages of the pandemic (Lucchini et
al. 2023), in line with the evidencee found for Global North countries.

Despite anecdotal evidence suggesting pandemic impacts on the patterns
of human mobility across the rural-urban hierarchy in the Global South,
lack of suitable data has not allowed us to identify and quantify the
magnitude of these impacts. In this paper, we use Meta-Facebook data to
provide evidence on the effect of the pandemic on the patterns of
internal population movements across four Latin American countries.

\subsection{Contemporary trends of human mobility across the rural-urban hierarchy in Latin America}

Until 1980, rural to urban migration dominated internal population
movements in Latin America with significant levels of population
redistribution, especially during the rapid industrialisation process
between the 1950s and 1970s when a critical mass of individuals moved
from villages and towns to urban areas (Firebaugh 1979; A. Lattes 1995;
J. Sobrino 2012). Urban growth due to internal mobility was primarily
concentrated in large urban centres resulting in a significant
population imbalance between the chief cities and other settlements
(Pinto da Cunha 2002; A. E. Lattes, Rodríguez, and Villa 2017). As a
consequence, Latin America shows high rates of urbanisation with more
than 81\% of the population living in cities, the highest figure after
North America, and about 46\% of urban residents settled in cities with
more than 1 million inhabitants (Nations" 2019).

Currently, internal population movements amongst cities dominate the
migratory system in Latin America, while rural to urban flows are of
less importance (Bernard et al. 2017; Rodríguez-Vignoli and Rowe 2018;
Nations" 2019). According to the 2010 census round, 80\% of internal
migrants moved between cities (Rodríguez-Vignoli and Rowe 2018).
Contrary to the period preceding the 1980s, medium-sizes cities (those
between 500K and 1 million inhabitants) experienced the highest net
migration rates, while large cities (\textgreater1 million) presented
balanced rates and small cities (\textgreater500K) lost individuals by
internal population movements (Rodríguez-Vignoli and Rowe 2018).
Changing patterns in human mobility within Latin American countries were
driven by a decline in internal migration since the 1980s crisis (Chávez
Galindo et al. 2016) and deconcentration dynamics in large urban
centres, such as Mexico City (Jaime Sobrino 2006) or Santiago de Chile
(González Ollino and Rodríguez Vignoli 2006), where long distance
inflows have declined over time. At the same time, there has been an
increasing domestic and foreign investment in export-oriented activities
or tourism industries in certain middle-sized cities, fostering the
geographic dispersal of employment opportunities and increasing the
attractiveness of these cities (Brea 2003; Pérez-Campuzano 2013; Chávez
Galindo et al. 2016).

Suburbanisation also represents an important type of internal population
movement and has increased over time due to the growing expansion of
Latin American cities (Graizbord and Acuña 2007; Chávez Galindo et al.
2016). These movements are underpinned by flows of middle- and
upper-class families moving away from central cities to surrounding
auto-segregated areas of wealthy individuals (Borsdorf 2003; Rodríguez
Vignoli and Rowe 2017). Low-income populations also settle in specific
areas within suburbs where the cost of living is usually cheaper than in
cities (Janoschka 2002; Rodríguez Vignoli and Rowe 2017). A large
proportion of both wealthy individuals and low-income residents commute
daily to central cities, mainly for work reasons (Chávez Galindo et al.
2016). Recently, reurbanisation dynamics have also been observed due to
gentrification processes in specific central areas (J. Sobrino 2012;
Chávez Galindo et al. 2016). In this paper, we analyse how the COVID-19
pandemic has changed patterns of internal population movements across
the rural-urban hierarchy in Mexico, Colombia, Perú, Chile and
Argentina.

\section{Data}

The analysis proposed in this paper is based on several data sets.
Firstly, Facebook Population and Facebook Movements are data sets based
on the Facebook users' location history. The data sets were created by
Facebook's owner company Meta and can be accessed through their Data for
Good programme (\url{https://dataforgood.facebook.com/}). Prior to
releasing the data sets, Meta applies three techniques to ensure privacy
and anonymisation. First, a small undisclosed amount of random noise is
added to ensure that precise location cannot be identified for small
population counts in sparsely populated areas. While removing small
counts may lead to an underrepresentation of the population in these
places, the geographic distribution of population is still reflected in
the data. Second, spatial smoothing is applied to produce a smooth
population count surface using inverse distance-weighted averaging.
Third, any remaining population counts of less than 10 are removed from
the final data set (see Maas et al. Maas et al. (2019) for details).

Both data sets Facebook Population and Facebook Movements contain data
corresponding to a time period comprising approximately two years,
starting from March 2020, and to four Latin American countries,
Argentina, Chile, Colombia and Mexico. The data in both data sets is
temporally aggregated into three 8-hour windows (00:00--08:00,
08:00--16:00 and 16:00--00:00) for every day in the aforementioned
two-year period. It is spatially aggregated into tiles according to the
Bing Maps Tile System. This geospatial indexing system was developed by
Microsoft and it partitions the world into square cells at various
levels of resolution. The Facebook Movements and Facebook Population
datasets are aggregated into various levels of resolution depending on
the country, and the length of the Bing tile sides vary also by country
according to the country's distance to the Equator, even when two
countries use tiles at the same level of resolution.

The Facebook Population data set provides information on the number of
active Facebook users in each tile. The data set Facebook Movements
captures the total number of Facebook users moving between pairs of
origin and destination Bing tiles. We note here that due to the nature
of the Facebook Movement data, we cannot distinguish between different
types of movements, for example, daily commutes to work or permanent
changes of address. However, we are still able to detect the evolution
of movements between origin-destination pairs of Bing tiles and hence,
we are able to capture the impact that COVID-19 has on mobility
patterns.

On top of the data for the two-year period, each entry in the Facebook
Population and Facebook Movements datasets include data for baseline
levels before COVID-19. The baseline values are computed based on a
45-day period ending on the 10th of March of 2020. The data sets also
include a `quality' score indicating the number of standard deviations
by which the observed data at specific locations and time windows differ
from the baseline values, hence highlighting statistically significant
changes.

An additional data set from WorldPop was used to capture the spatial
distribution of population density in the different countries analysed
here. The WorlPop dataset is in raster format and contains gridded
population data at 1 sqkm resolution.

\section{Method}

\subsection{Data pre-processing}

Ultimately, we work with a dataset where each entry represents a
population flow between an origin and a destination pair. For each
entry, we consider when the movement took place, the population density
at the origin and destination, the number of active Facebook users at
the origin and destination and?

\subsubsection{Spatial aggregation}

The working dataset is aggregated at the same level as the Bing tiles
used to track the movements in a given country. To each Bing tile, we
aggregate WorldPop population data and use it to estimate the population
density of each Bing tile.

\subsubsection{Temporal aggregation}

The raw data is aggregated into 8-hour windows for every day in the
two-year period. However, we are interested in longer-term trends and
not in how population flows vary throughout the day. For this reason, we
obtain estimates for the parameters in the models described in
subsection \ref{subsec-models} by splitting the total data set by month.
For the visualisations in section \ref{sec-results}, we also present the
data aggregated by month.

\subsection{Classifying Bing tiles according to their population density}

Here we aim to understand how the population density at the origin and
the destination might influence mobility behaviours. To help
characterise the population density between orign and destination, we
classify the Bing tiles into 10 discrete categories of population
density according to the Jenks natural breaks classification method,
hence obtaining a categorisation of Bing tiles a lot more detailed than
the traditional binary rural/urban classification.

\begin{figure}

{\centering \includegraphics[width=6.51042in,height=6.66667in]{images/Map_density_histogram.png}

}

\caption{Maps of population density categories by country with
histograms, showing that the distribution of Bing tiles in each density
category.}

\end{figure}

\subsection{Estimating the strength of the flows between tiles of different population density categories}

\ref{subsec-models}

In this section, we propose a statistical model to capture the extent to
which the strength of the population flows is systematically impacted by
the density classes of the origin and destination tiles, and how this
effect varies over time. Our approach here is very similar to that
proposed by \cite{Rowe}.

While we could obtain the population flows between different population
classes and their evolution over time, the results would fail to capture
specifically the effect of the density classes of the origin and
destination tiles. This is due to the fact that movement patterns are
influenced not only by the population density of the starting and ending
locations, but also by factors such as the distance, the population
size, the time of the week and day at which movements take place, etc.
These additional variables can act as confounding factors, obscuring the
actual impact of population density. In order to disentangle and obtain
a more accurate estimation of the importance of population density, we
chose to utilise a regression modeling approach. I AM HERE

We adopted a spatial interaction framework (Rowe, Lovelace, et al.,
2022). We used population flows between tiles by time window and day (as
described in Section 3), and modelled them as a function of
characteristics of the flow itself (Fij), the origin (Tilei) and
destination (Tilej) tiles and the temporal nature of the flow (Tw).
Crucially, we included indicator variables that capture the pair of
population classes (Figure 1) of the origin and destination tile. In
mathematical form:

\begin{enumerate}
\def\labelenumi{(\arabic{enumi})}
\setcounter{enumi}{1}
\tightlist
\item
  where µijw is the expectation of the flow of people from tile i to
  tile j in the time window w; ⍺ is an intercept; is a series of
  indicator variables that reflect the pair of population density
  classes of a given origin i (I) and destination tile j (J), resulting
  in 99 pairs (10 classes × 10 classes minus one so it is not collinear
  with ⍺); dij is the geographic distance between tiles i and j; qijw is
  a measure of the quality of the flow estimate provided by
  Meta-Facebook and related to the uncertainty behind the user count of
  the flow as described in Section 3; Popi,j represents the population
  at the origin (i) and destination (j) tiles; are parameters to
  estimate in the model linking their respective covariates to µijw; D
  is a trend tracking the day to which the flow relates to during the
  period in analysis; while Wk and H are indicator variables capturing
  day of the week (i.e., weekday or weekend) and hour window (i.e.,
  00:00--08:00, 08:00--16:00 and 16:00--00:00). Our focus in Equation
  (2) is centred on . Controlling for all other variables, these
  parameters capture the extent to which, the expected flow between a
  given origin-destination population density class pair of tiles (e.g.,
  a high-density origin to a low-density destination) is systematically
  higher or lower than if it occurred between a baseline
  origin-destination population density class pair of tiles (e.g., a
  low-density origin to a low-density destination). Additionally, we
  standardised continuous variables (dij, qijw, Popi j), α so that they
  can be interpreted as the expected flow on the first day (D = 0),
  during the first temporal window (H = 00:00--08:00), on a weekday
  (Wk = 0), for the baseline origin-population population density class
  pair, when all the other variables are at their mean value. In this
  context, each can also be seen as the `modulation factor' around that
  expectation associated with each pair of origin-destination classes.
  The baseline origin-destination population density class pair is the
  lowest population density class as origin and destination.
\end{enumerate}

We used a count data regression model. Specifically, we fitted a
generalised linear model (GLM) where the error term is assumed to be
distributed following a Poisson distribution, with a flow expectation of
(µijw) linked to the flow count (Fijw) through a log link: (3) (4) The
Poisson regression model (PRM) assumes that equidispersion, that is,
equality of mean and variance in the response variable (Cameron \&
Trivedi, 2013). In practice, the equidispersion property is commonly
violated because of overdispersion, that is, the variance exceeds the
mean. When this occurs, the PRM may produce biased parameter estimates,
causing the standard errors of the estimates to be underestimated, and
compromising the statistical inference process (Hilbe, 2011). To test
for overdispersion, we used a regression-based test based on an
auxiliary regression of the conditional variance as described in Cameron
and Trivedi (2013).

Following Gelman and Hill (2006), we used a quasi-PRM to address
overdispersion in our response variable. This is one of the most common
strategies to deal with overdispersion in count data models (Hilbe,
2011). Intuitively, this model adjusts the standard errors of the
estimates to account for the extra dispersion in the data. To implement
this, we estimated Equation (2) by using robust variance estimators. The
number of active Facebook users were used as a weight to account for the
variability of the observed count of population movement over time. This
strategy is also used to mitigate for any potential biases regarding the
variation in the observed number of active Facebook users changes over
time across Britain.

We fitted Equation (2) using iteratively reweighted least squares
(IWLS). We separately estimated models for individual months in our
data, resulting in 18 sets of estimates. Our key aim was to generate
estimates for ⍺ and , so that we focused on discussing the evolution of
these estimates in a grid of line plots with 10 rows and 10 columns,
each of them representing one of our population density classes. The
plot corresponding to the Ith row and Jth column displays the evolution
of the parameter that tracks the intensity of population flows from
tiles in population density class Ith to those in population density
class Jth.

\section{Results}\label{sec-results}

\section{Discussion}

\section{Conclusion}

\section{References}

\hypertarget{refs}{}
\begin{CSLReferences}{1}{0}
\leavevmode\vadjust pre{\hypertarget{ref-aromuxed2023}{}}%
Aromí, Daniel, María Paula Bonel, Julian Cristia, Martín Llada, Juan
Pereira, Xiomara Pulido, and Julieth Santamaria. 2023. {``\#StayAtHome:
Social Distancing Policies and Mobility in Latin America and the
Caribbean.''} \emph{Economía} 22 (1).
\url{https://doi.org/10.31389/eco.4}.

\leavevmode\vadjust pre{\hypertarget{ref-BBCnews2023}{}}%
BBC news. 2023. {``Unemployment: Who Are the Millions of Britons Not
Working?''} \url{https://www.bbc.co.uk/news/business-52660591}.

\leavevmode\vadjust pre{\hypertarget{ref-bell2015}{}}%
Bell, Martin, Elin Charles-Edwards, Philipp Ueffing, John Stillwell,
Marek Kupiszewski, and Dorota Kupiszewska. 2015. {``Internal Migration
and Development: Comparing Migration Intensities Around the World.''}
\emph{Population and Development Review} 41 (1): 33--58.
\url{https://doi.org/10.1111/j.1728-4457.2015.00025.x}.

\leavevmode\vadjust pre{\hypertarget{ref-bernard2017}{}}%
Bernard, Aude, Francisco Rowe, Martin Bell, Philipp Ueffing, and Elin
Charles-Edwards. 2017. {``Comparing Internal Migration Across the
Countries of Latin America: A Multidimensional Approach.''} Edited by
Osman Alimamy Sankoh. \emph{PLOS ONE} 12 (3): e0173895.
\url{https://doi.org/10.1371/journal.pone.0173895}.

\leavevmode\vadjust pre{\hypertarget{ref-bhadra2020}{}}%
Bhadra, Arunava, Arindam Mukherjee, and Kabita Sarkar. 2020. {``Impact
of Population Density on Covid-19 Infected and Mortality Rate in
India.''} \emph{Modeling Earth Systems and Environment} 7 (1): 623--29.
\url{https://doi.org/10.1007/s40808-020-00984-7}.

\leavevmode\vadjust pre{\hypertarget{ref-blustein2020}{}}%
Blustein, David L., Ryan Duffy, Joaquim A. Ferreira, Valerie
Cohen-Scali, Rachel Gali Cinamon, and Blake A. Allan. 2020.
{``Unemployment in the Time of COVID-19: A Research Agenda.''}
\emph{Journal of Vocational Behavior} 119 (June): 103436.
\url{https://doi.org/10.1016/j.jvb.2020.103436}.

\leavevmode\vadjust pre{\hypertarget{ref-borsdorf2003}{}}%
Borsdorf, Axel. 2003. {``Cómo Modelar El Desarrollo y La Dinámica de La
Ciudad Latinoamericana.''} \emph{EURE (Santiago)} 29 (86).
\url{https://doi.org/10.4067/s0250-71612003008600002}.

\leavevmode\vadjust pre{\hypertarget{ref-branduxe9n2020}{}}%
Brandén, Maria, Siddartha Aradhya, Martin Kolk, Juho Härkönen, Sven
Drefahl, Bo Malmberg, Mikael Rostila, Agneta Cederström, Gunnar
Andersson, and Eleonora Mussino. 2020. {``Residential Context and
COVID-19 Mortality Among Adults Aged 70 Years and Older in Stockholm: A
Population-Based, Observational Study Using Individual-Level Data.''}
\emph{The Lancet Healthy Longevity} 1 (2): e80--88.
\url{https://doi.org/10.1016/s2666-7568(20)30016-7}.

\leavevmode\vadjust pre{\hypertarget{ref-Brea03}{}}%
Brea, J. 2003. {``Population Dynamics in Latin America.''}
\emph{Population Bulletin} 58: 3--36.
\url{https://population.un.org/wup/Publications/Files/WUP2018-Report.pdf}.

\leavevmode\vadjust pre{\hypertarget{ref-chuxe1vezgalindo2016}{}}%
Chávez Galindo, Ana María, Jorge Rodríguez Vignoli, Mario Acuña, Jorge
Barquero, Daniel Macadar, José Marcos Pinto da Cunha, and Jaime Sobrino.
2016. {``Migración Interna y Cambios Metropolitanos.''} \emph{Revista
Latinoamericana de Población} 10 (18): 7--41.
\url{https://doi.org/10.31406/relap2016.v10.i1.n18.1}.

\leavevmode\vadjust pre{\hypertarget{ref-Chen2004}{}}%
Chen, Wenhong, and Barry Wellman. 2004. {``The Global Digital Divide -
Within and Between Countries.''} \emph{IT \& Society} 1: 39--45.
\url{http://www.ifs.tuwien.ac.at/~dieter/teaching/GmA/Chen2004.pdf}.

\leavevmode\vadjust pre{\hypertarget{ref-fielding2021}{}}%
Fielding, Tony, and Yoshitaka Ishikawa. 2021. {``COVID{-}19 and
Migration: A Research Note on the Effects of COVID{-}19 on Internal
Migration Rates and Patterns in Japan.''} \emph{Population, Space and
Place} 27 (6). \url{https://doi.org/10.1002/psp.2499}.

\leavevmode\vadjust pre{\hypertarget{ref-firebaugh1979}{}}%
Firebaugh, Glenn. 1979. {``Structural Determinants of Urbanization in
Asia and Latin America, 1950- 1970.''} \emph{American Sociological
Review} 44 (2): 199. \url{https://doi.org/10.2307/2094505}.

\leavevmode\vadjust pre{\hypertarget{ref-florida2021}{}}%
Florida, Richard, Andrés Rodríguez-Pose, and Michael Storper. 2021.
{``Critical Commentary: Cities in a Post-COVID World.''} \emph{Urban
Studies} 60 (8): 1509--31.
\url{https://doi.org/10.1177/00420980211018072}.

\leavevmode\vadjust pre{\hypertarget{ref-ghosh2020}{}}%
Ghosh, Somenath, Pallabi Seth, and Harsha Tiwary. 2020. {``How Does
Covid-19 Aggravate the Multidimensional Vulnerability of Slums in India?
A Commentary.''} \emph{Social Sciences \& Humanities Open} 2 (1):
100068. \url{https://doi.org/10.1016/j.ssaho.2020.100068}.

\leavevmode\vadjust pre{\hypertarget{ref-ginsburg2022}{}}%
Ginsburg, Carren, Mark A. Collinson, F. Xavier Gómez-Olivé, Sadson
Harawa, Chantel F. Pheiffer, and Michael J. White. 2022. {``The Impact
of COVID-19 on a Cohort of Origin Residents and Internal Migrants from
South Africa's Rural Northeast.''} \emph{SSM - Population Health} 17
(March): 101049. \url{https://doi.org/10.1016/j.ssmph.2022.101049}.

\leavevmode\vadjust pre{\hypertarget{ref-Glaeser20}{}}%
Glaeser, Edward L. 2020. {``Cities and Pandemics Have a Long History.
City Journal.''}
\url{https://www.city-journal.org/article/cities-and-pandemics-have-a-long-history}.

\leavevmode\vadjust pre{\hypertarget{ref-gonzuxe1lezollino2006}{}}%
González Ollino, Daniela, and Jorge Rodríguez Vignoli. 2006. {``Spatial
Redistribution and Internal Migration of the Populationin Chile over the
Past 35 Years (1965-20).''} \emph{Estudios Demográficos y Urbanos} 21
(2): 369. \url{https://doi.org/10.24201/edu.v21i2.1253}.

\leavevmode\vadjust pre{\hypertarget{ref-gonzuxe1lez-leonardo2022b}{}}%
González-Leonardo, Miguel, Antonio López-Gay, Niall Newsham, Joaquín
Recaño, and Francisco Rowe. 2022. {``Understanding Patterns of Internal
Migration During the COVID{-}19 Pandemic in Spain.''} \emph{Population,
Space and Place} 28 (6). \url{https://doi.org/10.1002/psp.2578}.

\leavevmode\vadjust pre{\hypertarget{ref-gonzuxe1lez-leonardo2023}{}}%
González-Leonardo, Miguel, Michaela Potančoková, Dilek Yildiz, and
Francisco Rowe. 2023. {``Quantifying the Impact of COVID-19 on
Immigration in Receiving High-Income Countries.''} Edited by Ricky Chee
Jiun Chia. \emph{PLOS ONE} 18 (1): e0280324.
\url{https://doi.org/10.1371/journal.pone.0280324}.

\leavevmode\vadjust pre{\hypertarget{ref-gonzuxe1lez-leonardo2022a}{}}%
González-Leonardo, Miguel, Francisco Rowe, and Alberto
Fresolone-Caparrós. 2022. {``Rural Revival? The Rise in Internal
Migration to Rural Areas During the COVID-19 Pandemic. Who Moved and
Where?''} \emph{Journal of Rural Studies} 96 (December): 332--42.
\url{https://doi.org/10.1016/j.jrurstud.2022.11.006}.

\leavevmode\vadjust pre{\hypertarget{ref-gonzuxe1lez-leonardo2022}{}}%
González-Leonardo, Miguel, and Jeroen Spijker. 2022. {``The Impact of
Covid-19 on Demographic Components in Spain, 2020{\textendash}31: A
Scenario Approach.''} \emph{Population Studies}, November, 1--17.
\url{https://doi.org/10.1080/00324728.2022.2138521}.

\leavevmode\vadjust pre{\hypertarget{ref-graizbord2007}{}}%
Graizbord, Boris, and Beatriz Acuña. 2007. {``Movilidad Residencial En
La Ciudad de México / Residential Mobility in Mexico City.''}
\emph{Estudios Demográficos y Urbanos} 22 (2): 291.
\url{https://doi.org/10.24201/edu.v22i2.1281}.

\leavevmode\vadjust pre{\hypertarget{ref-green2021}{}}%
Green, Mark, Frances Darlington Pollock, and Francisco Rowe. 2021.
{``New Forms of Data and New Forms of Opportunities to Monitor and
Tackle a Pandemic.''} In, 423--29. Springer International Publishing.
\url{https://doi.org/10.1007/978-3-030-70179-6_56}.

\leavevmode\vadjust pre{\hypertarget{ref-haslag2021}{}}%
Haslag, Peter H., and Daniel Weagley. 2021. {``From L.A. To Boise: How
Migration Has Changed During the COVID-19 Pandemic.''} \emph{SSRN
Electronic Journal}. \url{https://doi.org/10.2139/ssrn.3808326}.

\leavevmode\vadjust pre{\hypertarget{ref-huremovic2019}{}}%
Huremović, Damir. 2019. {``Brief History of Pandemics (Pandemics
Throughout History).''} In, 7--35. Springer International Publishing.
\url{https://doi.org/10.1007/978-3-030-15346-5_2}.

\leavevmode\vadjust pre{\hypertarget{ref-internat2021}{}}%
{``International Migration 2020: Highlights.''} 2021, January.
\url{https://doi.org/10.18356/9789210052689}.

\leavevmode\vadjust pre{\hypertarget{ref-irudayarajan2020}{}}%
Irudaya Rajan, S., P. Sivakumar, and Aditya Srinivasan. 2020. {``The
COVID-19 Pandemic and Internal Labour Migration in India: A {`}Crisis of
Mobility{'}.''} \emph{The Indian Journal of Labour Economics} 63 (4):
1021--39. \url{https://doi.org/10.1007/s41027-020-00293-8}.

\leavevmode\vadjust pre{\hypertarget{ref-janoschka2002}{}}%
Janoschka, Michael. 2002. {``El Nuevo Modelo de La Ciudad
Latinoamericana: Fragmentación y Privatización.''} \emph{EURE
(Santiago)} 28 (85).
\url{https://doi.org/10.4067/s0250-71612002008500002}.

\leavevmode\vadjust pre{\hypertarget{ref-kotsubo2022}{}}%
Kotsubo, Masaki, and Tomoki Nakaya. 2022. {``Trends in Internal
Migration in Japan, 2012{\textendash}2020: The Impact of the COVID{-}19
Pandemic.''} \emph{Population, Space and Place} 29 (4).
\url{https://doi.org/10.1002/psp.2634}.

\leavevmode\vadjust pre{\hypertarget{ref-lattes95}{}}%
Lattes, A. 1995. {``Urbanización, Crecimiento Urbano y Migraciones En
América Latina. Población y Desarrollo: Tendencias y Desafíos.''}
\emph{Notas de Población} 62: 211--60.
\url{https://repositorio.cepal.org/handle/11362/38594}.

\leavevmode\vadjust pre{\hypertarget{ref-lattes2017}{}}%
Lattes, Alfredo E., Jorge Rodríguez, and Miguel Villa. 2017.
{``Population Dynamics and Urbanization in Latin America: Concepts and
Data Limitations.''} In, 89--111. Routledge.
\url{https://doi.org/10.4324/9781315248073-5}.

\leavevmode\vadjust pre{\hypertarget{ref-lucchini2023}{}}%
Lucchini, Lorenzo, Ollin Langle-Chimal, Lorenzo Candeago, Lucio Melito,
Alex Chunet, Aleister Montfort, Bruno Lepri, Nancy Lozano-Gracia, and
Samuel P. Fraiberger. 2023. {``Socioeconomic Disparities in Mobility
Behavior During the COVID-19 Pandemic in Developing Countries.''}
\url{https://doi.org/10.48550/ARXIV.2305.06888}.

\leavevmode\vadjust pre{\hypertarget{ref-Maas19}{}}%
Maas, P., S. Iyer, A. Gros, W. Park, L. McGorman, C. Nayak, and P. A.
Dow. 2019. {``Facebook Disaster Maps: Aggregate Insights for Crisis
Response and Recovery.''} In \emph{16th International Conference on
Information Systems for Crisis Response and Management}, 836--47.

\leavevmode\vadjust pre{\hypertarget{ref-McKinsey21}{}}%
McKinsey \& Company. 2021. {``What Executives Are Saying about the
Future of Hybrid Work.''}
\url{https://www.mckinsey.com/capabilities/people-and-organizational-performance/our-insights/what-executives-are-saying-about-the-future-of-hybrid-work}.

\leavevmode\vadjust pre{\hypertarget{ref-ILO2020}{}}%
Nair, Shabari, and Divya Verma. 2020. {``A Policy Framework for India's
Covid-19 Migration. International Labour Organization.''}
\url{https://www.ilo.org/newdelhi/info/public/fs/WCMS_746654/lang--en/index.htm}.

\leavevmode\vadjust pre{\hypertarget{ref-nathan2020}{}}%
Nathan, Max, and Henry Overman. 2020. {``Will Coronavirus Cause a Big
City Exodus?''} \emph{Environment and Planning B: Urban Analytics and
City Science} 47 (9): 1537--42.
\url{https://doi.org/10.1177/2399808320971910}.

\leavevmode\vadjust pre{\hypertarget{ref-UNpopulation19}{}}%
Nations", "United. 2019. {``World Urbanization Prospects. The 2018
Revision.''}
\url{https://population.un.org/wup/Publications/Files/WUP2018-Report.pdf}.

\leavevmode\vadjust pre{\hypertarget{ref-nouvellet2021}{}}%
Nouvellet, Pierre, Sangeeta Bhatia, Anne Cori, Kylie E. C. Ainslie, Marc
Baguelin, Samir Bhatt, Adhiratha Boonyasiri, et al. 2021. {``Reduction
in Mobility and COVID-19 Transmission.''} \emph{Nature Communications}
12 (1). \url{https://doi.org/10.1038/s41467-021-21358-2}.

\leavevmode\vadjust pre{\hypertarget{ref-oecdreg}{}}%
\emph{OECD Regions and Cities at a Glance}. n.d. OECD.
\url{https://doi.org/10.1787/26173212}.

\leavevmode\vadjust pre{\hypertarget{ref-oecdsoc}{}}%
{``OECD Social, Employment and Migration Working Papers.''} n.d.
\url{https://doi.org/10.1787/1815199x}.

\leavevmode\vadjust pre{\hypertarget{ref-perales2022}{}}%
Perales, Francisco, and Aude Bernard. 2022. {``Continuity or Change? How
the Onset of COVID{-}19 Affected Internal Migration in Australia.''}
\emph{Population, Space and Place} 29 (2).
\url{https://doi.org/10.1002/psp.2626}.

\leavevmode\vadjust pre{\hypertarget{ref-Puxe9rez-Campuzano13}{}}%
Pérez-Campuzano, C., E. y Santos-Cerquera. 2013. {``Tendencias Recientes
de La Migración Interna En México.''} \emph{"Papeles de Población"} 19:
53--88.
\url{https://www.scielo.org.mx/scielo.php?script=sci_arttext\&pid=S1405-74252013000200003}.

\leavevmode\vadjust pre{\hypertarget{ref-Pinilla2021}{}}%
Pinilla, Vicente, and Luis Antonio Sáez. 2021. {``La Despoblación Rural
En España: Características, Causas e Implicaciones Para Las Políticas
Públicas.''} \emph{Presupuesto y Gasto Público} 102: 75--92.
\url{https://www.ief.es/docs/destacados/publicaciones/revistas/pgp/102.pdf\#page=75}.

\leavevmode\vadjust pre{\hypertarget{ref-PintoDaCunha02}{}}%
Pinto da Cunha, J. M. P. 2002. {``Urbanización, Redistribución Espacial
de La Población y Transformaciones Socioeconómicas En América Latina.''}
\emph{United Nations}.
\url{https://repositorio.cepal.org/bitstream/handle/11362/7168/1/S029663_es.pdf}.

\leavevmode\vadjust pre{\hypertarget{ref-Pomeroy2021}{}}%
Pomeroy, Robin, and Ross Chiney. 2021. {``Has COVID Killed Our Cities?
World Economic Forum.''}
\url{https://www.weforum.org/agenda/2020/11/cities-podcast-new-york-dead/}.

\leavevmode\vadjust pre{\hypertarget{ref-ramani2021}{}}%
Ramani, Arjun, and Nicholas Bloom. 2021. {``The Donut Effect of Covid-19
on Cities.''} \url{https://doi.org/10.3386/w28876}.

\leavevmode\vadjust pre{\hypertarget{ref-rodruxedguezvignoli2017}{}}%
Rodríguez Vignoli, Jorge, and Francisco Rowe. 2017. {``¿Contribuye La
Migración Interna a Reducir La Segregación Residencial?''} \emph{Revista
Latinoamericana de Población} 11 (21): 7--45.
\url{https://doi.org/10.31406/relap2017.v11.i2.n21.1}.

\leavevmode\vadjust pre{\hypertarget{ref-rodruxedguez-vignoli2018}{}}%
Rodríguez-Vignoli, Jorge, and Francisco Rowe. 2018. {``How Is Internal
Migration Reshaping Metropolitan Populations in Latin America? A New
Method and New Evidence.''} \emph{Population Studies} 72 (2): 253--73.
\url{https://doi.org/10.1080/00324728.2017.1416155}.

\leavevmode\vadjust pre{\hypertarget{ref-rowe2021}{}}%
Rowe, Francisco. 2021. {``Big Data and Human Geography.''}
\url{http://dx.doi.org/10.31235/osf.io/phz3e}.

\leavevmode\vadjust pre{\hypertarget{ref-rowe2022}{}}%
Rowe, Francisco, Alessia Calafiore, Daniel Arribas-Bel, Krasen
Samardzhiev, and Martin Fleischmann. 2022. {``Urban Exodus?
Understanding Human Mobility in Britain During the COVID-19 Pandemic
Using Facebook Data.''} \url{https://doi.org/10.48550/ARXIV.2206.03272}.

\leavevmode\vadjust pre{\hypertarget{ref-rowe2023}{}}%
Rowe, Francisco, Miguel González-Leonardo, and Tony Champion. 2023.
{``Virtual Special Issue: Internal Migration in Times of COVID{-}19.''}
\emph{Population, Space and Place}, March.
\url{https://doi.org/10.1002/psp.2652}.

\leavevmode\vadjust pre{\hypertarget{ref-OIT2018}{}}%
Salazar-Xirinachs, Juan, José Manuel \& Chacaltana. 2018. {``Políticas
de Formalización En América Latina: Avances y Desafíos. Organización
Internacional Del Trabajo.''}
\url{https://www.ilo.org/americas/publicaciones/WCMS_645159/lang--es/index.htm}.

\leavevmode\vadjust pre{\hypertarget{ref-sobrino12}{}}%
Sobrino, J. 2012. {``La Urbanización En El México Contemporáneo.''}
\emph{Notas de Población} 94: 93--122.
\url{https://repository.eclac.org/handle/11362/12898}.

\leavevmode\vadjust pre{\hypertarget{ref-sobrino2006}{}}%
Sobrino, Jaime. 2006. {``Comptetitiveness and Employment in the Largest
Metropolitan Areas of Mexico.''} In, 309--54. El Colegio de México.
\url{https://doi.org/10.2307/j.ctv3dnrw3.15}.

\leavevmode\vadjust pre{\hypertarget{ref-stawarz2022}{}}%
Stawarz, Nico, Matthias Rosenbaum-Feldbrügge, Nikola Sander, Harun
Sulak, and Vanessa Knobloch. 2022. {``The Impact of the COVID{-}19
Pandemic on Internal Migration in Germany: A Descriptive Analysis.''}
\emph{Population, Space and Place} 28 (6).
\url{https://doi.org/10.1002/psp.2566}.

\leavevmode\vadjust pre{\hypertarget{ref-storper2004}{}}%
Storper, M., and A. J. Venables. 2004. {``Buzz: Face-to-Face Contact and
the Urban Economy.''} \emph{Journal of Economic Geography} 4 (4):
351--70. \url{https://doi.org/10.1093/jnlecg/lbh027}.

\leavevmode\vadjust pre{\hypertarget{ref-TheGuardian2020}{}}%
The Guardian. 2020. {``Escape to the Country: How Covid Is Driving an
Exodus from Britain's Cities.''}
\url{https://www.theguardian.com/world/2020/sep/26/escape-country-covid-exodus-britain-cities-pandemic-urban-green-space}.

\leavevmode\vadjust pre{\hypertarget{ref-tuxf8nnessen2021}{}}%
Tønnessen, Marianne. 2021. {``Movers from the City in the First Year of
Covid.''} \emph{Nordic Journal of Urban Studies} 1 (2): 131--47.
\url{https://doi.org/10.18261/issn.2703-8866-2021-02-03}.

\leavevmode\vadjust pre{\hypertarget{ref-USBureauLabor2021}{}}%
US Bureau of Labor Statistics. 2021. {``Unemployment Rises in 2020, as
the Country Battles the COVID-19 Pandemic.''}
\url{https://www.bls.gov/opub/mlr/2021/article/unemployment-rises-in-2020-as-the-country-battles-the-covid-19-pandemic.htm}.

\leavevmode\vadjust pre{\hypertarget{ref-vogiazides2022}{}}%
Vogiazides, Louisa, and Juta Kawalerowicz. 2022. {``Internal Migration
in the Time of Covid: Who Moves Out of the Inner City of Stockholm and
Where Do They Go?''} \emph{Population, Space and Place} 29 (4).
\url{https://doi.org/10.1002/psp.2641}.

\leavevmode\vadjust pre{\hypertarget{ref-wang2022}{}}%
Wang, Yikang, Chen Zhong, Qili Gao, and Carmen Cabrera-Arnau. 2022.
{``Understanding Internal Migration in the UK Before and During the
COVID-19 Pandemic Using Twitter Data.''} \emph{Urban Informatics} 1 (1).
\url{https://doi.org/10.1007/s44212-022-00018-w}.

\leavevmode\vadjust pre{\hypertarget{ref-WorldBank2023}{}}%
World Bank", "The. 2023. {``LAC Equity Lab: Poverty - Poverty Rate.''}
\url{https://www.worldbank.org/en/topic/poverty/lac-equity-lab1/poverty/head-count}.

\end{CSLReferences}

% print the bibliography
\setlength{\bibsep}{0.00cm plus 0.05cm} % no space between items
\bibliographystyle{apalike}
\bibliography{sim_refs}

\end{document}

### 2024-01-17 manuscript/manuscript.qmd
file-restructure

--- 2023-10-02 manuscript-old/manuscript-latin-countries.qmd
@@
 ---
-title: "The mid-term impact of COVID-19 on human mobility patterns in Latin American countries"
+title: "The recovery patterns of human mobility in Latin American urban areas after COVID-19"
 author: "Carmen Cabrera-Arnau, Miguel González-Leonardo, Andrea Nasuto, Ruth Neville, Francisco Rowe"
@@
 <!--# Add comments using this -->
-
-The COVID-19 pandemic globally resulted in disruptive shocks to the human mobility system.
-Governments stringency measures and the economic downturn caused by the pandemic constrained international migration [@internat2021; @gonzález-leonardo2022; @gonzález-leonardo2023] and population movements within countries [@nouvellet2021; @gonzález-leonardo2022a; @wang2022; @rowe2023].
-During early stages of the pandemic, big cities became the main epicentres of COVID-19 infections and deaths [@Pomeroy2021] due to their higher connectivity via air travel, higher population density and concentration of public-facing jobs [@brandén2020; @bhadra2020].
-As a result, governments implemented non-pharmaceutical interventions, such as lockdowns and business closures, leading to increases in unemployment and removing the urban vibrancy [@blustein2020; @BBCnews2023].
-In addition, remote work and online teaching reduced the need for commuting, living close to work and education centers, and added pressure for individuals living in small and crowded spaces to move out of cities and look for more affordable and comfortable housing [@nathan2020; @USBureauLabor2021].
-These changes are believed to have reduced the attractiveness of cities during the pandemic, while suburbs and rural areas have gained popularity [@florida2021].
-As a result, a number of headlines have claimed the emergence of an 'urban exodus' to areas with lower population densities (e.g. [@TheGuardian2020]).
-
-Despite their potential rise in popularity, remote locations offer limited availability and diversity of job opportunities, and tend to lack infrastructure and services, such as schooling and health, to effectively accommodate a large population influx [@Pinilla2021; @oecdreg].
-They generally lack high-speed Internet connection which is vital for teleworking [@Chen2004].
-Additionally, most jobs cannot be done remotely [@florida2021, @oecdsoc], and some companies have returned to in-person office work or have implemented hybrid forms of work, following the relaxation of stringency measures [@McKinsey21].
-Schools and universities have also returned to face-to-face teaching.
-Therefore, living close to work or study places remains relevant.
-Businesses and leisure activities in cities have gradually returned to normal, reinvigorating the economic functioning and urban life.
-Collectively, these arguments suggest that a potential increase in counterurbanisation movements during the pandemic was concentrated during periods of stringency measures and is likely to be short term.
-Big cities have successfully weathered previous pandemics [@huremovic2019; @Glaeser20] and are likely to remain attractive places to live.
-Thanks to agglomeration economies, cities are normally regarded as centres for economic growth, clusters of talent, consumer bases and spaces for face‐to‐face interaction and diversity [@storper2004; @florida2021].
-
-Recent empirical evidence has shown a decline on human mobility during periods of severe stringency measures, coupled with a rise of movements from cities to less populated areas and a slowdown of urbanisation movements across some countries of the Global North: the United States [@ramani2021], United Kingdom [@rowe2023; @wang2022], Spain [@gonzález-leonardo2022b; @gonzález-leonardo2022a], Germany [\@; @stawarz2022], Sweden [@vogiazides2022], Norway [@tønnessen2021], Australia [@perales2022] and Japan [@fielding2021, @kotsubo2022].
-The majority of movements away from cities headed to their suburbs [@vogiazides2022; @stawarz2022] or rural areas in their close proximity [@gonzález-leonardo2022b; @gonzález-leonardo2022a] showing in some instances a "doghnut effect", with a decline in population movements directed to cities and a rise in the inflow of people to less dense surrounding areas [@ramani2021].
-Studies demonstrated that professionals [@tønnessen2021] and individuals with a high income [@haslag2021], those who are potentially able to practice remote working, underpinned movements away from large cities.
-In addition, research has shown that a large share of counterurbanisation movements is headed to touristic locations with a high concentration of second homes, suggesting that middle and high class individuals may have been key actors in movements from cities to less populated areas during the pandemic [@gonzález-leonardo2022a].
-However, most studies suggested that variations in internal population movements during the pandemic are likely to be temporary and have not altered previous macro-structures of human mobility across the rural-urban hierarchy [@rowe2023].
-
-In Latin America and the Caribbean, a reduction in internal population movements was documented during periods of high stringency [@aromí2023], but there is no evidence on how human mobility has changed across the rural-urban continuum.
-Anecdotal reports have shown that movements away from cities seem to have accelerated in two countries of the Global South, India [@irudayarajan2020] and South Africa [@ginsburg2022], while urbanisation trends have decelerated.
-However, lack of suitable up-to-date data has prevented us from properly assessing the "urban exodus" hypothesis in Global South countries, as well as the extent, spatial patterns, and durability of potential changes on internal population movements.
-The evidence for these two countries suggested that increasing counterurbanisation movements during the pandemic in India and South Africa could be underpinned by internal migrants residing in cities who returned to their hometowns due to a rise in unemployment [@irudayarajan2020; @ginsburg2022].
-However, recent studies have shown that urban residents from wealthy neighborhoods were more likely to move to rural areas than those from low income neighborhoods during the first wave of the pandemic in Brazil, Colombia, Mexico, Indonesia, Philippines and South Africa [@lucchini2023].
-
-It is known that human mobility within countries tends to vary in systematic ways with development levels [@bell2015].
-In Latin America and the Caribbean, 28.6% of the population was living on less than \$6.8 per day in 2020 [@WorldBank2023] and the informality rate was about 53% in 2016 [@OIT2018].
-Thus, we could expect different pandemic impacts on the patterns of internal population movements between countries in the Global North and the Global South.
-The way in which COVID-19 unfolded in each country, with varying levels of stringency measures, may have also produced different outcomes on human mobility.
-
-Traditionally, censuses have been used to analyse human mobility patterns within Global South countries, which lack register data [@bell2015; @bernard2017].
-Censuses, however, are normally updated every 10 years, lacking the temporal granularity to explore population movements over yearly, monthly or short-time intervals.
-
-Digital traces data from mobile phone applications, on the other hand, provide a unique opportunity to capture internal population movements with small spatial and temporal granularity in multiple countries [@green2021; @rowe2021].
-Drawing on geographically granular Meta-Facebook data from March 2020 to March 2022, we aim to analyse how the pandemic has changed the patterns of human mobility across the rural-urban hierarchy in four Latin American countries: Argentina, Chile, Colombia and Mexico.
-We aim to address the following research questions: 1) To what extent did people leave cities and move to less populated areas in these countries?
-2) Which are the spatial patterns of potential changes on internal population movements?
-3) How have the patterns of human mobility changed over the course of the pandemic?
-4) Are potential changes likely to endure the pandemic?
-
-The rest of the paper is structured as follows: ...
-
-\section{Background}
-
-\subsection{What do we know about internal population movements during the pandemic?}
-
-The COVID-19 pandemic led to an overall decline in human mobility within countries [@nouvellet2021].
-Research has shown that internal population movements across the rural-urban continuum decreased in the United States [@ramani2021], several European countries, Japan and Austria during 2020 [@rowe2023].
-Declines ranged from 2.5% in Spain to 8.5% in Australia.
-The largest drops occurred during periods of national lockdowns and the implementation of other stringency measures, such as mobility restrictions, work and school closing, while levels of internal population movements recovered or exceeded pre-pandemic figures when lockdowns were lifted.
-Declines were partially attributed to a reduction in involuntary migration during lockdowns, a loss of labour market dynamism due to the pandemic, fewer people changing jobs and increasing remote work [@perales2022].
-
-COVID-19 has also altered patterns of internal population movements across the rural-urban hierarchy [@rowe2022].
-In large metro areas across the United States, population movements, as well as households and businesses, shifted from central cities to suburbs and exurbs during 2020 [@ramani2021].
-These authors label this trend as a "doughnut effect", reflecting a decrease in population inflows and activities in city centers and a growth in suburban rings.
-They found a sizeable "doghnut effect" in large cities, a smaller effect in medium-sized cities and almost no effect for small cities.
-A similar trend was observed in Norway [@tønnessen2021], Germany [@stawarz2022], Sweden [@vogiazides2022] and Japan [@fielding2021; @kotsubo2022], where net migration rates in big cities declined, while population movement to their suburbs increased.
-In addition, this trend was accompanied by a deceleration of urbanisation trends and unusual population gains in rural areas due to an acceleration in counterurbanisation movements, mostly in rural locations which are known to be holiday destinations with a high concentration of second homes and are generally located in the close proximity to big cities.
-An increase of counterurbanisation was the main pandemic outcome on human mobility in Spain [@gonzález-leonardo2022b; @gonzález-leonardo2022a], the United Kingdom [@wang2022; @rowe2022] and Australia [@perales2022], where there is no evidence of a "doghnut effect", since substantial variations in suburbs were not found across these countries.
-
-However, studies suggest that variations are likely to be temporary [@rowe2023].
-In Spain, the urbanisation movement returned to pre-pandemic levels when the lockdown ended in mid-June 2020 [@gonzález-leonardo2022b] and unusually high levels of counterurbanisation persisted over 2021, although the majority of movements continued to occur between and within urban areas [@gonzález-leonardo2022a].
-In Australia, the pandemic has caused minor changes in spatial patterns of internal migration, and its effects were minimal by the end of 2020 [@perales2022].
-In the United Kingdom, mobility patterns returned to those registered prior to the pandemic after the easing of non-pharmaceutical interventions [@rowe2022; @wang2022].
-These findings suggest that COVID-19 generated shock waves leading to temporary changes in the patterns of internal population movement across the rural-urban continuum, but it has not significantly altered the prevalent macro-structures of population movement within countries.
-
-Despite some evidence in the Global North, lees is known about the impact of COVID-19 on human mobility in the Global South countries.
-In Latin America and the Caribbean, internal population movements declined by 10% during periods of severe stringent measures, from 16-19% in Bolivia, Ecuador and Argentina to less than 3% in Paraguay and Venezuela [@aromí2023].
-However, there is no evidence about pandemic outcomes on human mobility across the rural-urban continuum.
-To date, surveys carried out in India [@irudayarajan2020] and South Africa [@ginsburg2022] have shown anecdotal evidence suggesting that urbanisation trends declined during the pandemic, while movements from cities to rural areas increased.
-Both studies pointed out that a number of rural residents initiating a migration decreased during 2020 and some labourers in major cities seem to have returned to their hometowns due to job losses, since stringency measures have extensively impacted the functioning of the economy [@ghosh2020].
-Many internal migrants residing in Indian cities lost their jobs and were unable to afford basic necessities [@ILO2020].
-Collectively, these findings suggest that vulnerable populations seem to have played a role in increasing counterurbanisation movements in Global South countries.
-However, recent work has shown that high-wealth groups residing in cities across Mexico, Colombia, Brazil, Indonesia, Philippines and South Africa were on average 159% more likely to have moved to rural area compared to those of low-wealth groups during early stages of the pandemic [@lucchini2023], in line with the evidencee found for Global North countries.
-
-Despite anecdotal evidence suggesting pandemic impacts on the patterns of human mobility across the rural-urban hierarchy in the Global South, lack of suitable data has not allowed us to identify and quantify the magnitude of these impacts.
-In this paper, we use Meta-Facebook data to provide evidence on the effect of the pandemic on the patterns of internal population movements across four Latin American countries.
-
-\subsection{Contemporary trends of human mobility across the rural-urban hierarchy in Latin America}
-
-Until 1980, rural to urban migration dominated internal population movements in Latin America with significant levels of population redistribution, especially during the rapid industrialisation process between the 1950s and 1970s when a critical mass of individuals moved from villages and towns to urban areas [@firebaugh1979; @lattes95; @sobrino12].
-Urban growth due to internal mobility was primarily concentrated in large urban centres resulting in a significant population imbalance between the chief cities and other settlements [@PintoDaCunha02; @lattes2017].
-As a consequence, Latin America shows high rates of urbanisation with more than 81% of the population living in cities, the highest figure after North America, and about 46% of urban residents settled in cities with more than 1 million inhabitants [@UNpopulation19].
-
-Currently, internal population movements amongst cities dominate the migratory system in Latin America, while rural to urban flows are of less importance [@bernard2017; @rodríguez-vignoli2018; @UNpopulation19].
-According to the 2010 census round, 80% of internal migrants moved between cities [@rodríguez-vignoli2018].
-Contrary to the period preceding the 1980s, medium-sizes cities (those between 500K and 1 million inhabitants) experienced the highest net migration rates, while large cities (\>1 million) presented balanced rates and small cities (\>500K) lost individuals by internal population movements [@rodríguez-vignoli2018].
-Changing patterns in human mobility within Latin American countries were driven by a decline in internal migration since the 1980s crisis [@chávezgalindo2016] and deconcentration dynamics in large urban centres, such as Mexico City [@sobrino2006] or Santiago de Chile [@gonzálezollino2006], where long distance inflows have declined over time.
-At the same time, there has been an increasing domestic and foreign investment in export-oriented activities or tourism industries in certain middle-sized cities, fostering the geographic dispersal of employment opportunities and increasing the attractiveness of these cities [@Brea03; @Pérez-Campuzano13; @chávezgalindo2016].
-
-Suburbanisation also represents an important type of internal population movement and has increased over time due to the growing expansion of Latin American cities [@graizbord2007; @chávezgalindo2016].
-These movements are underpinned by flows of middle- and upper-class families moving away from central cities to surrounding auto-segregated areas of wealthy individuals [@borsdorf2003; @rodríguezvignoli2017].
-Low-income populations also settle in specific areas within suburbs where the cost of living is usually cheaper than in cities [@janoschka2002; @rodríguezvignoli2017].
-A large proportion of both wealthy individuals and low-income residents commute daily to central cities, mainly for work reasons [@chávezgalindo2016].
-Recently, reurbanisation dynamics have also been observed due to gentrification processes in specific central areas [@sobrino12; @chávezgalindo2016].
-In this paper, we analyse how the COVID-19 pandemic has changed patterns of internal population movements across the rural-urban hierarchy in Mexico, Colombia, Perú, Chile and Argentina.
 

### 2024-01-17 manuscript-1-old/manuscript-latin-countries.tex
compiling-manuscript

--- 2024-01-17 manuscript-latin-countries-old/manuscript-latin-countries.tex
@@
 	pdfauthor={Carmen Cabrera-Arnau},
-	pdftitle={The mid-term impact of COVID-19 on human mobility patterns in Latin American countries},
+	pdftitle={The influence of socioeconomic deprivation on the recovery of urban mobility after COVID-19 in Latin America},
 	pdfsubject={Title},
@@
 
-\title{The mid-term impact of COVID-19 on human mobility patterns in Latin American countries\footnote{\textbf{Citation}: Cabrera-Arnau, C., González-Leonardo, M., Nasuto, A., Neville, R., Rowe, F.
-(2023).
-The mid-term impact of COVID-19 on human mobility patterns in Latin American countries}}
+\title{The influence of socioeconomic deprivation on the recovery of urban mobility after COVID-19 in Latin America}
 \author[1]{Carmen Cabrera-Arnau \thanks{\textit{Corresponding author}: c.cabrera-arnau@liverpool.ac.uk}}
@@
 
-Recent empirical studies, predominantly from countries in the Global North, have shown that the COVID-19 pandemic disrupted human mobility at different spatial scales, but particularly in big cities, initially identified as epicentres of infections.
-Lockdowns, remote work, and online education decreased the demand for commuting and urban living, resulting in an "urban exodus".
-Despite the existing evidence, little is known about whether counterurbanisation movements have unfolded similarly in the Global South.
-In this study, we use anonymised, high-resolution location data from Facebook users to answer the following research questions: i) to what extent are people leaving cities?
-ii) are there variations across countries?
-iii) are there changes in mobility patterns over time?
-iv) how long do these changes last?
-The data is aggregated into spatial units according to the Bing Maps Tile System and temporally aggregated into 8-hour time windows for a period of approximately 2 years.
-The methodology involves three steps.
-Firstly, we classify the Bing tiles into different population density categories according to population data from WorldPop.
-Secondly, we quantify and analyse the number and characteristics of movements between different locations in each country.
-Finally, we use statistical modelling and spatial interaction models to assess how the population density category at the origin and destination affects the strength of the flows at different points in time.
-We observe that the strength of flows between Bing tiles belonging to different population density categories was systematically reduced in the early days of the pandemic and, even though by 2022 it seemed to have increased, it never returned to pre-pandemic levels.
-In the early days of COVID-19, we also find a systematically increase above pre-pandemic levels of the strength of the flows between Bing tiles belonding to the the same population density category.
-As of 2022, the frequency of these movements within population density classes have decreased with respect to 2020.
-
 %\vspace{1cm}
@@
 
-\begin{figure}
-
-{\centering \includegraphics[width=6.51042in,height=6.66667in]{images/Map_density_histogram.png}
+\begin{figure}[H]
+
+{\centering \includegraphics[width=6.51042in,height=6.66667in]{figures/Map_density_histogram.png}
 
@@
 
-\end{figure}
+\end{figure}%
 
@@
 
-\hypertarget{refs}{}
+\phantomsection\label{refs}
 \begin{CSLReferences}{1}{0}
-\leavevmode\vadjust pre{\hypertarget{ref-aromuxed2023}{}}%
+\bibitem[\citeproctext]{ref-aromuxed2023}
 Aromí, Daniel, María Paula Bonel, Julian Cristia, Martín Llada, Juan
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-BBCnews2023}{}}%
+\bibitem[\citeproctext]{ref-BBCnews2023}
 BBC news. 2023. {``Unemployment: Who Are the Millions of Britons Not
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-bell2015}{}}%
+\bibitem[\citeproctext]{ref-bell2015}
 Bell, Martin, Elin Charles-Edwards, Philipp Ueffing, John Stillwell,
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-bernard2017}{}}%
+\bibitem[\citeproctext]{ref-bernard2017}
 Bernard, Aude, Francisco Rowe, Martin Bell, Philipp Ueffing, and Elin
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-bhadra2020}{}}%
+\bibitem[\citeproctext]{ref-bhadra2020}
 Bhadra, Arunava, Arindam Mukherjee, and Kabita Sarkar. 2020. {``Impact
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-blustein2020}{}}%
+\bibitem[\citeproctext]{ref-blustein2020}
 Blustein, David L., Ryan Duffy, Joaquim A. Ferreira, Valerie
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-borsdorf2003}{}}%
+\bibitem[\citeproctext]{ref-borsdorf2003}
 Borsdorf, Axel. 2003. {``Cómo Modelar El Desarrollo y La Dinámica de La
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-branduxe9n2020}{}}%
+\bibitem[\citeproctext]{ref-branduxe9n2020}
 Brandén, Maria, Siddartha Aradhya, Martin Kolk, Juho Härkönen, Sven
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-Brea03}{}}%
+\bibitem[\citeproctext]{ref-Brea03}
 Brea, J. 2003. {``Population Dynamics in Latin America.''}
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-chuxe1vezgalindo2016}{}}%
+\bibitem[\citeproctext]{ref-chuxe1vezgalindo2016}
 Chávez Galindo, Ana María, Jorge Rodríguez Vignoli, Mario Acuña, Jorge
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-Chen2004}{}}%
+\bibitem[\citeproctext]{ref-Chen2004}
 Chen, Wenhong, and Barry Wellman. 2004. {``The Global Digital Divide -
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-fielding2021}{}}%
+\bibitem[\citeproctext]{ref-fielding2021}
 Fielding, Tony, and Yoshitaka Ishikawa. 2021. {``COVID{-}19 and
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-firebaugh1979}{}}%
+\bibitem[\citeproctext]{ref-firebaugh1979}
 Firebaugh, Glenn. 1979. {``Structural Determinants of Urbanization in
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-florida2021}{}}%
+\bibitem[\citeproctext]{ref-florida2021}
 Florida, Richard, Andrés Rodríguez-Pose, and Michael Storper. 2021.
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-ghosh2020}{}}%
+\bibitem[\citeproctext]{ref-ghosh2020}
 Ghosh, Somenath, Pallabi Seth, and Harsha Tiwary. 2020. {``How Does
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-ginsburg2022}{}}%
+\bibitem[\citeproctext]{ref-ginsburg2022}
 Ginsburg, Carren, Mark A. Collinson, F. Xavier Gómez-Olivé, Sadson
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-Glaeser20}{}}%
+\bibitem[\citeproctext]{ref-Glaeser20}
 Glaeser, Edward L. 2020. {``Cities and Pandemics Have a Long History.
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-gonzuxe1lezollino2006}{}}%
+\bibitem[\citeproctext]{ref-gonzuxe1lezollino2006}
 González Ollino, Daniela, and Jorge Rodríguez Vignoli. 2006. {``Spatial
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-gonzuxe1lez-leonardo2022b}{}}%
+\bibitem[\citeproctext]{ref-gonzuxe1lez-leonardo2022b}
 González-Leonardo, Miguel, Antonio López-Gay, Niall Newsham, Joaquín
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-gonzuxe1lez-leonardo2023}{}}%
+\bibitem[\citeproctext]{ref-gonzuxe1lez-leonardo2023}
 González-Leonardo, Miguel, Michaela Potančoková, Dilek Yildiz, and
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-gonzuxe1lez-leonardo2022a}{}}%
+\bibitem[\citeproctext]{ref-gonzuxe1lez-leonardo2022a}
 González-Leonardo, Miguel, Francisco Rowe, and Alberto
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-gonzuxe1lez-leonardo2022}{}}%
+\bibitem[\citeproctext]{ref-gonzuxe1lez-leonardo2022}
 González-Leonardo, Miguel, and Jeroen Spijker. 2022. {``The Impact of
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-graizbord2007}{}}%
+\bibitem[\citeproctext]{ref-graizbord2007}
 Graizbord, Boris, and Beatriz Acuña. 2007. {``Movilidad Residencial En
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-green2021}{}}%
+\bibitem[\citeproctext]{ref-green2021}
 Green, Mark, Frances Darlington Pollock, and Francisco Rowe. 2021.
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-haslag2021}{}}%
+\bibitem[\citeproctext]{ref-haslag2021}
 Haslag, Peter H., and Daniel Weagley. 2021. {``From L.A. To Boise: How
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-huremovic2019}{}}%
+\bibitem[\citeproctext]{ref-huremovic2019}
 Huremović, Damir. 2019. {``Brief History of Pandemics (Pandemics
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-internat2021}{}}%
+\bibitem[\citeproctext]{ref-internat2021}
 {``International Migration 2020: Highlights.''} 2021, January.
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-irudayarajan2020}{}}%
+\bibitem[\citeproctext]{ref-irudayarajan2020}
 Irudaya Rajan, S., P. Sivakumar, and Aditya Srinivasan. 2020. {``The
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-janoschka2002}{}}%
+\bibitem[\citeproctext]{ref-janoschka2002}
 Janoschka, Michael. 2002. {``El Nuevo Modelo de La Ciudad
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-kotsubo2022}{}}%
+\bibitem[\citeproctext]{ref-kotsubo2022}
 Kotsubo, Masaki, and Tomoki Nakaya. 2022. {``Trends in Internal
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-lattes95}{}}%
+\bibitem[\citeproctext]{ref-lattes95}
 Lattes, A. 1995. {``Urbanización, Crecimiento Urbano y Migraciones En
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-lattes2017}{}}%
+\bibitem[\citeproctext]{ref-lattes2017}
 Lattes, Alfredo E., Jorge Rodríguez, and Miguel Villa. 2017.
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-lucchini2023}{}}%
+\bibitem[\citeproctext]{ref-lucchini2023}
 Lucchini, Lorenzo, Ollin Langle-Chimal, Lorenzo Candeago, Lucio Melito,
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-Maas19}{}}%
+\bibitem[\citeproctext]{ref-Maas19}
 Maas, P., S. Iyer, A. Gros, W. Park, L. McGorman, C. Nayak, and P. A.
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-McKinsey21}{}}%
+\bibitem[\citeproctext]{ref-McKinsey21}
 McKinsey \& Company. 2021. {``What Executives Are Saying about the
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-ILO2020}{}}%
+\bibitem[\citeproctext]{ref-ILO2020}
 Nair, Shabari, and Divya Verma. 2020. {``A Policy Framework for India's
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-nathan2020}{}}%
+\bibitem[\citeproctext]{ref-nathan2020}
 Nathan, Max, and Henry Overman. 2020. {``Will Coronavirus Cause a Big
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-UNpopulation19}{}}%
+\bibitem[\citeproctext]{ref-UNpopulation19}
 Nations", "United. 2019. {``World Urbanization Prospects. The 2018
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-nouvellet2021}{}}%
+\bibitem[\citeproctext]{ref-nouvellet2021}
 Nouvellet, Pierre, Sangeeta Bhatia, Anne Cori, Kylie E. C. Ainslie, Marc
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-oecdreg}{}}%
+\bibitem[\citeproctext]{ref-oecdreg}
 \emph{OECD Regions and Cities at a Glance}. n.d. OECD.
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-oecdsoc}{}}%
+\bibitem[\citeproctext]{ref-oecdsoc}
 {``OECD Social, Employment and Migration Working Papers.''} n.d.
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-perales2022}{}}%
+\bibitem[\citeproctext]{ref-perales2022}
 Perales, Francisco, and Aude Bernard. 2022. {``Continuity or Change? How
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-Puxe9rez-Campuzano13}{}}%
+\bibitem[\citeproctext]{ref-Puxe9rez-Campuzano13}
 Pérez-Campuzano, C., E. y Santos-Cerquera. 2013. {``Tendencias Recientes
@@
 53--88.
-\url{https://www.scielo.org.mx/scielo.php?script=sci_arttext\&pid=S1405-74252013000200003}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-Pinilla2021}{}}%
+\url{https://www.scielo.org.mx/scielo.php?script=sci_arttext&pid=S1405-74252013000200003}.
+
+\bibitem[\citeproctext]{ref-Pinilla2021}
 Pinilla, Vicente, and Luis Antonio Sáez. 2021. {``La Despoblación Rural
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-PintoDaCunha02}{}}%
+\bibitem[\citeproctext]{ref-PintoDaCunha02}
 Pinto da Cunha, J. M. P. 2002. {``Urbanización, Redistribución Espacial
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-Pomeroy2021}{}}%
+\bibitem[\citeproctext]{ref-Pomeroy2021}
 Pomeroy, Robin, and Ross Chiney. 2021. {``Has COVID Killed Our Cities?
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-ramani2021}{}}%
+\bibitem[\citeproctext]{ref-ramani2021}
 Ramani, Arjun, and Nicholas Bloom. 2021. {``The Donut Effect of Covid-19
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-rodruxedguezvignoli2017}{}}%
+\bibitem[\citeproctext]{ref-rodruxedguezvignoli2017}
 Rodríguez Vignoli, Jorge, and Francisco Rowe. 2017. {``¿Contribuye La
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-rodruxedguez-vignoli2018}{}}%
+\bibitem[\citeproctext]{ref-rodruxedguez-vignoli2018}
 Rodríguez-Vignoli, Jorge, and Francisco Rowe. 2018. {``How Is Internal
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-rowe2021}{}}%
+\bibitem[\citeproctext]{ref-rowe2021}
 Rowe, Francisco. 2021. {``Big Data and Human Geography.''}
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-rowe2022}{}}%
+\bibitem[\citeproctext]{ref-rowe2022}
 Rowe, Francisco, Alessia Calafiore, Daniel Arribas-Bel, Krasen
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-rowe2023}{}}%
+\bibitem[\citeproctext]{ref-rowe2023}
 Rowe, Francisco, Miguel González-Leonardo, and Tony Champion. 2023.
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-OIT2018}{}}%
+\bibitem[\citeproctext]{ref-OIT2018}
 Salazar-Xirinachs, Juan, José Manuel \& Chacaltana. 2018. {``Políticas
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-sobrino12}{}}%
+\bibitem[\citeproctext]{ref-sobrino12}
 Sobrino, J. 2012. {``La Urbanización En El México Contemporáneo.''}
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-sobrino2006}{}}%
+\bibitem[\citeproctext]{ref-sobrino2006}
 Sobrino, Jaime. 2006. {``Comptetitiveness and Employment in the Largest
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-stawarz2022}{}}%
+\bibitem[\citeproctext]{ref-stawarz2022}
 Stawarz, Nico, Matthias Rosenbaum-Feldbrügge, Nikola Sander, Harun
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-storper2004}{}}%
+\bibitem[\citeproctext]{ref-storper2004}
 Storper, M., and A. J. Venables. 2004. {``Buzz: Face-to-Face Contact and
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-TheGuardian2020}{}}%
+\bibitem[\citeproctext]{ref-TheGuardian2020}
 The Guardian. 2020. {``Escape to the Country: How Covid Is Driving an
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-tuxf8nnessen2021}{}}%
+\bibitem[\citeproctext]{ref-tuxf8nnessen2021}
 Tønnessen, Marianne. 2021. {``Movers from the City in the First Year of
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-USBureauLabor2021}{}}%
+\bibitem[\citeproctext]{ref-USBureauLabor2021}
 US Bureau of Labor Statistics. 2021. {``Unemployment Rises in 2020, as
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-vogiazides2022}{}}%
+\bibitem[\citeproctext]{ref-vogiazides2022}
 Vogiazides, Louisa, and Juta Kawalerowicz. 2022. {``Internal Migration
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-wang2022}{}}%
+\bibitem[\citeproctext]{ref-wang2022}
 Wang, Yikang, Chen Zhong, Qili Gao, and Carmen Cabrera-Arnau. 2022.
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-WorldBank2023}{}}%
+\bibitem[\citeproctext]{ref-WorldBank2023}
 World Bank", "The. 2023. {``LAC Equity Lab: Poverty - Poverty Rate.''}
@@
 \bibliographystyle{apalike}
-\bibliography{sim_refs}
+\bibliography{refs}
 

### 2024-01-17 manuscript-3-old/manuscript-latin-countries.qmd
compiling-manuscript

--- 2023-10-02 manuscript-old/manuscript-latin-countries.qmd
@@
 
-![Maps of population density categories by country with histograms, showing that the distribution of Bing tiles in each density category.](figures/Map_density_histogram.png){fig-align="center" width="625" height="640"}
+![Maps of population density categories by country with histograms, showing that the distribution of Bing tiles in each density category.](images/Map_density_histogram.png){fig-align="center" width="625" height="640"}
 

### 2024-01-17 manuscript-3-old/manuscript-latin-countries.tex
compiling-manuscript

--- 2024-01-17 manuscript-latin-countries-old/manuscript-latin-countries.tex
@@
 downturn caused by the pandemic constrained international migration
-({``International Migration 2020: Highlights''} 2021; González-Leonardo
-and Spijker 2022; González-Leonardo et al. 2023) and population
-movements within countries (Nouvellet et al. 2021; González-Leonardo,
-Rowe, and Fresolone-Caparrós 2022; Wang et al. 2022; Rowe,
-González-Leonardo, and Champion 2023). During early stages of the
+(\textbf{internat2021?}; \textbf{gonzález-leonardo2022?};
+\textbf{gonzález-leonardo2023?}) and population movements within
+countries (\textbf{nouvellet2021?}; \textbf{gonzález-leonardo2022a?};
+Wang et al. 2022; \textbf{rowe2023?}). During early stages of the
 pandemic, big cities became the main epicentres of COVID-19 infections
-and deaths (Pomeroy and Chiney 2021) due to their higher connectivity
-via air travel, higher population density and concentration of
-public-facing jobs (Brandén et al. 2020; Bhadra, Mukherjee, and Sarkar
-2020). As a result, governments implemented non-pharmaceutical
-interventions, such as lockdowns and business closures, leading to
-increases in unemployment and removing the urban vibrancy (Blustein et
-al. 2020; BBC news 2023). In addition, remote work and online teaching
+and deaths (\textbf{Pomeroy2021?}) due to their higher connectivity via
+air travel, higher population density and concentration of public-facing
+jobs (\textbf{brandén2020?}; \textbf{bhadra2020?}). As a result,
+governments implemented non-pharmaceutical interventions, such as
+lockdowns and business closures, leading to increases in unemployment
+and removing the urban vibrancy (\textbf{blustein2020?};
+\textbf{BBCnews2023?}). In addition, remote work and online teaching
 reduced the need for commuting, living close to work and education
@@
 spaces to move out of cities and look for more affordable and
-comfortable housing (Nathan and Overman 2020; US Bureau of Labor
-Statistics 2021). These changes are believed to have reduced the
-attractiveness of cities during the pandemic, while suburbs and rural
-areas have gained popularity (Florida, Rodríguez-Pose, and Storper
-2021). As a result, a number of headlines have claimed the emergence of
-an `urban exodus' to areas with lower population densities (e.g. (The
-Guardian 2020)).
+comfortable housing (\textbf{nathan2020?}; \textbf{USBureauLabor2021?}).
+These changes are believed to have reduced the attractiveness of cities
+during the pandemic, while suburbs and rural areas have gained
+popularity (\textbf{florida2021?}). As a result, a number of headlines
+have claimed the emergence of an `urban exodus' to areas with lower
+population densities (e.g. (\textbf{TheGuardian2020?})).
 
@@
 lack infrastructure and services, such as schooling and health, to
-effectively accommodate a large population influx (Pinilla and Sáez
-2021; \emph{OECD Regions and Cities at a Glance}, n.d.). They generally
-lack high-speed Internet connection which is vital for teleworking (Chen
-and Wellman 2004). Additionally, most jobs cannot be done remotely
-{``OECD Social, Employment and Migration Working Papers''} (n.d.), and
-some companies have returned to in-person office work or have
-implemented hybrid forms of work, following the relaxation of stringency
-measures (McKinsey \& Company 2021). Schools and universities have also
-returned to face-to-face teaching. Therefore, living close to work or
-study places remains relevant. Businesses and leisure activities in
-cities have gradually returned to normal, reinvigorating the economic
-functioning and urban life. Collectively, these arguments suggest that a
-potential increase in counterurbanisation movements during the pandemic
-was concentrated during periods of stringency measures and is likely to
-be short term. Big cities have successfully weathered previous pandemics
-(Huremović 2019; Glaeser 2020) and are likely to remain attractive
-places to live. Thanks to agglomeration economies, cities are normally
-regarded as centres for economic growth, clusters of talent, consumer
-bases and spaces for face‐to‐face interaction and diversity (Storper and
-Venables 2004; Florida, Rodríguez-Pose, and Storper 2021).
+effectively accommodate a large population influx
+(\textbf{Pinilla2021?}; \textbf{oecdreg?}). They generally lack
+high-speed Internet connection which is vital for teleworking
+(\textbf{Chen2004?}). Additionally, most jobs cannot be done remotely
+(\textbf{oecdsoc?}), and some companies have returned to in-person
+office work or have implemented hybrid forms of work, following the
+relaxation of stringency measures (\textbf{McKinsey21?}). Schools and
+universities have also returned to face-to-face teaching. Therefore,
+living close to work or study places remains relevant. Businesses and
+leisure activities in cities have gradually returned to normal,
+reinvigorating the economic functioning and urban life. Collectively,
+these arguments suggest that a potential increase in counterurbanisation
+movements during the pandemic was concentrated during periods of
+stringency measures and is likely to be short term. Big cities have
+successfully weathered previous pandemics (\textbf{huremovic2019?};
+\textbf{Glaeser20?}) and are likely to remain attractive places to live.
+Thanks to agglomeration economies, cities are normally regarded as
+centres for economic growth, clusters of talent, consumer bases and
+spaces for face‐to‐face interaction and diversity
+(\textbf{storper2004?}; \textbf{florida2021?}).
 
@@
 movements across some countries of the Global North: the United States
-(Ramani and Bloom 2021), United Kingdom (Rowe, González-Leonardo, and
-Champion 2023; Wang et al. 2022), Spain (González-Leonardo et al. 2022;
-González-Leonardo, Rowe, and Fresolone-Caparrós 2022), Germany {[}@;
-Stawarz et al. (2022){]}, Sweden (Vogiazides and Kawalerowicz 2022),
-Norway (Tønnessen 2021), Australia (Perales and Bernard 2022) and Japan
-Kotsubo and Nakaya (2022). The majority of movements away from cities
-headed to their suburbs (Vogiazides and Kawalerowicz 2022; Stawarz et
-al. 2022) or rural areas in their close proximity (González-Leonardo et
-al. 2022; González-Leonardo, Rowe, and Fresolone-Caparrós 2022) showing
-in some instances a ``doghnut effect'', with a decline in population
-movements directed to cities and a rise in the inflow of people to less
-dense surrounding areas (Ramani and Bloom 2021). Studies demonstrated
-that professionals (Tønnessen 2021) and individuals with a high income
-(Haslag and Weagley 2021), those who are potentially able to practice
-remote working, underpinned movements away from large cities. In
-addition, research has shown that a large share of counterurbanisation
-movements is headed to touristic locations with a high concentration of
-second homes, suggesting that middle and high class individuals may have
-been key actors in movements from cities to less populated areas during
-the pandemic (González-Leonardo, Rowe, and Fresolone-Caparrós 2022).
-However, most studies suggested that variations in internal population
-movements during the pandemic are likely to be temporary and have not
-altered previous macro-structures of human mobility across the
-rural-urban hierarchy (Rowe, González-Leonardo, and Champion 2023).
+(\textbf{ramani2021?}), United Kingdom (\textbf{rowe2023?}; Wang et al.
+2022), Spain (\textbf{gonzález-leonardo2022b?};
+\textbf{gonzález-leonardo2022a?}), Germany {[}@;
+(\textbf{stawarz2022?}){]}, Sweden (\textbf{vogiazides2022?}), Norway
+(\textbf{tønnessen2021?}), Australia (\textbf{perales2022?}) and Japan
+(\textbf{kotsubo2022?}). The majority of movements away from cities
+headed to their suburbs (\textbf{vogiazides2022?};
+\textbf{stawarz2022?}) or rural areas in their close proximity
+(\textbf{gonzález-leonardo2022b?}; \textbf{gonzález-leonardo2022a?})
+showing in some instances a ``doghnut effect'', with a decline in
+population movements directed to cities and a rise in the inflow of
+people to less dense surrounding areas (\textbf{ramani2021?}). Studies
+demonstrated that professionals (\textbf{tønnessen2021?}) and
+individuals with a high income (\textbf{haslag2021?}), those who are
+potentially able to practice remote working, underpinned movements away
+from large cities. In addition, research has shown that a large share of
+counterurbanisation movements is headed to touristic locations with a
+high concentration of second homes, suggesting that middle and high
+class individuals may have been key actors in movements from cities to
+less populated areas during the pandemic
+(\textbf{gonzález-leonardo2022a?}). However, most studies suggested that
+variations in internal population movements during the pandemic are
+likely to be temporary and have not altered previous macro-structures of
+human mobility across the rural-urban hierarchy (\textbf{rowe2023?}).
 
 In Latin America and the Caribbean, a reduction in internal population
-movements was documented during periods of high stringency (Aromí et al.
-2023), but there is no evidence on how human mobility has changed across
-the rural-urban continuum. Anecdotal reports have shown that movements
-away from cities seem to have accelerated in two countries of the Global
-South, India (Irudaya Rajan, Sivakumar, and Srinivasan 2020) and South
-Africa (Ginsburg et al. 2022), while urbanisation trends have
+movements was documented during periods of high stringency
+(\textbf{aromí2023?}), but there is no evidence on how human mobility
+has changed across the rural-urban continuum. Anecdotal reports have
+shown that movements away from cities seem to have accelerated in two
+countries of the Global South, India (\textbf{irudayarajan2020?}) and
+South Africa (\textbf{ginsburg2022?}), while urbanisation trends have
 decelerated. However, lack of suitable up-to-date data has prevented us
@@
 underpinned by internal migrants residing in cities who returned to
-their hometowns due to a rise in unemployment (Irudaya Rajan, Sivakumar,
-and Srinivasan 2020; Ginsburg et al. 2022). However, recent studies have
-shown that urban residents from wealthy neighborhoods were more likely
-to move to rural areas than those from low income neighborhoods during
-the first wave of the pandemic in Brazil, Colombia, Mexico, Indonesia,
-Philippines and South Africa (Lucchini et al. 2023).
+their hometowns due to a rise in unemployment
+(\textbf{irudayarajan2020?}; \textbf{ginsburg2022?}). However, recent
+studies have shown that urban residents from wealthy neighborhoods were
+more likely to move to rural areas than those from low income
+neighborhoods during the first wave of the pandemic in Brazil, Colombia,
+Mexico, Indonesia, Philippines and South Africa
+(\textbf{lucchini2023?}).
 
 It is known that human mobility within countries tends to vary in
-systematic ways with development levels (Bell et al. 2015). In Latin
+systematic ways with development levels (\textbf{bell2015?}). In Latin
 America and the Caribbean, 28.6\% of the population was living on less
-than \$6.8 per day in 2020 (World Bank" 2023) and the informality rate
-was about 53\% in 2016 (Salazar-Xirinachs 2018). Thus, we could expect
+than \$6.8 per day in 2020 (\textbf{WorldBank2023?}) and the informality
+rate was about 53\% in 2016 (\textbf{OIT2018?}). Thus, we could expect
 different pandemic impacts on the patterns of internal population
@@
 Traditionally, censuses have been used to analyse human mobility
-patterns within Global South countries, which lack register data (Bell
-et al. 2015; Bernard et al. 2017). Censuses, however, are normally
-updated every 10 years, lacking the temporal granularity to explore
-population movements over yearly, monthly or short-time intervals.
+patterns within Global South countries, which lack register data
+(\textbf{bell2015?}; \textbf{bernard2017?}). Censuses, however, are
+normally updated every 10 years, lacking the temporal granularity to
+explore population movements over yearly, monthly or short-time
+intervals.
 
@@
 with small spatial and temporal granularity in multiple countries
-(Green, Pollock, and Rowe 2021; Rowe 2021). Drawing on geographically
+(\textbf{green2021?}; \textbf{rowe2021?}). Drawing on geographically
 granular Meta-Facebook data from March 2020 to March 2022, we aim to
@@
 The COVID-19 pandemic led to an overall decline in human mobility within
-countries (Nouvellet et al. 2021). Research has shown that internal
+countries (\textbf{nouvellet2021?}). Research has shown that internal
 population movements across the rural-urban continuum decreased in the
-United States (Ramani and Bloom 2021), several European countries, Japan
-and Austria during 2020 (Rowe, González-Leonardo, and Champion 2023).
-Declines ranged from 2.5\% in Spain to 8.5\% in Australia. The largest
-drops occurred during periods of national lockdowns and the
-implementation of other stringency measures, such as mobility
-restrictions, work and school closing, while levels of internal
-population movements recovered or exceeded pre-pandemic figures when
-lockdowns were lifted. Declines were partially attributed to a reduction
-in involuntary migration during lockdowns, a loss of labour market
-dynamism due to the pandemic, fewer people changing jobs and increasing
-remote work (Perales and Bernard 2022).
+United States (\textbf{ramani2021?}), several European countries, Japan
+and Austria during 2020 (\textbf{rowe2023?}). Declines ranged from 2.5\%
+in Spain to 8.5\% in Australia. The largest drops occurred during
+periods of national lockdowns and the implementation of other stringency
+measures, such as mobility restrictions, work and school closing, while
+levels of internal population movements recovered or exceeded
+pre-pandemic figures when lockdowns were lifted. Declines were partially
+attributed to a reduction in involuntary migration during lockdowns, a
+loss of labour market dynamism due to the pandemic, fewer people
+changing jobs and increasing remote work (\textbf{perales2022?}).
 
@@
 households and businesses, shifted from central cities to suburbs and
-exurbs during 2020 (Ramani and Bloom 2021). These authors label this
+exurbs during 2020 (\textbf{ramani2021?}). These authors label this
 trend as a ``doughnut effect'', reflecting a decrease in population
@@
 effect in medium-sized cities and almost no effect for small cities. A
-similar trend was observed in Norway (Tønnessen 2021), Germany (Stawarz
-et al. 2022), Sweden (Vogiazides and Kawalerowicz 2022) and Japan
-(Fielding and Ishikawa 2021; Kotsubo and Nakaya 2022), where net
-migration rates in big cities declined, while population movement to
-their suburbs increased. In addition, this trend was accompanied by a
-deceleration of urbanisation trends and unusual population gains in
-rural areas due to an acceleration in counterurbanisation movements,
-mostly in rural locations which are known to be holiday destinations
-with a high concentration of second homes and are generally located in
-the close proximity to big cities. An increase of counterurbanisation
-was the main pandemic outcome on human mobility in Spain
-(González-Leonardo et al. 2022; González-Leonardo, Rowe, and
-Fresolone-Caparrós 2022), the United Kingdom (Wang et al. 2022; Rowe et
-al. 2022) and Australia (Perales and Bernard 2022), where there is no
-evidence of a ``doghnut effect'', since substantial variations in
-suburbs were not found across these countries.
+similar trend was observed in Norway (\textbf{tønnessen2021?}), Germany
+(\textbf{stawarz2022?}), Sweden (\textbf{vogiazides2022?}) and Japan
+(\textbf{fielding2021?}; \textbf{kotsubo2022?}), where net migration
+rates in big cities declined, while population movement to their suburbs
+increased. In addition, this trend was accompanied by a deceleration of
+urbanisation trends and unusual population gains in rural areas due to
+an acceleration in counterurbanisation movements, mostly in rural
+locations which are known to be holiday destinations with a high
+concentration of second homes and are generally located in the close
+proximity to big cities. An increase of counterurbanisation was the main
+pandemic outcome on human mobility in Spain
+(\textbf{gonzález-leonardo2022b?}; \textbf{gonzález-leonardo2022a?}),
+the United Kingdom (Wang et al. 2022; Rowe et al. 2022) and Australia
+(\textbf{perales2022?}), where there is no evidence of a ``doghnut
+effect'', since substantial variations in suburbs were not found across
+these countries.
 
 However, studies suggest that variations are likely to be temporary
-(Rowe, González-Leonardo, and Champion 2023). In Spain, the urbanisation
-movement returned to pre-pandemic levels when the lockdown ended in
-mid-June 2020 (González-Leonardo et al. 2022) and unusually high levels
-of counterurbanisation persisted over 2021, although the majority of
+(\textbf{rowe2023?}). In Spain, the urbanisation movement returned to
+pre-pandemic levels when the lockdown ended in mid-June 2020
+(\textbf{gonzález-leonardo2022b?}) and unusually high levels of
+counterurbanisation persisted over 2021, although the majority of
 movements continued to occur between and within urban areas
-(González-Leonardo, Rowe, and Fresolone-Caparrós 2022). In Australia,
-the pandemic has caused minor changes in spatial patterns of internal
-migration, and its effects were minimal by the end of 2020 (Perales and
-Bernard 2022). In the United Kingdom, mobility patterns returned to
-those registered prior to the pandemic after the easing of
-non-pharmaceutical interventions (Rowe et al. 2022; Wang et al. 2022).
-These findings suggest that COVID-19 generated shock waves leading to
-temporary changes in the patterns of internal population movement across
-the rural-urban continuum, but it has not significantly altered the
-prevalent macro-structures of population movement within countries.
+(\textbf{gonzález-leonardo2022a?}). In Australia, the pandemic has
+caused minor changes in spatial patterns of internal migration, and its
+effects were minimal by the end of 2020 (\textbf{perales2022?}). In the
+United Kingdom, mobility patterns returned to those registered prior to
+the pandemic after the easing of non-pharmaceutical interventions (Rowe
+et al. 2022; Wang et al. 2022). These findings suggest that COVID-19
+generated shock waves leading to temporary changes in the patterns of
+internal population movement across the rural-urban continuum, but it
+has not significantly altered the prevalent macro-structures of
+population movement within countries.
 
@@
 Bolivia, Ecuador and Argentina to less than 3\% in Paraguay and
-Venezuela (Aromí et al. 2023). However, there is no evidence about
+Venezuela (\textbf{aromí2023?}). However, there is no evidence about
 pandemic outcomes on human mobility across the rural-urban continuum. To
-date, surveys carried out in India (Irudaya Rajan, Sivakumar, and
-Srinivasan 2020) and South Africa (Ginsburg et al. 2022) have shown
-anecdotal evidence suggesting that urbanisation trends declined during
-the pandemic, while movements from cities to rural areas increased. Both
-studies pointed out that a number of rural residents initiating a
-migration decreased during 2020 and some labourers in major cities seem
-to have returned to their hometowns due to job losses, since stringency
-measures have extensively impacted the functioning of the economy
-(Ghosh, Seth, and Tiwary 2020). Many internal migrants residing in
-Indian cities lost their jobs and were unable to afford basic
-necessities (Nair and Verma 2020). Collectively, these findings suggest
-that vulnerable populations seem to have played a role in increasing
-counterurbanisation movements in Global South countries. However, recent
-work has shown that high-wealth groups residing in cities across Mexico,
-Colombia, Brazil, Indonesia, Philippines and South Africa were on
-average 159\% more likely to have moved to rural area compared to those
-of low-wealth groups during early stages of the pandemic (Lucchini et
-al. 2023), in line with the evidencee found for Global North countries.
+date, surveys carried out in India (\textbf{irudayarajan2020?}) and
+South Africa (\textbf{ginsburg2022?}) have shown anecdotal evidence
+suggesting that urbanisation trends declined during the pandemic, while
+movements from cities to rural areas increased. Both studies pointed out
+that a number of rural residents initiating a migration decreased during
+2020 and some labourers in major cities seem to have returned to their
+hometowns due to job losses, since stringency measures have extensively
+impacted the functioning of the economy (\textbf{ghosh2020?}). Many
+internal migrants residing in Indian cities lost their jobs and were
+unable to afford basic necessities (\textbf{ILO2020?}). Collectively,
+these findings suggest that vulnerable populations seem to have played a
+role in increasing counterurbanisation movements in Global South
+countries. However, recent work has shown that high-wealth groups
+residing in cities across Mexico, Colombia, Brazil, Indonesia,
+Philippines and South Africa were on average 159\% more likely to have
+moved to rural area compared to those of low-wealth groups during early
+stages of the pandemic (\textbf{lucchini2023?}), in line with the
+evidencee found for Global North countries.
 
@@
 between the 1950s and 1970s when a critical mass of individuals moved
-from villages and towns to urban areas (Firebaugh 1979; A. Lattes 1995;
-J. Sobrino 2012). Urban growth due to internal mobility was primarily
-concentrated in large urban centres resulting in a significant
-population imbalance between the chief cities and other settlements
-(Pinto da Cunha 2002; A. E. Lattes, Rodríguez, and Villa 2017). As a
+from villages and towns to urban areas (\textbf{firebaugh1979?};
+\textbf{lattes95?}; \textbf{sobrino12?}). Urban growth due to internal
+mobility was primarily concentrated in large urban centres resulting in
+a significant population imbalance between the chief cities and other
+settlements (\textbf{PintoDaCunha02?}; \textbf{lattes2017?}). As a
 consequence, Latin America shows high rates of urbanisation with more
@@
 North America, and about 46\% of urban residents settled in cities with
-more than 1 million inhabitants (Nations" 2019).
+more than 1 million inhabitants (\textbf{UNpopulation19?}).
 
@@
 migratory system in Latin America, while rural to urban flows are of
-less importance (Bernard et al. 2017; Rodríguez-Vignoli and Rowe 2018;
-Nations" 2019). According to the 2010 census round, 80\% of internal
-migrants moved between cities (Rodríguez-Vignoli and Rowe 2018).
-Contrary to the period preceding the 1980s, medium-sizes cities (those
-between 500K and 1 million inhabitants) experienced the highest net
-migration rates, while large cities (\textgreater1 million) presented
-balanced rates and small cities (\textgreater500K) lost individuals by
-internal population movements (Rodríguez-Vignoli and Rowe 2018).
-Changing patterns in human mobility within Latin American countries were
-driven by a decline in internal migration since the 1980s crisis (Chávez
-Galindo et al. 2016) and deconcentration dynamics in large urban
-centres, such as Mexico City (Jaime Sobrino 2006) or Santiago de Chile
-(González Ollino and Rodríguez Vignoli 2006), where long distance
-inflows have declined over time. At the same time, there has been an
-increasing domestic and foreign investment in export-oriented activities
-or tourism industries in certain middle-sized cities, fostering the
-geographic dispersal of employment opportunities and increasing the
-attractiveness of these cities (Brea 2003; Pérez-Campuzano 2013; Chávez
-Galindo et al. 2016).
+less importance (\textbf{bernard2017?}; \textbf{rodríguez-vignoli2018?};
+\textbf{UNpopulation19?}). According to the 2010 census round, 80\% of
+internal migrants moved between cities
+(\textbf{rodríguez-vignoli2018?}). Contrary to the period preceding the
+1980s, medium-sizes cities (those between 500K and 1 million
+inhabitants) experienced the highest net migration rates, while large
+cities (\textgreater1 million) presented balanced rates and small cities
+(\textgreater500K) lost individuals by internal population movements
+(\textbf{rodríguez-vignoli2018?}). Changing patterns in human mobility
+within Latin American countries were driven by a decline in internal
+migration since the 1980s crisis (\textbf{chávezgalindo2016?}) and
+deconcentration dynamics in large urban centres, such as Mexico City
+(\textbf{sobrino2006?}) or Santiago de Chile
+(\textbf{gonzálezollino2006?}), where long distance inflows have
+declined over time. At the same time, there has been an increasing
+domestic and foreign investment in export-oriented activities or tourism
+industries in certain middle-sized cities, fostering the geographic
+dispersal of employment opportunities and increasing the attractiveness
+of these cities (\textbf{Brea03?}; \textbf{Pérez-Campuzano13?};
+\textbf{chávezgalindo2016?}).
 
@@
 movement and has increased over time due to the growing expansion of
-Latin American cities (Graizbord and Acuña 2007; Chávez Galindo et al.
-2016). These movements are underpinned by flows of middle- and
-upper-class families moving away from central cities to surrounding
-auto-segregated areas of wealthy individuals (Borsdorf 2003; Rodríguez
-Vignoli and Rowe 2017). Low-income populations also settle in specific
-areas within suburbs where the cost of living is usually cheaper than in
-cities (Janoschka 2002; Rodríguez Vignoli and Rowe 2017). A large
-proportion of both wealthy individuals and low-income residents commute
-daily to central cities, mainly for work reasons (Chávez Galindo et al.
-2016). Recently, reurbanisation dynamics have also been observed due to
-gentrification processes in specific central areas (J. Sobrino 2012;
-Chávez Galindo et al. 2016). In this paper, we analyse how the COVID-19
+Latin American cities (\textbf{graizbord2007?};
+\textbf{chávezgalindo2016?}). These movements are underpinned by flows
+of middle- and upper-class families moving away from central cities to
+surrounding auto-segregated areas of wealthy individuals
+(\textbf{borsdorf2003?}; \textbf{rodríguezvignoli2017?}). Low-income
+populations also settle in specific areas within suburbs where the cost
+of living is usually cheaper than in cities (\textbf{janoschka2002?};
+\textbf{rodríguezvignoli2017?}). A large proportion of both wealthy
+individuals and low-income residents commute daily to central cities,
+mainly for work reasons (\textbf{chávezgalindo2016?}). Recently,
+reurbanisation dynamics have also been observed due to gentrification
+processes in specific central areas (\textbf{sobrino12?};
+\textbf{chávezgalindo2016?}). In this paper, we analyse how the COVID-19
 pandemic has changed patterns of internal population movements across
@@
 
-\begin{figure}
+\begin{figure}[H]
 
@@
 
-\hypertarget{refs}{}
+\phantomsection\label{refs}
 \begin{CSLReferences}{1}{0}
-\leavevmode\vadjust pre{\hypertarget{ref-aromuxed2023}{}}%
-Aromí, Daniel, María Paula Bonel, Julian Cristia, Martín Llada, Juan
-Pereira, Xiomara Pulido, and Julieth Santamaria. 2023. {``\#StayAtHome:
-Social Distancing Policies and Mobility in Latin America and the
-Caribbean.''} \emph{Economía} 22 (1).
-\url{https://doi.org/10.31389/eco.4}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-BBCnews2023}{}}%
-BBC news. 2023. {``Unemployment: Who Are the Millions of Britons Not
-Working?''} \url{https://www.bbc.co.uk/news/business-52660591}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-bell2015}{}}%
-Bell, Martin, Elin Charles-Edwards, Philipp Ueffing, John Stillwell,
-Marek Kupiszewski, and Dorota Kupiszewska. 2015. {``Internal Migration
-and Development: Comparing Migration Intensities Around the World.''}
-\emph{Population and Development Review} 41 (1): 33--58.
-\url{https://doi.org/10.1111/j.1728-4457.2015.00025.x}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-bernard2017}{}}%
-Bernard, Aude, Francisco Rowe, Martin Bell, Philipp Ueffing, and Elin
-Charles-Edwards. 2017. {``Comparing Internal Migration Across the
-Countries of Latin America: A Multidimensional Approach.''} Edited by
-Osman Alimamy Sankoh. \emph{PLOS ONE} 12 (3): e0173895.
-\url{https://doi.org/10.1371/journal.pone.0173895}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-bhadra2020}{}}%
-Bhadra, Arunava, Arindam Mukherjee, and Kabita Sarkar. 2020. {``Impact
-of Population Density on Covid-19 Infected and Mortality Rate in
-India.''} \emph{Modeling Earth Systems and Environment} 7 (1): 623--29.
-\url{https://doi.org/10.1007/s40808-020-00984-7}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-blustein2020}{}}%
-Blustein, David L., Ryan Duffy, Joaquim A. Ferreira, Valerie
-Cohen-Scali, Rachel Gali Cinamon, and Blake A. Allan. 2020.
-{``Unemployment in the Time of COVID-19: A Research Agenda.''}
-\emph{Journal of Vocational Behavior} 119 (June): 103436.
-\url{https://doi.org/10.1016/j.jvb.2020.103436}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-borsdorf2003}{}}%
-Borsdorf, Axel. 2003. {``Cómo Modelar El Desarrollo y La Dinámica de La
-Ciudad Latinoamericana.''} \emph{EURE (Santiago)} 29 (86).
-\url{https://doi.org/10.4067/s0250-71612003008600002}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-branduxe9n2020}{}}%
-Brandén, Maria, Siddartha Aradhya, Martin Kolk, Juho Härkönen, Sven
-Drefahl, Bo Malmberg, Mikael Rostila, Agneta Cederström, Gunnar
-Andersson, and Eleonora Mussino. 2020. {``Residential Context and
-COVID-19 Mortality Among Adults Aged 70 Years and Older in Stockholm: A
-Population-Based, Observational Study Using Individual-Level Data.''}
-\emph{The Lancet Healthy Longevity} 1 (2): e80--88.
-\url{https://doi.org/10.1016/s2666-7568(20)30016-7}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-Brea03}{}}%
-Brea, J. 2003. {``Population Dynamics in Latin America.''}
-\emph{Population Bulletin} 58: 3--36.
-\url{https://population.un.org/wup/Publications/Files/WUP2018-Report.pdf}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-chuxe1vezgalindo2016}{}}%
-Chávez Galindo, Ana María, Jorge Rodríguez Vignoli, Mario Acuña, Jorge
-Barquero, Daniel Macadar, José Marcos Pinto da Cunha, and Jaime Sobrino.
-2016. {``Migración Interna y Cambios Metropolitanos.''} \emph{Revista
-Latinoamericana de Población} 10 (18): 7--41.
-\url{https://doi.org/10.31406/relap2016.v10.i1.n18.1}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-Chen2004}{}}%
-Chen, Wenhong, and Barry Wellman. 2004. {``The Global Digital Divide -
-Within and Between Countries.''} \emph{IT \& Society} 1: 39--45.
-\url{http://www.ifs.tuwien.ac.at/~dieter/teaching/GmA/Chen2004.pdf}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-fielding2021}{}}%
-Fielding, Tony, and Yoshitaka Ishikawa. 2021. {``COVID{-}19 and
-Migration: A Research Note on the Effects of COVID{-}19 on Internal
-Migration Rates and Patterns in Japan.''} \emph{Population, Space and
-Place} 27 (6). \url{https://doi.org/10.1002/psp.2499}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-firebaugh1979}{}}%
-Firebaugh, Glenn. 1979. {``Structural Determinants of Urbanization in
-Asia and Latin America, 1950- 1970.''} \emph{American Sociological
-Review} 44 (2): 199. \url{https://doi.org/10.2307/2094505}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-florida2021}{}}%
-Florida, Richard, Andrés Rodríguez-Pose, and Michael Storper. 2021.
-{``Critical Commentary: Cities in a Post-COVID World.''} \emph{Urban
-Studies} 60 (8): 1509--31.
-\url{https://doi.org/10.1177/00420980211018072}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-ghosh2020}{}}%
-Ghosh, Somenath, Pallabi Seth, and Harsha Tiwary. 2020. {``How Does
-Covid-19 Aggravate the Multidimensional Vulnerability of Slums in India?
-A Commentary.''} \emph{Social Sciences \& Humanities Open} 2 (1):
-100068. \url{https://doi.org/10.1016/j.ssaho.2020.100068}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-ginsburg2022}{}}%
-Ginsburg, Carren, Mark A. Collinson, F. Xavier Gómez-Olivé, Sadson
-Harawa, Chantel F. Pheiffer, and Michael J. White. 2022. {``The Impact
-of COVID-19 on a Cohort of Origin Residents and Internal Migrants from
-South Africa's Rural Northeast.''} \emph{SSM - Population Health} 17
-(March): 101049. \url{https://doi.org/10.1016/j.ssmph.2022.101049}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-Glaeser20}{}}%
-Glaeser, Edward L. 2020. {``Cities and Pandemics Have a Long History.
-City Journal.''}
-\url{https://www.city-journal.org/article/cities-and-pandemics-have-a-long-history}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-gonzuxe1lezollino2006}{}}%
-González Ollino, Daniela, and Jorge Rodríguez Vignoli. 2006. {``Spatial
-Redistribution and Internal Migration of the Populationin Chile over the
-Past 35 Years (1965-20).''} \emph{Estudios Demográficos y Urbanos} 21
-(2): 369. \url{https://doi.org/10.24201/edu.v21i2.1253}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-gonzuxe1lez-leonardo2022b}{}}%
-González-Leonardo, Miguel, Antonio López-Gay, Niall Newsham, Joaquín
-Recaño, and Francisco Rowe. 2022. {``Understanding Patterns of Internal
-Migration During the COVID{-}19 Pandemic in Spain.''} \emph{Population,
-Space and Place} 28 (6). \url{https://doi.org/10.1002/psp.2578}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-gonzuxe1lez-leonardo2023}{}}%
-González-Leonardo, Miguel, Michaela Potančoková, Dilek Yildiz, and
-Francisco Rowe. 2023. {``Quantifying the Impact of COVID-19 on
-Immigration in Receiving High-Income Countries.''} Edited by Ricky Chee
-Jiun Chia. \emph{PLOS ONE} 18 (1): e0280324.
-\url{https://doi.org/10.1371/journal.pone.0280324}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-gonzuxe1lez-leonardo2022a}{}}%
-González-Leonardo, Miguel, Francisco Rowe, and Alberto
-Fresolone-Caparrós. 2022. {``Rural Revival? The Rise in Internal
-Migration to Rural Areas During the COVID-19 Pandemic. Who Moved and
-Where?''} \emph{Journal of Rural Studies} 96 (December): 332--42.
-\url{https://doi.org/10.1016/j.jrurstud.2022.11.006}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-gonzuxe1lez-leonardo2022}{}}%
-González-Leonardo, Miguel, and Jeroen Spijker. 2022. {``The Impact of
-Covid-19 on Demographic Components in Spain, 2020{\textendash}31: A
-Scenario Approach.''} \emph{Population Studies}, November, 1--17.
-\url{https://doi.org/10.1080/00324728.2022.2138521}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-graizbord2007}{}}%
-Graizbord, Boris, and Beatriz Acuña. 2007. {``Movilidad Residencial En
-La Ciudad de México / Residential Mobility in Mexico City.''}
-\emph{Estudios Demográficos y Urbanos} 22 (2): 291.
-\url{https://doi.org/10.24201/edu.v22i2.1281}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-green2021}{}}%
-Green, Mark, Frances Darlington Pollock, and Francisco Rowe. 2021.
-{``New Forms of Data and New Forms of Opportunities to Monitor and
-Tackle a Pandemic.''} In, 423--29. Springer International Publishing.
-\url{https://doi.org/10.1007/978-3-030-70179-6_56}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-haslag2021}{}}%
-Haslag, Peter H., and Daniel Weagley. 2021. {``From L.A. To Boise: How
-Migration Has Changed During the COVID-19 Pandemic.''} \emph{SSRN
-Electronic Journal}. \url{https://doi.org/10.2139/ssrn.3808326}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-huremovic2019}{}}%
-Huremović, Damir. 2019. {``Brief History of Pandemics (Pandemics
-Throughout History).''} In, 7--35. Springer International Publishing.
-\url{https://doi.org/10.1007/978-3-030-15346-5_2}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-internat2021}{}}%
-{``International Migration 2020: Highlights.''} 2021, January.
-\url{https://doi.org/10.18356/9789210052689}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-irudayarajan2020}{}}%
-Irudaya Rajan, S., P. Sivakumar, and Aditya Srinivasan. 2020. {``The
-COVID-19 Pandemic and Internal Labour Migration in India: A {`}Crisis of
-Mobility{'}.''} \emph{The Indian Journal of Labour Economics} 63 (4):
-1021--39. \url{https://doi.org/10.1007/s41027-020-00293-8}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-janoschka2002}{}}%
-Janoschka, Michael. 2002. {``El Nuevo Modelo de La Ciudad
-Latinoamericana: Fragmentación y Privatización.''} \emph{EURE
-(Santiago)} 28 (85).
-\url{https://doi.org/10.4067/s0250-71612002008500002}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-kotsubo2022}{}}%
-Kotsubo, Masaki, and Tomoki Nakaya. 2022. {``Trends in Internal
-Migration in Japan, 2012{\textendash}2020: The Impact of the COVID{-}19
-Pandemic.''} \emph{Population, Space and Place} 29 (4).
-\url{https://doi.org/10.1002/psp.2634}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-lattes95}{}}%
-Lattes, A. 1995. {``Urbanización, Crecimiento Urbano y Migraciones En
-América Latina. Población y Desarrollo: Tendencias y Desafíos.''}
-\emph{Notas de Población} 62: 211--60.
-\url{https://repositorio.cepal.org/handle/11362/38594}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-lattes2017}{}}%
-Lattes, Alfredo E., Jorge Rodríguez, and Miguel Villa. 2017.
-{``Population Dynamics and Urbanization in Latin America: Concepts and
-Data Limitations.''} In, 89--111. Routledge.
-\url{https://doi.org/10.4324/9781315248073-5}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-lucchini2023}{}}%
-Lucchini, Lorenzo, Ollin Langle-Chimal, Lorenzo Candeago, Lucio Melito,
-Alex Chunet, Aleister Montfort, Bruno Lepri, Nancy Lozano-Gracia, and
-Samuel P. Fraiberger. 2023. {``Socioeconomic Disparities in Mobility
-Behavior During the COVID-19 Pandemic in Developing Countries.''}
-\url{https://doi.org/10.48550/ARXIV.2305.06888}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-Maas19}{}}%
+\bibitem[\citeproctext]{ref-Maas19}
 Maas, P., S. Iyer, A. Gros, W. Park, L. McGorman, C. Nayak, and P. A.
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-McKinsey21}{}}%
-McKinsey \& Company. 2021. {``What Executives Are Saying about the
-Future of Hybrid Work.''}
-\url{https://www.mckinsey.com/capabilities/people-and-organizational-performance/our-insights/what-executives-are-saying-about-the-future-of-hybrid-work}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-ILO2020}{}}%
-Nair, Shabari, and Divya Verma. 2020. {``A Policy Framework for India's
-Covid-19 Migration. International Labour Organization.''}
-\url{https://www.ilo.org/newdelhi/info/public/fs/WCMS_746654/lang--en/index.htm}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-nathan2020}{}}%
-Nathan, Max, and Henry Overman. 2020. {``Will Coronavirus Cause a Big
-City Exodus?''} \emph{Environment and Planning B: Urban Analytics and
-City Science} 47 (9): 1537--42.
-\url{https://doi.org/10.1177/2399808320971910}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-UNpopulation19}{}}%
-Nations", "United. 2019. {``World Urbanization Prospects. The 2018
-Revision.''}
-\url{https://population.un.org/wup/Publications/Files/WUP2018-Report.pdf}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-nouvellet2021}{}}%
-Nouvellet, Pierre, Sangeeta Bhatia, Anne Cori, Kylie E. C. Ainslie, Marc
-Baguelin, Samir Bhatt, Adhiratha Boonyasiri, et al. 2021. {``Reduction
-in Mobility and COVID-19 Transmission.''} \emph{Nature Communications}
-12 (1). \url{https://doi.org/10.1038/s41467-021-21358-2}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-oecdreg}{}}%
-\emph{OECD Regions and Cities at a Glance}. n.d. OECD.
-\url{https://doi.org/10.1787/26173212}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-oecdsoc}{}}%
-{``OECD Social, Employment and Migration Working Papers.''} n.d.
-\url{https://doi.org/10.1787/1815199x}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-perales2022}{}}%
-Perales, Francisco, and Aude Bernard. 2022. {``Continuity or Change? How
-the Onset of COVID{-}19 Affected Internal Migration in Australia.''}
-\emph{Population, Space and Place} 29 (2).
-\url{https://doi.org/10.1002/psp.2626}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-Puxe9rez-Campuzano13}{}}%
-Pérez-Campuzano, C., E. y Santos-Cerquera. 2013. {``Tendencias Recientes
-de La Migración Interna En México.''} \emph{"Papeles de Población"} 19:
-53--88.
-\url{https://www.scielo.org.mx/scielo.php?script=sci_arttext\&pid=S1405-74252013000200003}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-Pinilla2021}{}}%
-Pinilla, Vicente, and Luis Antonio Sáez. 2021. {``La Despoblación Rural
-En España: Características, Causas e Implicaciones Para Las Políticas
-Públicas.''} \emph{Presupuesto y Gasto Público} 102: 75--92.
-\url{https://www.ief.es/docs/destacados/publicaciones/revistas/pgp/102.pdf\#page=75}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-PintoDaCunha02}{}}%
-Pinto da Cunha, J. M. P. 2002. {``Urbanización, Redistribución Espacial
-de La Población y Transformaciones Socioeconómicas En América Latina.''}
-\emph{United Nations}.
-\url{https://repositorio.cepal.org/bitstream/handle/11362/7168/1/S029663_es.pdf}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-Pomeroy2021}{}}%
-Pomeroy, Robin, and Ross Chiney. 2021. {``Has COVID Killed Our Cities?
-World Economic Forum.''}
-\url{https://www.weforum.org/agenda/2020/11/cities-podcast-new-york-dead/}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-ramani2021}{}}%
-Ramani, Arjun, and Nicholas Bloom. 2021. {``The Donut Effect of Covid-19
-on Cities.''} \url{https://doi.org/10.3386/w28876}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-rodruxedguezvignoli2017}{}}%
-Rodríguez Vignoli, Jorge, and Francisco Rowe. 2017. {``¿Contribuye La
-Migración Interna a Reducir La Segregación Residencial?''} \emph{Revista
-Latinoamericana de Población} 11 (21): 7--45.
-\url{https://doi.org/10.31406/relap2017.v11.i2.n21.1}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-rodruxedguez-vignoli2018}{}}%
-Rodríguez-Vignoli, Jorge, and Francisco Rowe. 2018. {``How Is Internal
-Migration Reshaping Metropolitan Populations in Latin America? A New
-Method and New Evidence.''} \emph{Population Studies} 72 (2): 253--73.
-\url{https://doi.org/10.1080/00324728.2017.1416155}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-rowe2021}{}}%
-Rowe, Francisco. 2021. {``Big Data and Human Geography.''}
-\url{http://dx.doi.org/10.31235/osf.io/phz3e}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-rowe2022}{}}%
+\bibitem[\citeproctext]{ref-rowe2022}
 Rowe, Francisco, Alessia Calafiore, Daniel Arribas-Bel, Krasen
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-rowe2023}{}}%
-Rowe, Francisco, Miguel González-Leonardo, and Tony Champion. 2023.
-{``Virtual Special Issue: Internal Migration in Times of COVID{-}19.''}
-\emph{Population, Space and Place}, March.
-\url{https://doi.org/10.1002/psp.2652}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-OIT2018}{}}%
-Salazar-Xirinachs, Juan, José Manuel \& Chacaltana. 2018. {``Políticas
-de Formalización En América Latina: Avances y Desafíos. Organización
-Internacional Del Trabajo.''}
-\url{https://www.ilo.org/americas/publicaciones/WCMS_645159/lang--es/index.htm}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-sobrino12}{}}%
-Sobrino, J. 2012. {``La Urbanización En El México Contemporáneo.''}
-\emph{Notas de Población} 94: 93--122.
-\url{https://repository.eclac.org/handle/11362/12898}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-sobrino2006}{}}%
-Sobrino, Jaime. 2006. {``Comptetitiveness and Employment in the Largest
-Metropolitan Areas of Mexico.''} In, 309--54. El Colegio de México.
-\url{https://doi.org/10.2307/j.ctv3dnrw3.15}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-stawarz2022}{}}%
-Stawarz, Nico, Matthias Rosenbaum-Feldbrügge, Nikola Sander, Harun
-Sulak, and Vanessa Knobloch. 2022. {``The Impact of the COVID{-}19
-Pandemic on Internal Migration in Germany: A Descriptive Analysis.''}
-\emph{Population, Space and Place} 28 (6).
-\url{https://doi.org/10.1002/psp.2566}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-storper2004}{}}%
-Storper, M., and A. J. Venables. 2004. {``Buzz: Face-to-Face Contact and
-the Urban Economy.''} \emph{Journal of Economic Geography} 4 (4):
-351--70. \url{https://doi.org/10.1093/jnlecg/lbh027}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-TheGuardian2020}{}}%
-The Guardian. 2020. {``Escape to the Country: How Covid Is Driving an
-Exodus from Britain's Cities.''}
-\url{https://www.theguardian.com/world/2020/sep/26/escape-country-covid-exodus-britain-cities-pandemic-urban-green-space}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-tuxf8nnessen2021}{}}%
-Tønnessen, Marianne. 2021. {``Movers from the City in the First Year of
-Covid.''} \emph{Nordic Journal of Urban Studies} 1 (2): 131--47.
-\url{https://doi.org/10.18261/issn.2703-8866-2021-02-03}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-USBureauLabor2021}{}}%
-US Bureau of Labor Statistics. 2021. {``Unemployment Rises in 2020, as
-the Country Battles the COVID-19 Pandemic.''}
-\url{https://www.bls.gov/opub/mlr/2021/article/unemployment-rises-in-2020-as-the-country-battles-the-covid-19-pandemic.htm}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-vogiazides2022}{}}%
-Vogiazides, Louisa, and Juta Kawalerowicz. 2022. {``Internal Migration
-in the Time of Covid: Who Moves Out of the Inner City of Stockholm and
-Where Do They Go?''} \emph{Population, Space and Place} 29 (4).
-\url{https://doi.org/10.1002/psp.2641}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-wang2022}{}}%
+\bibitem[\citeproctext]{ref-wang2022}
 Wang, Yikang, Chen Zhong, Qili Gao, and Carmen Cabrera-Arnau. 2022.
@@
 \url{https://doi.org/10.1007/s44212-022-00018-w}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-WorldBank2023}{}}%
-World Bank", "The. 2023. {``LAC Equity Lab: Poverty - Poverty Rate.''}
-\url{https://www.worldbank.org/en/topic/poverty/lac-equity-lab1/poverty/head-count}.
 

### 2024-01-17 manuscript/manuscript.qmd
compiling-manuscript

@@
 ---
-title: "The recovery patterns of human mobility in Latin American urban areas after COVID-19"
-author: "Carmen Cabrera-Arnau, Miguel González-Leonardo, Andrea Nasuto, Ruth Neville, Francisco Rowe"
+title: "**The recovery patterns of human mobility in Latin American urban areas after COVID-19**"
+author:
+  - name: Carmen Cabrera-Arnau
+    affil-id: 1
+  - name: Francisco Rowe
+    affil-id: 1
+  - name: Miguel González-Leonardo
+    affil-id: 2
+  - name: Andrea Nasuto
+    affil-id: 1
+  - name: Ruth Neville
+    affil-id: 1
+affiliations: 
+  - id: 1
+    name: Geographic Data Science Lab, Department of Geography and Planning, University of Liverpool, Liverpool, UK
+  - id: 2
+    name: Centre for Demographic Urban and Environmental Studies, El Colegio de México, Ciudad de México, México
+#  - id: 3
+#    name: International Institute for Applied Systems Analysis, Wittgenstein Centre, Vienna, Austria
+abstract: |
+  The COVID‐19 pandemic has impacted the national systems of population movement around the world. 
 format: 
   pdf:
-    template: template.tex
+    fontsize: 11pt
+    geometry:
+      - left=2.54cm
+      - right=2.54cm
+    header-includes:
+      \usepackage{lipsum}
+      \usepackage{libertine}
+    keep-tex: true
+    template-partials: 
+      - title.tex
+    include-in-header:
+      text: |
+        \usepackage[noblocks]{authblk}
+        \renewcommand*{\Authsep}{, }
+        \renewcommand*{\Authand}{, }
+        \renewcommand*{\Authands}{, }
+        \renewcommand\Affilfont{\small}
+number-sections: true
 editor: visual
@@
 
-```{=html}
-<!--# The abstract has to be set up in the tex file, but I copy it below.
+# Introduction {#sec-intro}
 
-Recent empirical studies, predominantly from countries in the Global North, have shown that the COVID-19 pandemic disrupted human mobility at different spatial scales, but particularly in big cities, initially identified as epicentres of infections.
-Lockdowns, remote work, and online education decreased the demand for commuting and urban living, resulting in an "urban exodus".
-Despite the existing evidence, little is known about whether counterurbanisation movements have unfolded similarly in the Global South.
-In this study, we use anonymised, high-resolution location data from Facebook users to answer the following research questions: i) to what extent are people leaving cities?
-ii) are there variations across countries?
-iii) are there changes in mobility patterns over time?
-iv) how long do these changes last?
-The data is aggregated into spatial units according to the Bing Maps Tile System and temporally aggregated into 8-hour time windows for a period of approximately 2 years.
-The methodology involves three steps.
-Firstly, we classify the Bing tiles into different population density categories according to population data from WorldPop.
-Secondly, we quantify and analyse the number and characteristics of movements between different locations in each country.
-Finally, we use statistical modelling and spatial interaction models to assess how the population density category at the origin and destination affects the strength of the flows at different points in time.
-We observe that the strength of flows between Bing tiles belonging to different population density categories was systematically reduced in the early days of the pandemic and, even though by 2022 it seemed to have increased, it never returned to pre-pandemic levels.
-In the early days of COVID-19, we also find a systematically increase above pre-pandemic levels of the strength of the flows between Bing tiles belonding to the the same population density category.
-As of 2022, the frequency of these movements within population density classes have decreased with respect to 2020.
+<!--# Francisco to write this -->
 
--->
-```
-\section{Introduction}
+# Results {#sec-results}
 
-<!--# Add comments using this -->
+<!--# Carmen to write this -->
 
-\section{Data}
+The evolution of the levels of movement was measured with respect to a baseline period prior to the pandemic as described in @sec-methods.
+For the purposes of the analysis, we aggregate the raw movement data temporally into months and spatially into administrative units according to various GADM levels.
+The analysis focuses on administrative areas that are within the boundaries of functional urban areas as specified by the Global Human Settlement Layer.
+For each administrative area, we compute the Relative Deprivation Index based on data from NASA's Socioeconomic Data and Applications Centre (SEDAC). Further details are given in @sec-methods.
 
-The analysis proposed in this paper is based on several data sets.
-Firstly, Facebook Population and Facebook Movements are data sets based on the Facebook users' location history.
-The data sets were created by Facebook's owner company Meta and can be accessed through their Data for Good programme (\url{https://dataforgood.facebook.com/}).
-Prior to releasing the data sets, Meta applies three techniques to ensure privacy and anonymisation.
-First, a small undisclosed amount of random noise is added to ensure that precise location cannot be identified for small population counts in sparsely populated areas.
-While removing small counts may lead to an underrepresentation of the population in these places, the geographic distribution of population is still reflected in the data.
-Second, spatial smoothing is applied to produce a smooth population count surface using inverse distance-weighted averaging.
-Third, any remaining population counts of less than 10 are removed from the final data set (see Maas et al. @Maas19 for details).
+# Discussion {#sec-discussion}
 
-Both data sets Facebook Population and Facebook Movements contain data corresponding to a time period comprising approximately two years, starting from March 2020, and to four Latin American countries, Argentina, Chile, Colombia and Mexico.
-The data in both data sets is temporally aggregated into three 8-hour windows (00:00--08:00, 08:00--16:00 and 16:00--00:00) for every day in the aforementioned two-year period.
-It is spatially aggregated into tiles according to the Bing Maps Tile System.
-This geospatial indexing system was developed by Microsoft and it partitions the world into square cells at various levels of resolution.
-The Facebook Movements and Facebook Population datasets are aggregated into various levels of resolution depending on the country, and the length of the Bing tile sides vary also by country according to the country's distance to the Equator, even when two countries use tiles at the same level of resolution.
+<!--# Carmen to write this -->
 
-The Facebook Population data set provides information on the number of active Facebook users in each tile.
-The data set Facebook Movements captures the total number of Facebook users moving between pairs of origin and destination Bing tiles.
-We note here that due to the nature of the Facebook Movement data, we cannot distinguish between different types of movements, for example, daily commutes to work or permanent changes of address.
-However, we are still able to detect the evolution of movements between origin-destination pairs of Bing tiles and hence, we are able to capture the impact that COVID-19 has on mobility patterns.
+# Methods {#sec-methods}
 
-On top of the data for the two-year period, each entry in the Facebook Population and Facebook Movements datasets include data for baseline levels before COVID-19.
-The baseline values are computed based on a 45-day period ending on the 10th of March of 2020.
-The data sets also include a 'quality' score indicating the number of standard deviations by which the observed data at specific locations and time windows differ from the baseline values, hence highlighting statistically significant changes.
+<!--# Carmen to write this -->
 
-An additional data set from WorldPop was used to capture the spatial distribution of population density in the different countries analysed here.
-The WorlPop dataset is in raster format and contains gridded population data at 1 sqkm resolution.
-
-\section{Method}
-
-\subsection{Data pre-processing}
-
-Ultimately, we work with a dataset where each entry represents a population flow between an origin and a destination pair.
-For each entry, we consider when the movement took place, the population density at the origin and destination, the number of active Facebook users at the origin and destination and?
-
-\subsubsection{Spatial aggregation}
-
-The working dataset is aggregated at the same level as the Bing tiles used to track the movements in a given country.
-To each Bing tile, we aggregate WorldPop population data and use it to estimate the population density of each Bing tile.
-
-\subsubsection{Temporal aggregation}
-
-The raw data is aggregated into 8-hour windows for every day in the two-year period.
-However, we are interested in longer-term trends and not in how population flows vary throughout the day.
-For this reason, we obtain estimates for the parameters in the models described in subsection \ref{subsec-models} by splitting the total data set by month.
-For the visualisations in section \ref{sec-results}, we also present the data aggregated by month.
-
-\subsection{Classifying Bing tiles according to their population density}
-
-Here we aim to understand how the population density at the origin and the destination might influence mobility behaviours.
-To help characterise the population density between orign and destination, we classify the Bing tiles into 10 discrete categories of population density according to the Jenks natural breaks classification method, hence obtaining a categorisation of Bing tiles a lot more detailed than the traditional binary rural/urban classification.
-
-![Maps of population density categories by country with histograms, showing that the distribution of Bing tiles in each density category.](figures/Map_density_histogram.png){fig-align="center" width="625" height="640"}
-
-\subsection{Estimating the strength of the flows between tiles of different population density categories}
-
-\ref{subsec-models}
-
-In this section, we propose a statistical model to capture the extent to which the strength of the population flows is systematically impacted by the density classes of the origin and destination tiles, and how this effect varies over time.
-Our approach here is very similar to that proposed by \cite{Rowe}.
-
-While we could obtain the population flows between different population classes and their evolution over time, the results would fail to capture specifically the effect of the density classes of the origin and destination tiles.
-This is due to the fact that movement patterns are influenced not only by the population density of the starting and ending locations, but also by factors such as the distance, the population size, the time of the week and day at which movements take place, etc.
-These additional variables can act as confounding factors, obscuring the actual impact of population density.
-In order to disentangle and obtain a more accurate estimation of the importance of population density, we chose to utilise a regression modeling approach.
-I AM HERE
-
-We adopted a spatial interaction framework (Rowe, Lovelace, et al., 2022).
-We used population flows between tiles by time window and day (as described in Section 3), and modelled them as a function of characteristics of the flow itself (Fij), the origin (Tilei) and destination (Tilej) tiles and the temporal nature of the flow (Tw).
-Crucially, we included indicator variables that capture the pair of population classes (Figure 1) of the origin and destination tile.
-In mathematical form:
-
-(2) where µijw is the expectation of the flow of people from tile i to tile j in the time window w; ⍺ is an intercept; is a series of indicator variables that reflect the pair of population density classes of a given origin i (I) and destination tile j (J), resulting in 99 pairs (10 classes × 10 classes minus one so it is not collinear with ⍺); dij is the geographic distance between tiles i and j; qijw is a measure of the quality of the flow estimate provided by Meta-Facebook and related to the uncertainty behind the user count of the flow as described in Section 3; Popi,j represents the population at the origin (i) and destination (j) tiles; are parameters to estimate in the model linking their respective covariates to µijw; D is a trend tracking the day to which the flow relates to during the period in analysis; while Wk and H are indicator variables capturing day of the week (i.e., weekday or weekend) and hour window (i.e., 00:00--08:00, 08:00--16:00 and 16:00--00:00).
-Our focus in Equation (2) is centred on .
-Controlling for all other variables, these parameters capture the extent to which, the expected flow between a given origin-destination population density class pair of tiles (e.g., a high-density origin to a low-density destination) is systematically higher or lower than if it occurred between a baseline origin-destination population density class pair of tiles (e.g., a low-density origin to a low-density destination).
-Additionally, we standardised continuous variables (dij, qijw, Popi j), α so that they can be interpreted as the expected flow on the first day (D = 0), during the first temporal window (H = 00:00--08:00), on a weekday (Wk = 0), for the baseline origin-population population density class pair, when all the other variables are at their mean value.
-In this context, each can also be seen as the 'modulation factor' around that expectation associated with each pair of origin-destination classes.
-The baseline origin-destination population density class pair is the lowest population density class as origin and destination.
-
-We used a count data regression model.
-Specifically, we fitted a generalised linear model (GLM) where the error term is assumed to be distributed following a Poisson distribution, with a flow expectation of (µijw) linked to the flow count (Fijw) through a log link: (3) (4) The Poisson regression model (PRM) assumes that equidispersion, that is, equality of mean and variance in the response variable (Cameron & Trivedi, 2013).
-In practice, the equidispersion property is commonly violated because of overdispersion, that is, the variance exceeds the mean.
-When this occurs, the PRM may produce biased parameter estimates, causing the standard errors of the estimates to be underestimated, and compromising the statistical inference process (Hilbe, 2011).
-To test for overdispersion, we used a regression-based test based on an auxiliary regression of the conditional variance as described in Cameron and Trivedi (2013).
-
-Following Gelman and Hill (2006), we used a quasi-PRM to address overdispersion in our response variable.
-This is one of the most common strategies to deal with overdispersion in count data models (Hilbe, 2011).
-Intuitively, this model adjusts the standard errors of the estimates to account for the extra dispersion in the data.
-To implement this, we estimated Equation (2) by using robust variance estimators.
-The number of active Facebook users were used as a weight to account for the variability of the observed count of population movement over time.
-This strategy is also used to mitigate for any potential biases regarding the variation in the observed number of active Facebook users changes over time across Britain.
-
-We fitted Equation (2) using iteratively reweighted least squares (IWLS).
-We separately estimated models for individual months in our data, resulting in 18 sets of estimates.
-Our key aim was to generate estimates for ⍺ and , so that we focused on discussing the evolution of these estimates in a grid of line plots with 10 rows and 10 columns, each of them representing one of our population density classes.
-The plot corresponding to the Ith row and Jth column displays the evolution of the parameter that tracks the intensity of population flows from tiles in population density class Ith to those in population density class Jth.
-
-\section{Results}\label{sec-results}
-
-\section{Discussion}
-
-\section{Conclusion}
-
-\section{References}
+# References

### 2024-01-17 manuscript/manuscript.tex
compiling-manuscript

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
\PassOptionsToPackage{dvipsnames,svgnames,x11names}{xcolor}
%
\documentclass[
  11pt,
]{article}

\usepackage{amsmath,amssymb}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
\usepackage{lmodern}
\ifPDFTeX\else  
    % xetex/luatex font selection
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\usepackage[left=2.54cm,right=2.54cm]{geometry}
\setlength{\emergencystretch}{3em} % prevent overfull lines
\setcounter{secnumdepth}{5}
% Make \paragraph and \subparagraph free-standing
\ifx\paragraph\undefined\else
  \let\oldparagraph\paragraph
  \renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
  \let\oldsubparagraph\subparagraph
  \renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi

\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}\usepackage{longtable,booktabs,array}
\usepackage{calc} % for calculating minipage widths
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother

\usepackage[noblocks]{authblk}
\renewcommand*{\Authsep}{, }
\renewcommand*{\Authand}{, }
\renewcommand*{\Authands}{, }
\renewcommand\Affilfont{\small}
\usepackage{lipsum} \usepackage{libertine}
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\usepackage{bookmark}

\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={The recovery patterns of human mobility in Latin American urban areas after COVID-19},
  pdfauthor={Carmen Cabrera-Arnau; Francisco Rowe; Miguel González-Leonardo; Andrea Nasuto; Ruth Neville},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

\title{\textbf{The recovery patterns of human mobility in Latin American
urban areas after COVID-19}}

\author[1]{Carmen Cabrera-Arnau}
\author[1]{Francisco Rowe}
\author[2]{Miguel González-Leonardo}
\author[1]{Andrea Nasuto}
\author[1]{Ruth Neville}

\affil[1]{Geographic Data Science Lab, Department of Geography and
Planning, University of Liverpool, Liverpool, UK}
\affil[2]{Centre for Demographic Urban and Environmental Studies, El
Colegio de México, Ciudad de México, México}

\date{}
\begin{document}
\maketitle
\begin{abstract}
The COVID‐19 pandemic has impacted the national systems of population
movement around the world.
\end{abstract}

\section{Introduction}\label{sec-intro}

\section{Results}\label{sec-results}

The evolution of the levels of movement was measured with respect to a
baseline period prior to the pandemic as described in
Section~\ref{sec-methods}. For the purposes of the analysis, we
aggregate the raw movement data temporally into months and spatially
into administrative units according to various GADM levels. The analysis
focuses on administrative areas that are within the boundaries of
functional urban areas as specified by the Global Human Settlement
Layer. For each administrative area, we compute the Relative Deprivation
Index based on data from NASA's Socioeconomic Data and Applications
Centre (SEDAC). Further details are given in Section~\ref{sec-methods}.

\section{Discussion}\label{sec-discussion}

\section{Methods}\label{sec-methods}

\section{References}\label{references}

\end{document}

### 2024-01-17 manuscript/manuscript.qmd
results

@@
 ---
-title: "**The recovery patterns of human mobility in Latin American urban areas after COVID-19**"
+title: "**Urban mobility in Latin America after COVID-19**"
 author:
@@
 The analysis focuses on administrative areas that are within the boundaries of functional urban areas as specified by the Global Human Settlement Layer.
-For each administrative area, we compute the Relative Deprivation Index based on data from NASA's Socioeconomic Data and Applications Centre (SEDAC). Further details are given in @sec-methods.
+For each administrative area, we compute the Relative Deprivation Index based on data from NASA's Socioeconomic Data and Applications Centre (SEDAC).
+Figure X displays the administrative areas included in the study, coloured according to the Relative Index of Deprivation.
+Predictions about the evolution of the levels of movement are made using the Prophet forecasting procedure.
+Further details are provided in @sec-methods.
+
+<!--# Add Figure X. -->
+
+## The impact of COVID-19 on patterns of urban mobility
+
+We analysed the evolution of the relative intensity of urban mobility with respect to a baseline period prior to the pandemic.
+Specifically, movements covering a distance of at most 70 km are considered.
+For a movement to be classified as urban it needs to start or end within a functional urban area from Argentina, Chile, Colombia and Mexico.
+The observed data is available for a two-year period starting in April 2020, just after the first wave of COVID-19 pandemic cases, and ending in March 2022.
+After 2022 no observations are available, however, we generate a 12-month forecast up to March 2023 in order to gain a better understanding of the recovery trends.
+
+Figure X displays the patterns of recovery for the mobility levels in the administrative units belonging to functional urban areas in the countries of interest.
+The three lines in each panel represent the mean levels of mobility for administrative units grouped into one of three terciles, according to their average relative deprivation index.
+
+Generally, there was a drop in the levels of mobility with respect to the baseline period in all four countries.
+This drop was especially large for Argentina, Chile and Colombia, with Mexico displaying a smaller decrease in the number of movements with respect to the baseline.
+Following the initial drop in movement, all four countries evolve towards the recovery of baseline levels of urban mobility, with a generally increasing trend.
+There are fluctuations from the general trend, which manifest differently for each country.
+These fluctuations mirror each other in the case of Argentina and Colombia, where urban mobility sharply bounces back closer to pre-pandemic levels around July of 2020.
+Chile and Mexico display more progressive patterns of recovery, although Chile never reaches baseline levels.
+These fluctuations are unique to each country and can be attributed to local factors such as the effects of seasonality or the different stringency measures imposed by the national governments during the pandemic.
+
+From Figure X, we observe that there is a consistent tendency in how administrative units with varying levels of deprivation were affected by the pandemic.
+For all four countries, we observe that the administrative units in the most deprived tercile are the ones that experienced the smallest loss in levels of mobility at the beginning of the pandemic.
+Differences in the levels of mobility across relative deprivation terciles diminish with time.
+Argentina and Chile stand out as the countries with the largest differences in mobility levels for different relative deprivation terciles.
+
+## Socioeconomic deprivation and recovery of urban mobility
+
+In this section we explore further the role of socioeconomic deprivation in the evolution of the levels of urban mobility.
+For a given point in time (i.e.
+a month), we start by considering the relationship between the number of movements relative to the pre-pandemic baseline period and the average relative deprivation index, at the administrative unit level.
+We assume that this relationship is linear and we use a linear regression to estimate the slope and intercept characterising the line of best fit.
+This is shown for April 2020 and March 2022, in the right-hand side panels of Figure X.
+After obtaining the slope and intercept for every month, we are able to plot the evolution of these parameters for both the observed and forecasted data, as displayed on the left-hand-side panels of the same Figure.
+
+We find patterns in the evolution of the estimated parameters that characterise the relationship between the levels of urban mobility and RDI.
+In Argentina, Colombia and Mexico, we observe that the slope of this relationship evolves to become smaller over time.
+The tendency is not apparent in Chile, where the slope of the relationship remains approximately the same despite the temporary fluctuations.
+The slope captures the extent of differences in the level of urban mobility across administrative units with varying levels of socioeconomic deprivation.
+It can therefore be regarded as a measure of inequality.
+A slope equal to zero would mean that all administrative units display the same intensity of movement regardless of their socioeconomic deprivation levels.
+Given the patterns observed in Argentina, Colombia and Mexico, we find that at the beginning of the pandemic there were notable inequalities between socioeconomic groups in terms of the levels of urban mobility.
+While it has taken more than two years for Argentina and Mexico to close the gap (their slope is close to zero from spring 2022), inequalities persist in Chile and Colombia as of March 2023.
+
+The intercept of the relationship displays stronger patterns, which are consistent across the four countries.
+The intercept captures the urban mobility levels in administrative where relative deprivation is zero, or in other words, if there was no socioeconomic deprivation in an administrative area, the intercept would represent its level of mobility.
+The intercept was well below the baseline level at the early stages of the pandemic.
+As observed in Figure X, while there are some differences between countries in the evolution of the intercept, the general tendency is for the intercept to increase.
+While Argentina and Mexico reach values that are closer to the baseline towards the end of the forecast period, the intercept for Chile and Colombia remains lower.
+Therefore, if there were areas with no socioeconomic deprivation, we would have seen a recovery in the levels of mobility, although not quite back to baseline levels in the case of Chile and Colombia.
 

### 2024-01-17 manuscript/manuscript.tex
results

@@
 \hypersetup{
-  pdftitle={The recovery patterns of human mobility in Latin American urban areas after COVID-19},
+  pdftitle={Urban mobility in Latin America after COVID-19},
   pdfauthor={Carmen Cabrera-Arnau; Francisco Rowe; Miguel González-Leonardo; Andrea Nasuto; Ruth Neville},
@@
 
-\title{\textbf{The recovery patterns of human mobility in Latin American
-urban areas after COVID-19}}
+\title{\textbf{Urban mobility in Latin America after COVID-19}}
 
@@
 Index based on data from NASA's Socioeconomic Data and Applications
-Centre (SEDAC). Further details are given in Section~\ref{sec-methods}.
+Centre (SEDAC). Figure X displays the administrative areas included in
+the study, coloured according to the Relative Index of Deprivation.
+Predictions about the evolution of the levels of movement are made using
+the Prophet forecasting procedure. Further details are provided in
+Section~\ref{sec-methods}.
+
+\subsection{The impact of COVID-19 on patterns of urban
+mobility}\label{the-impact-of-covid-19-on-patterns-of-urban-mobility}
+
+We analysed the evolution of the relative intensity of urban mobility
+with respect to a baseline period prior to the pandemic. Specifically,
+movements covering a distance of at most 70 km are considered. For a
+movement to be classified as urban it needs to start or end within a
+functional urban area from Argentina, Chile, Colombia and Mexico. The
+observed data is available for a two-year period starting in April 2020,
+just after the first wave of COVID-19 pandemic cases, and ending in
+March 2022. After 2022 no observations are available, however, we
+generate a 12-month forecast up to March 2023 in order to gain a better
+understanding of the recovery trends.
+
+Figure X displays the patterns of recovery for the mobility levels in
+the administrative units belonging to functional urban areas in the
+countries of interest. The three lines in each panel represent the mean
+levels of mobility for administrative units grouped into one of three
+terciles, according to their average relative deprivation index.
+
+Generally, there was a drop in the levels of mobility with respect to
+the baseline period in all four countries. This drop was especially
+large for Argentina, Chile and Colombia, with Mexico displaying a
+smaller decrease in the number of movements with respect to the
+baseline. Following the initial drop in movement, all four countries
+evolve towards the recovery of baseline levels of urban mobility, with a
+generally increasing trend. There are fluctuations from the general
+trend, which manifest differently for each country. These fluctuations
+mirror each other in the case of Argentina and Colombia, where urban
+mobility sharply bounces back closer to pre-pandemic levels around July
+of 2020. Chile and Mexico display more progressive patterns of recovery,
+although Chile never reaches baseline levels. These fluctuations are
+unique to each country and can be attributed to local factors such as
+the effects of seasonality or the different stringency measures imposed
+by the national governments during the pandemic.
+
+From Figure X, we observe that there is a consistent tendency in how
+administrative units with varying levels of deprivation were affected by
+the pandemic. For all four countries, we observe that the administrative
+units in the most deprived tercile are the ones that experienced the
+smallest loss in levels of mobility at the beginning of the pandemic.
+Differences in the levels of mobility across relative deprivation
+terciles diminish with time. Argentina and Chile stand out as the
+countries with the largest differences in mobility levels for different
+relative deprivation terciles.
+
+\subsection{Socioeconomic deprivation and recovery of urban
+mobility}\label{socioeconomic-deprivation-and-recovery-of-urban-mobility}
+
+In this section we explore further the role of socioeconomic deprivation
+in the evolution of the levels of urban mobility. For a given point in
+time (i.e.~a month), we start by considering the relationship between
+the number of movements relative to the pre-pandemic baseline period and
+the average relative deprivation index, at the administrative unit
+level. We assume that this relationship is linear and we use a linear
+regression to estimate the slope and intercept characterising the line
+of best fit. This is shown for April 2020 and March 2022, in the
+right-hand side panels of Figure X. After obtaining the slope and
+intercept for every month, we are able to plot the evolution of these
+parameters for both the observed and forecasted data, as displayed on
+the left-hand-side panels of the same Figure.
+
+We find patterns in the evolution of the estimated parameters that
+characterise the relationship between the levels of urban mobility and
+RDI. In Argentina, Colombia and Mexico, we observe that the slope of
+this relationship evolves to become smaller over time. The tendency is
+not apparent in Chile, where the slope of the relationship remains
+approximately the same despite the temporary fluctuations. The slope
+captures the extent of differences in the level of urban mobility across
+administrative units with varying levels of socioeconomic deprivation.
+It can therefore be regarded as a measure of inequality. A slope equal
+to zero would mean that all administrative units display the same
+intensity of movement regardless of their socioeconomic deprivation
+levels. Given the patterns observed in Argentina, Colombia and Mexico,
+we find that at the beginning of the pandemic there were notable
+inequalities between socioeconomic groups in terms of the levels of
+urban mobility. While it has taken more than two years for Argentina and
+Mexico to close the gap (their slope is close to zero from spring 2022),
+inequalities persist in Chile and Colombia as of March 2023.
+
+The intercept of the relationship displays stronger patterns, which are
+consistent across the four countries. The intercept captures the urban
+mobility levels in administrative where relative deprivation is zero, or
+in other words, if there was no socioeconomic deprivation in an
+administrative area, the intercept would represent its level of
+mobility. The intercept was well below the baseline level at the early
+stages of the pandemic. As observed in Figure X, while there are some
+differences between countries in the evolution of the intercept, the
+general tendency is for the intercept to increase. While Argentina and
+Mexico reach values that are closer to the baseline towards the end of
+the forecast period, the intercept for Chile and Colombia remains lower.
+Therefore, if there were areas with no socioeconomic deprivation, we
+would have seen a recovery in the levels of mobility, although not quite
+back to baseline levels in the case of Chile and Colombia.
 

### 2024-01-18 manuscript/manuscript.qmd
methods

@@
 
+## Meta-Facebook movement data
+
+To capture population movements, we used anonymised aggregate mobile
+phone location data from Meta users for Argentina, Colombia, Chile and Mexico, covering a 24-month period from April 2020 to March 2022.
+We used the dataset Facebook Movements created by Meta and accessed through their Data for Good Initiative (https://dataforgood.facebook.com).
+The data are built from Facebook app users who have the location services setting turned on on their smartphone.
+Prior to releasing the datasets, Meta ensures privacy and anonymity by removing personal information and applying privacy-preserving techniques (Maas et al.
+2019).
+Small-count dropping is one of these techniques.
+A data entry is removed if the population or movement count for an area is lower than 10.
+The removal of
+small counts may mean that population counts in small sparsely populated areas are not captured.
+A second technique consists in adding a small undisclosed amount of random noise to ensure that it is not possible to ascertain precise, true counts for sparsely populated locations.
+Third, spatial smoothing using inverse distance-weighted averaging is also applied is applied to produce a smooth population count surface.
+The Facebook Movements dataset offers information on the total number of Facebook users moving between and within spatial units in the form of origin-destination matrices.
+The data is temporally aggregated into three daily 8-hour time windows (i.e.
+00:00-08:00, 08:00-16:00 and 16:00-00:00).
+The dataset includes a baseline capturing the number of movements before COVID-19 based on a 45-day period ending on March 10th 2020.
+The baseline is computed using an average for the same time of the day and day of the week in the period preceding March 10th.
+For instance,
+the baseline for Monday 00:00-08:00 time window is obtained by averaging over data collected on Mondays from 00:00 to 8:00 for the 45-day period.
+Details about the baseline can be found in Maas et al.
+(2019).
+The Bing Maps Tile System developed by Microsoft (Microsoft) is used a spatial framework to organise the data.
+The Tile System is a geospatial indexing system that partitions the world into tile cells in a hierarchical way, comprising 23 different levels of detail (Microsoft).
+At the lowest level of detail (Level 1), the world is divided into four tiles with a coarse spatial resolution.
+At each successive level, the resolution increases by a factor of two.
+The data that we used are spatially aggregated into Bing tile levels 13.
+That is about 4.9 x
+4.9km at the Equator (Maas et al. 2019).
+
+## Spatiotemporal data aggregation
+
+Since the focus of this work is on urban mobility, we focus the analysis on Functional Urban Areas (FUAs), defined by the Global Human Settlement Layer.
+The spatial extent of the FUAs is often large and may include several towns and neighbourhoods displaying a variety of socioeconomic characteristics, hence using FUAs as the spatial units of aggregation would considerably mask the heterogeneity in the mobility patterns.
+While the original unit of aggregation for the movement data, i.e.
+the tiles from the Bing Maps Tile System, offers the highest degree of spatial granularity available, the interpretation of findings from an analysis based on administrative units is often more valuable.
+For this reason, we perform a spatial join to aggregate the movement data into administrative units at the GADM level 2 or 3.
+Only flows of people starting or ending within the boundaries of a FUA are considered.
+
+While the original movement data is originally aggregated into 8-hour windows, this resolution is too fine for our analysis.
+Since the analysis is focused on the longer-term evolution of patterns of urban mobility, we aggregate the movement data by month.
+
+In our analysis, we use the Relative Deprivation Index (RDI) as a measure of socioeocnomic deprivation.
+The RDI data is made available via NASA's Socioeconomic Data and Applications Centre (SEDAC), with a spatial resolution of 1km pixels.
+We perform a spatial join of the gridded data and the administrative units and compute the average RDI within each of these units.
+
+## Time series analysis
+
+For the countries included in the analysis, the movement data is available until March 2022.
+In order to gain a better understanding of the recovery trends after the pandemic, we generate a 12-month forecast up to March 2023.
+This is done using Prophet, a procedure to forecast time series data based on an additive model where non-linear trends can be fit with seasonality effects, holiday effects and other external factors such as the stringency index.
+For the analysis, only seasonality effects are included as they yield the most realistic predictions.
+
+In Figures X and Y, outliers are removed.
+Outliers are defined as administrative units which, at any given month, have a z-score greater than 4 for the relative intensity of movement.
+Furthermore, a Savitzky-Golay filter is applied to smooth the time series data, using a length of 4 units for the filter window and order 2 polynomials to fit the samples.
+
 # References

### 2024-01-18 manuscript/manuscript.tex
methods

@@
 
+\subsection{Meta-Facebook movement
+data}\label{meta-facebook-movement-data}
+
+To capture population movements, we used anonymised aggregate mobile
+phone location data from Meta users for Argentina, Colombia, Chile and
+Mexico, covering a 24-month period from April 2020 to March 2022. We
+used the dataset Facebook Movements created by Meta and accessed through
+their Data for Good Initiative (https://dataforgood.facebook.com). The
+data are built from Facebook app users who have the location services
+setting turned on on their smartphone. Prior to releasing the datasets,
+Meta ensures privacy and anonymity by removing personal information and
+applying privacy-preserving techniques (Maas et al.~2019). Small-count
+dropping is one of these techniques. A data entry is removed if the
+population or movement count for an area is lower than 10. The removal
+of small counts may mean that population counts in small sparsely
+populated areas are not captured. A second technique consists in adding
+a small undisclosed amount of random noise to ensure that it is not
+possible to ascertain precise, true counts for sparsely populated
+locations. Third, spatial smoothing using inverse distance-weighted
+averaging is also applied is applied to produce a smooth population
+count surface. The Facebook Movements dataset offers information on the
+total number of Facebook users moving between and within spatial units
+in the form of origin-destination matrices. The data is temporally
+aggregated into three daily 8-hour time windows (i.e.~00:00-08:00,
+08:00-16:00 and 16:00-00:00). The dataset includes a baseline capturing
+the number of movements before COVID-19 based on a 45-day period ending
+on March 10th 2020. The baseline is computed using an average for the
+same time of the day and day of the week in the period preceding March
+10th. For instance, the baseline for Monday 00:00-08:00 time window is
+obtained by averaging over data collected on Mondays from 00:00 to 8:00
+for the 45-day period. Details about the baseline can be found in Maas
+et al.~(2019). The Bing Maps Tile System developed by Microsoft
+(Microsoft) is used a spatial framework to organise the data. The Tile
+System is a geospatial indexing system that partitions the world into
+tile cells in a hierarchical way, comprising 23 different levels of
+detail (Microsoft). At the lowest level of detail (Level 1), the world
+is divided into four tiles with a coarse spatial resolution. At each
+successive level, the resolution increases by a factor of two. The data
+that we used are spatially aggregated into Bing tile levels 13. That is
+about 4.9 x 4.9km at the Equator (Maas et al.~2019).
+
+\subsection{Spatiotemporal data
+aggregation}\label{spatiotemporal-data-aggregation}
+
+Since the focus of this work is on urban mobility, we focus the analysis
+on Functional Urban Areas (FUAs), defined by the Global Human Settlement
+Layer. The spatial extent of the FUAs is often large and may include
+several towns and neighbourhoods displaying a variety of socioeconomic
+characteristics, hence using FUAs as the spatial units of aggregation
+would considerably mask the heterogeneity in the mobility patterns.
+While the original unit of aggregation for the movement data, i.e.~the
+tiles from the Bing Maps Tile System, offers the highest degree of
+spatial granularity available, the interpretation of findings from an
+analysis based on administrative units is often more valuable. For this
+reason, we perform a spatial join to aggregate the movement data into
+administrative units at the GADM level 2 or 3. Only flows of people
+starting or ending within the boundaries of a FUA are considered.
+
+While the original movement data is originally aggregated into 8-hour
+windows, this resolution is too fine for our analysis. Since the
+analysis is focused on the longer-term evolution of patterns of urban
+mobility, we aggregate the movement data by month.
+
+In our analysis, we use the Relative Deprivation Index (RDI) as a
+measure of socioeocnomic deprivation. The RDI data is made available via
+NASA's Socioeconomic Data and Applications Centre (SEDAC), with a
+spatial resolution of 1km pixels. We perform a spatial join of the
+gridded data and the administrative units and compute the average RDI
+within each of these units.
+
+\subsection{Time series analysis}\label{time-series-analysis}
+
+For the countries included in the analysis, the movement data is
+available until March 2022. In order to gain a better understanding of
+the recovery trends after the pandemic, we generate a 12-month forecast
+up to March 2023. This is done using Prophet, a procedure to forecast
+time series data based on an additive model where non-linear trends can
+be fit with seasonality effects, holiday effects and other external
+factors such as the stringency index. For the analysis, only seasonality
+effects are included as they yield the most realistic predictions.
+
+In Figures X and Y, outliers are removed. Outliers are defined as
+administrative units which, at any given month, have a z-score greater
+than 4 for the relative intensity of movement. Furthermore, a
+Savitzky-Golay filter is applied to smooth the time series data, using a
+length of 4 units for the filter window and order 2 polynomials to fit
+the samples.
+
 \section{References}\label{references}

### 2024-01-19 manuscript/manuscript.qmd
rm old versions

@@
 <!--# Francisco to write this -->
+
+Make it general
+
+• Context and importance: Focus on urban spaces, livability, resilience and recov- ery - links to public health and climate change\
+• Previous work\
+• Gap: focus on medium - long term impacts of COVID-19 / recovery - explain why focus on South America (global south could be Africa too, or less developed countries in Asia).\
+• Aim:
 
@@
 
-To capture population movements, we used anonymised aggregate mobile
-phone location data from Meta users for Argentina, Colombia, Chile and Mexico, covering a 24-month period from April 2020 to March 2022.
+To capture population movements, we used anonymised aggregate mobile phone location data from Meta users for Argentina, Colombia, Chile and Mexico, covering a 24-month period from April 2020 to March 2022.
 We used the dataset Facebook Movements created by Meta and accessed through their Data for Good Initiative (https://dataforgood.facebook.com).
 The data are built from Facebook app users who have the location services setting turned on on their smartphone.
-Prior to releasing the datasets, Meta ensures privacy and anonymity by removing personal information and applying privacy-preserving techniques (Maas et al.
-2019).
+Prior to releasing the datasets, Meta ensures privacy and anonymity by removing personal information and applying privacy-preserving techniques (Maas et al. 2019).
 Small-count dropping is one of these techniques.
 A data entry is removed if the population or movement count for an area is lower than 10.
-The removal of
-small counts may mean that population counts in small sparsely populated areas are not captured.
+The removal of small counts may mean that population counts in small sparsely populated areas are not captured.
 A second technique consists in adding a small undisclosed amount of random noise to ensure that it is not possible to ascertain precise, true counts for sparsely populated locations.
@@
 The Facebook Movements dataset offers information on the total number of Facebook users moving between and within spatial units in the form of origin-destination matrices.
-The data is temporally aggregated into three daily 8-hour time windows (i.e.
-00:00-08:00, 08:00-16:00 and 16:00-00:00).
+The data is temporally aggregated into three daily 8-hour time windows (i.e. 00:00-08:00, 08:00-16:00 and 16:00-00:00).
 The dataset includes a baseline capturing the number of movements before COVID-19 based on a 45-day period ending on March 10th 2020.
 The baseline is computed using an average for the same time of the day and day of the week in the period preceding March 10th.
-For instance,
-the baseline for Monday 00:00-08:00 time window is obtained by averaging over data collected on Mondays from 00:00 to 8:00 for the 45-day period.
-Details about the baseline can be found in Maas et al.
-(2019).
+For instance, the baseline for Monday 00:00-08:00 time window is obtained by averaging over data collected on Mondays from 00:00 to 8:00 for the 45-day period.
+Details about the baseline can be found in Maas et al. (2019).
 The Bing Maps Tile System developed by Microsoft (Microsoft) is used a spatial framework to organise the data.
@@
 The data that we used are spatially aggregated into Bing tile levels 13.
-That is about 4.9 x
-4.9km at the Equator (Maas et al. 2019).
+That is about 4.9 x 4.9km at the Equator (Maas et al. 2019).
 

### 2024-01-19 manuscript/manuscript.qmd
add text for intro

--- 2024-01-18 manuscript/manuscript.qmd
@@
 
+<!--# Are sections supposed to be numbered? -->
+
 # Introduction {#sec-intro}
@@
 <!--# Francisco to write this -->
+
+Spatial human mobility is key to creating sustainable, liveable and inclusive cities.
+At the societal level, spatial mobility enables the transfer of knowledge, skills and labour to places they are needed.
+Spatial mobility also shapes service and transport demand across urban spaces, and enables the monitoring and control of transmissible diseases.
+At the individual level, spatial mobility enables people to access and achieve opportunities and aspirations.
+Understanding spatial human mobility is thus important to supporting appropriate policy responses to address societal challenges relating to carbon emissions, urban planning, service delivery, public health, disaster management and transport.
+
+The COVID-19 pandemic resulted in a notable reduction in mobility within cities.
+Nonpharmaceutical interventions to contain the spread of COVID‐19 reduced overall levels of mobility.
+Especially during lockdowns, mobility recorded reductions in the frequency, distance and time of trips in cities across the globe.
+Rises in teleworking, online schooling and remote shopping activity reduced the need to travel for work, education, shopping and leisure.
+Coupled with fears of crowded public spaces, nonpharmaceutical interventions prompted more geographically localised mobility patterns.
+Reductions in mobility were however socioeconomically highly unequal, with more affluent households recording greater drops in mobility levels.
+Wealthier individuals can afford to move less as they are more likely to work on jobs that can be done remotely, while individuals from more deprived background tend to engage in face-to-face activities requiring physical presence.
+
+* Context and importance of human mobility: public health , climate change\
+* The pandemic resulted in major changes in human mobility - unequal impacts across socioeconomic groups\
+* Gap 1: Little work on understanding the long-term changes in mobility - recovery\
+* Gap 2: little work on less developed countries - where inequalities are more pronounced\
+* Aim - focus on South American countries as case study.
 
@@
 We assume that this relationship is linear and we use a linear regression to estimate the slope and intercept characterising the line of best fit.
-This is shown for April 2020 and March 2022, in the right-hand side panels of Figure X.
+This is shown for April 2020 and March 2022 in the right-hand side panels of Figure X.
 After obtaining the slope and intercept for every month, we are able to plot the evolution of these parameters for both the observed and forecasted data, as displayed on the left-hand-side panels of the same Figure.
@@
 
+<!--# would sell a bit more why Prophet - what is new and can done "better" in comparison to existing approaches - and include citation to [the paper](https://doi.org/10.7287/peerj.preprints.3190v2) -->
+
 For the countries included in the analysis, the movement data is available until March 2022.

### 2024-01-19 manuscript/manuscript.tex
add text for intro

@@
 \makeatletter
+\makeatother
+\makeatletter
+\makeatother
+\makeatletter
 \@ifpackageloaded{caption}{}{\usepackage{caption}}
@@
 \makeatletter
+\@ifpackageloaded{caption}{}{\usepackage{caption}}
+\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
 \makeatother
 \makeatletter
-\@ifpackageloaded{caption}{}{\usepackage{caption}}
-\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
+\@ifpackageloaded{tcolorbox}{}{\usepackage[skins,breakable]{tcolorbox}}
+\makeatother
+\makeatletter
+\@ifundefined{shadecolor}{\definecolor{shadecolor}{rgb}{.97, .97, .97}}
+\makeatother
+\makeatletter
+\makeatother
+\makeatletter
 \makeatother
@@
 \fi
-\usepackage{bookmark}
-
+\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
 \IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
@@
 \end{abstract}
+\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[enhanced, borderline west={3pt}{0pt}{shadecolor}, frame hidden, boxrule=0pt, interior hidden, breakable, sharp corners]}{\end{tcolorbox}}\fi
 
-\section{Introduction}\label{sec-intro}
+\hypertarget{sec-intro}{%
+\section{Introduction}\label{sec-intro}}
 
-\section{Results}\label{sec-results}
+Spatial human mobility is key to creating sustainable, liveable and
+inclusive cities. At the societal level, spatial mobility enables the
+transfer of knowledge, skills and labour to places they are needed.
+Spatial mobility also shapes service and transport demand across urban
+spaces, and enables the monitoring and control of transmissible
+diseases. At the individual level, spatial mobility enables people to
+access and achieve opportunities and aspirations. Understanding spatial
+human mobility is thus important to supporting appropriate policy
+responses to address societal challenges relating to carbon emissions,
+urban planning, service delivery, public health, disaster management and
+transport.
+
+The COVID-19 pandemic resulted in a notable reduction in mobility within
+cities. Nonpharmaceutical interventions to contain the spread of
+COVID‐19 reduced overall levels of mobility. Especially during
+lockdowns, mobility recorded reductions in the frequency, distance and
+time of trips in cities across the globe. Rises in teleworking, online
+schooling and remote shopping activity reduced the need to travel for
+work, education, shopping and leisure. Coupled with fears of crowded
+public spaces, nonpharmaceutical interventions prompted more
+geographically localised mobility patterns. Reductions in mobility were
+however socioeconomically highly unequal, with more affluent households
+recording greater drops in mobility levels. Wealthier individuals can
+afford to move less as they are more likely to work on jobs that can be
+done remotely, while individuals from more deprived background tend to
+engage in face-to-face activities requiring physical presence.
+
+\begin{itemize}
+\tightlist
+\item
+  Context and importance of human mobility: public health , climate
+  change\\
+\item
+  The pandemic resulted in major changes in human mobility - unequal
+  impacts across socioeconomic groups\\
+\item
+  Gap 1: Little work on understanding the long-term changes in mobility
+  - recovery\\
+\item
+  Gap 2: little work on less developed countries - where inequalities
+  are more pronounced\\
+\item
+  Aim - focus on South American countries as case study.
+\end{itemize}
+
+\hypertarget{sec-results}{%
+\section{Results}\label{sec-results}}
 
@@
 
+\hypertarget{the-impact-of-covid-19-on-patterns-of-urban-mobility}{%
 \subsection{The impact of COVID-19 on patterns of urban
-mobility}\label{the-impact-of-covid-19-on-patterns-of-urban-mobility}
+mobility}\label{the-impact-of-covid-19-on-patterns-of-urban-mobility}}
 
@@
 
+\hypertarget{socioeconomic-deprivation-and-recovery-of-urban-mobility}{%
 \subsection{Socioeconomic deprivation and recovery of urban
-mobility}\label{socioeconomic-deprivation-and-recovery-of-urban-mobility}
+mobility}\label{socioeconomic-deprivation-and-recovery-of-urban-mobility}}
 
@@
 regression to estimate the slope and intercept characterising the line
-of best fit. This is shown for April 2020 and March 2022, in the
+of best fit. This is shown for April 2020 and March 2022 in the
 right-hand side panels of Figure X. After obtaining the slope and
@@
 
-\section{Discussion}\label{sec-discussion}
+\hypertarget{sec-discussion}{%
+\section{Discussion}\label{sec-discussion}}
 
-\section{Methods}\label{sec-methods}
+\hypertarget{sec-methods}{%
+\section{Methods}\label{sec-methods}}
 
+\hypertarget{meta-facebook-movement-data}{%
 \subsection{Meta-Facebook movement
-data}\label{meta-facebook-movement-data}
+data}\label{meta-facebook-movement-data}}
 
@@
 
+\hypertarget{spatiotemporal-data-aggregation}{%
 \subsection{Spatiotemporal data
-aggregation}\label{spatiotemporal-data-aggregation}
+aggregation}\label{spatiotemporal-data-aggregation}}
 
@@
 
-\subsection{Time series analysis}\label{time-series-analysis}
+\hypertarget{time-series-analysis}{%
+\subsection{Time series analysis}\label{time-series-analysis}}
 
@@
 
-\section{References}\label{references}
+\hypertarget{references}{%
+\section{References}\label{references}}
 

### 2024-01-20 manuscript/manuscript.qmd
update intro

@@
 Coupled with fears of crowded public spaces, nonpharmaceutical interventions prompted more geographically localised mobility patterns.
-Reductions in mobility were however socioeconomically highly unequal, with more affluent households recording greater drops in mobility levels.
-Wealthier individuals can afford to move less as they are more likely to work on jobs that can be done remotely, while individuals from more deprived background tend to engage in face-to-face activities requiring physical presence.
 
-* Context and importance of human mobility: public health , climate change\
-* The pandemic resulted in major changes in human mobility - unequal impacts across socioeconomic groups\
-* Gap 1: Little work on understanding the long-term changes in mobility - recovery\
-* Gap 2: little work on less developed countries - where inequalities are more pronounced\
-* Aim - focus on South American countries as case study.
+However, reductions in mobility levels were highly unequal reflecting existing socioeconomic inequalities.
+In most countries, affluent individuals tended to record the greatest drops in mobility levels as they are predominantly employed in knowledge-intensive jobs which can be done fully or partly remotely.
+During the COVID-19 pandemic, the adoption of remote work reduced the need of commuting for knowledge-intensive, non-public facing jobs.
+At the same time, individuals from less privileged socioeconomic backgrounds displayed less pronounced declines mirroring the nature of their work requiring public-facing, face-to-face interaction, and thus a requirement for daily work commutes.
+
+Thus, while a growing body of empirical evidence has contributed to advancing our understanding of the impacts of the COVID-19 pandemic on spatial mobility within cities, existing research has focused on more developed countries and the immediate effects of the pandemic during 2020.
+Less is known about the longer term patterns of resilience in urban mobility in less developed countries.
+Assessing the extent to which the intensity of movement has returned back to pre-pandemic levels across socioeconomic groups is important to understand the potentially unequal impacts of hybrid working in the society.
+
+A key barrier to continously monitor changes in geographic mobility patterns in less developed countries during and post the COVID-19 pandemic has been the lack of suitable data.
+Traditionally census and survey data have been employed to analyse human mobility patterns in these countries.
+Yet, these data streams are not frequently updated and suffer from slow releases, with census data for example
+being collected over intervals of ten years in most countries.
+These data streams thus lack the temporal granularity to analyse population movements over short-time periods.
+Data resulting from social interactions on digital platforms has emerged as an unique source of information to capture human population movement at scale in less developed countries.
+Particularly location data from mobile phone applications have become a prominent source to sense patterns of human mobility at higher geographical and temporal resolution over in real time.
+
+We aim to
+
+-   Context and importance of human mobility: public health , climate change\
+-   The pandemic resulted in major changes in human mobility - unequal impacts across socioeconomic groups\
+-   Gap 1: Little work on understanding the long-term changes in mobility - recovery\
+-   Gap 2: little work on less developed countries - where inequalities are more pronounced\
+-   Aim - focus on South American countries as case study.
 

### 2024-01-20 manuscript/manuscript.tex
update intro

@@
 \usepackage{amsmath,amssymb}
+\usepackage{lmodern}
 \usepackage{iftex}
@@
   \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
-\fi
-\usepackage{lmodern}
-\ifPDFTeX\else  
-    % xetex/luatex font selection
 \fi
@@
 \makeatletter
-\@ifpackageloaded{tcolorbox}{}{\usepackage[skins,breakable]{tcolorbox}}
+\@ifpackageloaded{tcolorbox}{}{\usepackage[many]{tcolorbox}}
 \makeatother
@@
 \@ifundefined{shadecolor}{\definecolor{shadecolor}{rgb}{.97, .97, .97}}
-\makeatother
-\makeatletter
 \makeatother
@@
 \end{abstract}
-\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[enhanced, borderline west={3pt}{0pt}{shadecolor}, frame hidden, boxrule=0pt, interior hidden, breakable, sharp corners]}{\end{tcolorbox}}\fi
+\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[boxrule=0pt, interior hidden, sharp corners, breakable, borderline west={3pt}{0pt}{shadecolor}, frame hidden, enhanced]}{\end{tcolorbox}}\fi
 
@@
 public spaces, nonpharmaceutical interventions prompted more
-geographically localised mobility patterns. Reductions in mobility were
-however socioeconomically highly unequal, with more affluent households
-recording greater drops in mobility levels. Wealthier individuals can
-afford to move less as they are more likely to work on jobs that can be
-done remotely, while individuals from more deprived background tend to
-engage in face-to-face activities requiring physical presence.
+geographically localised mobility patterns.
+
+However, reductions in mobility levels were highly unequal reflecting
+existing socioeconomic inequalities. In most countries, affluent
+individuals tended to record the greatest drops in mobility levels as
+they are predominantly employed in knowledge-intensive jobs which can be
+done fully or partly remotely. During the COVID-19 pandemic, the
+adoption of remote work reduced the need of commuting for
+knowledge-intensive, non-public facing jobs. At the same time,
+individuals from less privileged socioeconomic backgrounds displayed
+less pronounced declines mirroring the nature of their work requiring
+public-facing, face-to-face interaction, and thus a requirement for
+daily work commutes.
+
+Thus, while a growing body of empirical evidence has contributed to
+advancing our understanding of the impacts of the COVID-19 pandemic on
+spatial mobility within cities, existing research has focused on more
+developed countries and the immediate effects of the pandemic during
+2020. Less is known about the longer term patterns of resilience in
+urban mobility in less developed countries. Assessing the extent to
+which the intensity of movement has returned back to pre-pandemic levels
+across socioeconomic groups is important to understand the potentially
+unequal impacts of hybrid working in the society.
+
+A key barrier to continously monitor changes in geographic mobility
+patterns in less developed countries during and post the COVID-19
+pandemic has been the lack of suitable data. Traditionally census and
+survey data have been employed to analyse human mobility patterns in
+these countries. Yet, these data streams are not frequently updated and
+suffer from slow releases, with census data for example being collected
+over intervals of ten years in most countries. These data streams thus
+lack the temporal granularity to analyse population movements over
+short-time periods. Data resulting from social interactions on digital
+platforms has emerged as an unique source of information to capture
+human population movement at scale in less developed countries.
+Particularly location data from mobile phone applications have become a
+prominent source to sense patterns of human mobility at higher
+geographical and temporal resolution over in real time.
+
+We aim to
 

### 2024-01-20 manuscript/manuscript.tex
update intro

@@
 \end{abstract}
-\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[boxrule=0pt, interior hidden, sharp corners, breakable, borderline west={3pt}{0pt}{shadecolor}, frame hidden, enhanced]}{\end{tcolorbox}}\fi
+\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[enhanced, interior hidden, sharp corners, boxrule=0pt, borderline west={3pt}{0pt}{shadecolor}, frame hidden, breakable]}{\end{tcolorbox}}\fi
 
@@
 
-\hypertarget{the-impact-of-covid-19-on-patterns-of-urban-mobility}{%
-\subsection{The impact of COVID-19 on patterns of urban
-mobility}\label{the-impact-of-covid-19-on-patterns-of-urban-mobility}}
+\hypertarget{the-impact-of-covid-19-on-urban-mobility}{%
+\subsection{The impact of COVID-19 on urban
+mobility}\label{the-impact-of-covid-19-on-urban-mobility}}
 

### 2024-01-22 manuscript/manuscript.qmd
update intro

@@
 Spatial mobility also shapes service and transport demand across urban spaces, and enables the monitoring and control of transmissible diseases.
-At the individual level, spatial mobility enables people to access and achieve opportunities and aspirations.
+At the individual level, mobility enables people to access and achieve opportunities and aspirations in space.
 Understanding spatial human mobility is thus important to supporting appropriate policy responses to address societal challenges relating to carbon emissions, urban planning, service delivery, public health, disaster management and transport.
@@
 
-A key barrier to continously monitor changes in geographic mobility patterns in less developed countries during and post the COVID-19 pandemic has been the lack of suitable data.
+A key barrier to continuously monitor changes in geographic mobility patterns in less developed countries during and post the COVID-19 pandemic has been the lack of suitable data.
 Traditionally census and survey data have been employed to analyse human mobility patterns in these countries.
-Yet, these data streams are not frequently updated and suffer from slow releases, with census data for example
-being collected over intervals of ten years in most countries.
-These data streams thus lack the temporal granularity to analyse population movements over short-time periods.
-Data resulting from social interactions on digital platforms has emerged as an unique source of information to capture human population movement at scale in less developed countries.
-Particularly location data from mobile phone applications have become a prominent source to sense patterns of human mobility at higher geographical and temporal resolution over in real time.
+Yet, these data streams are not frequently updated and suffer from slow releases, with census data for example being collected over intervals of ten years in most countries.
+Traditional data streams thus lack the temporal granularity to analyse population movements over short-time periods and offer an up-to-date representation of the urban mobility system.
+Data resulting from social interactions on digital platforms have emerged as an unique source of information to deliver this representation and capture human population movement in less developed countries at scale.
+Particularly location data from mobile phone applications have become a prominent source to sense patterns of human mobility at higher geographical and temporal resolution in real time.
 
@@
 
-## The impact of COVID-19 on patterns of urban mobility
+## The impact of COVID-19 on urban mobility
 

### 2024-01-22 manuscript/manuscript.tex
update intro

@@
 \end{abstract}
-\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[enhanced, interior hidden, sharp corners, boxrule=0pt, borderline west={3pt}{0pt}{shadecolor}, frame hidden, breakable]}{\end{tcolorbox}}\fi
+\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[borderline west={3pt}{0pt}{shadecolor}, frame hidden, interior hidden, enhanced, breakable, sharp corners, boxrule=0pt]}{\end{tcolorbox}}\fi
 
@@
 spaces, and enables the monitoring and control of transmissible
-diseases. At the individual level, spatial mobility enables people to
-access and achieve opportunities and aspirations. Understanding spatial
+diseases. At the individual level, mobility enables people to access and
+achieve opportunities and aspirations in space. Understanding spatial
 human mobility is thus important to supporting appropriate policy
@@
 
-A key barrier to continously monitor changes in geographic mobility
+A key barrier to continuously monitor changes in geographic mobility
 patterns in less developed countries during and post the COVID-19
@@
 suffer from slow releases, with census data for example being collected
-over intervals of ten years in most countries. These data streams thus
-lack the temporal granularity to analyse population movements over
-short-time periods. Data resulting from social interactions on digital
-platforms has emerged as an unique source of information to capture
-human population movement at scale in less developed countries.
-Particularly location data from mobile phone applications have become a
-prominent source to sense patterns of human mobility at higher
-geographical and temporal resolution over in real time.
+over intervals of ten years in most countries. Traditional data streams
+thus lack the temporal granularity to analyse population movements over
+short-time periods and offer an up-to-date representation of the urban
+mobility system. Data resulting from social interactions on digital
+platforms have emerged as an unique source of information to deliver
+this representation and capture human population movement in less
+developed countries at scale. Particularly location data from mobile
+phone applications have become a prominent source to sense patterns of
+human mobility at higher geographical and temporal resolution in real
+time.
 

### 2024-01-22 manuscript/manuscript.qmd
update to intro

@@
 
-A key barrier to continuously monitor changes in geographic mobility patterns in less developed countries during and post the COVID-19 pandemic has been the lack of suitable data.
+A key barrier to monitor changes in geographic mobility patterns in less developed countries during and post the COVID-19 pandemic has been the lack of suitable data.
 Traditionally census and survey data have been employed to analyse human mobility patterns in these countries.
@@
 
-We aim to
+Drawing on a dataset of 213 million observations from Meta-Facebook users' mobile location data, we aim to assess socioeconomic differences in the extent and persistence of decline in urban mobility in Argentina, Chile, Colombia and Mexico during and after the COVID-19 pandemic from March 2020 to March 2023.
+We use Meta-Facebook observed data to measure origin-destination flows from March 2020 to May 2022, and Meta Prophet time-series forecasting machine learning algorithm to predict origin-destination flows from June 2022 to March 2023.
+<!--# Perhaps say how predictions have been validated --> We use Functional Urban Areas (FUAs) boundaries from the Global Human Settlement Layer (GHSL) to define urban areas, and the Global Gridded Relative Deprivation Index (GRDI) developed by NASA’s Socioeconomic Data and Applications Centre (SEDAC) from sociodemographic and satellite data inputs.
+Building on existing evidence, we hypothesised that (1) urban mobility has recovered returning to pre-pandemic levels of movement as nonpharmaceuthical restrictions were lifted; and, that (2) socioeconomic differences in urban mobility have endured the pandemic reflecting deep societal inequalities as knowledge-intensive business adopt hybrid working.
+<!--# Once we have a more definitive version of the results, highlight our novel contributions here.
+-->
 

### 2024-01-22 manuscript/manuscript.tex
update to intro

@@
 \usepackage{amsmath,amssymb}
-\usepackage{lmodern}
 \usepackage{iftex}
@@
   \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
+\fi
+\usepackage{lmodern}
+\ifPDFTeX\else  
+    % xetex/luatex font selection
 \fi
@@
 \makeatletter
-\@ifpackageloaded{tcolorbox}{}{\usepackage[many]{tcolorbox}}
+\@ifpackageloaded{tcolorbox}{}{\usepackage[skins,breakable]{tcolorbox}}
 \makeatother
@@
 \@ifundefined{shadecolor}{\definecolor{shadecolor}{rgb}{.97, .97, .97}}
+\makeatother
+\makeatletter
 \makeatother
@@
 \end{abstract}
-\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[borderline west={3pt}{0pt}{shadecolor}, frame hidden, interior hidden, enhanced, breakable, sharp corners, boxrule=0pt]}{\end{tcolorbox}}\fi
+\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[sharp corners, interior hidden, enhanced, breakable, borderline west={3pt}{0pt}{shadecolor}, boxrule=0pt, frame hidden]}{\end{tcolorbox}}\fi
 
@@
 
-A key barrier to continuously monitor changes in geographic mobility
-patterns in less developed countries during and post the COVID-19
-pandemic has been the lack of suitable data. Traditionally census and
-survey data have been employed to analyse human mobility patterns in
-these countries. Yet, these data streams are not frequently updated and
-suffer from slow releases, with census data for example being collected
-over intervals of ten years in most countries. Traditional data streams
-thus lack the temporal granularity to analyse population movements over
-short-time periods and offer an up-to-date representation of the urban
-mobility system. Data resulting from social interactions on digital
-platforms have emerged as an unique source of information to deliver
-this representation and capture human population movement in less
-developed countries at scale. Particularly location data from mobile
-phone applications have become a prominent source to sense patterns of
-human mobility at higher geographical and temporal resolution in real
-time.
+A key barrier to monitor changes in geographic mobility patterns in less
+developed countries during and post the COVID-19 pandemic has been the
+lack of suitable data. Traditionally census and survey data have been
+employed to analyse human mobility patterns in these countries. Yet,
+these data streams are not frequently updated and suffer from slow
+releases, with census data for example being collected over intervals of
+ten years in most countries. Traditional data streams thus lack the
+temporal granularity to analyse population movements over short-time
+periods and offer an up-to-date representation of the urban mobility
+system. Data resulting from social interactions on digital platforms
+have emerged as an unique source of information to deliver this
+representation and capture human population movement in less developed
+countries at scale. Particularly location data from mobile phone
+applications have become a prominent source to sense patterns of human
+mobility at higher geographical and temporal resolution in real time.
 
-We aim to
+Drawing on a dataset of 213 million observations from Meta-Facebook
+users' mobile location data, we aim to assess socioeconomic differences
+in the extent and persistence of decline in urban mobility in Argentina,
+Chile, Colombia and Mexico during and after the COVID-19 pandemic from
+March 2020 to March 2023. We use Meta-Facebook observed data to measure
+origin-destination flows from March 2020 to May 2022, and Meta Prophet
+time-series forecasting machine learning algorithm to predict
+origin-destination flows from June 2022 to March 2023. We use Functional
+Urban Areas (FUAs) boundaries from the Global Human Settlement Layer
+(GHSL) to define urban areas, and the Global Gridded Relative
+Deprivation Index (GRDI) developed by NASA's Socioeconomic Data and
+Applications Centre (SEDAC) from sociodemographic and satellite data
+inputs. Building on existing evidence, we hypothesised that (1) urban
+mobility has recovered returning to pre-pandemic levels of movement as
+nonpharmaceuthical restrictions were lifted; and, that (2) socioeconomic
+differences in urban mobility have endured the pandemic reflecting deep
+societal inequalities as knowledge-intensive business adopt hybrid
+working.
 

### 2024-01-23 manuscript/manuscript.tex
figure-prediction

@@
 \makeatletter
-\makeatother
-\makeatletter
-\makeatother
-\makeatletter
 \@ifpackageloaded{caption}{}{\usepackage{caption}}
@@
 \makeatletter
+\makeatother
+\makeatletter
 \@ifpackageloaded{caption}{}{\usepackage{caption}}
 \@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
-\makeatother
-\makeatletter
-\@ifpackageloaded{tcolorbox}{}{\usepackage[skins,breakable]{tcolorbox}}
-\makeatother
-\makeatletter
-\@ifundefined{shadecolor}{\definecolor{shadecolor}{rgb}{.97, .97, .97}}
-\makeatother
-\makeatletter
-\makeatother
-\makeatletter
 \makeatother
@@
 \fi
-\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
+\usepackage{bookmark}
+
 \IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
@@
 \end{abstract}
-\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[sharp corners, interior hidden, enhanced, breakable, borderline west={3pt}{0pt}{shadecolor}, boxrule=0pt, frame hidden]}{\end{tcolorbox}}\fi
 
-\hypertarget{sec-intro}{%
-\section{Introduction}\label{sec-intro}}
+\section{Introduction}\label{sec-intro}
 
@@
 
-\hypertarget{sec-results}{%
-\section{Results}\label{sec-results}}
+\section{Results}\label{sec-results}
 
@@
 
-\hypertarget{the-impact-of-covid-19-on-urban-mobility}{%
 \subsection{The impact of COVID-19 on urban
-mobility}\label{the-impact-of-covid-19-on-urban-mobility}}
+mobility}\label{the-impact-of-covid-19-on-urban-mobility}
 
@@
 
-\hypertarget{socioeconomic-deprivation-and-recovery-of-urban-mobility}{%
 \subsection{Socioeconomic deprivation and recovery of urban
-mobility}\label{socioeconomic-deprivation-and-recovery-of-urban-mobility}}
+mobility}\label{socioeconomic-deprivation-and-recovery-of-urban-mobility}
 
@@
 
-\hypertarget{sec-discussion}{%
-\section{Discussion}\label{sec-discussion}}
+\section{Discussion}\label{sec-discussion}
 
-\hypertarget{sec-methods}{%
-\section{Methods}\label{sec-methods}}
+\section{Methods}\label{sec-methods}
 
-\hypertarget{meta-facebook-movement-data}{%
 \subsection{Meta-Facebook movement
-data}\label{meta-facebook-movement-data}}
+data}\label{meta-facebook-movement-data}
 
@@
 
-\hypertarget{spatiotemporal-data-aggregation}{%
 \subsection{Spatiotemporal data
-aggregation}\label{spatiotemporal-data-aggregation}}
+aggregation}\label{spatiotemporal-data-aggregation}
 
@@
 
-\hypertarget{time-series-analysis}{%
-\subsection{Time series analysis}\label{time-series-analysis}}
+\subsection{Time series analysis}\label{time-series-analysis}
 
@@
 
-\hypertarget{references}{%
-\section{References}\label{references}}
+\section{References}\label{references}
 

### 2024-01-27 manuscript/manuscript.qmd
figure-prediction
manuscript

@@
 abstract: |
-  The COVID‐19 pandemic has impacted the national systems of population movement around the world. 
+  The movement of people between locations is crucial for sustainable and inclusive cities, facilitating the exchange of knowledge and resources.
+Due to restrictions, lockdowns and fear of crowded places, the COVID-19 pandemic significantly reduced the number and extent of people's movements.
+Existing research suggests that the impact of the pandemic on mobility was unequal across locations and socioeconomic groups.
+However, evidence and analysis of long-term changes in mobility is scarce, especially in less developed countries.
+In this study we use location data from Meta-Facebook users to analyse patterns of urban mobility in X Functional Urban Areas from South American countries.
+Findings reveal a general decrease in mobility during the pandemic, with gradual recovery trends up to April 2023.
+However, socioeconomic disparities persist through the period of analysis, with deprived areas experiencing smaller initial losses in mobility.
+Inequalities in mobility levels across socioeconomic groups generally diminish over time, although they persist in all the countries included in the analysis.
+This research highlights the importance of understanding the long-term effects of the pandemic on mobility and the need for equitable policy responses to address societal challenges in urban areas.
 format: 
@@
 
-<!--# Are sections supposed to be numbered? -->
-
 # Introduction {#sec-intro}
 
-<!--# Francisco to write this -->
-
-Spatial human mobility is key to creating sustainable, liveable and inclusive cities.
+Spatial human mobility is key to creating sustainable, livable and inclusive cities.
 At the societal level, spatial mobility enables the transfer of knowledge, skills and labour to places they are needed.
@@
 
-The COVID-19 pandemic resulted in a notable reduction in mobility within cities.
-Nonpharmaceutical interventions to contain the spread of COVID‐19 reduced overall levels of mobility.
-Especially during lockdowns, mobility recorded reductions in the frequency, distance and time of trips in cities across the globe.
-Rises in teleworking, online schooling and remote shopping activity reduced the need to travel for work, education, shopping and leisure.
-Coupled with fears of crowded public spaces, nonpharmaceutical interventions prompted more geographically localised mobility patterns.
+The COVID-19 pandemic resulted in a notable decrease in mobility, particularly in cities.
+This decrease in the overall levels of mobility was prompted by nonpharmaceutical interventions to contain the spread of COVID‐19, coupled with general fears of visiting crowded public spaces.Especially during lockdowns, mobility recorded reductions in the frequency, distance and time of trips in cities across the globe.
+Higher engagement with remote working, online schooling and shopping activity reduced the need to travel for work, education, shopping and leisure, hence giving rise to more geographically localised mobility patterns.
 
@@
 <!--# Perhaps say how predictions have been validated --> We use Functional Urban Areas (FUAs) boundaries from the Global Human Settlement Layer (GHSL) to define urban areas, and the Global Gridded Relative Deprivation Index (GRDI) developed by NASA’s Socioeconomic Data and Applications Centre (SEDAC) from sociodemographic and satellite data inputs.
-Building on existing evidence, we hypothesised that (1) urban mobility has recovered returning to pre-pandemic levels of movement as nonpharmaceuthical restrictions were lifted; and, that (2) socioeconomic differences in urban mobility have endured the pandemic reflecting deep societal inequalities as knowledge-intensive business adopt hybrid working.
-<!--# Once we have a more definitive version of the results, highlight our novel contributions here.
--->
+Building on existing evidence, we hypothesised that (1) urban mobility has recovered returning to pre-pandemic levels of movement as nonpharmaceuthical restrictions were lifted; and (2) socioeconomic differences in urban mobility have endured the pandemic reflecting deep societal inequalities as knowledge-intensive businesses adopt hybrid working.
+<!--# Once we have a more definitive version of the results, highlight our novel contributions here. -->
 
@@
 
-The evolution of the levels of movement was measured with respect to a baseline period prior to the pandemic as described in @sec-methods.
+The evolution of the levels of movement is measured with respect to a baseline period prior to the pandemic as described in @sec-methods.
 For the purposes of the analysis, we aggregate the raw movement data temporally into months and spatially into administrative units according to various GADM levels.
 The analysis focuses on administrative areas that are within the boundaries of functional urban areas as specified by the Global Human Settlement Layer.
-For each administrative area, we compute the Relative Deprivation Index based on data from NASA's Socioeconomic Data and Applications Centre (SEDAC).
-Figure X displays the administrative areas included in the study, coloured according to the Relative Index of Deprivation.
+For each administrative area, we compute the Relative Deprivation Index (RDI) based on data from NASA's Socioeconomic Data and Applications Centre (SEDAC).
+Figure 1 displays the administrative areas included in the study, coloured according to the Relative Index of Deprivation.
 Predictions about the evolution of the levels of movement are made using the Prophet forecasting procedure.
@@
 
-<!--# Add Figure X. -->
+![Administrative units within functional urban areas in Argentina, Chile, Colombia and Mexico, coloured according to the average Relative Deprivation Index.](figures/maps.pdf)
 
@@
 
-We analysed the evolution of the relative intensity of urban mobility with respect to a baseline period prior to the pandemic.
+We analyse the evolution of the relative intensity of urban mobility with respect to a baseline period prior to the pandemic.
 Specifically, movements covering a distance of at most 70 km are considered.
@@
 
-Figure X displays the patterns of recovery for the mobility levels in the administrative units belonging to functional urban areas in the countries of interest.
+Figure 2 displays the patterns of recovery for the mobility levels in the administrative units belonging to functional urban areas in the countries of interest.
 The three lines in each panel represent the mean levels of mobility for administrative units grouped into one of three terciles, according to their average relative deprivation index.
 
+![Patterns of recovery for urban mobility in administrative units from Argentina, Chile, Colombia and Mexico.](figures/prediction-rdi-band.pdf)
+
 Generally, there was a drop in the levels of mobility with respect to the baseline period in all four countries.
-This drop was especially large for Argentina, Chile and Colombia, with Mexico displaying a smaller decrease in the number of movements with respect to the baseline.
+This drop was especially large for Argentina, Chile and Colombia, with Mexico displaying a smaller decrease in the relative number of movements with respect to the baseline.
 Following the initial drop in movement, all four countries evolve towards the recovery of baseline levels of urban mobility, with a generally increasing trend.
-There are fluctuations from the general trend, which manifest differently for each country.
+There are, however, fluctuations from the general trend which manifest differently for each country.
 These fluctuations mirror each other in the case of Argentina and Colombia, where urban mobility sharply bounces back closer to pre-pandemic levels around July of 2020.
@@
 
-From Figure X, we observe that there is a consistent tendency in how administrative units with varying levels of deprivation were affected by the pandemic.
+From Figure 2, we observe that there is a consistent tendency in how administrative units with varying levels of deprivation were affected by the pandemic.
 For all four countries, we observe that the administrative units in the most deprived tercile are the ones that experienced the smallest loss in levels of mobility at the beginning of the pandemic.
@@
 We assume that this relationship is linear and we use a linear regression to estimate the slope and intercept characterising the line of best fit.
-This is shown for April 2020 and March 2022 in the right-hand side panels of Figure X.
+This is shown for April 2020 and March 2022 in the right-hand side panels of Figure 3.
 After obtaining the slope and intercept for every month, we are able to plot the evolution of these parameters for both the observed and forecasted data, as displayed on the left-hand-side panels of the same Figure.
+
+![Evolution of the parameters characterising the relationship between relative deprivation index and number of movements relative to the pre-pandemic baseline period.](figures/regression-evo-nobackground.pdf)
 
@@
 The slope captures the extent of differences in the level of urban mobility across administrative units with varying levels of socioeconomic deprivation.
-It can therefore be regarded as a measure of inequality.
+It can therefore be regarded as a measure of inequality in mobility patterns across socioeconomic groups.
 A slope equal to zero would mean that all administrative units display the same intensity of movement regardless of their socioeconomic deprivation levels.
@@
 The intercept of the relationship displays stronger patterns, which are consistent across the four countries.
-The intercept captures the urban mobility levels in administrative where relative deprivation is zero, or in other words, if there was no socioeconomic deprivation in an administrative area, the intercept would represent its level of mobility.
-The intercept was well below the baseline level at the early stages of the pandemic.
-As observed in Figure X, while there are some differences between countries in the evolution of the intercept, the general tendency is for the intercept to increase.
+The intercept estimates the urban mobility levels that would be observed in administrative units where relative deprivation is zero.
+The intercept was below the baseline level at the early stages of the pandemic.
+As observed in Figure 3, while there are some differences between countries in the evolution of the intercept, the general tendency is for the intercept to increase.
 While Argentina and Mexico reach values that are closer to the baseline towards the end of the forecast period, the intercept for Chile and Colombia remains lower.
-Therefore, if there were areas with no socioeconomic deprivation, we would have seen a recovery in the levels of mobility, although not quite back to baseline levels in the case of Chile and Colombia.
+Therefore, if there were areas with no socioeconomic deprivation, we would have seen a recovery in the levels of mobility, especially in Argentina and Mexico
 
@@
 
-<!--# Carmen to write this -->
+Using location data from Meta-Facebook users, our study aimed to examine the evolution of patterns of mobility across socioeconomic groups in functional urban areas from Argentina, Chile, Colombia and Mexico from April 2020 to March 2023, following the COVID-19 pandemic.
+We found a systematic decline in population movement levels during the implementation of non-pharmaceutical interventions in April 2020, with the largest reductions observed in the most affluent administrative units within functional urban areas from Argentina, Chile and Colombia.
+While mobility rebounded closer to pre-pandemic levels approximately two years later after COVID-19 restrictions eased, movement intensity remained below pre-pandemic levels in Chile and generally in areas where the relative deprivation index is lower.
+Furthermore, we found that at the beginning of the pandemic there were inequalities between socioeconomic groups in terms of the levels of urban mobility.
+While it has taken more than two years for Argentina and Mexico to gradually close the gap, inequalities persist in Chile and Colombia as of March 2023 according to our estimated data.
+
+<!--- Interpretation --->
+
+<!--- Limitations --->
+
+We focused the analysis on short-distance movements in urban areas, which are typically considered to capture local and routine mobility (Owen and Green 1992).
+However, due to the characteristics of the dataset, we are unable to distinguish movements of this nature from short-distance movements that involve a permanent change of place of residence.
+Future work could extend and validate our work identifying changes to moves relating permanent and
+temporary displacements, and triangulating alternative DFD and traditional data sources, such as censuses
+and representative surveys collecting information on residential location.
+Further research is also needed to establish the cause of the observed changes in the spatial patterns of
+internal population movement since the start of the COVID-19 pandemic.
+Understanding these causes can
+assist in anticipating long-term structural changes in the direction of mobility flows in the coming years.
+
+Additionally, accessing and working with DFD to measure human mobility and migration is challenging.
+DFD are known to suffer from biases and issues of statistical representation reflecting differences in technology usage and accessibility (Rowe 2023).
+While concerted efforts to develop data services and
+methodological frameworks to leverage on DFD exist in the Global North \[REF\], limited progress has been
+done in Latin America.
+National government and statistical agencies have a prime position to promote
+the creation of data sharing partnerships with private companies and create the digital infrastructure to
+ensure equal and secure access to DFD sources for the social good.
+With academia, these agencies could
+play a major role in promoting the creation of a methodological framework for the use and analysis of
+DFD, setting common definitions, metrics, and standard approaches to measure and correct data biases
+and validate data analyses.
+Location data have been proven key to monitor and develop appropriate policy
+responses to population displacement and movement during natural disasters, wars and epidemics (Green,
+Pollock, and Rowe 2021).
+We envision resources like those created by the European Commission Joint
+Research Centre (2022) and the UNSD (2019), identifying sources of non-traditional data and establishing
+methodological guidelines for the use of mobile phone data for official mobility statistics.
+Existing resources have a global character, we call for initiatives with a regional focus accounting for differences in the local landscape of data availability and digital technology usage.
 
 # Methods {#sec-methods}
-
-<!--# Carmen to write this -->
 

### 2024-01-27 manuscript/manuscript.tex
manuscript

--- 2024-01-22 manuscript/manuscript.tex
@@
 \begin{abstract}
-The COVID‐19 pandemic has impacted the national systems of population
-movement around the world.
+The movement of people between locations is crucial for sustainable and
+inclusive cities, facilitating the exchange of knowledge and resources.
+Due to restrictions, lockdowns and fear of crowded places, the COVID-19
+pandemic significantly reduced the number and extent of people's
+movements. Existing research suggests that the impact of the pandemic on
+mobility was unequal across locations and socioeconomic groups. However,
+evidence and analysis of long-term changes in mobility is scarce,
+especially in less developed countries. In this study we use location
+data from Meta-Facebook users to analyse patterns of urban mobility in X
+Functional Urban Areas from South American countries. Findings reveal a
+general decrease in mobility during the pandemic, with gradual recovery
+trends up to April 2023. However, socioeconomic disparities persist
+through the period of analysis, with deprived areas experiencing smaller
+initial losses in mobility. Inequalities in mobility levels across
+socioeconomic groups generally diminish over time, although they persist
+in all the countries included in the analysis. This research highlights
+the importance of understanding the long-term effects of the pandemic on
+mobility and the need for equitable policy responses to address societal
+challenges in urban areas.
 \end{abstract}
-\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[sharp corners, interior hidden, enhanced, breakable, borderline west={3pt}{0pt}{shadecolor}, boxrule=0pt, frame hidden]}{\end{tcolorbox}}\fi
+\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[enhanced, sharp corners, frame hidden, breakable, boxrule=0pt, interior hidden, borderline west={3pt}{0pt}{shadecolor}]}{\end{tcolorbox}}\fi
 
@@
 
-Spatial human mobility is key to creating sustainable, liveable and
+Spatial human mobility is key to creating sustainable, livable and
 inclusive cities. At the societal level, spatial mobility enables the
@@
 
-The COVID-19 pandemic resulted in a notable reduction in mobility within
-cities. Nonpharmaceutical interventions to contain the spread of
-COVID‐19 reduced overall levels of mobility. Especially during
-lockdowns, mobility recorded reductions in the frequency, distance and
-time of trips in cities across the globe. Rises in teleworking, online
-schooling and remote shopping activity reduced the need to travel for
-work, education, shopping and leisure. Coupled with fears of crowded
-public spaces, nonpharmaceutical interventions prompted more
-geographically localised mobility patterns.
+The COVID-19 pandemic resulted in a notable decrease in mobility,
+particularly in cities. This decrease in the overall levels of mobility
+was prompted by nonpharmaceutical interventions to contain the spread of
+COVID‐19, coupled with general fears of visiting crowded public
+spaces.Especially during lockdowns, mobility recorded reductions in the
+frequency, distance and time of trips in cities across the globe. Higher
+engagement with remote working, online schooling and shopping activity
+reduced the need to travel for work, education, shopping and leisure,
+hence giving rise to more geographically localised mobility patterns.
 
@@
 mobility has recovered returning to pre-pandemic levels of movement as
-nonpharmaceuthical restrictions were lifted; and, that (2) socioeconomic
+nonpharmaceuthical restrictions were lifted; and (2) socioeconomic
 differences in urban mobility have endured the pandemic reflecting deep
-societal inequalities as knowledge-intensive business adopt hybrid
+societal inequalities as knowledge-intensive businesses adopt hybrid
 working.
@@
 
-The evolution of the levels of movement was measured with respect to a
+The evolution of the levels of movement is measured with respect to a
 baseline period prior to the pandemic as described in
@@
 Layer. For each administrative area, we compute the Relative Deprivation
-Index based on data from NASA's Socioeconomic Data and Applications
-Centre (SEDAC). Figure X displays the administrative areas included in
-the study, coloured according to the Relative Index of Deprivation.
-Predictions about the evolution of the levels of movement are made using
-the Prophet forecasting procedure. Further details are provided in
-Section~\ref{sec-methods}.
+Index (RDI) based on data from NASA's Socioeconomic Data and
+Applications Centre (SEDAC). Figure 1 displays the administrative areas
+included in the study, coloured according to the Relative Index of
+Deprivation. Predictions about the evolution of the levels of movement
+are made using the Prophet forecasting procedure. Further details are
+provided in Section~\ref{sec-methods}.
+
+\begin{figure}
+
+{\centering \includegraphics{figures/maps.pdf}
+
+}
+
+\caption{Administrative units within functional urban areas in
+Argentina, Chile, Colombia and Mexico, coloured according to the average
+Relative Deprivation Index.}
+
+\end{figure}
 
@@
 
-We analysed the evolution of the relative intensity of urban mobility
+We analyse the evolution of the relative intensity of urban mobility
 with respect to a baseline period prior to the pandemic. Specifically,
@@
 
-Figure X displays the patterns of recovery for the mobility levels in
+Figure 2 displays the patterns of recovery for the mobility levels in
 the administrative units belonging to functional urban areas in the
@@
 
+\begin{figure}
+
+{\centering \includegraphics{figures/prediction-rdi-band.pdf}
+
+}
+
+\caption{Patterns of recovery for urban mobility in administrative units
+from Argentina, Chile, Colombia and Mexico.}
+
+\end{figure}
+
 Generally, there was a drop in the levels of mobility with respect to
@@
 large for Argentina, Chile and Colombia, with Mexico displaying a
-smaller decrease in the number of movements with respect to the
+smaller decrease in the relative number of movements with respect to the
 baseline. Following the initial drop in movement, all four countries
 evolve towards the recovery of baseline levels of urban mobility, with a
-generally increasing trend. There are fluctuations from the general
-trend, which manifest differently for each country. These fluctuations
-mirror each other in the case of Argentina and Colombia, where urban
-mobility sharply bounces back closer to pre-pandemic levels around July
-of 2020. Chile and Mexico display more progressive patterns of recovery,
-although Chile never reaches baseline levels. These fluctuations are
-unique to each country and can be attributed to local factors such as
-the effects of seasonality or the different stringency measures imposed
-by the national governments during the pandemic.
+generally increasing trend. There are, however, fluctuations from the
+general trend which manifest differently for each country. These
+fluctuations mirror each other in the case of Argentina and Colombia,
+where urban mobility sharply bounces back closer to pre-pandemic levels
+around July of 2020. Chile and Mexico display more progressive patterns
+of recovery, although Chile never reaches baseline levels. These
+fluctuations are unique to each country and can be attributed to local
+factors such as the effects of seasonality or the different stringency
+measures imposed by the national governments during the pandemic.
 
-From Figure X, we observe that there is a consistent tendency in how
+From Figure 2, we observe that there is a consistent tendency in how
 administrative units with varying levels of deprivation were affected by
@@
 of best fit. This is shown for April 2020 and March 2022 in the
-right-hand side panels of Figure X. After obtaining the slope and
+right-hand side panels of Figure 3. After obtaining the slope and
 intercept for every month, we are able to plot the evolution of these
@@
 the left-hand-side panels of the same Figure.
+
+\begin{figure}
+
+{\centering \includegraphics{figures/regression-evo-nobackground.pdf}
+
+}
+
+\caption{Evolution of the parameters characterising the relationship
+between relative deprivation index and number of movements relative to
+the pre-pandemic baseline period.}
+
+\end{figure}
 
@@
 administrative units with varying levels of socioeconomic deprivation.
-It can therefore be regarded as a measure of inequality. A slope equal
-to zero would mean that all administrative units display the same
-intensity of movement regardless of their socioeconomic deprivation
-levels. Given the patterns observed in Argentina, Colombia and Mexico,
-we find that at the beginning of the pandemic there were notable
-inequalities between socioeconomic groups in terms of the levels of
-urban mobility. While it has taken more than two years for Argentina and
-Mexico to close the gap (their slope is close to zero from spring 2022),
-inequalities persist in Chile and Colombia as of March 2023.
+It can therefore be regarded as a measure of inequality in mobility
+patterns across socioeconomic groups. A slope equal to zero would mean
+that all administrative units display the same intensity of movement
+regardless of their socioeconomic deprivation levels. Given the patterns
+observed in Argentina, Colombia and Mexico, we find that at the
+beginning of the pandemic there were notable inequalities between
+socioeconomic groups in terms of the levels of urban mobility. While it
+has taken more than two years for Argentina and Mexico to close the gap
+(their slope is close to zero from spring 2022), inequalities persist in
+Chile and Colombia as of March 2023.
 
 The intercept of the relationship displays stronger patterns, which are
-consistent across the four countries. The intercept captures the urban
-mobility levels in administrative where relative deprivation is zero, or
-in other words, if there was no socioeconomic deprivation in an
-administrative area, the intercept would represent its level of
-mobility. The intercept was well below the baseline level at the early
-stages of the pandemic. As observed in Figure X, while there are some
-differences between countries in the evolution of the intercept, the
-general tendency is for the intercept to increase. While Argentina and
-Mexico reach values that are closer to the baseline towards the end of
-the forecast period, the intercept for Chile and Colombia remains lower.
-Therefore, if there were areas with no socioeconomic deprivation, we
-would have seen a recovery in the levels of mobility, although not quite
-back to baseline levels in the case of Chile and Colombia.
+consistent across the four countries. The intercept estimates the urban
+mobility levels that would be observed in administrative units where
+relative deprivation is zero. The intercept was below the baseline level
+at the early stages of the pandemic. As observed in Figure 3, while
+there are some differences between countries in the evolution of the
+intercept, the general tendency is for the intercept to increase. While
+Argentina and Mexico reach values that are closer to the baseline
+towards the end of the forecast period, the intercept for Chile and
+Colombia remains lower. Therefore, if there were areas with no
+socioeconomic deprivation, we would have seen a recovery in the levels
+of mobility, especially in Argentina and Mexico
 
@@
 \section{Discussion}\label{sec-discussion}}
+
+Using location data from Meta-Facebook users, our study aimed to examine
+the evolution of patterns of mobility across socioeconomic groups in
+functional urban areas from Argentina, Chile, Colombia and Mexico from
+April 2020 to March 2023, following the COVID-19 pandemic. We found a
+systematic decline in population movement levels during the
+implementation of non-pharmaceutical interventions in April 2020, with
+the largest reductions observed in the most affluent administrative
+units within functional urban areas from Argentina, Chile and Colombia.
+While mobility rebounded closer to pre-pandemic levels approximately two
+years later after COVID-19 restrictions eased, movement intensity
+remained below pre-pandemic levels in Chile and generally in areas where
+the relative deprivation index is lower. Furthermore, we found that at
+the beginning of the pandemic there were inequalities between
+socioeconomic groups in terms of the levels of urban mobility. While it
+has taken more than two years for Argentina and Mexico to gradually
+close the gap, inequalities persist in Chile and Colombia as of March
+2023 according to our estimated data.
+
+We focused the analysis on short-distance movements in urban areas,
+which are typically considered to capture local and routine mobility
+(Owen and Green 1992). However, due to the characteristics of the
+dataset, we are unable to distinguish movements of this nature from
+short-distance movements that involve a permanent change of place of
+residence. Future work could extend and validate our work identifying
+changes to moves relating permanent and temporary displacements, and
+triangulating alternative DFD and traditional data sources, such as
+censuses and representative surveys collecting information on
+residential location. Further research is also needed to establish the
+cause of the observed changes in the spatial patterns of internal
+population movement since the start of the COVID-19 pandemic.
+Understanding these causes can assist in anticipating long-term
+structural changes in the direction of mobility flows in the coming
+years.
+
+Additionally, accessing and working with DFD to measure human mobility
+and migration is challenging. DFD are known to suffer from biases and
+issues of statistical representation reflecting differences in
+technology usage and accessibility (Rowe 2023). While concerted efforts
+to develop data services and methodological frameworks to leverage on
+DFD exist in the Global North {[}REF{]}, limited progress has been done
+in Latin America. National government and statistical agencies have a
+prime position to promote the creation of data sharing partnerships with
+private companies and create the digital infrastructure to ensure equal
+and secure access to DFD sources for the social good. With academia,
+these agencies could play a major role in promoting the creation of a
+methodological framework for the use and analysis of DFD, setting common
+definitions, metrics, and standard approaches to measure and correct
+data biases and validate data analyses. Location data have been proven
+key to monitor and develop appropriate policy responses to population
+displacement and movement during natural disasters, wars and epidemics
+(Green, Pollock, and Rowe 2021). We envision resources like those
+created by the European Commission Joint Research Centre (2022) and the
+UNSD (2019), identifying sources of non-traditional data and
+establishing methodological guidelines for the use of mobile phone data
+for official mobility statistics. Existing resources have a global
+character, we call for initiatives with a regional focus accounting for
+differences in the local landscape of data availability and digital
+technology usage.
 

### 2024-01-28 manuscript/manuscript.qmd
discussion-and-figs

@@
 Using location data from Meta-Facebook users, our study aimed to examine the evolution of patterns of mobility across socioeconomic groups in functional urban areas from Argentina, Chile, Colombia and Mexico from April 2020 to March 2023, following the COVID-19 pandemic.
-We found a systematic decline in population movement levels during the implementation of non-pharmaceutical interventions in April 2020, with the largest reductions observed in the most affluent administrative units within functional urban areas from Argentina, Chile and Colombia.
-While mobility rebounded closer to pre-pandemic levels approximately two years later after COVID-19 restrictions eased, movement intensity remained below pre-pandemic levels in Chile and generally in areas where the relative deprivation index is lower.
+We found a systematic drop in the number of population movements in April 2020, with the largest reductions observed in the most affluent administrative units within functional urban areas from Argentina, Chile and Colombia.
+While mobility rebounded closer to pre-pandemic levels approximately two years later, when COVID-19 restrictions eased, the number of movements remained below pre-pandemic in Chile.
 Furthermore, we found that at the beginning of the pandemic there were inequalities between socioeconomic groups in terms of the levels of urban mobility.
@@
 
-<!--- Interpretation --->
+We focused the analysis on short-distance movements in urban areas, especifically those covering 70 km or less.
+These journeys are typically considered to represent local and routine mobility (Owen and Green 1992).
+However, due to the characteristics of the Meta-Facebook movement data, we are unable to distinguish the purpose of these short-distance movements.
+Hence, some of our data could be capturing journeys that involve a permanent change of place of residence.
+Our work therefore motivates the need to answer questions regarding the validity of digital footprint data for the analysis of human mobility.
+Further research should focus on inferring more specific information about the nature of the journeys, for example following the approach in X and Y (cite JOana), and quantifying the extent to which the digital footprint data mirrors the true mobility patterns.
 
-<!--- Limitations --->
+Conducting research on urban mobility using digital footprint data is not straightforward, due to the challenges in accessing and working with unstructured data sets which are often subject to biases and statistical representation issues.
+These biases often arise from inequalities in access and usage of digital technologies across demographic groups (cite).
+Despite these challenges, the data and analysis that we used for this work provide evidence for non-trivial patterns that are consistent across four countries in Latin America and with other parts of the world.
+Our findings highlight the dynamic interplay between socioeconomic status and urban mobility, and shall be used to motivate and inform the public debate on the deep societal consecuences of how disparities in urban mobility may be affecting the wider socioeconomic landscape in Latin American countries.
 
-We focused the analysis on short-distance movements in urban areas, which are typically considered to capture local and routine mobility (Owen and Green 1992).
-However, due to the characteristics of the dataset, we are unable to distinguish movements of this nature from short-distance movements that involve a permanent change of place of residence.
-Future work could extend and validate our work identifying changes to moves relating permanent and
-temporary displacements, and triangulating alternative DFD and traditional data sources, such as censuses
-and representative surveys collecting information on residential location.
-Further research is also needed to establish the cause of the observed changes in the spatial patterns of
-internal population movement since the start of the COVID-19 pandemic.
-Understanding these causes can
-assist in anticipating long-term structural changes in the direction of mobility flows in the coming years.
-
-Additionally, accessing and working with DFD to measure human mobility and migration is challenging.
-DFD are known to suffer from biases and issues of statistical representation reflecting differences in technology usage and accessibility (Rowe 2023).
-While concerted efforts to develop data services and
-methodological frameworks to leverage on DFD exist in the Global North \[REF\], limited progress has been
-done in Latin America.
-National government and statistical agencies have a prime position to promote
-the creation of data sharing partnerships with private companies and create the digital infrastructure to
-ensure equal and secure access to DFD sources for the social good.
-With academia, these agencies could
-play a major role in promoting the creation of a methodological framework for the use and analysis of
-DFD, setting common definitions, metrics, and standard approaches to measure and correct data biases
-and validate data analyses.
-Location data have been proven key to monitor and develop appropriate policy
-responses to population displacement and movement during natural disasters, wars and epidemics (Green,
-Pollock, and Rowe 2021).
-We envision resources like those created by the European Commission Joint
-Research Centre (2022) and the UNSD (2019), identifying sources of non-traditional data and establishing
-methodological guidelines for the use of mobile phone data for official mobility statistics.
-Existing resources have a global character, we call for initiatives with a regional focus accounting for differences in the local landscape of data availability and digital technology usage.
+In conclusion, we argue that this work goes beyond the analysis of specific patterns by demonstrating the potential of digital footprint data for policy-relevant research on human mobility at an unprecedented level of spatiotemporal granularity.
+While we have seen a rise in initiatives to improve data services and methodological frameworks to facilitate the use of digital footprint data for social good, progress is still limited, especially in some parts of the world including Latin America.
+It is in the hands of governments and public organisations to prioritise maximiasing the societal benefits that digital footprint data has to offer.
+This includes engaging in activities such as building strategic partnerships with commercial data-holders and academic institutions to establish a unified framework for the use of digital footprint data in policy and research.
+In particular, we call for the creation of resources like those developed by the European Commission Joint Research Centre (2022) and the UNSD (2019), which identify sources of non-traditional data and set methodological protocols for incorporating mobile phone data into official mobility statistics.
+While current resources tend to have a global reach, we advocate for more tailored local initiatives that acknowledge disparities in regional data availability and adoption of digital technologies.
 

### 2024-01-28 manuscript/manuscript.tex
discussion-and-figs

@@
 \end{abstract}
-\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[enhanced, sharp corners, frame hidden, breakable, boxrule=0pt, interior hidden, borderline west={3pt}{0pt}{shadecolor}]}{\end{tcolorbox}}\fi
+\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[borderline west={3pt}{0pt}{shadecolor}, enhanced, breakable, boxrule=0pt, sharp corners, frame hidden, interior hidden]}{\end{tcolorbox}}\fi
 
@@
 April 2020 to March 2023, following the COVID-19 pandemic. We found a
-systematic decline in population movement levels during the
-implementation of non-pharmaceutical interventions in April 2020, with
-the largest reductions observed in the most affluent administrative
+systematic drop in the number of population movements in April 2020,
+with the largest reductions observed in the most affluent administrative
 units within functional urban areas from Argentina, Chile and Colombia.
 While mobility rebounded closer to pre-pandemic levels approximately two
-years later after COVID-19 restrictions eased, movement intensity
-remained below pre-pandemic levels in Chile and generally in areas where
-the relative deprivation index is lower. Furthermore, we found that at
-the beginning of the pandemic there were inequalities between
-socioeconomic groups in terms of the levels of urban mobility. While it
-has taken more than two years for Argentina and Mexico to gradually
-close the gap, inequalities persist in Chile and Colombia as of March
-2023 according to our estimated data.
+years later, when COVID-19 restrictions eased, the number of movements
+remained below pre-pandemic in Chile. Furthermore, we found that at the
+beginning of the pandemic there were inequalities between socioeconomic
+groups in terms of the levels of urban mobility. While it has taken more
+than two years for Argentina and Mexico to gradually close the gap,
+inequalities persist in Chile and Colombia as of March 2023 according to
+our estimated data.
 
 We focused the analysis on short-distance movements in urban areas,
-which are typically considered to capture local and routine mobility
-(Owen and Green 1992). However, due to the characteristics of the
-dataset, we are unable to distinguish movements of this nature from
-short-distance movements that involve a permanent change of place of
-residence. Future work could extend and validate our work identifying
-changes to moves relating permanent and temporary displacements, and
-triangulating alternative DFD and traditional data sources, such as
-censuses and representative surveys collecting information on
-residential location. Further research is also needed to establish the
-cause of the observed changes in the spatial patterns of internal
-population movement since the start of the COVID-19 pandemic.
-Understanding these causes can assist in anticipating long-term
-structural changes in the direction of mobility flows in the coming
-years.
+especifically those covering 70 km or less. These journeys are typically
+considered to represent local and routine mobility (Owen and Green
+1992). However, due to the characteristics of the Meta-Facebook movement
+data, we are unable to distinguish the purpose of these short-distance
+movements. Hence, some of our data could be capturing journeys that
+involve a permanent change of place of residence. Our work therefore
+motivates the need to answer questions regarding the validity of digital
+footprint data for the analysis of human mobility. Further research
+should focus on inferring more specific information about the nature of
+the journeys, for example following the approach in X and Y (cite
+JOana), and quantifying the extent to which the digital footprint data
+mirrors the true mobility patterns.
 
-Additionally, accessing and working with DFD to measure human mobility
-and migration is challenging. DFD are known to suffer from biases and
-issues of statistical representation reflecting differences in
-technology usage and accessibility (Rowe 2023). While concerted efforts
-to develop data services and methodological frameworks to leverage on
-DFD exist in the Global North {[}REF{]}, limited progress has been done
-in Latin America. National government and statistical agencies have a
-prime position to promote the creation of data sharing partnerships with
-private companies and create the digital infrastructure to ensure equal
-and secure access to DFD sources for the social good. With academia,
-these agencies could play a major role in promoting the creation of a
-methodological framework for the use and analysis of DFD, setting common
-definitions, metrics, and standard approaches to measure and correct
-data biases and validate data analyses. Location data have been proven
-key to monitor and develop appropriate policy responses to population
-displacement and movement during natural disasters, wars and epidemics
-(Green, Pollock, and Rowe 2021). We envision resources like those
-created by the European Commission Joint Research Centre (2022) and the
-UNSD (2019), identifying sources of non-traditional data and
-establishing methodological guidelines for the use of mobile phone data
-for official mobility statistics. Existing resources have a global
-character, we call for initiatives with a regional focus accounting for
-differences in the local landscape of data availability and digital
-technology usage.
+Conducting research on urban mobility using digital footprint data is
+not straightforward, due to the challenges in accessing and working with
+unstructured data sets which are often subject to biases and statistical
+representation issues. These biases often arise from inequalities in
+access and usage of digital technologies across demographic groups
+(cite). Despite these challenges, the data and analysis that we used for
+this work provide evidence for non-trivial patterns that are consistent
+across four countries in Latin America and with other parts of the
+world. Our findings highlight the dynamic interplay between
+socioeconomic status and urban mobility, and shall be used to motivate
+and inform the public debate on the deep societal consecuences of how
+disparities in urban mobility may be affecting the wider socioeconomic
+landscape in Latin American countries.
+
+In conclusion, we argue that this work goes beyond the analysis of
+specific patterns by demonstrating the potential of digital footprint
+data for policy-relevant research on human mobility at an unprecedented
+level of spatiotemporal granularity. While we have seen a rise in
+initiatives to improve data services and methodological frameworks to
+facilitate the use of digital footprint data for social good, progress
+is still limited, especially in some parts of the world including Latin
+America. It is in the hands of governments and public organisations to
+prioritise maximiasing the societal benefits that digital footprint data
+has to offer. This includes engaging in activities such as building
+strategic partnerships with commercial data-holders and academic
+institutions to establish a unified framework for the use of digital
+footprint data in policy and research. In particular, we call for the
+creation of resources like those developed by the European Commission
+Joint Research Centre (2022) and the UNSD (2019), which identify sources
+of non-traditional data and set methodological protocols for
+incorporating mobile phone data into official mobility statistics. While
+current resources tend to have a global reach, we advocate for more
+tailored local initiatives that acknowledge disparities in regional data
+availability and adoption of digital technologies.
 

### 2024-01-30 manuscript/manuscript.qmd
manuscript-refs

@@
 ---
-title: "**Urban mobility in Latin America after COVID-19**"
+title: "**Socioeconomic deprivation and urban mobility in Latin America after COVID-19**"
 author:
@@
 However, evidence and analysis of long-term changes in mobility is scarce, especially in less developed countries.
-In this study we use location data from Meta-Facebook users to analyse patterns of urban mobility in X Functional Urban Areas from South American countries.
+In this study, we use location data from Meta-Facebook users to analyse patterns of urban mobility in X Functional Urban Areas from South American countries.
 Findings reveal a general decrease in mobility during the pandemic, with gradual recovery trends up to April 2023.
-However, socioeconomic disparities persist through the period of analysis, with deprived areas experiencing smaller initial losses in mobility.
-Inequalities in mobility levels across socioeconomic groups generally diminish over time, although they persist in all the countries included in the analysis.
-This research highlights the importance of understanding the long-term effects of the pandemic on mobility and the need for equitable policy responses to address societal challenges in urban areas.
+However, socioeconomic disparities are present through the period of analysis, with deprived areas experiencing smaller initial losses in mobility.
+These disparities generally diminish over time, although they persist in all the countries included in the analysis.
+Our research highlights the importance of timely mobility data with high spatiotemporal resolution to understand the long-term effects of the pandemic and to inform equitable policy responses that address societal challenges in urban areas.
 format: 
@@
 Yet, these data streams are not frequently updated and suffer from slow releases, with census data for example being collected over intervals of ten years in most countries.
-Traditional data streams thus lack the temporal granularity to analyse population movements over short-time periods and offer an up-to-date representation of the urban mobility system.
+Traditional data streams thus lack the temporal granularity to analyse population movements over short-time periods and to offer an up-to-date representation of the urban mobility system.
 Data resulting from social interactions on digital platforms have emerged as an unique source of information to deliver this representation and capture human population movement in less developed countries at scale.
@@
 Drawing on a dataset of 213 million observations from Meta-Facebook users' mobile location data, we aim to assess socioeconomic differences in the extent and persistence of decline in urban mobility in Argentina, Chile, Colombia and Mexico during and after the COVID-19 pandemic from March 2020 to March 2023.
-We use Meta-Facebook observed data to measure origin-destination flows from March 2020 to May 2022, and Meta Prophet time-series forecasting machine learning algorithm to predict origin-destination flows from June 2022 to March 2023.
-<!--# Perhaps say how predictions have been validated --> We use Functional Urban Areas (FUAs) boundaries from the Global Human Settlement Layer (GHSL) to define urban areas, and the Global Gridded Relative Deprivation Index (GRDI) developed by NASA’s Socioeconomic Data and Applications Centre (SEDAC) from sociodemographic and satellite data inputs.
-Building on existing evidence, we hypothesised that (1) urban mobility has recovered returning to pre-pandemic levels of movement as nonpharmaceuthical restrictions were lifted; and (2) socioeconomic differences in urban mobility have endured the pandemic reflecting deep societal inequalities as knowledge-intensive businesses adopt hybrid working.
+We use Meta-Facebook data to measure origin-destination flows from March 2020 to May 2022, and Meta Prophet time-series forecasting machine learning algorithm to predict origin-destination flows from June 2022 to March 2023.
+<!--# Perhaps say how predictions have been validated --> We use Functional Urban Areas (FUAs) boundaries from the Global Human Settlement Layer (GHSL), developed by the European Comission's Joint Research Centre [@GHSL19] to define urban areas; and the Global Gridded Relative Deprivation Index (GRDI) developed by NASA’s Socioeconomic Data and Applications Centre [@RDI2022] from sociodemographic and satellite data inputs.
+Building on existing evidence (e.g.
+[@roweCalafiore2022; @wang2022]), we hypothesised that (1) urban mobility has recovered returning to pre-pandemic levels of movement as nonpharmaceuthical restrictions were lifted; and (2) socioeconomic differences in urban mobility have endured the pandemic reflecting deep societal inequalities as knowledge-intensive businesses adopt hybrid working.
 <!--# Once we have a more definitive version of the results, highlight our novel contributions here. -->
@@
 -   Gap 2: little work on less developed countries - where inequalities are more pronounced\
--   Aim - focus on South American countries as case study.
+-   Aim - focus on Latin American countries as case study.
+-   Aim - focus on urban areas, as the future of humanity is the future of cities and Latin Ameriaca is highly urbanised
 
@@
 
-<!--# Carmen to write this -->
-
 The evolution of the levels of movement is measured with respect to a baseline period prior to the pandemic as described in @sec-methods.
-For the purposes of the analysis, we aggregate the raw movement data temporally into months and spatially into administrative units according to various GADM levels.
-The analysis focuses on administrative areas that are within the boundaries of functional urban areas as specified by the Global Human Settlement Layer.
-For each administrative area, we compute the Relative Deprivation Index (RDI) based on data from NASA's Socioeconomic Data and Applications Centre (SEDAC).
-Figure 1 displays the administrative areas included in the study, coloured according to the Relative Index of Deprivation.
+For the purposes of the analysis, we aggregate the raw movement data temporally into months and spatially into administrative units at various levels according to GADM, the Database of Global Administrative Areas [@GADM24].
+The analysis focuses on administrative units that are within the boundaries of functional urban areas as specified by the Global Human Settlement Layer (GHSL).
+For each administrative unit, we compute the Relative Deprivation Index (RDI) based on data from NASA's Socioeconomic Data and Applications Centre (SEDAC).
+Figure 1 displays the administrative units included in the study, coloured according to their average RDI.
 Predictions about the evolution of the levels of movement are made using the Prophet forecasting procedure.
@@
 
-## The impact of COVID-19 on urban mobility
+## The heterogeneous impact of COVID-19 on urban mobility
 
 We analyse the evolution of the relative intensity of urban mobility with respect to a baseline period prior to the pandemic.
-Specifically, movements covering a distance of at most 70 km are considered.
+Specifically, we focus the analysis on short-distance movements in urban areas to represent local and routine mobility [@owen1992migration], so only movements covering a distance of at most 70 km are considered.
 For a movement to be classified as urban it needs to start or end within a functional urban area from Argentina, Chile, Colombia and Mexico.
@@
 Figure 2 displays the patterns of recovery for the mobility levels in the administrative units belonging to functional urban areas in the countries of interest.
-The three lines in each panel represent the mean levels of mobility for administrative units grouped into one of three terciles, according to their average relative deprivation index.
+The three lines in each panel represent the mean levels of mobility for administrative units grouped into one of three terciles, according to their average RDI.
 
@@
 This drop was especially large for Argentina, Chile and Colombia, with Mexico displaying a smaller decrease in the relative number of movements with respect to the baseline.
-Following the initial drop in movement, all four countries evolve towards the recovery of baseline levels of urban mobility, with a generally increasing trend.
+Following the initial drop in mobility, all four countries evolve towards the recovery of baseline levels, with a generally increasing trend.
 There are, however, fluctuations from the general trend which manifest differently for each country.
@@
 
-## Socioeconomic deprivation and recovery of urban mobility
+## The most deprived areas experienced the smallest drop in mobility relative to pre-pandemic levels
 
@@
 For a given point in time (i.e.
-a month), we start by considering the relationship between the number of movements relative to the pre-pandemic baseline period and the average relative deprivation index, at the administrative unit level.
+a month), we start by considering the relationship between the number of movements relative to the pre-pandemic baseline period and the average RDI, at the administrative unit level.
 We assume that this relationship is linear and we use a linear regression to estimate the slope and intercept characterising the line of best fit.
@@
 In Argentina, Colombia and Mexico, we observe that the slope of this relationship evolves to become smaller over time.
-The tendency is not apparent in Chile, where the slope of the relationship remains approximately the same despite the temporary fluctuations.
+The tendency is not apparent in Chile, where the slope of the relationship remains approximately the same over time despite the temporary fluctuations.
 The slope captures the extent of differences in the level of urban mobility across administrative units with varying levels of socioeconomic deprivation.
 It can therefore be regarded as a measure of inequality in mobility patterns across socioeconomic groups.
-A slope equal to zero would mean that all administrative units display the same intensity of movement regardless of their socioeconomic deprivation levels.
+A slope equal to zero would mean that all administrative units display the same intensity of movement regardless of their average RDI.
 Given the patterns observed in Argentina, Colombia and Mexico, we find that at the beginning of the pandemic there were notable inequalities between socioeconomic groups in terms of the levels of urban mobility.
@@
 The intercept of the relationship displays stronger patterns, which are consistent across the four countries.
-The intercept estimates the urban mobility levels that would be observed in administrative units where relative deprivation is zero.
+The intercept estimates the urban mobility levels that would be observed in administrative units where the RDI is zero.
 The intercept was below the baseline level at the early stages of the pandemic.
@@
 Using location data from Meta-Facebook users, our study aimed to examine the evolution of patterns of mobility across socioeconomic groups in functional urban areas from Argentina, Chile, Colombia and Mexico from April 2020 to March 2023, following the COVID-19 pandemic.
-We found a systematic drop in the number of population movements in April 2020, with the largest reductions observed in the most affluent administrative units within functional urban areas from Argentina, Chile and Colombia.
+We found a systematic drop in the number of population movements in April 2020, with the largest reductions observed in the most affluent administrative units within functional urban areas (FUAs) from Argentina, Chile and Colombia.
 While mobility rebounded closer to pre-pandemic levels approximately two years later, when COVID-19 restrictions eased, the number of movements remained below pre-pandemic in Chile.
 Furthermore, we found that at the beginning of the pandemic there were inequalities between socioeconomic groups in terms of the levels of urban mobility.
-While it has taken more than two years for Argentina and Mexico to gradually close the gap, inequalities persist in Chile and Colombia as of March 2023 according to our estimated data.
+While it has taken more than two years for Argentina and Mexico to gradually reduce gap, inequalities persist as of March 2023, especially in Chile and Colombia according to our estimated data.
 
 We focused the analysis on short-distance movements in urban areas, especifically those covering 70 km or less.
-These journeys are typically considered to represent local and routine mobility (Owen and Green 1992).
+These journeys are typically considered to represent local and routine mobility [@owen1992migration] .
 However, due to the characteristics of the Meta-Facebook movement data, we are unable to distinguish the purpose of these short-distance movements.
@@
 Our work therefore motivates the need to answer questions regarding the validity of digital footprint data for the analysis of human mobility.
-Further research should focus on inferring more specific information about the nature of the journeys, for example following the approach in X and Y (cite JOana), and quantifying the extent to which the digital footprint data mirrors the true mobility patterns.
+Further research should focus on inferring more specific information about the nature of the journeys, following similar approaches to those proposed by [@LaurenAlexander15; @MengChuishi17; @NiluferSariAslam21 or @Cabrera-Arnau2023], and quantifying the extent to which the digital footprint data mirrors the true mobility patterns.
 
 Conducting research on urban mobility using digital footprint data is not straightforward, due to the challenges in accessing and working with unstructured data sets which are often subject to biases and statistical representation issues.
-These biases often arise from inequalities in access and usage of digital technologies across demographic groups (cite).
+These biases often arise from inequalities in access and usage of digital technologies across demographic groups [@RoweBigData23].
 Despite these challenges, the data and analysis that we used for this work provide evidence for non-trivial patterns that are consistent across four countries in Latin America and with other parts of the world.
-Our findings highlight the dynamic interplay between socioeconomic status and urban mobility, and shall be used to motivate and inform the public debate on the deep societal consecuences of how disparities in urban mobility may be affecting the wider socioeconomic landscape in Latin American countries.
+Our findings highlight the dynamic interplay between socioeconomic status and urban mobility, and shall be used to motivate and inform the public debate regarding the deep societal consequences of urban mobility disparities on the wider socioeconomic landscape of Latin American countries.
 
@@
 While we have seen a rise in initiatives to improve data services and methodological frameworks to facilitate the use of digital footprint data for social good, progress is still limited, especially in some parts of the world including Latin America.
-It is in the hands of governments and public organisations to prioritise maximiasing the societal benefits that digital footprint data has to offer.
+It is in the hands of governments and public organisations to prioritise the maximisation the societal benefits that digital footprint data has to offer.
 This includes engaging in activities such as building strategic partnerships with commercial data-holders and academic institutions to establish a unified framework for the use of digital footprint data in policy and research.
-In particular, we call for the creation of resources like those developed by the European Commission Joint Research Centre (2022) and the UNSD (2019), which identify sources of non-traditional data and set methodological protocols for incorporating mobile phone data into official mobility statistics.
+In particular, we call for the creation of resources like those developed by the European Commission Joint Research Centre [@ECJRC22] and the UN Statistics Division [@UNSD], which identify sources of non-traditional data and set methodological protocols for incorporating mobile phone data into official mobility statistics.
 While current resources tend to have a global reach, we advocate for more tailored local initiatives that acknowledge disparities in regional data availability and adoption of digital technologies.
@@
 To capture population movements, we used anonymised aggregate mobile phone location data from Meta users for Argentina, Colombia, Chile and Mexico, covering a 24-month period from April 2020 to March 2022.
-We used the dataset Facebook Movements created by Meta and accessed through their Data for Good Initiative (https://dataforgood.facebook.com).
+We used the dataset Facebook Movements created by Meta and accessed through their Data for Good Initiative (<https://dataforgood.facebook.com>).
 The data are built from Facebook app users who have the location services setting turned on on their smartphone.
-Prior to releasing the datasets, Meta ensures privacy and anonymity by removing personal information and applying privacy-preserving techniques (Maas et al. 2019).
+Prior to releasing the datasets, Meta ensures privacy and anonymity by removing personal information and applying privacy-preserving techniques [@Maas19].
 Small-count dropping is one of these techniques.
@@
 For instance, the baseline for Monday 00:00-08:00 time window is obtained by averaging over data collected on Mondays from 00:00 to 8:00 for the 45-day period.
-Details about the baseline can be found in Maas et al. (2019).
+Details about the baseline can be found in [@Maas19].
 The Bing Maps Tile System developed by Microsoft (Microsoft) is used a spatial framework to organise the data.
@@
 The data that we used are spatially aggregated into Bing tile levels 13.
-That is about 4.9 x 4.9km at the Equator (Maas et al. 2019).
+That is about 4.9 x 4.9km at the Equator [@Maas19].
 
@@
 
-Since the focus of this work is on urban mobility, we focus the analysis on Functional Urban Areas (FUAs), defined by the Global Human Settlement Layer.
+Since the focus of this work is on urban mobility, we focus the analysis on Functional Urban Areas (FUAs), defined by the Global Human Settlement Layer (GHSL).
 The spatial extent of the FUAs is often large and may include several towns and neighbourhoods displaying a variety of socioeconomic characteristics, hence using FUAs as the spatial units of aggregation would considerably mask the heterogeneity in the mobility patterns.
@@
 The RDI data is made available via NASA's Socioeconomic Data and Applications Centre (SEDAC), with a spatial resolution of 1km pixels.
-We perform a spatial join of the gridded data and the administrative units and compute the average RDI within each of these units.
+We perform a spatial join of the gridded data and the administrative units and compute the average RDI within each of these administrative units.
 
@@
 
-In Figures X and Y, outliers are removed.
+In Figures 2 and 3, outliers are removed.
 Outliers are defined as administrative units which, at any given month, have a z-score greater than 4 for the relative intensity of movement.

### 2024-01-30 manuscript/manuscript.tex
manuscript-refs

@@
 \makeatother
+% definitions for citeproc citations
+\NewDocumentCommand\citeproctext{}{}
+\NewDocumentCommand\citeproc{mm}{%
+  \begingroup\def\citeproctext{#2}\cite{#1}\endgroup}
+\makeatletter
+ % allow citations to break across lines
+ \let\@cite@ofmt\@firstofone
+ % avoid brackets around text for \cite:
+ \def\@biblabel#1{}
+ \def\@cite#1#2{{#1\if@tempswa , #2\fi}}
+\makeatother
+\newlength{\cslhangindent}
+\setlength{\cslhangindent}{1.5em}
+\newlength{\csllabelwidth}
+\setlength{\csllabelwidth}{3em}
+\newenvironment{CSLReferences}[2] % #1 hanging-indent, #2 entry-spacing
+ {\begin{list}{}{%
+  \setlength{\itemindent}{0pt}
+  \setlength{\leftmargin}{0pt}
+  \setlength{\parsep}{0pt}
+  % turn on hanging indent if param 1 is 1
+  \ifodd #1
+   \setlength{\leftmargin}{\cslhangindent}
+   \setlength{\itemindent}{-1\cslhangindent}
+  \fi
+  % set entry spacing
+  \setlength{\itemsep}{#2\baselineskip}}}
+ {\end{list}}
+\usepackage{calc}
+\newcommand{\CSLBlock}[1]{\hfill\break\parbox[t]{\linewidth}{\strut\ignorespaces#1\strut}}
+\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{\strut#1\strut}}
+\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{\strut#1\strut}}
+\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
 
@@
 \usepackage{lipsum} \usepackage{libertine}
-\makeatletter
-\makeatother
-\makeatletter
-\makeatother
 \makeatletter
@@
 \hypersetup{
-  pdftitle={Urban mobility in Latin America after COVID-19},
+  pdftitle={Socioeconomic deprivation and urban mobility in Latin America after COVID-19},
   pdfauthor={Carmen Cabrera-Arnau; Francisco Rowe; Miguel González-Leonardo; Andrea Nasuto; Ruth Neville},
@@
 
-\title{\textbf{Urban mobility in Latin America after COVID-19}}
+\title{\textbf{Socioeconomic deprivation and urban mobility in Latin
+America after COVID-19}}
 
@@
 evidence and analysis of long-term changes in mobility is scarce,
-especially in less developed countries. In this study we use location
+especially in less developed countries. In this study, we use location
 data from Meta-Facebook users to analyse patterns of urban mobility in X
@@
 general decrease in mobility during the pandemic, with gradual recovery
-trends up to April 2023. However, socioeconomic disparities persist
+trends up to April 2023. However, socioeconomic disparities are present
 through the period of analysis, with deprived areas experiencing smaller
-initial losses in mobility. Inequalities in mobility levels across
-socioeconomic groups generally diminish over time, although they persist
-in all the countries included in the analysis. This research highlights
-the importance of understanding the long-term effects of the pandemic on
-mobility and the need for equitable policy responses to address societal
-challenges in urban areas.
+initial losses in mobility. These disparities generally diminish over
+time, although they persist in all the countries included in the
+analysis. Our research highlights the importance of timely mobility data
+with high spatiotemporal resolution to understand the long-term effects
+of the pandemic and to inform equitable policy responses that address
+societal challenges in urban areas.
 \end{abstract}
-\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[borderline west={3pt}{0pt}{shadecolor}, enhanced, breakable, boxrule=0pt, sharp corners, frame hidden, interior hidden]}{\end{tcolorbox}}\fi
 
-\hypertarget{sec-intro}{%
-\section{Introduction}\label{sec-intro}}
+\section{Introduction}\label{sec-intro}
 
@@
 temporal granularity to analyse population movements over short-time
-periods and offer an up-to-date representation of the urban mobility
+periods and to offer an up-to-date representation of the urban mobility
 system. Data resulting from social interactions on digital platforms
@@
 Chile, Colombia and Mexico during and after the COVID-19 pandemic from
-March 2020 to March 2023. We use Meta-Facebook observed data to measure
+March 2020 to March 2023. We use Meta-Facebook data to measure
 origin-destination flows from March 2020 to May 2022, and Meta Prophet
@@
 Urban Areas (FUAs) boundaries from the Global Human Settlement Layer
-(GHSL) to define urban areas, and the Global Gridded Relative
-Deprivation Index (GRDI) developed by NASA's Socioeconomic Data and
-Applications Centre (SEDAC) from sociodemographic and satellite data
-inputs. Building on existing evidence, we hypothesised that (1) urban
-mobility has recovered returning to pre-pandemic levels of movement as
+(GHSL), developed by the European Comission's Joint Research Centre
+(Schiavina M. 2019) to define urban areas; and the Global Gridded
+Relative Deprivation Index (GRDI) developed by NASA's Socioeconomic Data
+and Applications Centre (Columbia University 2022) from sociodemographic
+and satellite data inputs. Building on existing evidence (e.g. (Rowe et
+al. 2022; Wang et al. 2022)), we hypothesised that (1) urban mobility
+has recovered returning to pre-pandemic levels of movement as
 nonpharmaceuthical restrictions were lifted; and (2) socioeconomic
@@
 \item
-  Aim - focus on South American countries as case study.
+  Aim - focus on Latin American countries as case study.
+\item
+  Aim - focus on urban areas, as the future of humanity is the future of
+  cities and Latin Ameriaca is highly urbanised
 \end{itemize}
 
-\hypertarget{sec-results}{%
-\section{Results}\label{sec-results}}
+\section{Results}\label{sec-results}
 
@@
 aggregate the raw movement data temporally into months and spatially
-into administrative units according to various GADM levels. The analysis
-focuses on administrative areas that are within the boundaries of
-functional urban areas as specified by the Global Human Settlement
-Layer. For each administrative area, we compute the Relative Deprivation
-Index (RDI) based on data from NASA's Socioeconomic Data and
-Applications Centre (SEDAC). Figure 1 displays the administrative areas
-included in the study, coloured according to the Relative Index of
-Deprivation. Predictions about the evolution of the levels of movement
-are made using the Prophet forecasting procedure. Further details are
-provided in Section~\ref{sec-methods}.
+into administrative units at various levels according to GADM, the
+Database of Global Administrative Areas (GADM 2024). The analysis
+focuses on administrative units that are within the boundaries of
+functional urban areas as specified by the Global Human Settlement Layer
+(GHSL). For each administrative unit, we compute the Relative
+Deprivation Index (RDI) based on data from NASA's Socioeconomic Data and
+Applications Centre (SEDAC). Figure 1 displays the administrative units
+included in the study, coloured according to their average RDI.
+Predictions about the evolution of the levels of movement are made using
+the Prophet forecasting procedure. Further details are provided in
+Section~\ref{sec-methods}.
 
-\begin{figure}
+\begin{figure}[H]
 
@@
 
-\end{figure}
+\end{figure}%
 
-\hypertarget{the-impact-of-covid-19-on-urban-mobility}{%
-\subsection{The impact of COVID-19 on urban
-mobility}\label{the-impact-of-covid-19-on-urban-mobility}}
+\subsection{The heterogeneous impact of COVID-19 on urban
+mobility}\label{the-heterogeneous-impact-of-covid-19-on-urban-mobility}
 
@@
 with respect to a baseline period prior to the pandemic. Specifically,
+we focus the analysis on short-distance movements in urban areas to
+represent local and routine mobility (Owen and Green 1992), so only
 movements covering a distance of at most 70 km are considered. For a
@@
 levels of mobility for administrative units grouped into one of three
-terciles, according to their average relative deprivation index.
+terciles, according to their average RDI.
 
-\begin{figure}
+\begin{figure}[H]
 
@@
 smaller decrease in the relative number of movements with respect to the
-baseline. Following the initial drop in movement, all four countries
-evolve towards the recovery of baseline levels of urban mobility, with a
-generally increasing trend. There are, however, fluctuations from the
-general trend which manifest differently for each country. These
-fluctuations mirror each other in the case of Argentina and Colombia,
-where urban mobility sharply bounces back closer to pre-pandemic levels
-around July of 2020. Chile and Mexico display more progressive patterns
-of recovery, although Chile never reaches baseline levels. These
-fluctuations are unique to each country and can be attributed to local
-factors such as the effects of seasonality or the different stringency
-measures imposed by the national governments during the pandemic.
+baseline. Following the initial drop in mobility, all four countries
+evolve towards the recovery of baseline levels, with a generally
+increasing trend. There are, however, fluctuations from the general
+trend which manifest differently for each country. These fluctuations
+mirror each other in the case of Argentina and Colombia, where urban
+mobility sharply bounces back closer to pre-pandemic levels around July
+of 2020. Chile and Mexico display more progressive patterns of recovery,
+although Chile never reaches baseline levels. These fluctuations are
+unique to each country and can be attributed to local factors such as
+the effects of seasonality or the different stringency measures imposed
+by the national governments during the pandemic.
 
@@
 
-\hypertarget{socioeconomic-deprivation-and-recovery-of-urban-mobility}{%
-\subsection{Socioeconomic deprivation and recovery of urban
-mobility}\label{socioeconomic-deprivation-and-recovery-of-urban-mobility}}
+\subsection{The most deprived areas experienced the smallest drop in
+mobility relative to pre-pandemic
+levels}\label{the-most-deprived-areas-experienced-the-smallest-drop-in-mobility-relative-to-pre-pandemic-levels}
 
@@
 the number of movements relative to the pre-pandemic baseline period and
-the average relative deprivation index, at the administrative unit
-level. We assume that this relationship is linear and we use a linear
-regression to estimate the slope and intercept characterising the line
-of best fit. This is shown for April 2020 and March 2022 in the
-right-hand side panels of Figure 3. After obtaining the slope and
-intercept for every month, we are able to plot the evolution of these
-parameters for both the observed and forecasted data, as displayed on
-the left-hand-side panels of the same Figure.
+the average RDI, at the administrative unit level. We assume that this
+relationship is linear and we use a linear regression to estimate the
+slope and intercept characterising the line of best fit. This is shown
+for April 2020 and March 2022 in the right-hand side panels of Figure 3.
+After obtaining the slope and intercept for every month, we are able to
+plot the evolution of these parameters for both the observed and
+forecasted data, as displayed on the left-hand-side panels of the same
+Figure.
 
-\begin{figure}
+\begin{figure}[H]
 
@@
 not apparent in Chile, where the slope of the relationship remains
-approximately the same despite the temporary fluctuations. The slope
-captures the extent of differences in the level of urban mobility across
-administrative units with varying levels of socioeconomic deprivation.
-It can therefore be regarded as a measure of inequality in mobility
-patterns across socioeconomic groups. A slope equal to zero would mean
-that all administrative units display the same intensity of movement
-regardless of their socioeconomic deprivation levels. Given the patterns
-observed in Argentina, Colombia and Mexico, we find that at the
-beginning of the pandemic there were notable inequalities between
-socioeconomic groups in terms of the levels of urban mobility. While it
-has taken more than two years for Argentina and Mexico to close the gap
-(their slope is close to zero from spring 2022), inequalities persist in
-Chile and Colombia as of March 2023.
+approximately the same over time despite the temporary fluctuations. The
+slope captures the extent of differences in the level of urban mobility
+across administrative units with varying levels of socioeconomic
+deprivation. It can therefore be regarded as a measure of inequality in
+mobility patterns across socioeconomic groups. A slope equal to zero
+would mean that all administrative units display the same intensity of
+movement regardless of their average RDI. Given the patterns observed in
+Argentina, Colombia and Mexico, we find that at the beginning of the
+pandemic there were notable inequalities between socioeconomic groups in
+terms of the levels of urban mobility. While it has taken more than two
+years for Argentina and Mexico to close the gap (their slope is close to
+zero from spring 2022), inequalities persist in Chile and Colombia as of
+March 2023.
 
@@
 consistent across the four countries. The intercept estimates the urban
-mobility levels that would be observed in administrative units where
-relative deprivation is zero. The intercept was below the baseline level
-at the early stages of the pandemic. As observed in Figure 3, while
-there are some differences between countries in the evolution of the
-intercept, the general tendency is for the intercept to increase. While
-Argentina and Mexico reach values that are closer to the baseline
-towards the end of the forecast period, the intercept for Chile and
-Colombia remains lower. Therefore, if there were areas with no
-socioeconomic deprivation, we would have seen a recovery in the levels
-of mobility, especially in Argentina and Mexico
+mobility levels that would be observed in administrative units where the
+RDI is zero. The intercept was below the baseline level at the early
+stages of the pandemic. As observed in Figure 3, while there are some
+differences between countries in the evolution of the intercept, the
+general tendency is for the intercept to increase. While Argentina and
+Mexico reach values that are closer to the baseline towards the end of
+the forecast period, the intercept for Chile and Colombia remains lower.
+Therefore, if there were areas with no socioeconomic deprivation, we
+would have seen a recovery in the levels of mobility, especially in
+Argentina and Mexico
 
-\hypertarget{sec-discussion}{%
-\section{Discussion}\label{sec-discussion}}
+\section{Discussion}\label{sec-discussion}
 
@@
 with the largest reductions observed in the most affluent administrative
-units within functional urban areas from Argentina, Chile and Colombia.
-While mobility rebounded closer to pre-pandemic levels approximately two
-years later, when COVID-19 restrictions eased, the number of movements
-remained below pre-pandemic in Chile. Furthermore, we found that at the
-beginning of the pandemic there were inequalities between socioeconomic
-groups in terms of the levels of urban mobility. While it has taken more
-than two years for Argentina and Mexico to gradually close the gap,
-inequalities persist in Chile and Colombia as of March 2023 according to
-our estimated data.
+units within functional urban areas (FUAs) from Argentina, Chile and
+Colombia. While mobility rebounded closer to pre-pandemic levels
+approximately two years later, when COVID-19 restrictions eased, the
+number of movements remained below pre-pandemic in Chile. Furthermore,
+we found that at the beginning of the pandemic there were inequalities
+between socioeconomic groups in terms of the levels of urban mobility.
+While it has taken more than two years for Argentina and Mexico to
+gradually reduce gap, inequalities persist as of March 2023, especially
+in Chile and Colombia according to our estimated data.
 
@@
 especifically those covering 70 km or less. These journeys are typically
-considered to represent local and routine mobility (Owen and Green
-1992). However, due to the characteristics of the Meta-Facebook movement
+considered to represent local and routine mobility (Owen and Green 1992)
+. However, due to the characteristics of the Meta-Facebook movement
 data, we are unable to distinguish the purpose of these short-distance
@@
 should focus on inferring more specific information about the nature of
-the journeys, for example following the approach in X and Y (cite
-JOana), and quantifying the extent to which the digital footprint data
-mirrors the true mobility patterns.
+the journeys, following similar approaches to those proposed by
+Cabrera-Arnau et al. (2023), and quantifying the extent to which the
+digital footprint data mirrors the true mobility patterns.
 
@@
 representation issues. These biases often arise from inequalities in
-access and usage of digital technologies across demographic groups
-(cite). Despite these challenges, the data and analysis that we used for
+access and usage of digital technologies across demographic groups (Rowe
+2023). Despite these challenges, the data and analysis that we used for
 this work provide evidence for non-trivial patterns that are consistent
@@
 socioeconomic status and urban mobility, and shall be used to motivate
-and inform the public debate on the deep societal consecuences of how
-disparities in urban mobility may be affecting the wider socioeconomic
-landscape in Latin American countries.
+and inform the public debate regarding the deep societal consequences of
+urban mobility disparities on the wider socioeconomic landscape of Latin
+American countries.
 
@@
 America. It is in the hands of governments and public organisations to
-prioritise maximiasing the societal benefits that digital footprint data
-has to offer. This includes engaging in activities such as building
+prioritise the maximisation the societal benefits that digital footprint
+data has to offer. This includes engaging in activities such as building
 strategic partnerships with commercial data-holders and academic
@@
 creation of resources like those developed by the European Commission
-Joint Research Centre (2022) and the UNSD (2019), which identify sources
-of non-traditional data and set methodological protocols for
-incorporating mobile phone data into official mobility statistics. While
-current resources tend to have a global reach, we advocate for more
-tailored local initiatives that acknowledge disparities in regional data
-availability and adoption of digital technologies.
+Joint Research Centre (Commission et al. 2022) and the UN Statistics
+Division (Division 2019), which identify sources of non-traditional data
+and set methodological protocols for incorporating mobile phone data
+into official mobility statistics. While current resources tend to have
+a global reach, we advocate for more tailored local initiatives that
+acknowledge disparities in regional data availability and adoption of
+digital technologies.
 
-\hypertarget{sec-methods}{%
-\section{Methods}\label{sec-methods}}
+\section{Methods}\label{sec-methods}
 
-\hypertarget{meta-facebook-movement-data}{%
 \subsection{Meta-Facebook movement
-data}\label{meta-facebook-movement-data}}
+data}\label{meta-facebook-movement-data}
 
@@
 used the dataset Facebook Movements created by Meta and accessed through
-their Data for Good Initiative (https://dataforgood.facebook.com). The
-data are built from Facebook app users who have the location services
-setting turned on on their smartphone. Prior to releasing the datasets,
-Meta ensures privacy and anonymity by removing personal information and
-applying privacy-preserving techniques (Maas et al.~2019). Small-count
-dropping is one of these techniques. A data entry is removed if the
-population or movement count for an area is lower than 10. The removal
-of small counts may mean that population counts in small sparsely
-populated areas are not captured. A second technique consists in adding
-a small undisclosed amount of random noise to ensure that it is not
-possible to ascertain precise, true counts for sparsely populated
+their Data for Good Initiative (\url{https://dataforgood.facebook.com}).
+The data are built from Facebook app users who have the location
+services setting turned on on their smartphone. Prior to releasing the
+datasets, Meta ensures privacy and anonymity by removing personal
+information and applying privacy-preserving techniques (Maas et al.
+2019). Small-count dropping is one of these techniques. A data entry is
+removed if the population or movement count for an area is lower than
+10. The removal of small counts may mean that population counts in small
+sparsely populated areas are not captured. A second technique consists
+in adding a small undisclosed amount of random noise to ensure that it
+is not possible to ascertain precise, true counts for sparsely populated
 locations. Third, spatial smoothing using inverse distance-weighted
@@
 obtained by averaging over data collected on Mondays from 00:00 to 8:00
-for the 45-day period. Details about the baseline can be found in Maas
-et al.~(2019). The Bing Maps Tile System developed by Microsoft
+for the 45-day period. Details about the baseline can be found in (Maas
+et al. 2019). The Bing Maps Tile System developed by Microsoft
 (Microsoft) is used a spatial framework to organise the data. The Tile
@@
 that we used are spatially aggregated into Bing tile levels 13. That is
-about 4.9 x 4.9km at the Equator (Maas et al.~2019).
+about 4.9 x 4.9km at the Equator (Maas et al. 2019).
 
-\hypertarget{spatiotemporal-data-aggregation}{%
 \subsection{Spatiotemporal data
-aggregation}\label{spatiotemporal-data-aggregation}}
+aggregation}\label{spatiotemporal-data-aggregation}
 
@@
 on Functional Urban Areas (FUAs), defined by the Global Human Settlement
-Layer. The spatial extent of the FUAs is often large and may include
-several towns and neighbourhoods displaying a variety of socioeconomic
-characteristics, hence using FUAs as the spatial units of aggregation
-would considerably mask the heterogeneity in the mobility patterns.
-While the original unit of aggregation for the movement data, i.e.~the
-tiles from the Bing Maps Tile System, offers the highest degree of
-spatial granularity available, the interpretation of findings from an
+Layer (GHSL). The spatial extent of the FUAs is often large and may
+include several towns and neighbourhoods displaying a variety of
+socioeconomic characteristics, hence using FUAs as the spatial units of
+aggregation would considerably mask the heterogeneity in the mobility
+patterns. While the original unit of aggregation for the movement data,
+i.e.~the tiles from the Bing Maps Tile System, offers the highest degree
+of spatial granularity available, the interpretation of findings from an
 analysis based on administrative units is often more valuable. For this
@@
 gridded data and the administrative units and compute the average RDI
-within each of these units.
+within each of these administrative units.
 
-\hypertarget{time-series-analysis}{%
-\subsection{Time series analysis}\label{time-series-analysis}}
+\subsection{Time series analysis}\label{time-series-analysis}
 
@@
 
-In Figures X and Y, outliers are removed. Outliers are defined as
+In Figures 2 and 3, outliers are removed. Outliers are defined as
 administrative units which, at any given month, have a z-score greater
@@
 
-\hypertarget{references}{%
-\section{References}\label{references}}
+\section*{References}\label{references}
+\addcontentsline{toc}{section}{References}
+
+\phantomsection\label{refs}
+\begin{CSLReferences}{1}{0}
+\bibitem[\citeproctext]{ref-LaurenAlexander15}
+Alexander, Lauren, Shan Jiang, Mikel Murga, and Marta C. González. 2015.
+{``Origin--Destination Trips by Purpose and Time of Day Inferred from
+Mobile Phone Data.''} \emph{Transportation Research Part C: Emerging
+Technologies} 58: 240--50.
+https://doi.org/\url{https://doi.org/10.1016/j.trc.2015.02.018}.
+
+\bibitem[\citeproctext]{ref-Cabrera-Arnau2023}
+Cabrera-Arnau, Carmen, Chen Zhong, Michael Batty, Ricardo Silva, and
+Soong Moon Kang. 2023. {``Inferring Urban Polycentricity from the
+Variability in Human Mobility Patterns.''} \emph{Scientific Reports} 13
+(1): 5751. \url{https://doi.org/10.1038/s41598-023-33003-7}.
+
+\bibitem[\citeproctext]{ref-RDI2022}
+Columbia University, Center for International Earth Science Information
+Network. CIESIN. 2022. {``Global Gridded Relative Deprivation Index
+(GRDI), Version 1.''} Palisades, New York: NASA Socioeconomic Data;
+Applications Center (SEDAC). \url{https://doi.org/10.7927/3xxe-ap97}.
+
+\bibitem[\citeproctext]{ref-ECJRC22}
+Commission, European, Joint Research Centre, C Bosco, S
+Grubanov-Boskovic, S Iacus, U Minora, F Sermi, and S Spyratos. 2022.
+\emph{Data Innovation in Demography, Migration and Human Mobility}.
+Publications Office of the European Union.
+\url{https://doi.org/doi/10.2760/027157}.
+
+\bibitem[\citeproctext]{ref-UNSD}
+Division, United Nations Statistical. 2019. {``Handbook on the Use of
+Mobile Phone Data for Official Statistics.''}
+
+\bibitem[\citeproctext]{ref-GADM24}
+GADM. 2024. {``GADM Maps and Data.''}
+
+\bibitem[\citeproctext]{ref-Maas19}
+Maas, P., S. Iyer, A. Gros, W. Park, L. McGorman, C. Nayak, and P. A.
+Dow. 2019. {``Facebook Disaster Maps: Aggregate Insights for Crisis
+Response and Recovery.''} In \emph{16th International Conference on
+Information Systems for Crisis Response and Management}, 836--47.
+
+\bibitem[\citeproctext]{ref-MengChuishi17}
+Meng, Chuishi, Yu Cui, Qing He, Lu Su, and Jing Gao. 2017. {``Travel
+Purpose Inference with GPS Trajectories, POIs, and Geo-Tagged Social
+Media Data.''} In \emph{2017 IEEE International Conference on Big Data
+(Big Data)}, 1319--24.
+\url{https://doi.org/10.1109/BigData.2017.8258062}.
+
+\bibitem[\citeproctext]{ref-NiluferSariAslam21}
+Nilufer Sari Aslam, Tao Cheng, Mohamed R. Ibrahim, and Yang Zhang. 2021.
+{``ActivityNET: Neural Networks to Predict Public Transport Trip
+Purposes from Individual Smart Card Data and POIs.''} \emph{Geo-Spatial
+Information Science} 24 (4): 711--21.
+\url{https://doi.org/10.1080/10095020.2021.1985943}.
+
+\bibitem[\citeproctext]{ref-owen1992migration}
+Owen, D., and A. Green. 1992. {``Migration Patterns and Trends.''} In
+\emph{Migration Processes and Patterns: Research Progress and
+Prospects}, edited by T. Champion and T. Fielding. Belhaven Press.
+
+\bibitem[\citeproctext]{ref-RoweBigData23}
+Rowe, Francisco. 2023. {``9.: Big Data.''} In \emph{Concise Encyclopedia
+of Human Geography}, 42--47. Cheltenham, UK: Edward Elgar Publishing.
+\url{https://doi.org/10.4337/9781800883499.ch09}.
+
+\bibitem[\citeproctext]{ref-roweCalafiore2022}
+Rowe, Francisco, Alessia Calafiore, Daniel Arribas-Bel, Krasen
+Samardzhiev, and Martin Fleischmann. 2022. {``Urban Exodus?
+Understanding Human Mobility in Britain During the COVID-19 Pandemic
+Using Facebook Data.''} \url{https://doi.org/10.48550/ARXIV.2206.03272}.
+
+\bibitem[\citeproctext]{ref-GHSL19}
+Schiavina M., Maffenini L., Moreno-Monroy A. 2019. {``GHS-FUA R2019A -
+GHS Functional Urban Areas, Derived from GHS-UCDB R2019A, (2015),
+R2019A.''} European Commission, Joint Research Centre (JRC).
+
+\bibitem[\citeproctext]{ref-wang2022}
+Wang, Yikang, Chen Zhong, Qili Gao, and Carmen Cabrera-Arnau. 2022.
+{``Understanding Internal Migration in the UK Before and During the
+COVID-19 Pandemic Using Twitter Data.''} \emph{Urban Informatics} 1 (1).
+\url{https://doi.org/10.1007/s44212-022-00018-w}.
+
+\end{CSLReferences}
 

### 2024-02-01 manuscript/manuscript.qmd
manuscript-prophet

@@
 ---
-title: "**Socioeconomic deprivation and urban mobility in Latin America after COVID-19**"
+title: "**Urban mobility and socioeconomic deprivation in Latin America after COVID-19**"
 author:
@@
 Less is known about the longer term patterns of resilience in urban mobility in less developed countries.
-Assessing the extent to which the intensity of movement has returned back to pre-pandemic levels across socioeconomic groups is important to understand the potentially unequal impacts of hybrid working in the society.
+Assessing the extent to which the level of mobility has returned back to the pre-pandemic baseline level across socioeconomic groups is important to understand the potentially unequal impacts of hybrid working in the society.
 
@@
 Building on existing evidence (e.g.
-[@roweCalafiore2022; @wang2022]), we hypothesised that (1) urban mobility has recovered returning to pre-pandemic levels of movement as nonpharmaceuthical restrictions were lifted; and (2) socioeconomic differences in urban mobility have endured the pandemic reflecting deep societal inequalities as knowledge-intensive businesses adopt hybrid working.
+[@roweCalafiore2022; @wang2022]), we hypothesised that (1) urban mobility has recovered returning to the pre-pandemic baseline level of movement as nonpharmaceuthical restrictions were lifted; and (2) socioeconomic differences in urban mobility have endured the pandemic reflecting deep societal inequalities as knowledge-intensive businesses adopt hybrid working.
 <!--# Once we have a more definitive version of the results, highlight our novel contributions here. -->
@@
 
-The evolution of the levels of movement is measured with respect to a baseline period prior to the pandemic as described in @sec-methods.
+The evolution of the percentage change in the number of movements is measured with respect to a baseline period prior to the pandemic as described in @sec-methods.
 For the purposes of the analysis, we aggregate the raw movement data temporally into months and spatially into administrative units at various levels according to GADM, the Database of Global Administrative Areas [@GADM24].
@@
 Figure 1 displays the administrative units included in the study, coloured according to their average RDI.
-Predictions about the evolution of the levels of movement are made using the Prophet forecasting procedure.
+Predictions about the evolution of the percentage change in the number of movements are made using the Prophet forecasting procedure.
 Further details are provided in @sec-methods.
@@
 
-We analyse the evolution of the relative intensity of urban mobility with respect to a baseline period prior to the pandemic.
+We analyse the evolution of the percentage change in the number of movements with respect to a baseline period prior to the pandemic.
 Specifically, we focus the analysis on short-distance movements in urban areas to represent local and routine mobility [@owen1992migration], so only movements covering a distance of at most 70 km are considered.
@@
 Generally, there was a drop in the levels of mobility with respect to the baseline period in all four countries.
-This drop was especially large for Argentina, Chile and Colombia, with Mexico displaying a smaller decrease in the relative number of movements with respect to the baseline.
-Following the initial drop in mobility, all four countries evolve towards the recovery of baseline levels, with a generally increasing trend.
+This drop was especially large for Argentina, Chile and Colombia, with Mexico displaying a smaller decrease in the percentage change in the number of movements with respect to the baseline.
+Following the initial drop in mobility, all four countries evolve towards the recovery of baseline levels of mobility, with a generally increasing trend.
 There are, however, fluctuations from the general trend which manifest differently for each country.
@@
 For a given point in time (i.e.
-a month), we start by considering the relationship between the number of movements relative to the pre-pandemic baseline period and the average RDI, at the administrative unit level.
+a month), we start by considering the relationship between the percentage change in the number of movements relative to the pre-pandemic baseline period and the average RDI, at the administrative unit level.
 We assume that this relationship is linear and we use a linear regression to estimate the slope and intercept characterising the line of best fit.
@@
 
-![Evolution of the parameters characterising the relationship between relative deprivation index and number of movements relative to the pre-pandemic baseline period.](figures/regression-evo-nobackground.pdf)
+![Evolution of the parameters characterising the relationship between relative deprivation index and the percentage change in the number of movements relative to the pre-pandemic baseline level of mobility.](figures/regression-evo-nobackground.pdf)
 
@@
 It can therefore be regarded as a measure of inequality in mobility patterns across socioeconomic groups.
-A slope equal to zero would mean that all administrative units display the same intensity of movement regardless of their average RDI.
+A slope equal to zero would mean that all administrative units display the same level of mobility regardless of their average RDI.
 Given the patterns observed in Argentina, Colombia and Mexico, we find that at the beginning of the pandemic there were notable inequalities between socioeconomic groups in terms of the levels of urban mobility.
@@
 
-<!--# would sell a bit more why Prophet - what is new and can done "better" in comparison to existing approaches - and include citation to [the paper](https://doi.org/10.7287/peerj.preprints.3190v2) -->
-
 For the countries included in the analysis, the movement data is available until March 2022.
 In order to gain a better understanding of the recovery trends after the pandemic, we generate a 12-month forecast up to March 2023.
-This is done using Prophet, a procedure to forecast time series data based on an additive model where non-linear trends can be fit with seasonality effects, holiday effects and other external factors such as the stringency index.
-For the analysis, only seasonality effects are included as they yield the most realistic predictions.
+This is done using Prophet [@prophet17], a procedure developed by Meta's Core Data Science team to forecast time series data based on an additive model.
+Prophet stands out from other forecasting methods due to its ability to capture recurring seasonal trends inherent in mobility data, such as fluctuations during holidays, or changes in the general trend due to the impact of particular circumstances, such as the COVID-19 pandemic.
+It is particularly intuitive to use compared to other models due to its automated trend-detection capabilities, making it accessible to users with no expertise on the underlying model, while being robust to missing data and outliers.
+For the analysis in this paper, yearly seasonality effects are included as they yield the most realistic predictions.
 
-In Figures 2 and 3, outliers are removed.
-Outliers are defined as administrative units which, at any given month, have a z-score greater than 4 for the relative intensity of movement.
+Figures 2 and 3 have been created by removing statistical outliers.
+We define these as the administrative units which, at any given month, have a z-score greater than 4 for the percentage change in the number of movements.
 Furthermore, a Savitzky-Golay filter is applied to smooth the time series data, using a length of 4 units for the filter window and order 2 polynomials to fit the samples.

### 2024-02-01 manuscript/manuscript.tex
manuscript-prophet

@@
 \makeatother
-% definitions for citeproc citations
-\NewDocumentCommand\citeproctext{}{}
-\NewDocumentCommand\citeproc{mm}{%
-  \begingroup\def\citeproctext{#2}\cite{#1}\endgroup}
-\makeatletter
- % allow citations to break across lines
- \let\@cite@ofmt\@firstofone
- % avoid brackets around text for \cite:
- \def\@biblabel#1{}
- \def\@cite#1#2{{#1\if@tempswa , #2\fi}}
-\makeatother
 \newlength{\cslhangindent}
@@
 \setlength{\csllabelwidth}{3em}
-\newenvironment{CSLReferences}[2] % #1 hanging-indent, #2 entry-spacing
- {\begin{list}{}{%
-  \setlength{\itemindent}{0pt}
-  \setlength{\leftmargin}{0pt}
-  \setlength{\parsep}{0pt}
+\newlength{\cslentryspacingunit} % times entry-spacing
+\setlength{\cslentryspacingunit}{\parskip}
+\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
+ {% don't indent paragraphs
+  \setlength{\parindent}{0pt}
   % turn on hanging indent if param 1 is 1
   \ifodd #1
-   \setlength{\leftmargin}{\cslhangindent}
-   \setlength{\itemindent}{-1\cslhangindent}
+  \let\oldpar\par
+  \def\par{\hangindent=\cslhangindent\oldpar}
   \fi
   % set entry spacing
-  \setlength{\itemsep}{#2\baselineskip}}}
- {\end{list}}
+  \setlength{\parskip}{#2\cslentryspacingunit}
+ }%
+ {}
 \usepackage{calc}
-\newcommand{\CSLBlock}[1]{\hfill\break\parbox[t]{\linewidth}{\strut\ignorespaces#1\strut}}
-\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{\strut#1\strut}}
-\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{\strut#1\strut}}
+\newcommand{\CSLBlock}[1]{#1\hfill\break}
+\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
+\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
 \newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
@@
 \usepackage{lipsum} \usepackage{libertine}
+\makeatletter
+\makeatother
+\makeatletter
+\makeatother
 \makeatletter
@@
 \hypersetup{
-  pdftitle={Socioeconomic deprivation and urban mobility in Latin America after COVID-19},
+  pdftitle={Urban mobility and socioeconomic deprivation in Latin America after COVID-19},
   pdfauthor={Carmen Cabrera-Arnau; Francisco Rowe; Miguel González-Leonardo; Andrea Nasuto; Ruth Neville},
@@
 
-\title{\textbf{Socioeconomic deprivation and urban mobility in Latin
+\title{\textbf{Urban mobility and socioeconomic deprivation in Latin
 America after COVID-19}}
@@
 \end{abstract}
+\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[boxrule=0pt, frame hidden, breakable, enhanced, interior hidden, sharp corners, borderline west={3pt}{0pt}{shadecolor}]}{\end{tcolorbox}}\fi
 
-\section{Introduction}\label{sec-intro}
+\hypertarget{sec-intro}{%
+\section{Introduction}\label{sec-intro}}
 
@@
 urban mobility in less developed countries. Assessing the extent to
-which the intensity of movement has returned back to pre-pandemic levels
-across socioeconomic groups is important to understand the potentially
-unequal impacts of hybrid working in the society.
+which the level of mobility has returned back to the pre-pandemic
+baseline level across socioeconomic groups is important to understand
+the potentially unequal impacts of hybrid working in the society.
 
@@
 al. 2022; Wang et al. 2022)), we hypothesised that (1) urban mobility
-has recovered returning to pre-pandemic levels of movement as
-nonpharmaceuthical restrictions were lifted; and (2) socioeconomic
+has recovered returning to the pre-pandemic baseline level of movement
+as nonpharmaceuthical restrictions were lifted; and (2) socioeconomic
 differences in urban mobility have endured the pandemic reflecting deep
@@
 
-\section{Results}\label{sec-results}
+\hypertarget{sec-results}{%
+\section{Results}\label{sec-results}}
 
-The evolution of the levels of movement is measured with respect to a
-baseline period prior to the pandemic as described in
-Section~\ref{sec-methods}. For the purposes of the analysis, we
-aggregate the raw movement data temporally into months and spatially
-into administrative units at various levels according to GADM, the
-Database of Global Administrative Areas (GADM 2024). The analysis
+The evolution of the percentage change in the number of movements is
+measured with respect to a baseline period prior to the pandemic as
+described in Section~\ref{sec-methods}. For the purposes of the
+analysis, we aggregate the raw movement data temporally into months and
+spatially into administrative units at various levels according to GADM,
+the Database of Global Administrative Areas (GADM 2024). The analysis
 focuses on administrative units that are within the boundaries of
@@
 included in the study, coloured according to their average RDI.
-Predictions about the evolution of the levels of movement are made using
-the Prophet forecasting procedure. Further details are provided in
-Section~\ref{sec-methods}.
+Predictions about the evolution of the percentage change in the number
+of movements are made using the Prophet forecasting procedure. Further
+details are provided in Section~\ref{sec-methods}.
 
-\begin{figure}[H]
+\begin{figure}
 
@@
 
-\end{figure}%
+\end{figure}
 
+\hypertarget{the-heterogeneous-impact-of-covid-19-on-urban-mobility}{%
 \subsection{The heterogeneous impact of COVID-19 on urban
-mobility}\label{the-heterogeneous-impact-of-covid-19-on-urban-mobility}
+mobility}\label{the-heterogeneous-impact-of-covid-19-on-urban-mobility}}
 
-We analyse the evolution of the relative intensity of urban mobility
-with respect to a baseline period prior to the pandemic. Specifically,
-we focus the analysis on short-distance movements in urban areas to
-represent local and routine mobility (Owen and Green 1992), so only
-movements covering a distance of at most 70 km are considered. For a
-movement to be classified as urban it needs to start or end within a
+We analyse the evolution of the percentage change in the number of
+movements with respect to a baseline period prior to the pandemic.
+Specifically, we focus the analysis on short-distance movements in urban
+areas to represent local and routine mobility (Owen and Green 1992), so
+only movements covering a distance of at most 70 km are considered. For
+a movement to be classified as urban it needs to start or end within a
 functional urban area from Argentina, Chile, Colombia and Mexico. The
@@
 
-\begin{figure}[H]
+\begin{figure}
 
@@
 
-\end{figure}%
+\end{figure}
 
@@
 large for Argentina, Chile and Colombia, with Mexico displaying a
-smaller decrease in the relative number of movements with respect to the
-baseline. Following the initial drop in mobility, all four countries
-evolve towards the recovery of baseline levels, with a generally
-increasing trend. There are, however, fluctuations from the general
-trend which manifest differently for each country. These fluctuations
-mirror each other in the case of Argentina and Colombia, where urban
-mobility sharply bounces back closer to pre-pandemic levels around July
-of 2020. Chile and Mexico display more progressive patterns of recovery,
-although Chile never reaches baseline levels. These fluctuations are
-unique to each country and can be attributed to local factors such as
-the effects of seasonality or the different stringency measures imposed
-by the national governments during the pandemic.
+smaller decrease in the percentage change in the number of movements
+with respect to the baseline. Following the initial drop in mobility,
+all four countries evolve towards the recovery of baseline levels of
+mobility, with a generally increasing trend. There are, however,
+fluctuations from the general trend which manifest differently for each
+country. These fluctuations mirror each other in the case of Argentina
+and Colombia, where urban mobility sharply bounces back closer to
+pre-pandemic levels around July of 2020. Chile and Mexico display more
+progressive patterns of recovery, although Chile never reaches baseline
+levels. These fluctuations are unique to each country and can be
+attributed to local factors such as the effects of seasonality or the
+different stringency measures imposed by the national governments during
+the pandemic.
 
@@
 
+\hypertarget{the-most-deprived-areas-experienced-the-smallest-drop-in-mobility-relative-to-pre-pandemic-levels}{%
 \subsection{The most deprived areas experienced the smallest drop in
 mobility relative to pre-pandemic
-levels}\label{the-most-deprived-areas-experienced-the-smallest-drop-in-mobility-relative-to-pre-pandemic-levels}
+levels}\label{the-most-deprived-areas-experienced-the-smallest-drop-in-mobility-relative-to-pre-pandemic-levels}}
 
@@
 time (i.e.~a month), we start by considering the relationship between
-the number of movements relative to the pre-pandemic baseline period and
-the average RDI, at the administrative unit level. We assume that this
-relationship is linear and we use a linear regression to estimate the
-slope and intercept characterising the line of best fit. This is shown
-for April 2020 and March 2022 in the right-hand side panels of Figure 3.
-After obtaining the slope and intercept for every month, we are able to
-plot the evolution of these parameters for both the observed and
-forecasted data, as displayed on the left-hand-side panels of the same
-Figure.
+the percentage change in the number of movements relative to the
+pre-pandemic baseline period and the average RDI, at the administrative
+unit level. We assume that this relationship is linear and we use a
+linear regression to estimate the slope and intercept characterising the
+line of best fit. This is shown for April 2020 and March 2022 in the
+right-hand side panels of Figure 3. After obtaining the slope and
+intercept for every month, we are able to plot the evolution of these
+parameters for both the observed and forecasted data, as displayed on
+the left-hand-side panels of the same Figure.
 
-\begin{figure}[H]
+\begin{figure}
 
@@
 \caption{Evolution of the parameters characterising the relationship
-between relative deprivation index and number of movements relative to
-the pre-pandemic baseline period.}
+between relative deprivation index and the percentage change in the
+number of movements relative to the pre-pandemic baseline level of
+mobility.}
 
-\end{figure}%
+\end{figure}
 
@@
 mobility patterns across socioeconomic groups. A slope equal to zero
-would mean that all administrative units display the same intensity of
-movement regardless of their average RDI. Given the patterns observed in
+would mean that all administrative units display the same level of
+mobility regardless of their average RDI. Given the patterns observed in
 Argentina, Colombia and Mexico, we find that at the beginning of the
@@
 
-\section{Discussion}\label{sec-discussion}
+\hypertarget{sec-discussion}{%
+\section{Discussion}\label{sec-discussion}}
 
@@
 
-\section{Methods}\label{sec-methods}
+\hypertarget{sec-methods}{%
+\section{Methods}\label{sec-methods}}
 
+\hypertarget{meta-facebook-movement-data}{%
 \subsection{Meta-Facebook movement
-data}\label{meta-facebook-movement-data}
+data}\label{meta-facebook-movement-data}}
 
@@
 the recovery trends after the pandemic, we generate a 12-month forecast
-up to March 2023. This is done using Prophet, a procedure to forecast
-time series data based on an additive model where non-linear trends can
-be fit with seasonality effects, holiday effects and other external
-factors such as the stringency index. For the analysis, only seasonality
-effects are included as they yield the most realistic predictions.
+up to March 2023. This is done using Prophet (Taylor and Letham 2017), a
+procedure developed by Meta's Core Data Science team to forecast time
+series data based on an additive model. Prophet stands out from other
+forecasting methods due to its ability to capture recurring seasonal
+trends inherent in mobility data, such as fluctuations during holidays,
+or changes in the general trend due to the impact of particular
+circumstances, such as the COVID-19 pandemic. It is particularly
+intuitive to use compared to other models due to its automated
+trend-detection capabilities, making it accessible to users with no
+expertise on the underlying model, while being robust to missing data
+and outliers. For the analysis in this paper, yearly seasonality effects
+are included as they yield the most realistic predictions.
 
-In Figures 2 and 3, outliers are removed. Outliers are defined as
-administrative units which, at any given month, have a z-score greater
-than 4 for the relative intensity of movement. Furthermore, a
-Savitzky-Golay filter is applied to smooth the time series data, using a
-length of 4 units for the filter window and order 2 polynomials to fit
-the samples.
+Figures 2 and 3 have been created by removing statistical outliers. We
+define these as the administrative units which, at any given month, have
+a z-score greater than 4 for the percentage change in the number of
+movements. Furthermore, a Savitzky-Golay filter is applied to smooth the
+time series data, using a length of 4 units for the filter window and
+order 2 polynomials to fit the samples.
 
-\section*{References}\label{references}
+\hypertarget{references}{%
+\section*{References}\label{references}}
 \addcontentsline{toc}{section}{References}
 
-\phantomsection\label{refs}
+\hypertarget{refs}{}
 \begin{CSLReferences}{1}{0}
-\bibitem[\citeproctext]{ref-LaurenAlexander15}
+\leavevmode\vadjust pre{\hypertarget{ref-LaurenAlexander15}{}}%
 Alexander, Lauren, Shan Jiang, Mikel Murga, and Marta C. González. 2015.
@@
 
-\bibitem[\citeproctext]{ref-Cabrera-Arnau2023}
+\leavevmode\vadjust pre{\hypertarget{ref-Cabrera-Arnau2023}{}}%
 Cabrera-Arnau, Carmen, Chen Zhong, Michael Batty, Ricardo Silva, and
@@
 
-\bibitem[\citeproctext]{ref-RDI2022}
+\leavevmode\vadjust pre{\hypertarget{ref-RDI2022}{}}%
 Columbia University, Center for International Earth Science Information
@@
 
-\bibitem[\citeproctext]{ref-ECJRC22}
+\leavevmode\vadjust pre{\hypertarget{ref-ECJRC22}{}}%
 Commission, European, Joint Research Centre, C Bosco, S
@@
 
-\bibitem[\citeproctext]{ref-UNSD}
+\leavevmode\vadjust pre{\hypertarget{ref-UNSD}{}}%
 Division, United Nations Statistical. 2019. {``Handbook on the Use of
@@
 
-\bibitem[\citeproctext]{ref-GADM24}
+\leavevmode\vadjust pre{\hypertarget{ref-GADM24}{}}%
 GADM. 2024. {``GADM Maps and Data.''}
 
-\bibitem[\citeproctext]{ref-Maas19}
+\leavevmode\vadjust pre{\hypertarget{ref-Maas19}{}}%
 Maas, P., S. Iyer, A. Gros, W. Park, L. McGorman, C. Nayak, and P. A.
@@
 
-\bibitem[\citeproctext]{ref-MengChuishi17}
+\leavevmode\vadjust pre{\hypertarget{ref-MengChuishi17}{}}%
 Meng, Chuishi, Yu Cui, Qing He, Lu Su, and Jing Gao. 2017. {``Travel
@@
 
-\bibitem[\citeproctext]{ref-NiluferSariAslam21}
+\leavevmode\vadjust pre{\hypertarget{ref-NiluferSariAslam21}{}}%
 Nilufer Sari Aslam, Tao Cheng, Mohamed R. Ibrahim, and Yang Zhang. 2021.
@@
 
-\bibitem[\citeproctext]{ref-owen1992migration}
+\leavevmode\vadjust pre{\hypertarget{ref-owen1992migration}{}}%
 Owen, D., and A. Green. 1992. {``Migration Patterns and Trends.''} In
@@
 
-\bibitem[\citeproctext]{ref-RoweBigData23}
+\leavevmode\vadjust pre{\hypertarget{ref-RoweBigData23}{}}%
 Rowe, Francisco. 2023. {``9.: Big Data.''} In \emph{Concise Encyclopedia
@@
 
-\bibitem[\citeproctext]{ref-roweCalafiore2022}
+\leavevmode\vadjust pre{\hypertarget{ref-roweCalafiore2022}{}}%
 Rowe, Francisco, Alessia Calafiore, Daniel Arribas-Bel, Krasen
@@
 
-\bibitem[\citeproctext]{ref-GHSL19}
+\leavevmode\vadjust pre{\hypertarget{ref-GHSL19}{}}%
 Schiavina M., Maffenini L., Moreno-Monroy A. 2019. {``GHS-FUA R2019A -
@@
 
-\bibitem[\citeproctext]{ref-wang2022}
+\leavevmode\vadjust pre{\hypertarget{ref-prophet17}{}}%
+Taylor, Sean J, and Benjamin Letham. 2017. {``Forecasting at Scale.''}
+\emph{PeerJ Preprints} 5 (September): e3190v2.
+\url{https://doi.org/10.7287/peerj.preprints.3190v2}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-wang2022}{}}%
 Wang, Yikang, Chen Zhong, Qili Gao, and Carmen Cabrera-Arnau. 2022.

### 2024-02-06 manuscript/manuscript.qmd
add refs to intro

@@
     name: Centre for Demographic Urban and Environmental Studies, El Colegio de México, Ciudad de México, México
-#  - id: 3
-#    name: International Institute for Applied Systems Analysis, Wittgenstein Centre, Vienna, Austria
 abstract: |
@@
 However, socioeconomic disparities are present through the period of analysis, with deprived areas experiencing smaller initial losses in mobility.
-These disparities generally diminish over time, although they persist in all the countries included in the analysis.
-Our research highlights the importance of timely mobility data with high spatiotemporal resolution to understand the long-term effects of the pandemic and to inform equitable policy responses that address societal challenges in urban areas.
+These disparities generally diminish over time, although they persist in all the countries included in the analysis <!--# They seem to converge in Argentina and Mexico -->.
+Our research highlights the importance of timely mobility data with high spatiotemporal resolution to understand the long-term effects of the pandemic and inform equitable policy responses that address societal challenges in urban areas.
 format: 
@@
 editor: visual
+csl: nature.csl
 bibliography: refs.bib
@@
 Spatial human mobility is key to creating sustainable, livable and inclusive cities.
-At the societal level, spatial mobility enables the transfer of knowledge, skills and labour to places they are needed.
-Spatial mobility also shapes service and transport demand across urban spaces, and enables the monitoring and control of transmissible diseases.
-At the individual level, mobility enables people to access and achieve opportunities and aspirations in space.
-Understanding spatial human mobility is thus important to supporting appropriate policy responses to address societal challenges relating to carbon emissions, urban planning, service delivery, public health, disaster management and transport.
+At the societal level, spatial mobility enables the transfer of knowledge, skills and labour to places they are needed [@ackers2005].
+Spatial mobility also shapes service and transport demand across urban spaces [@chen2016], and enables the monitoring and control of transmissible diseases [@belik2011].
+At the individual level, mobility enables people to access and achieve opportunities and aspirations in space [@klugman2009human].
+Understanding spatial human mobility is thus important to supporting appropriate policy responses to address societal challenges relating to carbon emissions, urban planning, service delivery, public health, disaster management and transport [@barbosa2018; @chinazzi2020].
 
-The COVID-19 pandemic resulted in a notable decrease in mobility, particularly in cities.
-This decrease in the overall levels of mobility was prompted by nonpharmaceutical interventions to contain the spread of COVID‐19, coupled with general fears of visiting crowded public spaces.Especially during lockdowns, mobility recorded reductions in the frequency, distance and time of trips in cities across the globe.
-Higher engagement with remote working, online schooling and shopping activity reduced the need to travel for work, education, shopping and leisure, hence giving rise to more geographically localised mobility patterns.
+The COVID-19 pandemic resulted in a notable decrease in mobility, particularly in cities [@nouvellet2021].
+Coupled with fears of contagion in crowded public spaces, nonpharmaceutical interventions to contain the spread of COVID‐19 prompted this decrease in the overall levels of mobility [@nouvellet2021; @rowe2023].
+Especially during lockdowns, mobility recorded reductions in the frequency, distance and time of trips in cities across the globe [@abdullah2020; @bonaccorsi2020; @abu-rayash2020; @lee2023].
+Higher engagement with remote working, online schooling and shopping activity reduced the need to travel for work, education, shopping and leisure, hence giving rise to more geographically localised mobility patterns [@borkowski2021].
 
-However, reductions in mobility levels were highly unequal reflecting existing socioeconomic inequalities.
-In most countries, affluent individuals tended to record the greatest drops in mobility levels as they are predominantly employed in knowledge-intensive jobs which can be done fully or partly remotely.
-During the COVID-19 pandemic, the adoption of remote work reduced the need of commuting for knowledge-intensive, non-public facing jobs.
-At the same time, individuals from less privileged socioeconomic backgrounds displayed less pronounced declines mirroring the nature of their work requiring public-facing, face-to-face interaction, and thus a requirement for daily work commutes.
+However, reductions in mobility levels were highly unequal reflecting existing socioeconomic inequalities in our societies [@chang2020].
+In most countries, affluent individuals tended to record the greatest drops in mobility levels as they are predominantly employed in knowledge-intensive jobs which can be done fully or partly remotely [@fraiberger2020; @bonaccorsi2020; @weill2020; @dueñas2021; @santana2023].
+During the COVID-19 pandemic, the adoption of remote work reduced the need of commuting for knowledge-intensive, non-public facing jobs [@florida2021].
+At the same time, individuals from less privileged socioeconomic backgrounds displayed less pronounced declines mirroring the nature of their work requiring public-facing, face-to-face interaction, and thus a requirement for daily work commutes [@dueñas2021; @santana2023].
 
 Thus, while a growing body of empirical evidence has contributed to advancing our understanding of the impacts of the COVID-19 pandemic on spatial mobility within cities, existing research has focused on more developed countries and the immediate effects of the pandemic during 2020.
-Less is known about the longer term patterns of resilience in urban mobility in less developed countries.
-Assessing the extent to which the level of mobility has returned back to the pre-pandemic baseline level across socioeconomic groups is important to understand the potentially unequal impacts of hybrid working in the society.
+Less is known about the longer term patterns of resilience in urban mobility in less developed countries extending beyond this period [@rowe2023a].
+Urban spaces have changed considerably since then, from going through waves of high COVID-19 fatality, infections, school and business closures to the removal of all COVID-19 restrictions as the UN World Health Organization (WHO) declared an end to the pandemic as a public health emergency; yet, different configurations of hybrid working have remained in the norm across many sector of the economy [@barrero2021; @aksoy2022].
+Thus, assessing the extent to which the level of mobility has returned back to the pre-pandemic baseline level across socioeconomic groups is important to understand the potentially unequal long-term impacts of hybrid working.
 
-A key barrier to monitor changes in geographic mobility patterns in less developed countries during and post the COVID-19 pandemic has been the lack of suitable data.
-Traditionally census and survey data have been employed to analyse human mobility patterns in these countries.
-Yet, these data streams are not frequently updated and suffer from slow releases, with census data for example being collected over intervals of ten years in most countries.
-Traditional data streams thus lack the temporal granularity to analyse population movements over short-time periods and to offer an up-to-date representation of the urban mobility system.
-Data resulting from social interactions on digital platforms have emerged as an unique source of information to deliver this representation and capture human population movement in less developed countries at scale.
-Particularly location data from mobile phone applications have become a prominent source to sense patterns of human mobility at higher geographical and temporal resolution in real time.
+A key barrier to monitor changes in geographic mobility patterns in less developed countries during and post the COVID-19 pandemic has been the lack of suitable data [@rowe2023a].
+Traditionally census and survey data have been employed to analyse human mobility patterns in these countries [@green2021].
+Yet, these data streams are not frequently updated and suffer from slow releases, with census data for example being collected over intervals of ten years in most countries [@bell2014].
+Traditional data streams thus lack the temporal granularity to analyse population movements over short-time periods and to offer an up-to-date representation of the urban mobility system [@rowe2023big].
+Data resulting from social interactions on digital platforms have emerged as an unique source of information to deliver this representation and capture human population movement in less developed countries at scale [@rowe2023big].
+Particularly location data from mobile phone applications have become a prominent source to sense patterns of human mobility at higher geographical and temporal resolution in real time [@calafiore2023].
 
 Drawing on a dataset of 213 million observations from Meta-Facebook users' mobile location data, we aim to assess socioeconomic differences in the extent and persistence of decline in urban mobility in Argentina, Chile, Colombia and Mexico during and after the COVID-19 pandemic from March 2020 to March 2023.
-We use Meta-Facebook data to measure origin-destination flows from March 2020 to May 2022, and Meta Prophet time-series forecasting machine learning algorithm to predict origin-destination flows from June 2022 to March 2023.
-<!--# Perhaps say how predictions have been validated --> We use Functional Urban Areas (FUAs) boundaries from the Global Human Settlement Layer (GHSL), developed by the European Comission's Joint Research Centre [@GHSL19] to define urban areas; and the Global Gridded Relative Deprivation Index (GRDI) developed by NASA’s Socioeconomic Data and Applications Centre [@RDI2022] from sociodemographic and satellite data inputs.
-Building on existing evidence (e.g.
-[@roweCalafiore2022; @wang2022]), we hypothesised that (1) urban mobility has recovered returning to the pre-pandemic baseline level of movement as nonpharmaceuthical restrictions were lifted; and (2) socioeconomic differences in urban mobility have endured the pandemic reflecting deep societal inequalities as knowledge-intensive businesses adopt hybrid working.
-<!--# Once we have a more definitive version of the results, highlight our novel contributions here. -->
+We use Meta-Facebook data to measure origin-destination flows from March 2020 to May 2022, and Meta Prophet time-series forecasting machine learning algorithm [@prophet17] to predict origin-destination flows from June 2022 to March 2023.
+<!--# Perhaps say how predictions have been validated --> We use Functional Urban Areas (FUAs) boundaries from the Global Human Settlement Layer (GHSL), developed by the European Comission's Joint Research Centre [@GHSL19] to define urban areas; and the Global Gridded Relative Deprivation Index (GRDI) developed by NASA's Socioeconomic Data and Applications Centre [@RDI2022] from sociodemographic and satellite data inputs.
+Building on existing evidence [e.g.
+@roweCalafiore2022; @wang2022], we hypothesised that (1) urban mobility has recovered returning to the pre-pandemic baseline level of movement as nonpharmaceuthical restrictions were lifted; and, that (2) socioeconomic differences in urban mobility have endured the pandemic reflecting deep societal inequalities as knowledge-intensive businesses adopt hybrid working.
+<!--# Once we have a more definitive version of the results, highlight our novel contributions here. -->\
 
--   Context and importance of human mobility: public health , climate change\
--   The pandemic resulted in major changes in human mobility - unequal impacts across socioeconomic groups\
--   Gap 1: Little work on understanding the long-term changes in mobility - recovery\
--   Gap 2: little work on less developed countries - where inequalities are more pronounced\
 -   Aim - focus on Latin American countries as case study.
@@
 ![Patterns of recovery for urban mobility in administrative units from Argentina, Chile, Colombia and Mexico.](figures/prediction-rdi-band.pdf)
+
+<!--# please add a description of what the bands mean. -->
 

### 2024-02-06 manuscript/manuscript.tex
add refs to intro

@@
 time, although they persist in all the countries included in the
-analysis. Our research highlights the importance of timely mobility data
-with high spatiotemporal resolution to understand the long-term effects
-of the pandemic and to inform equitable policy responses that address
-societal challenges in urban areas.
+analysis . Our research highlights the importance of timely mobility
+data with high spatiotemporal resolution to understand the long-term
+effects of the pandemic and inform equitable policy responses that
+address societal challenges in urban areas.
 \end{abstract}
-\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[boxrule=0pt, frame hidden, breakable, enhanced, interior hidden, sharp corners, borderline west={3pt}{0pt}{shadecolor}]}{\end{tcolorbox}}\fi
+\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[enhanced, frame hidden, breakable, sharp corners, interior hidden, borderline west={3pt}{0pt}{shadecolor}, boxrule=0pt]}{\end{tcolorbox}}\fi
 
@@
 inclusive cities. At the societal level, spatial mobility enables the
-transfer of knowledge, skills and labour to places they are needed.
-Spatial mobility also shapes service and transport demand across urban
-spaces, and enables the monitoring and control of transmissible
-diseases. At the individual level, mobility enables people to access and
-achieve opportunities and aspirations in space. Understanding spatial
-human mobility is thus important to supporting appropriate policy
-responses to address societal challenges relating to carbon emissions,
-urban planning, service delivery, public health, disaster management and
-transport.
+transfer of knowledge, skills and labour to places they are
+needed\textsuperscript{1}. Spatial mobility also shapes service and
+transport demand across urban spaces\textsuperscript{2}, and enables the
+monitoring and control of transmissible diseases\textsuperscript{3}. At
+the individual level, mobility enables people to access and achieve
+opportunities and aspirations in space\textsuperscript{4}. Understanding
+spatial human mobility is thus important to supporting appropriate
+policy responses to address societal challenges relating to carbon
+emissions, urban planning, service delivery, public health, disaster
+management and transport\textsuperscript{5,6}.
 
 The COVID-19 pandemic resulted in a notable decrease in mobility,
-particularly in cities. This decrease in the overall levels of mobility
-was prompted by nonpharmaceutical interventions to contain the spread of
-COVID‐19, coupled with general fears of visiting crowded public
-spaces.Especially during lockdowns, mobility recorded reductions in the
-frequency, distance and time of trips in cities across the globe. Higher
+particularly in cities\textsuperscript{7}. Coupled with fears of
+contagion in crowded public spaces, nonpharmaceutical interventions to
+contain the spread of COVID‐19 prompted this decrease in the overall
+levels of mobility\textsuperscript{7,8}. Especially during lockdowns,
+mobility recorded reductions in the frequency, distance and time of
+trips in cities across the globe\textsuperscript{9--12}. Higher
 engagement with remote working, online schooling and shopping activity
 reduced the need to travel for work, education, shopping and leisure,
-hence giving rise to more geographically localised mobility patterns.
+hence giving rise to more geographically localised mobility
+patterns\textsuperscript{13}.
 
 However, reductions in mobility levels were highly unequal reflecting
-existing socioeconomic inequalities. In most countries, affluent
-individuals tended to record the greatest drops in mobility levels as
-they are predominantly employed in knowledge-intensive jobs which can be
-done fully or partly remotely. During the COVID-19 pandemic, the
-adoption of remote work reduced the need of commuting for
-knowledge-intensive, non-public facing jobs. At the same time,
-individuals from less privileged socioeconomic backgrounds displayed
-less pronounced declines mirroring the nature of their work requiring
-public-facing, face-to-face interaction, and thus a requirement for
-daily work commutes.
+existing socioeconomic inequalities in our
+societies\textsuperscript{14}. In most countries, affluent individuals
+tended to record the greatest drops in mobility levels as they are
+predominantly employed in knowledge-intensive jobs which can be done
+fully or partly remotely\textsuperscript{10,15--18}. During the COVID-19
+pandemic, the adoption of remote work reduced the need of commuting for
+knowledge-intensive, non-public facing jobs\textsuperscript{19}. At the
+same time, individuals from less privileged socioeconomic backgrounds
+displayed less pronounced declines mirroring the nature of their work
+requiring public-facing, face-to-face interaction, and thus a
+requirement for daily work commutes\textsuperscript{17,18}.
 
@@
 2020. Less is known about the longer term patterns of resilience in
-urban mobility in less developed countries. Assessing the extent to
-which the level of mobility has returned back to the pre-pandemic
-baseline level across socioeconomic groups is important to understand
-the potentially unequal impacts of hybrid working in the society.
+urban mobility in less developed countries extending beyond this
+period\textsuperscript{20}. Urban spaces have changed considerably since
+then, from going through waves of high COVID-19 fatality, infections,
+school and business closures to the removal of all COVID-19 restrictions
+as the UN World Health Organization (WHO) declared an end to the
+pandemic as a public health emergency; yet, different configurations of
+hybrid working have remained in the norm across many sector of the
+economy\textsuperscript{21,22}. Thus, assessing the extent to which the
+level of mobility has returned back to the pre-pandemic baseline level
+across socioeconomic groups is important to understand the potentially
+unequal long-term impacts of hybrid working.
 
@@
 developed countries during and post the COVID-19 pandemic has been the
-lack of suitable data. Traditionally census and survey data have been
-employed to analyse human mobility patterns in these countries. Yet,
-these data streams are not frequently updated and suffer from slow
-releases, with census data for example being collected over intervals of
-ten years in most countries. Traditional data streams thus lack the
+lack of suitable data\textsuperscript{20}. Traditionally census and
+survey data have been employed to analyse human mobility patterns in
+these countries\textsuperscript{23}. Yet, these data streams are not
+frequently updated and suffer from slow releases, with census data for
+example being collected over intervals of ten years in most
+countries\textsuperscript{24}. Traditional data streams thus lack the
 temporal granularity to analyse population movements over short-time
 periods and to offer an up-to-date representation of the urban mobility
-system. Data resulting from social interactions on digital platforms
-have emerged as an unique source of information to deliver this
-representation and capture human population movement in less developed
-countries at scale. Particularly location data from mobile phone
-applications have become a prominent source to sense patterns of human
-mobility at higher geographical and temporal resolution in real time.
+system\textsuperscript{25}. Data resulting from social interactions on
+digital platforms have emerged as an unique source of information to
+deliver this representation and capture human population movement in
+less developed countries at scale\textsuperscript{25}. Particularly
+location data from mobile phone applications have become a prominent
+source to sense patterns of human mobility at higher geographical and
+temporal resolution in real time\textsuperscript{26}.
 
@@
 origin-destination flows from March 2020 to May 2022, and Meta Prophet
-time-series forecasting machine learning algorithm to predict
-origin-destination flows from June 2022 to March 2023. We use Functional
-Urban Areas (FUAs) boundaries from the Global Human Settlement Layer
-(GHSL), developed by the European Comission's Joint Research Centre
-(Schiavina M. 2019) to define urban areas; and the Global Gridded
-Relative Deprivation Index (GRDI) developed by NASA's Socioeconomic Data
-and Applications Centre (Columbia University 2022) from sociodemographic
-and satellite data inputs. Building on existing evidence (e.g. (Rowe et
-al. 2022; Wang et al. 2022)), we hypothesised that (1) urban mobility
-has recovered returning to the pre-pandemic baseline level of movement
-as nonpharmaceuthical restrictions were lifted; and (2) socioeconomic
-differences in urban mobility have endured the pandemic reflecting deep
-societal inequalities as knowledge-intensive businesses adopt hybrid
-working.
+time-series forecasting machine learning algorithm\textsuperscript{27}
+to predict origin-destination flows from June 2022 to March 2023. We use
+Functional Urban Areas (FUAs) boundaries from the Global Human
+Settlement Layer (GHSL), developed by the European Comission's Joint
+Research Centre\textsuperscript{28} to define urban areas; and the
+Global Gridded Relative Deprivation Index (GRDI) developed by NASA's
+Socioeconomic Data and Applications Centre\textsuperscript{29} from
+sociodemographic and satellite data inputs. Building on existing
+evidence\textsuperscript{e.g. 30,31}, we hypothesised that (1) urban
+mobility has recovered returning to the pre-pandemic baseline level of
+movement as nonpharmaceuthical restrictions were lifted; and, that (2)
+socioeconomic differences in urban mobility have endured the pandemic
+reflecting deep societal inequalities as knowledge-intensive businesses
+adopt hybrid working. \\
 
@@
 \tightlist
-\item
-  Context and importance of human mobility: public health , climate
-  change\\
-\item
-  The pandemic resulted in major changes in human mobility - unequal
-  impacts across socioeconomic groups\\
-\item
-  Gap 1: Little work on understanding the long-term changes in mobility
-  - recovery\\
-\item
-  Gap 2: little work on less developed countries - where inequalities
-  are more pronounced\\
 \item
@@
 spatially into administrative units at various levels according to GADM,
-the Database of Global Administrative Areas (GADM 2024). The analysis
-focuses on administrative units that are within the boundaries of
-functional urban areas as specified by the Global Human Settlement Layer
-(GHSL). For each administrative unit, we compute the Relative
+the Database of Global Administrative Areas\textsuperscript{32}. The
+analysis focuses on administrative units that are within the boundaries
+of functional urban areas as specified by the Global Human Settlement
+Layer (GHSL). For each administrative unit, we compute the Relative
 Deprivation Index (RDI) based on data from NASA's Socioeconomic Data and
@@
 Specifically, we focus the analysis on short-distance movements in urban
-areas to represent local and routine mobility (Owen and Green 1992), so
+areas to represent local and routine mobility\textsuperscript{33}, so
 only movements covering a distance of at most 70 km are considered. For
@@
 especifically those covering 70 km or less. These journeys are typically
-considered to represent local and routine mobility (Owen and Green 1992)
-. However, due to the characteristics of the Meta-Facebook movement
-data, we are unable to distinguish the purpose of these short-distance
+considered to represent local and routine mobility\textsuperscript{33} .
+However, due to the characteristics of the Meta-Facebook movement data,
+we are unable to distinguish the purpose of these short-distance
 movements. Hence, some of our data could be capturing journeys that
@@
 should focus on inferring more specific information about the nature of
-the journeys, following similar approaches to those proposed by
-Cabrera-Arnau et al. (2023), and quantifying the extent to which the
-digital footprint data mirrors the true mobility patterns.
+the journeys, following similar approaches to those proposed
+by\textsuperscript{37}, and quantifying the extent to which the digital
+footprint data mirrors the true mobility patterns.
 
@@
 representation issues. These biases often arise from inequalities in
-access and usage of digital technologies across demographic groups (Rowe
-2023). Despite these challenges, the data and analysis that we used for
-this work provide evidence for non-trivial patterns that are consistent
-across four countries in Latin America and with other parts of the
-world. Our findings highlight the dynamic interplay between
-socioeconomic status and urban mobility, and shall be used to motivate
-and inform the public debate regarding the deep societal consequences of
-urban mobility disparities on the wider socioeconomic landscape of Latin
-American countries.
+access and usage of digital technologies across demographic
+groups\textsuperscript{38}. Despite these challenges, the data and
+analysis that we used for this work provide evidence for non-trivial
+patterns that are consistent across four countries in Latin America and
+with other parts of the world. Our findings highlight the dynamic
+interplay between socioeconomic status and urban mobility, and shall be
+used to motivate and inform the public debate regarding the deep
+societal consequences of urban mobility disparities on the wider
+socioeconomic landscape of Latin American countries.
 
@@
 creation of resources like those developed by the European Commission
-Joint Research Centre (Commission et al. 2022) and the UN Statistics
-Division (Division 2019), which identify sources of non-traditional data
-and set methodological protocols for incorporating mobile phone data
-into official mobility statistics. While current resources tend to have
-a global reach, we advocate for more tailored local initiatives that
-acknowledge disparities in regional data availability and adoption of
-digital technologies.
+Joint Research Centre\textsuperscript{39} and the UN Statistics
+Division\textsuperscript{40}, which identify sources of non-traditional
+data and set methodological protocols for incorporating mobile phone
+data into official mobility statistics. While current resources tend to
+have a global reach, we advocate for more tailored local initiatives
+that acknowledge disparities in regional data availability and adoption
+of digital technologies.
 
@@
 datasets, Meta ensures privacy and anonymity by removing personal
-information and applying privacy-preserving techniques (Maas et al.
-2019). Small-count dropping is one of these techniques. A data entry is
-removed if the population or movement count for an area is lower than
-10. The removal of small counts may mean that population counts in small
-sparsely populated areas are not captured. A second technique consists
-in adding a small undisclosed amount of random noise to ensure that it
-is not possible to ascertain precise, true counts for sparsely populated
-locations. Third, spatial smoothing using inverse distance-weighted
-averaging is also applied is applied to produce a smooth population
-count surface. The Facebook Movements dataset offers information on the
-total number of Facebook users moving between and within spatial units
-in the form of origin-destination matrices. The data is temporally
-aggregated into three daily 8-hour time windows (i.e.~00:00-08:00,
-08:00-16:00 and 16:00-00:00). The dataset includes a baseline capturing
-the number of movements before COVID-19 based on a 45-day period ending
-on March 10th 2020. The baseline is computed using an average for the
-same time of the day and day of the week in the period preceding March
-10th. For instance, the baseline for Monday 00:00-08:00 time window is
-obtained by averaging over data collected on Mondays from 00:00 to 8:00
-for the 45-day period. Details about the baseline can be found in (Maas
-et al. 2019). The Bing Maps Tile System developed by Microsoft
-(Microsoft) is used a spatial framework to organise the data. The Tile
-System is a geospatial indexing system that partitions the world into
-tile cells in a hierarchical way, comprising 23 different levels of
-detail (Microsoft). At the lowest level of detail (Level 1), the world
-is divided into four tiles with a coarse spatial resolution. At each
-successive level, the resolution increases by a factor of two. The data
-that we used are spatially aggregated into Bing tile levels 13. That is
-about 4.9 x 4.9km at the Equator (Maas et al. 2019).
+information and applying privacy-preserving
+techniques\textsuperscript{41}. Small-count dropping is one of these
+techniques. A data entry is removed if the population or movement count
+for an area is lower than 10. The removal of small counts may mean that
+population counts in small sparsely populated areas are not captured. A
+second technique consists in adding a small undisclosed amount of random
+noise to ensure that it is not possible to ascertain precise, true
+counts for sparsely populated locations. Third, spatial smoothing using
+inverse distance-weighted averaging is also applied is applied to
+produce a smooth population count surface. The Facebook Movements
+dataset offers information on the total number of Facebook users moving
+between and within spatial units in the form of origin-destination
+matrices. The data is temporally aggregated into three daily 8-hour time
+windows (i.e.~00:00-08:00, 08:00-16:00 and 16:00-00:00). The dataset
+includes a baseline capturing the number of movements before COVID-19
+based on a 45-day period ending on March 10th 2020. The baseline is
+computed using an average for the same time of the day and day of the
+week in the period preceding March 10th. For instance, the baseline for
+Monday 00:00-08:00 time window is obtained by averaging over data
+collected on Mondays from 00:00 to 8:00 for the 45-day period. Details
+about the baseline can be found in\textsuperscript{41}. The Bing Maps
+Tile System developed by Microsoft (Microsoft) is used a spatial
+framework to organise the data. The Tile System is a geospatial indexing
+system that partitions the world into tile cells in a hierarchical way,
+comprising 23 different levels of detail (Microsoft). At the lowest
+level of detail (Level 1), the world is divided into four tiles with a
+coarse spatial resolution. At each successive level, the resolution
+increases by a factor of two. The data that we used are spatially
+aggregated into Bing tile levels 13. That is about 4.9 x 4.9km at the
+Equator\textsuperscript{41}.
 
@@
 the recovery trends after the pandemic, we generate a 12-month forecast
-up to March 2023. This is done using Prophet (Taylor and Letham 2017), a
+up to March 2023. This is done using Prophet\textsuperscript{27}, a
 procedure developed by Meta's Core Data Science team to forecast time
@@
 \hypertarget{refs}{}
-\begin{CSLReferences}{1}{0}
+\begin{CSLReferences}{0}{0}
+\leavevmode\vadjust pre{\hypertarget{ref-ackers2005}{}}%
+\CSLLeftMargin{1. }%
+\CSLRightInline{Ackers, L.
+\href{https://doi.org/10.1111/j.1468-2435.2005.00343.x}{Moving People
+and Knowledge: Scientific Mobility in the European Union1}.
+\emph{International Migration} \textbf{43}, 99--131 (2005).}
+
+\leavevmode\vadjust pre{\hypertarget{ref-chen2016}{}}%
+\CSLLeftMargin{2. }%
+\CSLRightInline{Chen, C., Ma, J., Susilo, Y., Liu, Y. \& Wang, M.
+\href{https://doi.org/10.1016/j.trc.2016.04.005}{The promises of big
+data and small data for travel behavior (aka human mobility) analysis}.
+\emph{Transportation Research Part C: Emerging Technologies}
+\textbf{68}, 285--299 (2016).}
+
+\leavevmode\vadjust pre{\hypertarget{ref-belik2011}{}}%
+\CSLLeftMargin{3. }%
+\CSLRightInline{Belik, V., Geisel, T. \& Brockmann, D.
+\href{https://doi.org/10.1103/physrevx.1.011001}{Natural Human Mobility
+Patterns and Spatial Spread of Infectious Diseases}. \emph{Physical
+Review X} \textbf{1}, (2011).}
+
+\leavevmode\vadjust pre{\hypertarget{ref-klugman2009human}{}}%
+\CSLLeftMargin{4. }%
+\CSLRightInline{Klugman, J. Human development report 2009. Overcoming
+barriers: Human mobility and development. \emph{Overcoming Barriers:
+Human Mobility and Development (October 5, 2009). UNDP-HDRO Human
+Development Reports} (2009).}
+
+\leavevmode\vadjust pre{\hypertarget{ref-barbosa2018}{}}%
+\CSLLeftMargin{5. }%
+\CSLRightInline{Barbosa, H. \emph{et al.}
+\href{https://doi.org/10.1016/j.physrep.2018.01.001}{Human mobility:
+Models and applications}. \emph{Physics Reports} \textbf{734}, 1--74
+(2018).}
+
+\leavevmode\vadjust pre{\hypertarget{ref-chinazzi2020}{}}%
+\CSLLeftMargin{6. }%
+\CSLRightInline{Chinazzi, M. \emph{et al.}
+\href{https://doi.org/10.1126/science.aba9757}{The effect of travel
+restrictions on the spread of the 2019 novel coronavirus (COVID-19)
+outbreak}. \emph{Science} \textbf{368}, 395--400 (2020).}
+
+\leavevmode\vadjust pre{\hypertarget{ref-nouvellet2021}{}}%
+\CSLLeftMargin{7. }%
+\CSLRightInline{Nouvellet, P. \emph{et al.}
+\href{https://doi.org/10.1038/s41467-021-21358-2}{Reduction in mobility
+and COVID-19 transmission}. \emph{Nature Communications} \textbf{12},
+(2021).}
+
+\leavevmode\vadjust pre{\hypertarget{ref-rowe2023}{}}%
+\CSLLeftMargin{8. }%
+\CSLRightInline{Rowe, F., González-Leonardo, M. \& Champion, T. Virtual
+special issue: Internal migration in times of COVID{-}19.
+\emph{Population, Space and Place} (2023)
+doi:\href{https://doi.org/10.1002/psp.2652}{10.1002/psp.2652}.}
+
+\leavevmode\vadjust pre{\hypertarget{ref-abdullah2020}{}}%
+\CSLLeftMargin{9. }%
+\CSLRightInline{Abdullah, M., Dias, C., Muley, D. \& Shahin, Md.
+\href{https://doi.org/10.1016/j.trip.2020.100255}{Exploring the impacts
+of COVID-19 on travel behavior and mode preferences}.
+\emph{Transportation Research Interdisciplinary Perspectives}
+\textbf{8}, 100255 (2020).}
+
+\leavevmode\vadjust pre{\hypertarget{ref-bonaccorsi2020}{}}%
+\CSLLeftMargin{10. }%
+\CSLRightInline{Bonaccorsi, G. \emph{et al.}
+\href{https://doi.org/10.1073/pnas.2007658117}{Economic and social
+consequences of human mobility restrictions under COVID-19}.
+\emph{Proceedings of the National Academy of Sciences} \textbf{117},
+15530--15535 (2020).}
+
+\leavevmode\vadjust pre{\hypertarget{ref-abu-rayash2020}{}}%
+\CSLLeftMargin{11. }%
+\CSLRightInline{Abu-Rayash, A. \& Dincer, I.
+\href{https://doi.org/10.1016/j.erss.2020.101693}{Analysis of mobility
+trends during the COVID-19 coronavirus pandemic: Exploring the impacts
+on global aviation and travel in selected cities}. \emph{Energy Research
+\& Social Science} \textbf{68}, 101693 (2020).}
+
+\leavevmode\vadjust pre{\hypertarget{ref-lee2023}{}}%
+\CSLLeftMargin{12. }%
+\CSLRightInline{Lee, S., Ko, E., Jang, K. \& Kim, S.
+\href{https://doi.org/10.1016/j.cities.2023.104223}{Understanding
+individual-level travel behavior changes due to COVID-19: Trip
+frequency, trip regularity, and trip distance}. \emph{Cities}
+\textbf{135}, 104223 (2023).}
+
+\leavevmode\vadjust pre{\hypertarget{ref-borkowski2021}{}}%
+\CSLLeftMargin{13. }%
+\CSLRightInline{Borkowski, P., Jażdżewska-Gutta, M. \& Szmelter-Jarosz,
+A. \href{https://doi.org/10.1016/j.jtrangeo.2020.102906}{Lockdowned:
+Everyday mobility changes in response to COVID-19}. \emph{Journal of
+Transport Geography} \textbf{90}, 102906 (2021).}
+
+\leavevmode\vadjust pre{\hypertarget{ref-chang2020}{}}%
+\CSLLeftMargin{14. }%
+\CSLRightInline{Chang, S. \emph{et al.}
+\href{https://doi.org/10.1038/s41586-020-2923-3}{Mobility network models
+of COVID-19 explain inequities and inform reopening}. \emph{Nature}
+\textbf{589}, 82--87 (2020).}
+
+\leavevmode\vadjust pre{\hypertarget{ref-fraiberger2020}{}}%
+\CSLLeftMargin{15. }%
+\CSLRightInline{Fraiberger, S. P. \emph{et al.} Uncovering socioeconomic
+gaps in mobility reduction during the COVID-19 pandemic using location
+data. (2020)
+doi:\href{https://doi.org/10.48550/ARXIV.2006.15195}{10.48550/ARXIV.2006.15195}.}
+
+\leavevmode\vadjust pre{\hypertarget{ref-weill2020}{}}%
+\CSLLeftMargin{16. }%
+\CSLRightInline{Weill, J. A., Stigler, M., Deschenes, O. \& Springborn,
+M. R. \href{https://doi.org/10.1073/pnas.2009412117}{Social distancing
+responses to COVID-19 emergency declarations strongly differentiated by
+income}. \emph{Proceedings of the National Academy of Sciences}
+\textbf{117}, 19658--19660 (2020).}
+
+\leavevmode\vadjust pre{\hypertarget{ref-dueuxf1as2021}{}}%
+\CSLLeftMargin{17. }%
+\CSLRightInline{Dueñas, M., Campi, M. \& Olmos, L. E.
+\href{https://doi.org/10.1057/s41599-021-00775-0}{Changes in mobility
+and socioeconomic conditions during the COVID-19 outbreak}.
+\emph{Humanities and Social Sciences Communications} \textbf{8},
+(2021).}
+
+\leavevmode\vadjust pre{\hypertarget{ref-santana2023}{}}%
+\CSLLeftMargin{18. }%
+\CSLRightInline{Santana, C. \emph{et al.} COVID-19 is linked to changes
+in the time{\textendash}space dimension of human mobility. \emph{Nature
+Human Behaviour} (2023)
+doi:\href{https://doi.org/10.1038/s41562-023-01660-3}{10.1038/s41562-023-01660-3}.}
+
+\leavevmode\vadjust pre{\hypertarget{ref-florida2021}{}}%
+\CSLLeftMargin{19. }%
+\CSLRightInline{Florida, R., Rodríguez-Pose, A. \& Storper, M.
+\href{https://doi.org/10.1177/00420980211018072}{Critical Commentary:
+Cities in a post-COVID world}. \emph{Urban Studies} \textbf{60},
+1509--1531 (2021).}
+
+\leavevmode\vadjust pre{\hypertarget{ref-rowe2023a}{}}%
+\CSLLeftMargin{20. }%
+\CSLRightInline{Rowe, F., Cabrera-Arnau, C., González-Leonardo, M.,
+Nasuto, A. \& Neville, R. Reduced mobility? Urban exodus? Medium-term
+impacts of the COVID-19 pandemic on internal population movements in
+latin american countries. (2023)
+doi:\href{https://doi.org/10.48550/ARXIV.2311.01464}{10.48550/ARXIV.2311.01464}.}
+
+\leavevmode\vadjust pre{\hypertarget{ref-barrero2021}{}}%
+\CSLLeftMargin{21. }%
+\CSLRightInline{Barrero, J. M., Bloom, N. \& Davis, S. \emph{Why working
+from home will stick}. \url{http://dx.doi.org/10.3386/w28731} (2021)
+doi:\href{https://doi.org/10.3386/w28731}{10.3386/w28731}.}
+
+\leavevmode\vadjust pre{\hypertarget{ref-aksoy2022}{}}%
+\CSLLeftMargin{22. }%
+\CSLRightInline{Aksoy, C. G. \emph{et al.} \emph{Working from home
+around the world}. \url{http://dx.doi.org/10.3386/w30446} (2022)
+doi:\href{https://doi.org/10.3386/w30446}{10.3386/w30446}.}
+
+\leavevmode\vadjust pre{\hypertarget{ref-green2021}{}}%
+\CSLLeftMargin{23. }%
+\CSLRightInline{Green, M., Pollock, F. D. \& Rowe, F. New forms of data
+and new forms of opportunities to monitor and tackle a pandemic. in
+423--429 (Springer International Publishing, 2021).
+doi:\href{https://doi.org/10.1007/978-3-030-70179-6_56}{10.1007/978-3-030-70179-6\_56}.}
+
+\leavevmode\vadjust pre{\hypertarget{ref-bell2014}{}}%
+\CSLLeftMargin{24. }%
+\CSLRightInline{Bell, M. \emph{et al.}
+\href{https://doi.org/10.1002/psp.1848}{Internal Migration Data Around
+the World: Assessing Contemporary Practice}. \emph{Population, Space and
+Place} \textbf{21}, 1--17 (2014).}
+
+\leavevmode\vadjust pre{\hypertarget{ref-rowe2023big}{}}%
+\CSLLeftMargin{25. }%
+\CSLRightInline{Rowe, F. Big data. in \emph{Concise encyclopedia of
+human geography} 42--47 (Edward Elgar Publishing, 2023).}
+
+\leavevmode\vadjust pre{\hypertarget{ref-calafiore2023}{}}%
+\CSLLeftMargin{26. }%
+\CSLRightInline{Calafiore, A., Samardzhiev, K., Rowe, F., Fleischmann,
+M. \& Arribas-Bel, D. Inequalities in experiencing urban functions. An
+exploration of human digital (geo-)footprints. \emph{Environment and
+Planning B: Urban Analytics and City Science} (2023)
+doi:\href{https://doi.org/10.1177/23998083231208507}{10.1177/23998083231208507}.}
+
+\leavevmode\vadjust pre{\hypertarget{ref-prophet17}{}}%
+\CSLLeftMargin{27. }%
+\CSLRightInline{Taylor, S. J. \& Letham, B.
+\href{https://doi.org/10.7287/peerj.preprints.3190v2}{Forecasting at
+scale}. \emph{PeerJ Preprints} \textbf{5}, e3190v2 (2017).}
+
+\leavevmode\vadjust pre{\hypertarget{ref-GHSL19}{}}%
+\CSLLeftMargin{28. }%
+\CSLRightInline{Schiavina M., M. L., Moreno-Monroy A. GHS-FUA R2019A -
+GHS functional urban areas, derived from GHS-UCDB R2019A, (2015),
+R2019A. (2019).}
+
+\leavevmode\vadjust pre{\hypertarget{ref-RDI2022}{}}%
+\CSLLeftMargin{29. }%
+\CSLRightInline{Columbia University, C. for I. E. S. I. Network. C.
+Global gridded relative deprivation index (GRDI), version 1. (2022)
+doi:\href{https://doi.org/10.7927/3xxe-ap97}{10.7927/3xxe-ap97}.}
+
+\leavevmode\vadjust pre{\hypertarget{ref-roweCalafiore2022}{}}%
+\CSLLeftMargin{30. }%
+\CSLRightInline{Rowe, F., Calafiore, A., Arribas-Bel, D., Samardzhiev,
+K. \& Fleischmann, M. Urban exodus? Understanding human mobility in
+britain during the COVID-19 pandemic using facebook data. (2022)
+doi:\href{https://doi.org/10.48550/ARXIV.2206.03272}{10.48550/ARXIV.2206.03272}.}
+
+\leavevmode\vadjust pre{\hypertarget{ref-wang2022}{}}%
+\CSLLeftMargin{31. }%
+\CSLRightInline{Wang, Y., Zhong, C., Gao, Q. \& Cabrera-Arnau, C.
+\href{https://doi.org/10.1007/s44212-022-00018-w}{Understanding internal
+migration in the UK before and during the COVID-19 pandemic using
+twitter data}. \emph{Urban Informatics} \textbf{1}, (2022).}
+
+\leavevmode\vadjust pre{\hypertarget{ref-GADM24}{}}%
+\CSLLeftMargin{32. }%
+\CSLRightInline{GADM. GADM maps and data. (2024).}
+
+\leavevmode\vadjust pre{\hypertarget{ref-owen1992migration}{}}%
+\CSLLeftMargin{33. }%
+\CSLRightInline{Owen, D. \& Green, A. Migration patterns and trends. in
+\emph{Migration processes and patterns: Research progress and prospects}
+(eds. Champion, T. \& Fielding, T.) (Belhaven Press, 1992).}
+
 \leavevmode\vadjust pre{\hypertarget{ref-LaurenAlexander15}{}}%
-Alexander, Lauren, Shan Jiang, Mikel Murga, and Marta C. González. 2015.
-{``Origin--Destination Trips by Purpose and Time of Day Inferred from
-Mobile Phone Data.''} \emph{Transportation Research Part C: Emerging
-Technologies} 58: 240--50.
-https://doi.org/\url{https://doi.org/10.1016/j.trc.2015.02.018}.
+\CSLLeftMargin{34. }%
+\CSLRightInline{Alexander, L., Jiang, S., Murga, M. \& González, M. C.
+\href{https://doi.org/10.1016/j.trc.2015.02.018}{Origin--destination
+trips by purpose and time of day inferred from mobile phone data}.
+\emph{Transportation Research Part C: Emerging Technologies}
+\textbf{58}, 240--250 (2015).}
+
+\leavevmode\vadjust pre{\hypertarget{ref-MengChuishi17}{}}%
+\CSLLeftMargin{35. }%
+\CSLRightInline{Meng, C., Cui, Y., He, Q., Su, L. \& Gao, J. Travel
+purpose inference with GPS trajectories, POIs, and geo-tagged social
+media data. in \emph{2017 IEEE international conference on big data (big
+data)} 1319--1324 (2017).
+doi:\href{https://doi.org/10.1109/BigData.2017.8258062}{10.1109/BigData.2017.8258062}.}
+
+\leavevmode\vadjust pre{\hypertarget{ref-NiluferSariAslam21}{}}%
+\CSLLeftMargin{36. }%
+\CSLRightInline{Nilufer Sari Aslam, T. C., Mohamed R. Ibrahim \& Zhang,
+Y. \href{https://doi.org/10.1080/10095020.2021.1985943}{ActivityNET:
+Neural networks to predict public transport trip purposes from
+individual smart card data and POIs}. \emph{Geo-spatial Information
+Science} \textbf{24}, 711--721 (2021).}
 
 \leavevmode\vadjust pre{\hypertarget{ref-Cabrera-Arnau2023}{}}%
-Cabrera-Arnau, Carmen, Chen Zhong, Michael Batty, Ricardo Silva, and
-Soong Moon Kang. 2023. {``Inferring Urban Polycentricity from the
-Variability in Human Mobility Patterns.''} \emph{Scientific Reports} 13
-(1): 5751. \url{https://doi.org/10.1038/s41598-023-33003-7}.
+\CSLLeftMargin{37. }%
+\CSLRightInline{Cabrera-Arnau, C., Zhong, C., Batty, M., Silva, R. \&
+Kang, S. M. \href{https://doi.org/10.1038/s41598-023-33003-7}{Inferring
+urban polycentricity from the variability in human mobility patterns}.
+\emph{Scientific Reports} \textbf{13}, 5751 (2023).}
 
-\leavevmode\vadjust pre{\hypertarget{ref-RDI2022}{}}%
-Columbia University, Center for International Earth Science Information
-Network. CIESIN. 2022. {``Global Gridded Relative Deprivation Index
-(GRDI), Version 1.''} Palisades, New York: NASA Socioeconomic Data;
-Applications Center (SEDAC). \url{https://doi.org/10.7927/3xxe-ap97}.
+\leavevmode\vadjust pre{\hypertarget{ref-RoweBigData23}{}}%
+\CSLLeftMargin{38. }%
+\CSLRightInline{Rowe, F. 9.: Big data. in \emph{Concise encyclopedia of
+human geography} 42--47 (Edward Elgar Publishing, 2023).
+doi:\href{https://doi.org/10.4337/9781800883499.ch09}{10.4337/9781800883499.ch09}.}
 
 \leavevmode\vadjust pre{\hypertarget{ref-ECJRC22}{}}%
-Commission, European, Joint Research Centre, C Bosco, S
-Grubanov-Boskovic, S Iacus, U Minora, F Sermi, and S Spyratos. 2022.
-\emph{Data Innovation in Demography, Migration and Human Mobility}.
-Publications Office of the European Union.
-\url{https://doi.org/doi/10.2760/027157}.
+\CSLLeftMargin{39. }%
+\CSLRightInline{Commission, E. \emph{et al.} \emph{Data innovation in
+demography, migration and human mobility}. (Publications Office of the
+European Union, 2022).
+doi:\href{https://doi.org/doi/10.2760/027157}{doi/10.2760/027157}.}
 
 \leavevmode\vadjust pre{\hypertarget{ref-UNSD}{}}%
-Division, United Nations Statistical. 2019. {``Handbook on the Use of
-Mobile Phone Data for Official Statistics.''}
-
-\leavevmode\vadjust pre{\hypertarget{ref-GADM24}{}}%
-GADM. 2024. {``GADM Maps and Data.''}
+\CSLLeftMargin{40. }%
+\CSLRightInline{Division, U. N. S. Handbook on the use of mobile phone
+data for official statistics. (2019).}
 
 \leavevmode\vadjust pre{\hypertarget{ref-Maas19}{}}%
-Maas, P., S. Iyer, A. Gros, W. Park, L. McGorman, C. Nayak, and P. A.
-Dow. 2019. {``Facebook Disaster Maps: Aggregate Insights for Crisis
-Response and Recovery.''} In \emph{16th International Conference on
-Information Systems for Crisis Response and Management}, 836--47.
-
-\leavevmode\vadjust pre{\hypertarget{ref-MengChuishi17}{}}%
-Meng, Chuishi, Yu Cui, Qing He, Lu Su, and Jing Gao. 2017. {``Travel
-Purpose Inference with GPS Trajectories, POIs, and Geo-Tagged Social
-Media Data.''} In \emph{2017 IEEE International Conference on Big Data
-(Big Data)}, 1319--24.
-\url{https://doi.org/10.1109/BigData.2017.8258062}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-NiluferSariAslam21}{}}%
-Nilufer Sari Aslam, Tao Cheng, Mohamed R. Ibrahim, and Yang Zhang. 2021.
-{``ActivityNET: Neural Networks to Predict Public Transport Trip
-Purposes from Individual Smart Card Data and POIs.''} \emph{Geo-Spatial
-Information Science} 24 (4): 711--21.
-\url{https://doi.org/10.1080/10095020.2021.1985943}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-owen1992migration}{}}%
-Owen, D., and A. Green. 1992. {``Migration Patterns and Trends.''} In
-\emph{Migration Processes and Patterns: Research Progress and
-Prospects}, edited by T. Champion and T. Fielding. Belhaven Press.
-
-\leavevmode\vadjust pre{\hypertarget{ref-RoweBigData23}{}}%
-Rowe, Francisco. 2023. {``9.: Big Data.''} In \emph{Concise Encyclopedia
-of Human Geography}, 42--47. Cheltenham, UK: Edward Elgar Publishing.
-\url{https://doi.org/10.4337/9781800883499.ch09}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-roweCalafiore2022}{}}%
-Rowe, Francisco, Alessia Calafiore, Daniel Arribas-Bel, Krasen
-Samardzhiev, and Martin Fleischmann. 2022. {``Urban Exodus?
-Understanding Human Mobility in Britain During the COVID-19 Pandemic
-Using Facebook Data.''} \url{https://doi.org/10.48550/ARXIV.2206.03272}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-GHSL19}{}}%
-Schiavina M., Maffenini L., Moreno-Monroy A. 2019. {``GHS-FUA R2019A -
-GHS Functional Urban Areas, Derived from GHS-UCDB R2019A, (2015),
-R2019A.''} European Commission, Joint Research Centre (JRC).
-
-\leavevmode\vadjust pre{\hypertarget{ref-prophet17}{}}%
-Taylor, Sean J, and Benjamin Letham. 2017. {``Forecasting at Scale.''}
-\emph{PeerJ Preprints} 5 (September): e3190v2.
-\url{https://doi.org/10.7287/peerj.preprints.3190v2}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-wang2022}{}}%
-Wang, Yikang, Chen Zhong, Qili Gao, and Carmen Cabrera-Arnau. 2022.
-{``Understanding Internal Migration in the UK Before and During the
-COVID-19 Pandemic Using Twitter Data.''} \emph{Urban Informatics} 1 (1).
-\url{https://doi.org/10.1007/s44212-022-00018-w}.
+\CSLLeftMargin{41. }%
+\CSLRightInline{Maas, P. \emph{et al.} Facebook disaster maps: Aggregate
+insights for crisis response and recovery. in \emph{16th international
+conference on information systems for crisis response and management}
+836--847 (2019).}
 

### 2024-02-12 manuscript/manuscript.qmd
checking-representativeness

@@
 abstract: |
-  The movement of people between locations is crucial for sustainable and inclusive cities, facilitating the exchange of knowledge and resources.
-Due to restrictions, lockdowns and fear of crowded places, the COVID-19 pandemic significantly reduced the number and extent of people's movements.
-Existing research suggests that the impact of the pandemic on mobility was unequal across locations and socioeconomic groups.
-However, evidence and analysis of long-term changes in mobility is scarce, especially in less developed countries.
-In this study, we use location data from Meta-Facebook users to analyse patterns of urban mobility in X Functional Urban Areas from South American countries.
-Findings reveal a general decrease in mobility during the pandemic, with gradual recovery trends up to April 2023.
-However, socioeconomic disparities are present through the period of analysis, with deprived areas experiencing smaller initial losses in mobility.
-These disparities generally diminish over time, although they persist in all the countries included in the analysis <!--# They seem to converge in Argentina and Mexico -->.
-Our research highlights the importance of timely mobility data with high spatiotemporal resolution to understand the long-term effects of the pandemic and inform equitable policy responses that address societal challenges in urban areas.
+  COVID-19 triggered a reduction in the frequency and extent of people's movement.
+Existing evidence suggests that while the impact of the pandemic on mobility was widespread, the extent of this impact was unequally felt across socioeconomic groups in the early stages of the pandemic.
+Here, we find that the most deprived locations have experienced a more accelerated recovery towards pre-pandemic levels of mobility in the long term.
+Furthermore, the socioeconomic disparities in the patterns of mobility triggered by the first outbreak of COVID-19, have persisted as of April 2023.
+These findings are based on the analysis of time-series mobility data corresponding to X urban areas from Latin American countries collected from Meta-Facebook users upon their consent.
+Our research highlights the importance of timely mobility data with high spatiotemporal resolution to understand the long-term effects of the pandemic and to inform equitable policy responses that address societal challenges in urban areas.
 format: 
@@
 ---
+
+<!--# FR: A few questions to think about: Have we validated the data?
+i.e.
+tested their extent of representativeness?
+Shall we do something quick using census data for each country comparing overall populations?
+-->\
 
@@
 
--   Aim - focus on Latin American countries as case study.
--   Aim - focus on urban areas, as the future of humanity is the future of cities and Latin Ameriaca is highly urbanised
+Latin America provides an ideal test-bed for testing these hypotheses because of its exceptionally high levels of inequalities [@de2004inequality; @carranza2023wealth] and urbanisation [@united2023world].
+Half of the 20 most unequal countries in the planet are in this region.
+The average income Gini index of the region is 4 percentage points higher than that of Africa and 11 higher than China [@milanovic2016global], and cities display some of the starkest inequalities [@habitat2022world].
+Currently, over 80% of the population in Latin America live in urban areas.
+By 2050, this share is predicted to reach 89%, with the largest share concentrating in a few megacities [@habitat2022world].
+Developing an understanding of human mobility in Latin America is thus important to support sustainable and inclusive spaces [@habitat2022world].
 

### 2024-02-12 manuscript/manuscript.tex
checking-representativeness

@@
 \makeatother
+% definitions for citeproc citations
+\NewDocumentCommand\citeproctext{}{}
+\NewDocumentCommand\citeproc{mm}{%
+  \begingroup\def\citeproctext{#2}\cite{#1}\endgroup}
+\makeatletter
+ % allow citations to break across lines
+ \let\@cite@ofmt\@firstofone
+ % avoid brackets around text for \cite:
+ \def\@biblabel#1{}
+ \def\@cite#1#2{{#1\if@tempswa , #2\fi}}
+\makeatother
 \newlength{\cslhangindent}
@@
 \setlength{\csllabelwidth}{3em}
-\newlength{\cslentryspacingunit} % times entry-spacing
-\setlength{\cslentryspacingunit}{\parskip}
-\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
- {% don't indent paragraphs
-  \setlength{\parindent}{0pt}
+\newenvironment{CSLReferences}[2] % #1 hanging-indent, #2 entry-spacing
+ {\begin{list}{}{%
+  \setlength{\itemindent}{0pt}
+  \setlength{\leftmargin}{0pt}
+  \setlength{\parsep}{0pt}
   % turn on hanging indent if param 1 is 1
   \ifodd #1
-  \let\oldpar\par
-  \def\par{\hangindent=\cslhangindent\oldpar}
+   \setlength{\leftmargin}{\cslhangindent}
+   \setlength{\itemindent}{-1\cslhangindent}
   \fi
   % set entry spacing
-  \setlength{\parskip}{#2\cslentryspacingunit}
- }%
- {}
+  \setlength{\itemsep}{#2\baselineskip}}}
+ {\end{list}}
 \usepackage{calc}
-\newcommand{\CSLBlock}[1]{#1\hfill\break}
-\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
-\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
+\newcommand{\CSLBlock}[1]{\hfill\break\parbox[t]{\linewidth}{\strut\ignorespaces#1\strut}}
+\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{\strut#1\strut}}
+\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{\strut#1\strut}}
 \newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
@@
 \begin{abstract}
-The movement of people between locations is crucial for sustainable and
-inclusive cities, facilitating the exchange of knowledge and resources.
-Due to restrictions, lockdowns and fear of crowded places, the COVID-19
-pandemic significantly reduced the number and extent of people's
-movements. Existing research suggests that the impact of the pandemic on
-mobility was unequal across locations and socioeconomic groups. However,
-evidence and analysis of long-term changes in mobility is scarce,
-especially in less developed countries. In this study, we use location
-data from Meta-Facebook users to analyse patterns of urban mobility in X
-Functional Urban Areas from South American countries. Findings reveal a
-general decrease in mobility during the pandemic, with gradual recovery
-trends up to April 2023. However, socioeconomic disparities are present
-through the period of analysis, with deprived areas experiencing smaller
-initial losses in mobility. These disparities generally diminish over
-time, although they persist in all the countries included in the
-analysis . Our research highlights the importance of timely mobility
-data with high spatiotemporal resolution to understand the long-term
-effects of the pandemic and inform equitable policy responses that
-address societal challenges in urban areas.
+COVID-19 triggered a reduction in the number and extent of peoples
+movements. Existing evidence suggests that while the impact of the
+pandemic on mobility was widespread, the extent of this impact was
+unequally felt across socioeconomic groups in the early stages of the
+pandemic. Here, we find that in the long term, the most deprived
+locations have experienced a more accelerated recovery towards
+pre-pandemic levels of mobility. Furthermore, the socioeconomic
+disparities in the patterns of mobility triggered by the first outbreak
+of COVID-19, have persisted as of April 2023. These findings are based
+on the analysis of time-series mobility data corresponding to X urban
+areas from Latin American countries collected from Meta-Facebook users
+upon their consent. Our research highlights the importance of timely
+mobility data with high spatiotemporal resolution to understand the
+long-term effects of the pandemic and to inform equitable policy
+responses that address societal challenges in urban areas.
 \end{abstract}
-\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[enhanced, frame hidden, breakable, sharp corners, interior hidden, borderline west={3pt}{0pt}{shadecolor}, boxrule=0pt]}{\end{tcolorbox}}\fi
 
-\hypertarget{sec-intro}{%
-\section{Introduction}\label{sec-intro}}
+\hfill\break
+
+\section{Introduction}\label{sec-intro}
 
@@
 
-\begin{itemize}
-\tightlist
-\item
-  Aim - focus on Latin American countries as case study.
-\item
-  Aim - focus on urban areas, as the future of humanity is the future of
-  cities and Latin Ameriaca is highly urbanised
-\end{itemize}
+Latin America provides an ideal test-bed for testing these hypotheses
+because of its exceptionally high levels of
+inequalities\textsuperscript{32,33} and
+urbanisation\textsuperscript{34}. Half of the 20 most unequal countries
+in the planet are in this region. The average income Gini index of the
+region is 4 percentage points higher than that of Africa and 11 higher
+than China\textsuperscript{35}, and cities display some of the starkest
+inequalities\textsuperscript{36}. Currently, over 80\% of the population
+in Latin America live in urban areas. By 2050, this share is predicted
+to reach 89\%, with the largest share concentrating in a few
+megacities\textsuperscript{36}. Developing an understanding of human
+mobility in Latin America is thus important to support sustainable and
+inclusive spaces\textsuperscript{36}.
 
-\hypertarget{sec-results}{%
-\section{Results}\label{sec-results}}
+\section{Results}\label{sec-results}
 
@@
 spatially into administrative units at various levels according to GADM,
-the Database of Global Administrative Areas\textsuperscript{32}. The
+the Database of Global Administrative Areas\textsuperscript{37}. The
 analysis focuses on administrative units that are within the boundaries
@@
 
-\end{figure}
+\end{figure}%
 
-\hypertarget{the-heterogeneous-impact-of-covid-19-on-urban-mobility}{%
 \subsection{The heterogeneous impact of COVID-19 on urban
-mobility}\label{the-heterogeneous-impact-of-covid-19-on-urban-mobility}}
+mobility}\label{the-heterogeneous-impact-of-covid-19-on-urban-mobility}
 
@@
 Specifically, we focus the analysis on short-distance movements in urban
-areas to represent local and routine mobility\textsuperscript{33}, so
+areas to represent local and routine mobility\textsuperscript{38}, so
 only movements covering a distance of at most 70 km are considered. For
@@
 
-\hypertarget{the-most-deprived-areas-experienced-the-smallest-drop-in-mobility-relative-to-pre-pandemic-levels}{%
 \subsection{The most deprived areas experienced the smallest drop in
 mobility relative to pre-pandemic
-levels}\label{the-most-deprived-areas-experienced-the-smallest-drop-in-mobility-relative-to-pre-pandemic-levels}}
+levels}\label{the-most-deprived-areas-experienced-the-smallest-drop-in-mobility-relative-to-pre-pandemic-levels}
 
@@
 
-\hypertarget{sec-discussion}{%
-\section{Discussion}\label{sec-discussion}}
+\section{Discussion}\label{sec-discussion}
 
@@
 especifically those covering 70 km or less. These journeys are typically
-considered to represent local and routine mobility\textsuperscript{33} .
+considered to represent local and routine mobility\textsuperscript{38} .
 However, due to the characteristics of the Meta-Facebook movement data,
@@
 the journeys, following similar approaches to those proposed
-by\textsuperscript{37}, and quantifying the extent to which the digital
+by\textsuperscript{42}, and quantifying the extent to which the digital
 footprint data mirrors the true mobility patterns.
@@
 access and usage of digital technologies across demographic
-groups\textsuperscript{38}. Despite these challenges, the data and
+groups\textsuperscript{43}. Despite these challenges, the data and
 analysis that we used for this work provide evidence for non-trivial
@@
 creation of resources like those developed by the European Commission
-Joint Research Centre\textsuperscript{39} and the UN Statistics
-Division\textsuperscript{40}, which identify sources of non-traditional
+Joint Research Centre\textsuperscript{44} and the UN Statistics
+Division\textsuperscript{45}, which identify sources of non-traditional
 data and set methodological protocols for incorporating mobile phone
@@
 
-\hypertarget{sec-methods}{%
-\section{Methods}\label{sec-methods}}
+\section{Methods}\label{sec-methods}
 
-\hypertarget{meta-facebook-movement-data}{%
 \subsection{Meta-Facebook movement
-data}\label{meta-facebook-movement-data}}
+data}\label{meta-facebook-movement-data}
 
@@
 information and applying privacy-preserving
-techniques\textsuperscript{41}. Small-count dropping is one of these
+techniques\textsuperscript{46}. Small-count dropping is one of these
 techniques. A data entry is removed if the population or movement count
@@
 collected on Mondays from 00:00 to 8:00 for the 45-day period. Details
-about the baseline can be found in\textsuperscript{41}. The Bing Maps
+about the baseline can be found in\textsuperscript{46}. The Bing Maps
 Tile System developed by Microsoft (Microsoft) is used a spatial
@@
 aggregated into Bing tile levels 13. That is about 4.9 x 4.9km at the
-Equator\textsuperscript{41}.
+Equator\textsuperscript{46}.
 
-\hypertarget{spatiotemporal-data-aggregation}{%
 \subsection{Spatiotemporal data
-aggregation}\label{spatiotemporal-data-aggregation}}
+aggregation}\label{spatiotemporal-data-aggregation}
 
@@
 
-\hypertarget{references}{%
-\section*{References}\label{references}}
+\section*{References}\label{references}
 \addcontentsline{toc}{section}{References}
 
-\hypertarget{refs}{}
+\phantomsection\label{refs}
 \begin{CSLReferences}{0}{0}
-\leavevmode\vadjust pre{\hypertarget{ref-ackers2005}{}}%
+\bibitem[\citeproctext]{ref-ackers2005}
 \CSLLeftMargin{1. }%
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-chen2016}{}}%
+\bibitem[\citeproctext]{ref-chen2016}
 \CSLLeftMargin{2. }%
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-belik2011}{}}%
+\bibitem[\citeproctext]{ref-belik2011}
 \CSLLeftMargin{3. }%
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-klugman2009human}{}}%
+\bibitem[\citeproctext]{ref-klugman2009human}
 \CSLLeftMargin{4. }%
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-barbosa2018}{}}%
+\bibitem[\citeproctext]{ref-barbosa2018}
 \CSLLeftMargin{5. }%
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-chinazzi2020}{}}%
+\bibitem[\citeproctext]{ref-chinazzi2020}
 \CSLLeftMargin{6. }%
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-nouvellet2021}{}}%
+\bibitem[\citeproctext]{ref-nouvellet2021}
 \CSLLeftMargin{7. }%
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-rowe2023}{}}%
+\bibitem[\citeproctext]{ref-rowe2023}
 \CSLLeftMargin{8. }%
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-abdullah2020}{}}%
+\bibitem[\citeproctext]{ref-abdullah2020}
 \CSLLeftMargin{9. }%
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-bonaccorsi2020}{}}%
+\bibitem[\citeproctext]{ref-bonaccorsi2020}
 \CSLLeftMargin{10. }%
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-abu-rayash2020}{}}%
+\bibitem[\citeproctext]{ref-abu-rayash2020}
 \CSLLeftMargin{11. }%
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-lee2023}{}}%
+\bibitem[\citeproctext]{ref-lee2023}
 \CSLLeftMargin{12. }%
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-borkowski2021}{}}%
+\bibitem[\citeproctext]{ref-borkowski2021}
 \CSLLeftMargin{13. }%
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-chang2020}{}}%
+\bibitem[\citeproctext]{ref-chang2020}
 \CSLLeftMargin{14. }%
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-fraiberger2020}{}}%
+\bibitem[\citeproctext]{ref-fraiberger2020}
 \CSLLeftMargin{15. }%
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-weill2020}{}}%
+\bibitem[\citeproctext]{ref-weill2020}
 \CSLLeftMargin{16. }%
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-dueuxf1as2021}{}}%
+\bibitem[\citeproctext]{ref-dueuxf1as2021}
 \CSLLeftMargin{17. }%
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-santana2023}{}}%
+\bibitem[\citeproctext]{ref-santana2023}
 \CSLLeftMargin{18. }%
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-florida2021}{}}%
+\bibitem[\citeproctext]{ref-florida2021}
 \CSLLeftMargin{19. }%
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-rowe2023a}{}}%
+\bibitem[\citeproctext]{ref-rowe2023a}
 \CSLLeftMargin{20. }%
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-barrero2021}{}}%
+\bibitem[\citeproctext]{ref-barrero2021}
 \CSLLeftMargin{21. }%
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-aksoy2022}{}}%
+\bibitem[\citeproctext]{ref-aksoy2022}
 \CSLLeftMargin{22. }%
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-green2021}{}}%
+\bibitem[\citeproctext]{ref-green2021}
 \CSLLeftMargin{23. }%
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-bell2014}{}}%
+\bibitem[\citeproctext]{ref-bell2014}
 \CSLLeftMargin{24. }%
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-rowe2023big}{}}%
+\bibitem[\citeproctext]{ref-rowe2023big}
 \CSLLeftMargin{25. }%
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-calafiore2023}{}}%
+\bibitem[\citeproctext]{ref-calafiore2023}
 \CSLLeftMargin{26. }%
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-prophet17}{}}%
+\bibitem[\citeproctext]{ref-prophet17}
 \CSLLeftMargin{27. }%
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-GHSL19}{}}%
+\bibitem[\citeproctext]{ref-GHSL19}
 \CSLLeftMargin{28. }%
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-RDI2022}{}}%
+\bibitem[\citeproctext]{ref-RDI2022}
 \CSLLeftMargin{29. }%
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-roweCalafiore2022}{}}%
+\bibitem[\citeproctext]{ref-roweCalafiore2022}
 \CSLLeftMargin{30. }%
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-wang2022}{}}%
+\bibitem[\citeproctext]{ref-wang2022}
 \CSLLeftMargin{31. }%
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-GADM24}{}}%
+\bibitem[\citeproctext]{ref-de2004inequality}
 \CSLLeftMargin{32. }%
+\CSLRightInline{De Ferranti, D. M. \emph{Inequality in latin america:
+Breaking with history?} (World Bank publications, 2004).}
+
+\bibitem[\citeproctext]{ref-carranza2023wealth}
+\CSLLeftMargin{33. }%
+\CSLRightInline{Carranza, R., De Rosa, M. \& Flores, I. Wealth
+inequality in latin america. (2023).}
+
+\bibitem[\citeproctext]{ref-united2023world}
+\CSLLeftMargin{34. }%
+\CSLRightInline{United Nations, D. for E. \& Affairs., S. \emph{World
+population prospects 2022: Summary of results}. (UN, 2023).}
+
+\bibitem[\citeproctext]{ref-milanovic2016global}
+\CSLLeftMargin{35. }%
+\CSLRightInline{Milanovic, B. \emph{Global inequality: A new approach
+for the age of globalization}. (Harvard University Press, 2016).}
+
+\bibitem[\citeproctext]{ref-habitat2022world}
+\CSLLeftMargin{36. }%
+\CSLRightInline{Habitat, U. World cities report 2022: Envisaging the
+future of cities. \emph{United Nations Human Settlements Programme:
+Nairobi, Kenya} 41--44 (2022).}
+
+\bibitem[\citeproctext]{ref-GADM24}
+\CSLLeftMargin{37. }%
 \CSLRightInline{GADM. GADM maps and data. (2024).}
 
-\leavevmode\vadjust pre{\hypertarget{ref-owen1992migration}{}}%
-\CSLLeftMargin{33. }%
+\bibitem[\citeproctext]{ref-owen1992migration}
+\CSLLeftMargin{38. }%
 \CSLRightInline{Owen, D. \& Green, A. Migration patterns and trends. in
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-LaurenAlexander15}{}}%
-\CSLLeftMargin{34. }%
+\bibitem[\citeproctext]{ref-LaurenAlexander15}
+\CSLLeftMargin{39. }%
 \CSLRightInline{Alexander, L., Jiang, S., Murga, M. \& González, M. C.
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-MengChuishi17}{}}%
-\CSLLeftMargin{35. }%
+\bibitem[\citeproctext]{ref-MengChuishi17}
+\CSLLeftMargin{40. }%
 \CSLRightInline{Meng, C., Cui, Y., He, Q., Su, L. \& Gao, J. Travel
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-NiluferSariAslam21}{}}%
-\CSLLeftMargin{36. }%
+\bibitem[\citeproctext]{ref-NiluferSariAslam21}
+\CSLLeftMargin{41. }%
 \CSLRightInline{Nilufer Sari Aslam, T. C., Mohamed R. Ibrahim \& Zhang,
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-Cabrera-Arnau2023}{}}%
-\CSLLeftMargin{37. }%
+\bibitem[\citeproctext]{ref-Cabrera-Arnau2023}
+\CSLLeftMargin{42. }%
 \CSLRightInline{Cabrera-Arnau, C., Zhong, C., Batty, M., Silva, R. \&
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-RoweBigData23}{}}%
-\CSLLeftMargin{38. }%
+\bibitem[\citeproctext]{ref-RoweBigData23}
+\CSLLeftMargin{43. }%
 \CSLRightInline{Rowe, F. 9.: Big data. in \emph{Concise encyclopedia of
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-ECJRC22}{}}%
-\CSLLeftMargin{39. }%
+\bibitem[\citeproctext]{ref-ECJRC22}
+\CSLLeftMargin{44. }%
 \CSLRightInline{Commission, E. \emph{et al.} \emph{Data innovation in
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-UNSD}{}}%
-\CSLLeftMargin{40. }%
+\bibitem[\citeproctext]{ref-UNSD}
+\CSLLeftMargin{45. }%
 \CSLRightInline{Division, U. N. S. Handbook on the use of mobile phone
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-Maas19}{}}%
-\CSLLeftMargin{41. }%
+\bibitem[\citeproctext]{ref-Maas19}
+\CSLLeftMargin{46. }%
 \CSLRightInline{Maas, P. \emph{et al.} Facebook disaster maps: Aggregate

### 2024-02-12 manuscript/superseded_12022024/manuscript.qmd
checking-representativeness

--- 2024-02-06 manuscript/manuscript.qmd
@@
 However, socioeconomic disparities are present through the period of analysis, with deprived areas experiencing smaller initial losses in mobility.
-These disparities generally diminish over time, although they persist in all the countries included in the analysis <!--# They seem to converge in Argentina and Mexico -->.
+These disparities generally diminish over time, although they persist in all the countries included in the analysis <!--# FR: They seem to converge in Argentina and Mexico -->.
 Our research highlights the importance of timely mobility data with high spatiotemporal resolution to understand the long-term effects of the pandemic and inform equitable policy responses that address societal challenges in urban areas.

### 2024-02-12 manuscript/superseded_12022024/manuscript.tex
checking-representativeness

--- 2024-02-12 manuscript/manuscript.tex
@@
 \begin{abstract}
-COVID-19 triggered a reduction in the number and extent of peoples
-movements. Existing evidence suggests that while the impact of the
-pandemic on mobility was widespread, the extent of this impact was
-unequally felt across socioeconomic groups in the early stages of the
-pandemic. Here, we find that in the long term, the most deprived
-locations have experienced a more accelerated recovery towards
-pre-pandemic levels of mobility. Furthermore, the socioeconomic
-disparities in the patterns of mobility triggered by the first outbreak
-of COVID-19, have persisted as of April 2023. These findings are based
-on the analysis of time-series mobility data corresponding to X urban
-areas from Latin American countries collected from Meta-Facebook users
-upon their consent. Our research highlights the importance of timely
-mobility data with high spatiotemporal resolution to understand the
-long-term effects of the pandemic and to inform equitable policy
-responses that address societal challenges in urban areas.
+The movement of people between locations is crucial for sustainable and
+inclusive cities, facilitating the exchange of knowledge and resources.
+Due to restrictions, lockdowns and fear of crowded places, the COVID-19
+pandemic significantly reduced the number and extent of people's
+movements. Existing research suggests that the impact of the pandemic on
+mobility was unequal across locations and socioeconomic groups. However,
+evidence and analysis of long-term changes in mobility is scarce,
+especially in less developed countries. In this study, we use location
+data from Meta-Facebook users to analyse patterns of urban mobility in X
+Functional Urban Areas from South American countries. Findings reveal a
+general decrease in mobility during the pandemic, with gradual recovery
+trends up to April 2023. However, socioeconomic disparities are present
+through the period of analysis, with deprived areas experiencing smaller
+initial losses in mobility. These disparities generally diminish over
+time, although they persist in all the countries included in the
+analysis . Our research highlights the importance of timely mobility
+data with high spatiotemporal resolution to understand the long-term
+effects of the pandemic and inform equitable policy responses that
+address societal challenges in urban areas.
 \end{abstract}

### 2024-10-23 code/analysis/06-lme-with-R.qmd
transfer

---
title: "lme-r"
format: html
editor: visual
---

```{r}
# data wrangling
library(tidyverse)
# data utilities
library(timetk)
# estimating mixed effects models
library(lme4)
library(merTools)
library(glmmTMB)
library(nlme)
# correlograms
library(ggcorrplot)
library(corrplot)
# data visualisation
library(viridis)
library(ggthemes)
library(ggpubr)
library(zoo)
library(showtext)
library(patchwork)
library(ggrepel)
# display regression equation
library(equatiomatic)
# standardise input variables
  #library(arm)
  # reporting regression results
library(broom.mixed)
library(gtsummary)
library(sjPlot)
```

```{r}
# Set working directory
wd <- '/Users/carmen/Library/CloudStorage/OneDrive-TheUniversityofLiverpool/research/RECAST/latin-mobility-covid-local-files'

# Set the country
country <- 'Mexico'

# Initialize variables for country_short and country_code
country_short <- ""
country_code <- ""

# Assign country_short and country_code based on the country
if (country == 'Argentina') {
    country_short <- 'ARG'
    country_code <- 'AR'
} else if (country == 'Chile') {
    country_short <- 'CHL'
    country_code <- 'CL'
} else if (country == 'Colombia') {
    country_short <- 'COL'
    country_code <- 'CO'
} else if (country == 'Mexico') {
    country_short <- 'MEX'
    country_code <- 'MX'
}
```

```{r}
category <- "rdi"
```

```{r}
# Read the CSV file
file_path <- paste0(wd, '/data/outputs/', country_short, '/mov-analysis/by-', category, '/data_trend.csv')
df <- read.csv(file_path)

df$time2 <- df$time^2
```

```{r}

eq1 <- y ~ time
m1 <- glmmTMB(eq1, data = df)
summary(m1)

eq2 <- y ~ time + O + D
m2 <- glmmTMB(eq2, data = df)
summary(m2)

eq3 <- y ~ time + (1|O) + (1|D)
m3 <- glmmTMB(eq3, data = df)
m3_re <- plot_model(m3, type ="re")
summary(m3)

df$Of <- factor(df$O)
df$Df <- factor(df$D)
eq4 <- y ~ time + (time|Of) + (time|Df)
m4 <- glmmTMB(eq4, data = df)
m4_re <- plot_model(m4, type ="re")
summary(m4)

eq5 <- y ~ time + time2 + (time|O) + (time|D)
m5 <- glmmTMB(eq5, data = df)
m5_re <- plot_model(m5, type ="re")
summary(m5)

```

|     |
|-----|
|     |

```{r}
m4_re[[1]]$data
m4_re[[2]]$data
m5_re[[2]]$data
m5_re[[2]]$data
m4_re
m5_re
```

```{r}
p1_m1me <- plot_model(m, terms = c("1", "time")) %>% 
  .$data %>% ggplot(aes(x=estimate, y=term, color=group)) +
  geom_point(size = 3.5,  colour = "black") +
  geom_errorbar(aes(xmin = conf.low, xmax = conf.high), 
                width=.05, color= "black") +
  geom_pointrange(aes(xmin = conf.low, xmax = conf.high)) +
  scale_color_manual(values=c("#88ccee")) +
  theme(legend.position = "none",
        axis.title.y = element_blank(),
        axis.text = element_text(size = 10.5)
  ) +
  scale_y_discrete(labels=c("time" = "Week", 
                            "1" = "Intercept")) +
  geom_vline(xintercept=0, 
             linetype="solid", 
             color = "grey20",
             linewidth=1,
             alpha =.2) +
  labs(title = "Main fixed effects")

p1_m1me
```

```{r}
m2_re <- plot_model(m,
           type ="re", terms = "O")

m2_re <- m2_re[[1]]$data

p2_m2re <- m2_re %>% filter(facet == "O (Intercept)") %>% 
  ggplot(aes(x=estimate, y=term, color=group)) +
  geom_point(size = 3.5,  colour = "black") +
  geom_errorbar(aes(xmin = conf.low, xmax = conf.high), 
                width=.8, color= "black") +
  geom_pointrange(aes(xmin = conf.low, xmax = conf.high)) +
  scale_color_manual(values=c("#ddcc77", "#88ccee")) +
  theme(legend.position = "none",
        axis.title.y = element_blank(),
        axis.text = element_text(size = 10.5)
        ) +
  geom_vline(xintercept=0, 
             linetype="solid", 
             color = "grey20",
             size=1,
             alpha =.2) +
  labs(title = "d")

p2_m2re
```

```{r}
eq <- y ~  time + (time | O) + (time | D)
model <- glmmTMB(eq, data = df)
```

```{r}
summary(model)
```

```{r}
p1_m <- plot_model(model, terms = c("time", "O", "D")) %>% 
  .$df %>% ggplot(aes(x=estimate, y=term, color=group)) +
  geom_point(size = 3.5,  colour = "black") +
  geom_errorbar(aes(xmin = conf.low, xmax = conf.high), 
                width=.05, color= "black") +
  geom_pointrange(aes(xmin = conf.low, xmax = conf.high)) +
  scale_color_manual(values=c("#88ccee")) +
  theme(legend.position = "none",
        axis.title.y = element_blank(),
        axis.text = element_text(size = 10.5)
  ) +
  scale_y_discrete(labels=c("z_stringency_index" = "Stringency t", 
                            "z_cases" = "New cases t",
                            "lag(z_stringency_index, 1)" = "Stringency t-1")) +
  geom_vline(xintercept=0, 
             linetype="solid", 
             color = "grey20",
             linewidth=1,
             alpha =.2) +
  labs(title = "c",
       subtitle = " Main fixed effects")

p1_m
```

```{r}
coef_model <- coef(model)
head(coef_model$O)
```

```{r}
ranef_model <- ranef(model)
head(ranef_model$O)
```

```{r}
# obtain estimates
merTools::REsim(model) %>% head(10)
```

```{r}
plotREsim(REsim(model)) 
```

```{r}
## regression table
tab_model(mode, 
          collapse.ci = TRUE,
          p.style = "stars",
          show.aic = TRUE,
          #terms = c(), # show selected coefficients
          #rm.terms = c("time [1st degree]"), #remove selected coefficients
          #pred.labels = c("Intercept", "Stringency t", "New Cases t", "Stringency t-1",
          #                "Spline 1st degree", "Spline 2nd degree", 
          #                "Spline 3rd degree", "New cases t-1"),
          #dv.labels = c("Autoregressive", "Stringency random effects", "New COVID-19 cases random effects"),
          #file = "../outputs/modelling/estimates/reg_table.txt"
)
```

### 2024-10-23 manuscript/manuscript-231024/manuscript.qmd
transfer

--- 2024-02-12 manuscript/manuscript.qmd
@@
 Shall we do something quick using census data for each country comparing overall populations?
--->\
+-->
+
+\
 
@@
 
-![Administrative units within functional urban areas in Argentina, Chile, Colombia and Mexico, coloured according to the average Relative Deprivation Index.](figures/maps.pdf)
+<!--![Administrative units within functional urban areas in Argentina, Chile, Colombia and Mexico, coloured according to the average Relative Deprivation Index.](figures/maps.pdf)!-->
 
@@
 
-![Patterns of recovery for urban mobility in administrative units from Argentina, Chile, Colombia and Mexico.](figures/prediction-rdi-band.pdf)
+\<!-
+--[Patterns of recovery for urban mobility in administrative units from Argentina, Chile, Colombia and Mexico.](figures/prediction-rdi-band.pdf)
+---!\>
 
@@
 
-![Evolution of the parameters characterising the relationship between relative deprivation index and the percentage change in the number of movements relative to the pre-pandemic baseline level of mobility.](figures/regression-evo-nobackground.pdf)
+<!---![Evolution of the parameters characterising the relationship between relative deprivation index and the percentage change in the number of movements relative to the pre-pandemic baseline level of mobility.](figures/regression-evo-nobackground.pdf)!--->
 
@@
 
-# Methods {#sec-methods}
+# Data and Methods {#sec-methods}
 
-## Meta-Facebook movement data
+## Meta-Facebook data
+
+<!-- Include here population data too !-->
 
@@
 That is about 4.9 x 4.9km at the Equator [@Maas19].
+
+## WorlPpop population data
+
+We used data from WorldPop [@tatem2017] to classify the spatial units of analysis according to their level of urbanisation, and to estimate missing baseline values in the Facebook population data.
+WorldPop offers open access gridded population estimates at a resolution as small as 3 arc-seconds approximately 100m and 1km at the Equator, respectively.
+WorldPop produces these gridded datasets using top-down (i.e.
+dissagregating administrative area counts into smaller grid cells) or bottom-up (i.e.
+interpolating data from counts from sample locations into grid cells) approaches.
+For the purposes of this work, we use gridded population data at a resolution of 1km$^2$ in raster format.
+We perform a spatial join of the Facebook spatial units (Bing tiles) with the gridded population data and compute the sum of Worldpop populations corresponding to each of the Facebook spatial units.
+
+## Socioeconomic deprivation data
+
+In our analysis, we use the Global Gridded Relative Deprivation Index (GRDI), Version 1 (GRDIv1) data set as a measure of socioeocnomic deprivation.
+The GRDI data is made available via NASA's Socioeconomic Data and Applications Centre (SEDAC), at a spatial resolution of 30 arc-seconds, or 1 km$^2$ approximately.
+The index quantifies the relative levels of multidimensional deprivation and poverty, where a value of 100 represents the highest level of deprivation and a value of 0 the lowest.
+We perform a spatial join of the Facebook spatial units and the gridded relative deprivation data and compute the average RDI corresponding to each of the Facebook spatial units.
+
+## Classification of tiles according to level of urbanisation and socioeconomic deprivation
+
+<!-- Decide whether I'm going for natural breaks or my own splitting. The disadvantage of own splitting is that there is no common logic for all four countries !-->
+
+## Processing Facebook data
+
+To ensure the privacy and anonymity of the users' data, Meta removes information corresponding to data entries where the population or movement count is less than 10 for a specific time or day, retaining only information about the percentage change in the number of counts with respect to the baseline period is retained [@Maas19].
+Consequently, we observe many tiles where the population or movement count for either the baseline or crisis period are blank.
+<!-- Justify why we do the imputation.
+Spatial bias (maybe show Moran's I for missing values on a selected day?
+Or everyday?).
+Small number bias, underrespresentation of dataentries with low counts that we know for sure exist !-->
+
+### Facebook population data
+
+To imput Facebook population baseline values, we first identify all the baseline values that are available for each spatial unit and weekday (of the three available time windows, we only consider one per day).
+We then estimate the missing baseline values for each weekday, based on a linear model for the relationship between the Worldpop population and the Facebook population counts, which are fitted using ordinary least squares.
+This is illustrated in Supplementary Figure \ref{fig:imput-baseline-population}.
+
+We then use the complete baseline Facebook population values to compute missing Facebook population counts during the crisis period.
+This is possible because, as mentioned above, Meta reports the percentage change in the number of counts with respect to the baseline, even if the counts are not reported due to low value.
+
+### Facebook movement data
+
+The imputation of Facebook movement baseline values is done according to a spatial interaction model (see e.g. [@rowe2022a]).
+We considered the population flow between an origin and a destination tile, and model this variable as a function of the Facebook population count at the origin tile on the same weekday, the Facebook population count at the destination on the same weekday and the distance between origin and destination; we also included indicator variables to capture the day of the week and the pair of population density classes corresponding to the origin and destination tiles.
+Mathematically, this model can be expressed as \begin{equation}
+\mu_{ijw} = \beta_0 + \beta_1pop_{iw} +\beta_2pop_{jw} + \beta_3d_{ij} + \beta_4W + \beta_5C + \varepsilon
+\end{equation}
+
+where $\mu_{ijw}$ is the expectation of the flow of people from tile $i$ to tile $j$ on the weekday $w$; $\beta_0$ is an intercept $pop_{iw}$ and $pop_{jw}$ are the Facebook population counts at the origin and destination on weekday $w$ during the baseline period, $d_{ij}$ is the distance between origin and destination, $W$ is a series of indicator variables capturing the day of the week and similarly $C$ is a series of indicator variables reflecting the pair of population density classes for the origin and destination tiles, resulting in X pairs (10 classes × 10 classes minus one so it is not collinear with $\beta_0$), $\beta_{0,1,2,3,4,5}$ are model parameters to be estimated from the observed data.
+The error term is denoted by \varepsilon.
+
+To estimate the model parameters, we used a count data regression model.
+Specifically, we used a negative binomial regression which is a generalised linear model (GLM) where overdispersion of the error term is assumed, i.e. the variance exceeds the mean.
+
+We compute missing Facebook movement counts during the crisis period by considering the complete Facebook movements baseline and the percentage change in the number of counts with respect to the baseline, which is reported in the Facebook Movements datasets even when the count is not reported due to its low value.
+
+### Correction factors??? For representativeness
+
+## Data analysis
+
+```{=html}
+<!-- EXPLAIN - might include the things I had already written below
 
@@
 
+ !-->
+```
 # References

### 2024-10-23 code/analysis/np_percentile.py

@@ -1,0 +1,5718 @@
+import builtins
+import collections.abc
+import functools
+import re
+import sys
+import warnings
+
+import numpy as np
+
+array_function_dispatch = functools.partial(
+    overrides.array_function_dispatch, module='numpy')
+
+__all__ = [
+    'select', 'piecewise', 'trim_zeros', 'copy', 'iterable', 'percentile',
+    'diff', 'gradient', 'angle', 'unwrap', 'sort_complex', 'flip',
+    'rot90', 'extract', 'place', 'vectorize', 'asarray_chkfinite', 'average',
+    'bincount', 'digitize', 'cov', 'corrcoef',
+    'median', 'sinc', 'hamming', 'hanning', 'bartlett',
+    'blackman', 'kaiser', 'trapezoid', 'trapz', 'i0',
+    'meshgrid', 'delete', 'insert', 'append', 'interp',
+    'quantile'
+    ]
+
+# _QuantileMethods is a dictionary listing all the supported methods to
+# compute quantile/percentile.
+#
+# Below virtual_index refers to the index of the element where the percentile
+# would be found in the sorted sample.
+# When the sample contains exactly the percentile wanted, the virtual_index is
+# an integer to the index of this element.
+# When the percentile wanted is in between two elements, the virtual_index
+# is made of a integer part (a.k.a 'i' or 'left') and a fractional part
+# (a.k.a 'g' or 'gamma')
+#
+# Each method in _QuantileMethods has two properties
+# get_virtual_index : Callable
+#   The function used to compute the virtual_index.
+# fix_gamma : Callable
+#   A function used for discrete methods to force the index to a specific value.
+_QuantileMethods = dict(
+    # --- HYNDMAN and FAN METHODS
+    # Discrete methods
+    inverted_cdf=dict(
+        get_virtual_index=lambda n, quantiles: _inverted_cdf(n, quantiles),
+        fix_gamma=None,  # should never be called
+    ),
+    averaged_inverted_cdf=dict(
+        get_virtual_index=lambda n, quantiles: (n * quantiles) - 1,
+        fix_gamma=lambda gamma, _: _get_gamma_mask(
+            shape=gamma.shape,
+            default_value=1.,
+            conditioned_value=0.5,
+            where=gamma == 0),
+    ),
+    closest_observation=dict(
+        get_virtual_index=lambda n, quantiles: _closest_observation(n,
[…]

### 2024-10-23 code/plots.py

@@ -1,87 +1,0 @@
-
-import matplotlib
-import matplotlib.pyplot as plt
-import contextily as cx
-import geopandas as gpd
-import pandas as pd
-import numpy as np
-from mycolorpy import colorlist as mcp
-from matplotlib import colors as mcolors
-from math import pi, ceil, floor
-import matplotlib.lines as mlines
-import seaborn as sns
-
-
-
-
-
-
-# Create function to plot radar chart of cluster centroid
-def plot_radar_centroid(cl_centroid, average_centroid):
-
-    '''
-    Plot radar chart of cluster centroid
-
-    :param cl_centroid: df of cluster centroids
-    :return the radar chart of cluster centroid
-    '''
-
-    # Number of variables
-    var = cl_centroid.columns
-    n_var = len(var)
-
-    # Angle of axis
-    angles = [n / float(n_var) * 2 * pi for n in range(n_var)]
-    angles += angles[:1]
-
-    # Initialise the spider plot
-    fig, ax = plt.subplots(figsize=(15,15))
-    ax = plt.subplot(111, polar=True)
-    ax.set_theta_offset(pi / 2)
-    ax.set_theta_direction(-1)
-
-    # Draw ylabels
-    y_min = floor(min(cl_centroid.min()))
-    y_max = ceil(max(cl_centroid.max()))
-    plt.ylim(y_min, y_max)
-
-    # Plot average
-    values = list(average_centroid)
-    values += values[:1]
-    ax.plot(angles, values, linewidth=4, color='dimgrey',
-            linestyle=':', label="Average")
-    
-    # Plot cluster centroid
-    
-    colors=mcp.gen_color(cmap='viridis',n=len(cl_centroid))
-    for cl in range(len(cl_centroid)):
-        values = cl_centroid.iloc[cl].values.flatten().tolist()
-        values += values[:1]
-        ax.plot(angles, values,  
-                linewidth=5, marker="o", markersize=15,  color=colors[cl], label=f"Cluster {cl+1}")
-        
-    # Draw one axe per variable + add labels
-    ax.tick_params(labelsize=24)
-    ax.set_rticks([1, 2, 3])
-    ax.set_rlabel_position(250)
-    ax.set_xticks(angles[:-1], var)
-    
-
-    labels = []
-    for label, angle in zip(ax.get_xticklabels(), angles):
-        x,y = label.get_position()
[…]

### 2024-10-23 manuscript/manuscript/manuscript.qmd
manuscript-section-results-i

--- 2024-10-23 manuscript/manuscript-231024/manuscript.qmd
@@
 @roweCalafiore2022; @wang2022], we hypothesised that (1) urban mobility has recovered returning to the pre-pandemic baseline level of movement as nonpharmaceuthical restrictions were lifted; and, that (2) socioeconomic differences in urban mobility have endured the pandemic reflecting deep societal inequalities as knowledge-intensive businesses adopt hybrid working.
-<!--# Once we have a more definitive version of the results, highlight our novel contributions here. -->\
+<!--# Once we have a more definitive version of the results, highlight our novel contributions here. -->
 
@@
 
+## Analysing the variability in the impact of COVID-19 on mobility across population groups
+
+The focus of this work is on capturing how mobility patterns for different population groups have been impacted by COVID-19 in Latin America.
+For this purpose, we consider populations groups by classifying their local areas into categories according to their population density and their socioeconomic level.
+These two criteria for classification are chosen due to their relevance for ...
+
+We use five population density categories and three socioeconomic categories defined according to the relative deprivation index.
+These categories are used to classify the spatial units of analysis, following the methodology defined in section X.
+The geographic distribution of categories for each of the countries is displayed in Fig X.a), Fig X.a) and Fig X.a) for Argentina, Chile and Colombia respectively.
+Fig X.b), Fig X.b) and Fig.b) show, for each country, the population distribution into the categories of analysis, considering, on the one hand, Worldpop population estimations and on the other hand, Facebook population, based on user counts during the pre-pandemic baseline period (see section X of the methodology for further details).
+For the three countries we observe discrepancies in the population distribution according to WorldPop and Facebook data, which reflect the extent to which the Facebook data tends to overrepresent certain population groups with respect to the Worldpop data whilst underrepresenting others.
+For example, for Argentina and Chile, the proportion of population living in areas of low population density is higher according to Worldpop data, hence indicating that Facebook data underrepresents populations in these regions.
+Similarly, in those countries, Facebook data shows and overrepresentation of the most affluent socioeconomic group.
+
+Biases in Facebook data arise for a variety of reasons.
+EXPLAIN.
+Finally, Meta shares Facebook data for active user population and flows that has been spatially and temporally aggregated.
+Their data aggregation methodology is designed so that records corresponding to low counts are not included in the dataset in order to preserve the privacy of users.
+Analysing aggregated Facebook data in its raw format can therefore result in an underrepresentation of groups that produce low data counts.
+To counter these source of bias in the Facebook data, we propose a methodology to process the data, which is fully described in the section X.
+
+The analysis of mobility is done using Facebook data that has been pre-processed according to the proposed methodology for bias correction.
+Fig X.c), Fig X.c) and X.c) show, for Argentina, Chile and Colombia respectively, changes in the intensity of movement relative to pre-pandemic baseline levels, at two points during the period of analysis, May 2020 and March 2022.
+The patterns are consistent across the three countries in both dimensions of I'M HERE!!!!!
+
+## COVID-19 unevenly impacted human mobility across population categories in Latin America
+
 The evolution of the percentage change in the number of movements is measured with respect to a baseline period prior to the pandemic as described in @sec-methods.
@@
 
-<!--![Administrative units within functional urban areas in Argentina, Chile, Colombia and Mexico, coloured according to the average Relative Deprivation Index.](figures/maps.pdf)!-->
-
-## The heterogeneous impact of COVID-19 on urban mobility
-
 We analyse the evolution of the percentage change in the number of movements with respect to a baseline period prior to the pandemic.
 Specifically, we focus the analysis on short-distance movements in urban areas to represent local and routine mobility [@owen1992migration], so only movements covering a distance of at most 70 km are considered.
-For a movement to be classified as urban it needs to start or end within a functional urban area from Argentina, Chile, Colombia and Mexico.
+For a movement to be classified as urban it needs to start or end within a functional urban area from Argentina, Chile and Colombia.
 The observed data is available for a two-year period starting in April 2020, just after the first wave of COVID-19 pandemic cases, and ending in March 2022.
@@
 
-## The most deprived areas experienced the smallest drop in mobility relative to pre-pandemic levels
+## Recovery trajectories towards baseline mobility levels are heterogeneous across population categories in Latin America
 
@@
 
-## WorlPpop population data
+## WorldPpop population data
 
@@
 
-In our analysis, we use the Global Gridded Relative Deprivation Index (GRDI), Version 1 (GRDIv1) data set as a measure of socioeocnomic deprivation.
+We use the Global Gridded Relative Deprivation Index (GRDI), Version 1 (GRDIv1) data set as a measure of socioeocnomic deprivation.
 The GRDI data is made available via NASA's Socioeconomic Data and Applications Centre (SEDAC), at a spatial resolution of 30 arc-seconds, or 1 km$^2$ approximately.

### 2024-10-23 manuscript/manuscript/manuscript.tex
manuscript-section-results-i

--- 2024-02-12 manuscript/manuscript.tex
@@
 % Make \paragraph and \subparagraph free-standing
+\makeatletter
 \ifx\paragraph\undefined\else
   \let\oldparagraph\paragraph
-  \renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
+  \renewcommand{\paragraph}{
+    \@ifstar
+      \xxxParagraphStar
+      \xxxParagraphNoStar
+  }
+  \newcommand{\xxxParagraphStar}[1]{\oldparagraph*{#1}\mbox{}}
+  \newcommand{\xxxParagraphNoStar}[1]{\oldparagraph{#1}\mbox{}}
 \fi
@@
   \let\oldsubparagraph\subparagraph
-  \renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
+  \renewcommand{\subparagraph}{
+    \@ifstar
+      \xxxSubParagraphStar
+      \xxxSubParagraphNoStar
+  }
+  \newcommand{\xxxSubParagraphStar}[1]{\oldsubparagraph*{#1}\mbox{}}
+  \newcommand{\xxxSubParagraphNoStar}[1]{\oldsubparagraph{#1}\mbox{}}
 \fi
+\makeatother
 
@@
 \makeatother
+
 \ifLuaTeX
@@
 \begin{abstract}
-COVID-19 triggered a reduction in the number and extent of peoples
-movements. Existing evidence suggests that while the impact of the
+COVID-19 triggered a reduction in the frequency and extent of people's
+movement. Existing evidence suggests that while the impact of the
 pandemic on mobility was widespread, the extent of this impact was
 unequally felt across socioeconomic groups in the early stages of the
-pandemic. Here, we find that in the long term, the most deprived
-locations have experienced a more accelerated recovery towards
-pre-pandemic levels of mobility. Furthermore, the socioeconomic
-disparities in the patterns of mobility triggered by the first outbreak
-of COVID-19, have persisted as of April 2023. These findings are based
-on the analysis of time-series mobility data corresponding to X urban
-areas from Latin American countries collected from Meta-Facebook users
-upon their consent. Our research highlights the importance of timely
-mobility data with high spatiotemporal resolution to understand the
-long-term effects of the pandemic and to inform equitable policy
-responses that address societal challenges in urban areas.
+pandemic. Here, we find that the most deprived locations have
+experienced a more accelerated recovery towards pre-pandemic levels of
+mobility in the long term. Furthermore, the socioeconomic disparities in
+the patterns of mobility triggered by the first outbreak of COVID-19,
+have persisted as of April 2023. These findings are based on the
+analysis of time-series mobility data corresponding to X urban areas
+from Latin American countries collected from Meta-Facebook users upon
+their consent. Our research highlights the importance of timely mobility
+data with high spatiotemporal resolution to understand the long-term
+effects of the pandemic and to inform equitable policy responses that
+address societal challenges in urban areas.
 \end{abstract}
@@
 
-\begin{figure}[H]
-
-{\centering \includegraphics{figures/maps.pdf}
-
-}
-
-\caption{Administrative units within functional urban areas in
-Argentina, Chile, Colombia and Mexico, coloured according to the average
-Relative Deprivation Index.}
-
-\end{figure}%
-
 \subsection{The heterogeneous impact of COVID-19 on urban
@@
 
-\begin{figure}[H]
-
-{\centering \includegraphics{figures/prediction-rdi-band.pdf}
-
-}
-
-\caption{Patterns of recovery for urban mobility in administrative units
-from Argentina, Chile, Colombia and Mexico.}
-
-\end{figure}%
+\textless!- --\href{figures/prediction-rdi-band.pdf}{Patterns of
+recovery for urban mobility in administrative units from Argentina,
+Chile, Colombia and Mexico.} ---!\textgreater{}
 
@@
 the left-hand-side panels of the same Figure.
-
-\begin{figure}[H]
-
-{\centering \includegraphics{figures/regression-evo-nobackground.pdf}
-
-}
-
-\caption{Evolution of the parameters characterising the relationship
-between relative deprivation index and the percentage change in the
-number of movements relative to the pre-pandemic baseline level of
-mobility.}
-
-\end{figure}%
 
@@
 
-\section{Methods}\label{sec-methods}
+\section{Data and Methods}\label{sec-methods}
 
-\subsection{Meta-Facebook movement
-data}\label{meta-facebook-movement-data}
+\subsection{Meta-Facebook data}\label{meta-facebook-data}
 
@@
 
-\subsection{Spatiotemporal data
-aggregation}\label{spatiotemporal-data-aggregation}
+\subsection{WorlPpop population data}\label{worlppop-population-data}
 
-Since the focus of this work is on urban mobility, we focus the analysis
-on Functional Urban Areas (FUAs), defined by the Global Human Settlement
-Layer (GHSL). The spatial extent of the FUAs is often large and may
-include several towns and neighbourhoods displaying a variety of
-socioeconomic characteristics, hence using FUAs as the spatial units of
-aggregation would considerably mask the heterogeneity in the mobility
-patterns. While the original unit of aggregation for the movement data,
-i.e.~the tiles from the Bing Maps Tile System, offers the highest degree
-of spatial granularity available, the interpretation of findings from an
-analysis based on administrative units is often more valuable. For this
-reason, we perform a spatial join to aggregate the movement data into
-administrative units at the GADM level 2 or 3. Only flows of people
-starting or ending within the boundaries of a FUA are considered.
+We used data from WorldPop\textsuperscript{47} to classify the spatial
+units of analysis according to their level of urbanisation, and to
+estimate missing baseline values in the Facebook population data.
+WorldPop offers open access gridded population estimates at a resolution
+as small as 3 arc-seconds approximately 100m and 1km at the Equator,
+respectively. WorldPop produces these gridded datasets using top-down
+(i.e.~dissagregating administrative area counts into smaller grid cells)
+or bottom-up (i.e.~interpolating data from counts from sample locations
+into grid cells) approaches. For the purposes of this work, we use
+gridded population data at a resolution of 1km\(^2\) in raster format.
+We perform a spatial join of the Facebook spatial units (Bing tiles)
+with the gridded population data and compute the sum of Worldpop
+populations corresponding to each of the Facebook spatial units.
 
-While the original movement data is originally aggregated into 8-hour
-windows, this resolution is too fine for our analysis. Since the
-analysis is focused on the longer-term evolution of patterns of urban
-mobility, we aggregate the movement data by month.
+\subsection{Socioeconomic deprivation
+data}\label{socioeconomic-deprivation-data}
 
-In our analysis, we use the Relative Deprivation Index (RDI) as a
-measure of socioeocnomic deprivation. The RDI data is made available via
-NASA's Socioeconomic Data and Applications Centre (SEDAC), with a
-spatial resolution of 1km pixels. We perform a spatial join of the
-gridded data and the administrative units and compute the average RDI
-within each of these administrative units.
+In our analysis, we use the Global Gridded Relative Deprivation Index
+(GRDI), Version 1 (GRDIv1) data set as a measure of socioeocnomic
+deprivation. The GRDI data is made available via NASA's Socioeconomic
+Data and Applications Centre (SEDAC), at a spatial resolution of 30
+arc-seconds, or 1 km\(^2\) approximately. The index quantifies the
+relative levels of multidimensional deprivation and poverty, where a
+value of 100 represents the highest level of deprivation and a value of
+0 the lowest. We perform a spatial join of the Facebook spatial units
+and the gridded relative deprivation data and compute the average RDI
+corresponding to each of the Facebook spatial units.
 
-\subsection{Time series analysis}\label{time-series-analysis}
+\subsection{Classification of tiles according to level of urbanisation
+and socioeconomic
+deprivation}\label{classification-of-tiles-according-to-level-of-urbanisation-and-socioeconomic-deprivation}
 
-For the countries included in the analysis, the movement data is
-available until March 2022. In order to gain a better understanding of
-the recovery trends after the pandemic, we generate a 12-month forecast
-up to March 2023. This is done using Prophet\textsuperscript{27}, a
-procedure developed by Meta's Core Data Science team to forecast time
-series data based on an additive model. Prophet stands out from other
-forecasting methods due to its ability to capture recurring seasonal
-trends inherent in mobility data, such as fluctuations during holidays,
-or changes in the general trend due to the impact of particular
-circumstances, such as the COVID-19 pandemic. It is particularly
-intuitive to use compared to other models due to its automated
-trend-detection capabilities, making it accessible to users with no
-expertise on the underlying model, while being robust to missing data
-and outliers. For the analysis in this paper, yearly seasonality effects
-are included as they yield the most realistic predictions.
+\subsection{Processing Facebook data}\label{processing-facebook-data}
 
-Figures 2 and 3 have been created by removing statistical outliers. We
-define these as the administrative units which, at any given month, have
-a z-score greater than 4 for the percentage change in the number of
-movements. Furthermore, a Savitzky-Golay filter is applied to smooth the
-time series data, using a length of 4 units for the filter window and
-order 2 polynomials to fit the samples.
+To ensure the privacy and anonymity of the users' data, Meta removes
+information corresponding to data entries where the population or
+movement count is less than 10 for a specific time or day, retaining
+only information about the percentage change in the number of counts
+with respect to the baseline period is retained\textsuperscript{46}.
+Consequently, we observe many tiles where the population or movement
+count for either the baseline or crisis period are blank.
+
+\subsubsection{Facebook population data}\label{facebook-population-data}
+
+To imput Facebook population baseline values, we first identify all the
+baseline values that are available for each spatial unit and weekday (of
+the three available time windows, we only consider one per day). We then
+estimate the missing baseline values for each weekday, based on a linear
+model for the relationship between the Worldpop population and the
+Facebook population counts, which are fitted using ordinary least
+squares. This is illustrated in Supplementary Figure
+\ref{fig:imput-baseline-population}.
+
+We then use the complete baseline Facebook population values to compute
+missing Facebook population counts during the crisis period. This is
+possible because, as mentioned above, Meta reports the percentage change
+in the number of counts with respect to the baseline, even if the counts
+are not reported due to low value.
+
+\subsubsection{Facebook movement data}\label{facebook-movement-data}
+
+The imputation of Facebook movement baseline values is done according to
+a spatial interaction model (see e.g.\textsuperscript{48}). We
+considered the population flow between an origin and a destination tile,
+and model this variable as a function of the Facebook population count
+at the origin tile on the same weekday, the Facebook population count at
+the destination on the same weekday and the distance between origin and
+destination; we also included indicator variables to capture the day of
+the week and the pair of population density classes corresponding to the
+origin and destination tiles. Mathematically, this model can be
+expressed as \begin{equation}
+\mu_{ijw} = \beta_0 + \beta_1pop_{iw} +\beta_2pop_{jw} + \beta_3d_{ij} + \beta_4W + \beta_5C + \varepsilon
+\end{equation}
+
+where \(\mu_{ijw}\) is the expectation of the flow of people from tile
+\(i\) to tile \(j\) on the weekday \(w\); \(\beta_0\) is an intercept
+\(pop_{iw}\) and \(pop_{jw}\) are the Facebook population counts at the
+origin and destination on weekday \(w\) during the baseline period,
+\(d_{ij}\) is the distance between origin and destination, \(W\) is a
+series of indicator variables capturing the day of the week and
+similarly \(C\) is a series of indicator variables reflecting the pair
+of population density classes for the origin and destination tiles,
+resulting in X pairs (10 classes × 10 classes minus one so it is not
+collinear with \(\beta_0\)), \(\beta_{0,1,2,3,4,5}\) are model
+parameters to be estimated from the observed data. The error term is
+denoted by \varepsilon.
+
+To estimate the model parameters, we used a count data regression model.
+Specifically, we used a negative binomial regression which is a
+generalised linear model (GLM) where overdispersion of the error term is
+assumed, i.e.~the variance exceeds the mean.
+
+We compute missing Facebook movement counts during the crisis period by
+considering the complete Facebook movements baseline and the percentage
+change in the number of counts with respect to the baseline, which is
+reported in the Facebook Movements datasets even when the count is not
+reported due to its low value.
+
+\subsubsection{Correction factors??? For
+representativeness}\label{correction-factors-for-representativeness}
+
+\subsection{Data analysis}\label{data-analysis}
 
@@
 
+\bibitem[\citeproctext]{ref-tatem2017}
+\CSLLeftMargin{47. }%
+\CSLRightInline{Tatem, A. J.
+\href{https://doi.org/10.1038/sdata.2017.4}{WorldPop, open data for
+spatial demography}. \emph{Scientific Data} \textbf{4}, (2017).}
+
+\bibitem[\citeproctext]{ref-rowe2022a}
+\CSLLeftMargin{48. }%
+\CSLRightInline{Rowe, F., Lovelace, R. \& Dennett, A.
+\href{http://dx.doi.org/10.31219/osf.io/xcdms}{Spatial interaction
+modelling: A manifesto}. (2022).}
+
 \end{CSLReferences}

### 2024-10-23 manuscript/manuscript/supplementary.qmd
manuscript-section-results-i

---
title: "**Supplementary material**"
author:
  - name: Carmen Cabrera-Arnau
    affil-id: 1
  - name: Francisco Rowe
    affil-id: 1
  - name: Miguel González-Leonardo
    affil-id: 2
  - name: Andrea Nasuto
    affil-id: 1
  - name: Ruth Neville
    affil-id: 1
affiliations: 
  - id: 1
    name: Geographic Data Science Lab, Department of Geography and Planning, University of Liverpool, Liverpool, UK
  - id: 2
    name: Centre for Demographic Urban and Environmental Studies, El Colegio de México, Ciudad de México, México

format: 
  pdf:
    fontsize: 11pt
    geometry:
      - left=2.54cm
      - right=2.54cm
    header-includes:
      \usepackage{lipsum}
      \usepackage{libertine}
    keep-tex: true
    template-partials: 
      - title.tex
    include-in-header:
      text: |
        \usepackage[noblocks]{authblk}
        \renewcommand*{\Authsep}{, }
        \renewcommand*{\Authand}{, }
        \renewcommand*{\Authands}{, }
        \renewcommand\Affilfont{\small}
        \usepackage{multirow}
        \usepackage{makecell}
number-sections: true
editor: visual
csl: nature.csl
bibliography: refs.bib
editor_options:
  markdown:
    wrap: sentence
---

<!--# FR: A few questions to think about: Have we validated the data? i.e. tested their extent of representativeness? Shall we do something quick using census data for each country comparing overall populations?  -->

# Proportion of missing records across population density and socioeconomic deprivation classes,

## Facebook Population baseline data

```{=tex}
\begin{table}[ht]
 \centering 
 \captionsetup{width=\textwidth}
 \caption{Percentage of missing Facebook Population data records during the baseline period, across population density and socioeconomic deprivation classes (class 1 is the most sparsely populated and the most deprived).}

\begin{tabular}{cccccc}
  \multirow{2}{*}{\textbf{Country}} & \multirow{2}{*}{\textbf{Records count}} & \multirow{2}{*}{\textbf{Variable}} & \multirow{2}{*}{\textbf{Class 1}} & \multirow{2}{*}{\textbf{Class 2}} & \multirow{2}{*}{\textbf{Class 3}}\\
  & & & & \\
  \Xhline{1.5pt}

  \multirow{5}{*}{ \;\; Argentina \;\; } & \;\;Total\;\; & & & & \\ 
  \cline{2-6}
  & \multirow{2}{*}{Missing} & \;\;Pop. density\;\; & & & \\
  & & \;\;Socioec. deprivation\;\; & & & \\
  \cline{2-6}
  & \multirow{2}{*}{Percentage missing} &  \;\;Pop. density\;\; & & & \\
  & & \;\;Socioec. deprivation\;\; & & & \\
  \Xhline{1.5pt}

  \multirow{5}{*}{ \;\; Chile \;\; } & \;\;Total\;\; & & & & \\ 
  \cline{2-6}
  & \multirow{2}{*}{Missing} & \;\;Pop. density\;\; & & & \\
  & & \;\;Socioec. deprivation\;\; & & & \\
  \cline{2-6}
  & \multirow{2}{*}{Percentage missing} &  \;\;Pop. density\;\; & & & \\
  & & \;\;Socioec. deprivation\;\; & & & \\
  \Xhline{1.5pt}

  \multirow{5}{*}{ \;\; Colombia \;\; } & \;\;Total\;\; & & & & \\ 
  \cline{2-6}
  & \multirow{2}{*}{Missing} & \;\;Pop. density\;\; & & & \\
  & & \;\;Socioec. deprivation\;\; & & & \\
  \cline{2-6}
  & \multirow{2}{*}{Percentage missing} &  \;\;Pop. density\;\; & & & \\
  & & \;\;Socioec. deprivation\;\; & & & \\
  \Xhline{1.5pt}

  \multirow{5}{*}{ \;\; Mexico \;\; } & \;\;Total\;\; & & & & \\ 
  \cline{2-6}
  & \multirow{2}{*}{Missing} & \;\;Pop. density\;\; & & & \\
  & & \;\;Socioec. deprivation\;\; & & & \\
  \cline{2-6}
  & \multirow{2}{*}{Percentage missing} &  \;\;Pop. density\;\; & & & \\
  & & \;\;Socioec. deprivation\;\; & & & \\

 \end{tabular}
\vspace{5mm}
\end{table}
```
## For Facebook Movements baseline data

```{=tex}
\begin{table}[ht]
 \centering 
 \captionsetup{width=\textwidth}
 \caption{Percentage of missing Facebook Movement data records during the baseline period, across population density and socioeconomic deprivation classes of origin tile (class 1 is the most sparsely populated and the most deprived).}

\begin{tabular}{cccccc}
  \multirow{2}{*}{\textbf{Country}} & \multirow{2}{*}{\textbf{Records count}} & \multirow{2}{*}{\textbf{Variable}} & \multirow{2}{*}{\textbf{Class 1}} & \multirow{2}{*}{\textbf{Class 2}} & \multirow{2}{*}{\textbf{Class 3}}\\
  & & & & \\
  \Xhline{1.5pt}

  \multirow{5}{*}{ \;\; Argentina \;\; } & \;\;Total\;\; & & & & \\ 
  \cline{2-6}
  & \multirow{2}{*}{Missing} & \;\;Pop. density\;\; & & & \\
  & & \;\;Socioec. deprivation\;\; & & & \\
  \cline{2-6}
  & \multirow{2}{*}{Percentage missing} &  \;\;Pop. density\;\; & & & \\
  & & \;\;Socioec. deprivation\;\; & & & \\
  \Xhline{1.5pt}

  \multirow{5}{*}{ \;\; Chile \;\; } & \;\;Total\;\; & & & & \\ 
  \cline{2-6}
  & \multirow{2}{*}{Missing} & \;\;Pop. density\;\; & & & \\
  & & \;\;Socioec. deprivation\;\; & & & \\
  \cline{2-6}
  & \multirow{2}{*}{Percentage missing} &  \;\;Pop. density\;\; & & & \\
  & & \;\;Socioec. deprivation\;\; & & & \\
  \Xhline{1.5pt}

  \multirow{5}{*}{ \;\; Colombia \;\; } & \;\;Total\;\; & & & & \\ 
  \cline{2-6}
  & \multirow{2}{*}{Missing} & \;\;Pop. density\;\; & & & \\
  & & \;\;Socioec. deprivation\;\; & & & \\
  \cline{2-6}
  & \multirow{2}{*}{Percentage missing} &  \;\;Pop. density\;\; & & & \\
  & & \;\;Socioec. deprivation\;\; & & & \\
  \Xhline{1.5pt}

  \multirow{5}{*}{ \;\; Mexico \;\; } & \;\;Total\;\; & & & & \\ 
  \cline{2-6}
  & \multirow{2}{*}{Missing} & \;\;Pop. density\;\; & & & \\
  & & \;\;Socioec. deprivation\;\; & & & \\
  \cline{2-6}
  & \multirow{2}{*}{Percentage missing} &  \;\;Pop. density\;\; & & & \\
  & & \;\;Socioec. deprivation\;\; & & & \\

 \end{tabular}
\vspace{5mm}
\end{table}
```
## For Facebook Movements during crisis

```{=tex}

\begin{table}[ht]
 \centering 
 \captionsetup{width=\textwidth}
 \caption{Percentage of missing Facebook Movement data records during the crisis period, across population density and socioeconomic deprivation classes of origin tile (class 1 is the most sparsely populated and the most deprived).}

\begin{tabular}{cccccc}
  \multirow{2}{*}{\textbf{Country}} & \multirow{2}{*}{\textbf{Records count}} & \multirow{2}{*}{\textbf{Variable}} & \multirow{2}{*}{\textbf{Class 1}} & \multirow{2}{*}{\textbf{Class 2}} & \multirow{2}{*}{\textbf{Class 3}}\\
  & & & & \\
  \Xhline{1.5pt}

  \multirow{5}{*}{ \;\; Argentina \;\; } & \;\;Total\;\; & & & & \\ 
  \cline{2-6}
  & \multirow{2}{*}{Missing} & \;\;Pop. density\;\; & & & \\
  & & \;\;Socioec. deprivation\;\; & & & \\
  \cline{2-6}
  & \multirow{2}{*}{Percentage missing} &  \;\;Pop. density\;\; & & & \\
  & & \;\;Socioec. deprivation\;\; & & & \\
  \Xhline{1.5pt}

  \multirow{5}{*}{ \;\; Chile \;\; } & \;\;Total\;\; & & & & \\ 
  \cline{2-6}
  & \multirow{2}{*}{Missing} & \;\;Pop. density\;\; & & & \\
  & & \;\;Socioec. deprivation\;\; & & & \\
  \cline{2-6}
  & \multirow{2}{*}{Percentage missing} &  \;\;Pop. density\;\; & & & \\
  & & \;\;Socioec. deprivation\;\; & & & \\
  \Xhline{1.5pt}

  \multirow{5}{*}{ \;\; Colombia \;\; } & \;\;Total\;\; & & & & \\ 
  \cline{2-6}
  & \multirow{2}{*}{Missing} & \;\;Pop. density\;\; & & & \\
  & & \;\;Socioec. deprivation\;\; & & & \\
  \cline{2-6}
  & \multirow{2}{*}{Percentage missing} &  \;\;Pop. density\;\; & & & \\
  & & \;\;Socioec. deprivation\;\; & & & \\
  \Xhline{1.5pt}

  \multirow{5}{*}{ \;\; Mexico \;\; } & \;\;Total\;\; & & & & \\ 
  \cline{2-6}
  & \multirow{2}{*}{Missing} & \;\;Pop. density\;\; & & & \\
  & & \;\;Socioec. deprivation\;\; & & & \\
  \cline{2-6}
  & \multirow{2}{*}{Percentage missing} &  \;\;Pop. density\;\; & & & \\
  & & \;\;Socioec. deprivation\;\; & & & \\

 \end{tabular}
\vspace{5mm}
\end{table}
```
# References

### 2024-10-26 manuscript/manuscript-nature/manuscript-nature.qmd
results1-fig1

--- 2024-10-23 manuscript/manuscript-231024/manuscript.qmd
@@
 ---
-title: "**Urban mobility and socioeconomic deprivation in Latin America after COVID-19**"
+title: Data Descriptor for Nature Scientific Data
 author: 
+## listed alphabetically
     - name: Carmen Cabrera-Arnau
-    affil-id: 1
+      affiliation: 
+        name: "1,* "
     - name: Francisco Rowe
-    affil-id: 1
+      affiliation: 
+        name: "1"
     - name: Miguel González-Leonardo
-    affil-id: 2
+      affiliation: 
+        name: "2"
     - name: Andrea Nasuto
-    affil-id: 1
+      affiliation: 
+        name: "1"
     - name: Ruth Neville
-    affil-id: 1
-affiliations: 
-  - id: 1
-    name: Geographic Data Science Lab, Department of Geography and Planning, University of Liverpool, Liverpool, UK
-  - id: 2
-    name: Centre for Demographic Urban and Environmental Studies, El Colegio de México, Ciudad de México, México
+      affiliation: 
+        name: "1"
+authornotes:
+    - note: "1"
+      text: Geographic Data Science Lab, Department of Geography and Planning, University of Liverpool, Liverpool, UK
+    - note: "2"
+      text: Centre for Demographic Urban and Environmental Studies, El Colegio de México, Ciudad de México, México
+    - note: "*"
+      text: "`c.cabrera-arnau@liverpool.ac.uk`"
+format:
+  scientific-data-pdf: default
+date: last-modified
 abstract: |
@@
 Our research highlights the importance of timely mobility data with high spatiotemporal resolution to understand the long-term effects of the pandemic and to inform equitable policy responses that address societal challenges in urban areas.
-format: 
-  pdf:
-    fontsize: 11pt
-    geometry:
-      - left=2.54cm
-      - right=2.54cm
-    header-includes:
-      \usepackage{lipsum}
-      \usepackage{libertine}
-    keep-tex: true
-    template-partials: 
-      - title.tex
-    include-in-header:
-      text: |
-        \usepackage[noblocks]{authblk}
-        \renewcommand*{\Authsep}{, }
-        \renewcommand*{\Authand}{, }
-        \renewcommand*{\Authands}{, }
-        \renewcommand\Affilfont{\small}
-number-sections: true
-editor: visual
-csl: nature.csl
-bibliography: refs.bib
-editor_options:
-  markdown:
-    wrap: sentence
+
+bibliography: bibliography.bib
 ---
 
-<!--# FR: A few questions to think about: Have we validated the data?
-i.e.
-tested their extent of representativeness?
-Shall we do something quick using census data for each country comparing overall populations?
--->
+# Main {.unnumbered}
 
-\
-
-# Introduction {#sec-intro}
+<!--# CCA: Fr to rewrite section -->
 
@@
 In most countries, affluent individuals tended to record the greatest drops in mobility levels as they are predominantly employed in knowledge-intensive jobs which can be done fully or partly remotely [@fraiberger2020; @bonaccorsi2020; @weill2020; @dueñas2021; @santana2023].
-During the COVID-19 pandemic, the adoption of remote work reduced the need of commuting for knowledge-intensive, non-public facing jobs [@florida2021].
-At the same time, individuals from less privileged socioeconomic backgrounds displayed less pronounced declines mirroring the nature of their work requiring public-facing, face-to-face interaction, and thus a requirement for daily work commutes [@dueñas2021; @santana2023].
+During the COVID-19 pandemic, the adoption of remote work reduced the need of commuting for knowledge-intensive, non-public facing jobs [@florida2021].At the same time, individuals from less privileged socioeconomic backgrounds displayed less pronounced declines mirroring the nature of their work requiring public-facing, face-to-face interaction, and thus a requirement for daily work commutes [@dueñas2021; @santana2023].
 
@@
 
-# Results {#sec-results}
+# Results
 
-The evolution of the percentage change in the number of movements is measured with respect to a baseline period prior to the pandemic as described in @sec-methods.
+## Analysing the variability in the impact of COVID-19 on mobility across population groups
+
+We focus on capturing how mobility patterns for different population groups have been impacted by COVID-19 in Latin America.
+We consider populations groups by classifying the spatial units of analysis into categories according to their population density and their socioeconomic level.
+More details for the classification method are provided in section X.
+The two criteria for classification are chosen due to their relevance for ...
+We use five population density categories and three socioeconomic categories defined according to the relative deprivation index.
+The geographic distribution of categories for each of the countries is displayed in Figures @Fig1-ARG a), @Fig1-CHL a) and @Fig1-COL a) for Argentina, Chile and Colombia respectively.
+
+A key challenge in the in the analysis of population counts and movements with Facebook user data is the absence of small count records, which is a result of privacy-protection techniques applied to ensure that the location of individuals or small groups cannot be identified.
+As a consequence, the countries of the analysis have locations with null values for Facebook population counts and flows.
+These missing data are not distributed randomly.
+For example, the missing values for FB population counts display high spatial autocorrelation as shown in Supplementary Figure SF1.
+Therefore, simply removing the missing records from the analysis could lead to geographically biased results (Afghari et al., 2019).
+To address this, we designed a data processing method to estimate the missing values, both for Facebook population counts and flows.
+Furthermore, this method also applies a correction factor to eliminate the fluctuations in the daily number of observations, assuming that the representativeness of Facebook data across spatial units remains stable during the study period.
+The data processing method is described in section X.
+
+Even after applying the imputation method to estimate missing values, we acknowledge that Facebook data may still overrepresent certain population groups while underrepresenting others.
+For example, Figures @Fig1-ARG b), @Fig1-CHL b) and @Fig1-COL b) show, for each country, the proportion of the population across the various analysis categories.
+These comparisons use WorldPop population estimates and Facebook population counts, with the latter reflecting active users on an average day during a pre-pandemic baseline period (see Section X of the methodology for details).
+In all three countries, notable discrepancies appear between the population distributions according to WorldPop and Facebook data.
+For instance, in Argentina and Chile, WorldPop data indicates a higher proportion of people living in low-density areas, suggesting that Facebook data underrepresents populations in these regions.
+Similarly, in these countries, Facebook data shows an overrepresentation of the most affluent socioeconomic group.
+While addressing this kind of representativity bias is beyond the scope of this paper, we recognise its significance and the importance of addressing it in future analyses.
+This issue is revisited in the Discussion section.
+
+![](figures/Fig1.png)
+
+The analysis of mobility is therefore done using Facebook movement data that has been pre-processed according to our methodology.
+Figures @Fig1-ARG c), @Fig1-CHL c) and @Fig1-COL c) show, for Argentina, Chile and Colombia respectively, changes in the intensity of movement relative to pre-pandemic baseline levels, at two points during the period of analysis, May 2020 and March 2022.
+The patterns show remarkable consistency across countries for both classification criteria.
+
+## COVID-19 unevenly impacted human mobility across population categories in Latin America
+
+The evolution of the percentage change in the number of movements is measured with respect to a baseline period prior to the pandemic as described in Methods.
 For the purposes of the analysis, we aggregate the raw movement data temporally into months and spatially into administrative units at various levels according to GADM, the Database of Global Administrative Areas [@GADM24].
@@
 Predictions about the evolution of the percentage change in the number of movements are made using the Prophet forecasting procedure.
-Further details are provided in @sec-methods.
-
-<!--![Administrative units within functional urban areas in Argentina, Chile, Colombia and Mexico, coloured according to the average Relative Deprivation Index.](figures/maps.pdf)!-->
-
-## The heterogeneous impact of COVID-19 on urban mobility
+Further details are provided in Methods.
 
@@
 Specifically, we focus the analysis on short-distance movements in urban areas to represent local and routine mobility [@owen1992migration], so only movements covering a distance of at most 70 km are considered.
-For a movement to be classified as urban it needs to start or end within a functional urban area from Argentina, Chile, Colombia and Mexico.
+For a movement to be classified as urban it needs to start or end within a functional urban area from Argentina, Chile and Colombia.
 The observed data is available for a two-year period starting in April 2020, just after the first wave of COVID-19 pandemic cases, and ending in March 2022.
@@
 
-# Data and Methods {#sec-methods}
+# Data and methods {.unnumbered}
 
@@
 ```
+# Code availability {.unnumbered}
+
+For all studies using custom code in the generation or processing of datasets, a statement must be included under the heading "Code availability", indicating whether and how the code can be accessed, including any restrictions to access.
+This section should also include information on the versions of any software used, if relevant, and any specific variables or parameters used to generate, test, or process the current dataset.
+
 # References
+
+::: {#refs}
+:::
+
+# Acknowledgements {.unnumbered}
+
+The work was supported by Research England..., project no. ...
+, ‘RECAST’.
+
+# Author information {.unnumbered}
+
+## Authors and affiliations
+
+## Contributions
+
+A.A.
+conceived the experiment(s), A.A.
+and B.A.
+conducted the experiment(s), C.A.
+and D.A.
+analysed the results.
+All authors reviewed the manuscript.
+
+# Ethics declarations {.unnumbered}
+
+## Competing interests
+
+The authors declare no competing interests.
+
+# Supplementary information {.unnumbered}
+
+# Rights and permissions {.unnumbered}
+
+Springer Nature or its licensor (e.g.
+a society or other partner) holds exclusive rights to this article under a publishing agreement with the author(s) or other rightsholder(s); author self-archiving of the accepted manuscript version of this article is solely governed by the terms of such publishing agreement and applicable law.

### 2024-10-26 manuscript/manuscript/manuscript.qmd
results1-fig1

@@
 Our research highlights the importance of timely mobility data with high spatiotemporal resolution to understand the long-term effects of the pandemic and to inform equitable policy responses that address societal challenges in urban areas.
+
 format: 
+  nature-pdf: default
   pdf:
@@
 These two criteria for classification are chosen due to their relevance for ...
-
 We use five population density categories and three socioeconomic categories defined according to the relative deprivation index.
@@
 The geographic distribution of categories for each of the countries is displayed in Fig X.a), Fig X.a) and Fig X.a) for Argentina, Chile and Colombia respectively.
-Fig X.b), Fig X.b) and Fig.b) show, for each country, the population distribution into the categories of analysis, considering, on the one hand, Worldpop population estimations and on the other hand, Facebook population, based on user counts during the pre-pandemic baseline period (see section X of the methodology for further details).
-For the three countries we observe discrepancies in the population distribution according to WorldPop and Facebook data, which reflect the extent to which the Facebook data tends to overrepresent certain population groups with respect to the Worldpop data whilst underrepresenting others.
-For example, for Argentina and Chile, the proportion of population living in areas of low population density is higher according to Worldpop data, hence indicating that Facebook data underrepresents populations in these regions.
-Similarly, in those countries, Facebook data shows and overrepresentation of the most affluent socioeconomic group.
 
-Biases in Facebook data arise for a variety of reasons.
-EXPLAIN.
-Finally, Meta shares Facebook data for active user population and flows that has been spatially and temporally aggregated.
-Their data aggregation methodology is designed so that records corresponding to low counts are not included in the dataset in order to preserve the privacy of users.
-Analysing aggregated Facebook data in its raw format can therefore result in an underrepresentation of groups that produce low data counts.
-To counter these source of bias in the Facebook data, we propose a methodology to process the data, which is fully described in the section X.
+A key challenge in the in the analysis of population counts and movements with Facebook user data is the absence of small count records, which is a result of privacy-protection techniques applied to ensure that the location of individuals or small groups cannot be identified.
+As a consequence, the countries of the analysis have locations with null values for population counts and flows.
+Given that these missing data are not distributed randomly (Supplementary Table A1), simply deleting the incomplete records could lead to biased results (Afghari et al., 2019).
+Therefore, we designed an imputation method to estimate these missing values.
+The overview of this missing data imputation for Facebook user count is presented in Fig.
+1.
 
-The analysis of mobility is done using Facebook data that has been pre-processed according to the proposed methodology for bias correction.
+Even after applying the imputation method for the estimation of missing values, Facebook data tends to overrepresent certain population groups with respect to Wordlpop data whilst underrepresenting others.
+Fig X.b), Fig X.b) and Fig.b) show, for each country, the share of population in each of the categories of analysis, considering, on the one hand, Worldpop population estimations and on the other hand, Facebook population, based on user counts during the pre-pandemic baseline period (see section X of the methodology for further details).
+For the three countries we observe discrepancies in the population distribution according to WorldPop and Facebook data.
+For example, for Argentina and Chile, the share of population living in areas of low population density is higher according to Worldpop data, hence indicating that Facebook data underrepresents populations in these regions.
+Similarly, in those countries, Facebook data shows an overrepresentation of the most affluent socioeconomic group.
+"Addressing this type of representativity bias is beyond the scope of this paper, but we are aware of it and we recognise the importance of addressing this in future analysis".
+
+The analysis of mobility is therefore done using Facebook movementdata that has been pre-processed for imputation of low counts.
 Fig X.c), Fig X.c) and X.c) show, for Argentina, Chile and Colombia respectively, changes in the intensity of movement relative to pre-pandemic baseline levels, at two points during the period of analysis, May 2020 and March 2022.

### 2024-10-27 manuscript/manuscript-nature/manuscript-nature.qmd
manuscript-figures

@@
 Even after applying the imputation method to estimate missing values, we acknowledge that Facebook data may still overrepresent certain population groups while underrepresenting others.
-For example, Figures @Fig1-ARG b), @Fig1-CHL b) and @Fig1-COL b) show, for each country, the proportion of the population across the various analysis categories.
+For example, Figure @Fig1 shows, for each country, the proportion of the population across the various analysis categories.
 These comparisons use WorldPop population estimates and Facebook population counts, with the latter reflecting active users on an average day during a pre-pandemic baseline period (see Section X of the methodology for details).
@@
 
-![](figures/Fig1.png)
+![Classification of spatial units into categories by population density and by relative deprivation index.
+Spatial distribution of categories and population share in each category, by country.](figures/Fig1.png){#Fig1}
 
-The analysis of mobility is therefore done using Facebook movement data that has been pre-processed according to our methodology.
-Figures @Fig1-ARG c), @Fig1-CHL c) and @Fig1-COL c) show, for Argentina, Chile and Colombia respectively, changes in the intensity of movement relative to pre-pandemic baseline levels, at two points during the period of analysis, May 2020 and March 2022.
-The patterns show remarkable consistency across countries for both classification criteria.
+## COVID-19’s impact on human mobility displays consistently uneven patterns across population groups in Latin American countries
 
-## COVID-19 unevenly impacted human mobility across population categories in Latin America
+The analysis of mobility is done using Facebook movement data that has been pre-processed according to our methodology.
+Figure @Fig2 shows changes in the intensity of movement measured as the percentage change in the number of inflows relative to pre-pandemic baseline levels, at two points during the period of analysis, May 2020 and March 2022.
+These changes show remarkably consistent patterns across countries for both types of classification.
 
-The evolution of the percentage change in the number of movements is measured with respect to a baseline period prior to the pandemic as described in Methods.
-For the purposes of the analysis, we aggregate the raw movement data temporally into months and spatially into administrative units at various levels according to GADM, the Database of Global Administrative Areas [@GADM24].
-The analysis focuses on administrative units that are within the boundaries of functional urban areas as specified by the Global Human Settlement Layer (GHSL).
-For each administrative unit, we compute the Relative Deprivation Index (RDI) based on data from NASA's Socioeconomic Data and Applications Centre (SEDAC).
-Figure 1 displays the administrative units included in the study, coloured according to their average RDI.
-Predictions about the evolution of the percentage change in the number of movements are made using the Prophet forecasting procedure.
-Further details are provided in Methods.
+Focusing first on the early pandemic days, we observe a decrease in the number of inflows for all countries and most categories of analysis.
+This decrease displays a systematic gradient across categories of analysis whereby the more densely populated or the more affluent an area is, the higher the losses in the number of inflows.
+Differences in the levels of mobility across population density or RDI categories tend to diminish with time, with a less pronounced gradient across categories of analysis approximately two years later.
 
-We analyse the evolution of the percentage change in the number of movements with respect to a baseline period prior to the pandemic.
-Specifically, we focus the analysis on short-distance movements in urban areas to represent local and routine mobility [@owen1992migration], so only movements covering a distance of at most 70 km are considered.
-For a movement to be classified as urban it needs to start or end within a functional urban area from Argentina, Chile and Colombia.
-The observed data is available for a two-year period starting in April 2020, just after the first wave of COVID-19 pandemic cases, and ending in March 2022.
-After 2022 no observations are available, however, we generate a 12-month forecast up to March 2023 in order to gain a better understanding of the recovery trends.
+These patterns can be attributed to...
+remote work, movement to rural areas, affluent people being more able to abide to lockdowns, etc...
+and the patterns are consistent with what has been found in ...
 
-Figure 2 displays the patterns of recovery for the mobility levels in the administrative units belonging to functional urban areas in the countries of interest.
-The three lines in each panel represent the mean levels of mobility for administrative units grouped into one of three terciles, according to their average RDI.
-
-\<!-
---[Patterns of recovery for urban mobility in administrative units from Argentina, Chile, Colombia and Mexico.](figures/prediction-rdi-band.pdf)
----!\>
-
-<!--# please add a description of what the bands mean. -->
-
-Generally, there was a drop in the levels of mobility with respect to the baseline period in all four countries.
-This drop was especially large for Argentina, Chile and Colombia, with Mexico displaying a smaller decrease in the percentage change in the number of movements with respect to the baseline.
-Following the initial drop in mobility, all four countries evolve towards the recovery of baseline levels of mobility, with a generally increasing trend.
-There are, however, fluctuations from the general trend which manifest differently for each country.
-These fluctuations mirror each other in the case of Argentina and Colombia, where urban mobility sharply bounces back closer to pre-pandemic levels around July of 2020.
-Chile and Mexico display more progressive patterns of recovery, although Chile never reaches baseline levels.
-These fluctuations are unique to each country and can be attributed to local factors such as the effects of seasonality or the different stringency measures imposed by the national governments during the pandemic.
-
-From Figure 2, we observe that there is a consistent tendency in how administrative units with varying levels of deprivation were affected by the pandemic.
-For all four countries, we observe that the administrative units in the most deprived tercile are the ones that experienced the smallest loss in levels of mobility at the beginning of the pandemic.
-Differences in the levels of mobility across relative deprivation terciles diminish with time.
-Argentina and Chile stand out as the countries with the largest differences in mobility levels for different relative deprivation terciles.
+![Changes in the intensity of inflows as the percentage of change with respect to the pre-pandemic baseline levels, for each country and category of analysis.](figures/Fig2.png)
 
@@
 
-In this section we explore further the role of socioeconomic deprivation in the evolution of the levels of urban mobility.
-For a given point in time (i.e.
-a month), we start by considering the relationship between the percentage change in the number of movements relative to the pre-pandemic baseline period and the average RDI, at the administrative unit level.
-We assume that this relationship is linear and we use a linear regression to estimate the slope and intercept characterising the line of best fit.
-This is shown for April 2020 and March 2022 in the right-hand side panels of Figure 3.
-After obtaining the slope and intercept for every month, we are able to plot the evolution of these parameters for both the observed and forecasted data, as displayed on the left-hand-side panels of the same Figure.
+In this section we explore trends in the intensity of movements over time across population groups.
+Generally, all categories of analysis experienced an initial drop in the number of movements followed by an upwards general trend of recovery towards baseline levels of mobility.
+The extent of the initial drop and the speed of recovery varied by country and by analysis category.
+These general patterns can be visually explored in Figure @Fig3 and @Fig4 for a classification by population density category and by RDI respectively.
+Below we further explore these patterns by following a three-step method consisting in i) time series decomposition into general trend, seasonality and noise, ii) modelling the general trend , iii) analysing the factors that influence the values of the parameters estimated in ii) to characterise the general trend.
 
-<!---![Evolution of the parameters characterising the relationship between relative deprivation index and the percentage change in the number of movements relative to the pre-pandemic baseline level of mobility.](figures/regression-evo-nobackground.pdf)!--->
+![](figures/Fig3.png)
 
-We find patterns in the evolution of the estimated parameters that characterise the relationship between the levels of urban mobility and RDI.
-In Argentina, Colombia and Mexico, we observe that the slope of this relationship evolves to become smaller over time.
-The tendency is not apparent in Chile, where the slope of the relationship remains approximately the same over time despite the temporary fluctuations.
-The slope captures the extent of differences in the level of urban mobility across administrative units with varying levels of socioeconomic deprivation.
-It can therefore be regarded as a measure of inequality in mobility patterns across socioeconomic groups.
-A slope equal to zero would mean that all administrative units display the same level of mobility regardless of their average RDI.
-Given the patterns observed in Argentina, Colombia and Mexico, we find that at the beginning of the pandemic there were notable inequalities between socioeconomic groups in terms of the levels of urban mobility.
-While it has taken more than two years for Argentina and Mexico to close the gap (their slope is close to zero from spring 2022), inequalities persist in Chile and Colombia as of March 2023.
-
-The intercept of the relationship displays stronger patterns, which are consistent across the four countries.
-The intercept estimates the urban mobility levels that would be observed in administrative units where the RDI is zero.
-The intercept was below the baseline level at the early stages of the pandemic.
-As observed in Figure 3, while there are some differences between countries in the evolution of the intercept, the general tendency is for the intercept to increase.
-While Argentina and Mexico reach values that are closer to the baseline towards the end of the forecast period, the intercept for Chile and Colombia remains lower.
-Therefore, if there were areas with no socioeconomic deprivation, we would have seen a recovery in the levels of mobility, especially in Argentina and Mexico
+![](figures/Fig4.png)
 

### 2024-10-28 code/analysis/06-lme-with-R.qmd
models-trend

@@
 # Set the country
-country <- 'Mexico'
+country <- 'Argentina'
 
@@
 
+###
+
 eq2 <- y ~ time + O + D
@@
 summary(m2)
+
+eq2o <- y ~ time + O
+m2o <- glmmTMB(eq2o, data = df)
+summary(m2o)
+
+###
 
@@
 
+eq3o <- y ~ time + (1|O)
+m3o <- glmmTMB(eq3o, data = df)
+m3o_re <- plot_model(m3o, type ="re")
+summary(m3o)
+
+###
+
 df$Of <- factor(df$O)
 df$Df <- factor(df$D)
-eq4 <- y ~ time + (time|Of) + (time|Df)
+eq4 <- y ~ time + (0 + time|Of) + (0 + time|Df)
 m4 <- glmmTMB(eq4, data = df)
@@
 
-eq5 <- y ~ time + time2 + (time|O) + (time|D)
+df$Of <- factor(df$O)
+eq4o <- y ~ time + (0 + time|Of) 
+m4o <- glmmTMB(eq4o, data = df)
+m4o_re <- plot_model(m4o, type ="re")
+summary(m4o)
+
+###
+
+df$Of <- factor(df$O)
+df$Df <- factor(df$D)
+eq5 <- y ~ time + (time|Of) + (time|Df)
 m5 <- glmmTMB(eq5, data = df)
@@
 summary(m5)
+
+df$Of <- factor(df$O)
+eq5o <- y ~ time + (time|Of) 
+m5o <- glmmTMB(eq5o, data = df)
+m5o_re <- plot_model(m5o, type ="re")
+summary(m5o)
+
+###
+
+eq6 <- y ~ time + time2 + (time|O) + (time|D)
+m6 <- glmmTMB(eq6, data = df)
+m6_re <- plot_model(m6, type ="re")
+summary(m6)
 
@@
 ```{r}
-m4_re[[1]]$data
-m4_re[[2]]$data
-m5_re[[2]]$data
-m5_re[[2]]$data
-m4_re
-m5_re
+m3o$fit$par
+m4o$fit$par
+m5o$fit$par
+m3o_re
+m4o_re
+m5o_re
 ```
@@
 ```{r}
-p1_m1me <- plot_model(m, terms = c("1", "time")) %>% 
+
+file_m3o <- paste0(wd, '/data/outputs/', country_short, '/mov-analysis/by-', category, '/data_trend_param_m3o.csv')
+write.csv(m3o$fit$par, file_m3o, row.names = FALSE)
+
+file_m3o_re <- paste0(wd, '/data/outputs/', country_short, '/mov-analysis/by-', category, '/data_trend_param_m3o_re.csv')
+write.csv(m3o_re[[1]], file_m3o_re, row.names = FALSE)
+
+###
+
+file_m4o <- paste0(wd, '/data/outputs/', country_short, '/mov-analysis/by-', category, '/data_trend_param_m4o.csv')
+write.csv(m4o$fit$par, file_m4o, row.names = FALSE)
+
+file_m4o_re <- paste0(wd, '/data/outputs/', country_short, '/mov-analysis/by-', category, '/data_trend_param_m4o_re.csv')
+write.csv(m4o_re[[1]], file_m4o_re, row.names = FALSE)
+
+###
+
+file_m5o <- paste0(wd, '/data/outputs/', country_short, '/mov-analysis/by-', category, '/data_trend_param_m5o.csv')
+write.csv(m5o$fit$par, file_m5o, row.names = FALSE)
+
+file_m5o_re <- paste0(wd, '/data/outputs/', country_short, '/mov-analysis/by-', category, '/data_trend_param_m5o_re.csv')
+write.csv(m5o_re[[1]], file_m5o_re, row.names = FALSE)
+
+```
+
+```{r}
+p1_m1me <- plot_model(m3o_re, terms = c("1", "time")) %>% 
   .$data %>% ggplot(aes(x=estimate, y=term, color=group)) +
@@
 ```
+`
 
 ```{r}
-summary(model)
-```
-
-```{r}
-p1_m <- plot_model(model, terms = c("time", "O", "D")) %>% 
+pm3o <- plot_model(m3o, terms = c("time", "O")) %>% 
   .$df %>% ggplot(aes(x=estimate, y=term, color=group)) +
@@
 ```
-
-```{r}
-coef_model <- coef(model)
-head(coef_model$O)
-```
-
-```{r}
-ranef_model <- ranef(model)
-head(ranef_model$O)
-```
-
-```{r}
-# obtain estimates
-merTools::REsim(model) %>% head(10)
-```
-
-```{r}
-plotREsim(REsim(model)) 
-```
-
-```{r}
-## regression table
-tab_model(mode, 
-          collapse.ci = TRUE,
-          p.style = "stars",
-          show.aic = TRUE,
-          #terms = c(), # show selected coefficients
-          #rm.terms = c("time [1st degree]"), #remove selected coefficients
-          #pred.labels = c("Intercept", "Stringency t", "New Cases t", "Stringency t-1",
-          #                "Spline 1st degree", "Spline 2nd degree", 
-          #                "Spline 3rd degree", "New cases t-1"),
-          #dv.labels = c("Autoregressive", "Stringency random effects", "New COVID-19 cases random effects"),
-          #file = "../outputs/modelling/estimates/reg_table.txt"
-)
-```

### 2024-10-29 manuscript/manuscript-nature/manuscript-nature.qmd
headlines-and-writeup-results-part1

@@
 We use five population density categories and three socioeconomic categories defined according to the relative deprivation index.
-The geographic distribution of categories for each of the countries is displayed in Figures @Fig1-ARG a), @Fig1-CHL a) and @Fig1-COL a) for Argentina, Chile and Colombia respectively.
-
-A key challenge in the in the analysis of population counts and movements with Facebook user data is the absence of small count records, which is a result of privacy-protection techniques applied to ensure that the location of individuals or small groups cannot be identified.
-As a consequence, the countries of the analysis have locations with null values for Facebook population counts and flows.
-These missing data are not distributed randomly.
-For example, the missing values for FB population counts display high spatial autocorrelation as shown in Supplementary Figure SF1.
-Therefore, simply removing the missing records from the analysis could lead to geographically biased results (Afghari et al., 2019).
-To address this, we designed a data processing method to estimate the missing values, both for Facebook population counts and flows.
-Furthermore, this method also applies a correction factor to eliminate the fluctuations in the daily number of observations, assuming that the representativeness of Facebook data across spatial units remains stable during the study period.
-The data processing method is described in section X.
-
-Even after applying the imputation method to estimate missing values, we acknowledge that Facebook data may still overrepresent certain population groups while underrepresenting others.
-For example, Figure @Fig1 shows, for each country, the proportion of the population across the various analysis categories.
-These comparisons use WorldPop population estimates and Facebook population counts, with the latter reflecting active users on an average day during a pre-pandemic baseline period (see Section X of the methodology for details).
-In all three countries, notable discrepancies appear between the population distributions according to WorldPop and Facebook data.
-For instance, in Argentina and Chile, WorldPop data indicates a higher proportion of people living in low-density areas, suggesting that Facebook data underrepresents populations in these regions.
-Similarly, in these countries, Facebook data shows an overrepresentation of the most affluent socioeconomic group.
-While addressing this kind of representativity bias is beyond the scope of this paper, we recognise its significance and the importance of addressing it in future analyses.
-This issue is revisited in the Discussion section.
-
-![Classification of spatial units into categories by population density and by relative deprivation index.
-Spatial distribution of categories and population share in each category, by country.](figures/Fig1.png){#Fig1}
-
-## COVID-19’s impact on human mobility displays consistently uneven patterns across population groups in Latin American countries
 
 The analysis of mobility is done using Facebook movement data that has been pre-processed according to our methodology.
-Figure @Fig2 shows changes in the intensity of movement measured as the percentage change in the number of inflows relative to pre-pandemic baseline levels, at two points during the period of analysis, May 2020 and March 2022.
+
+## Affluent and densely populated areas saw steepest decline in daily mobility, but all population groups progress towards baseline levels
+
+We first focus the analysis on the variability in COVID-19's impact on daily mobility levels across different population groups.
+Figure @Fig-impact shows changes in the intensity of movement measured as the percentage change in the number of inflows relative to pre-pandemic baseline levels, at two key points during the period of analysis, May 2020 and March 2022.
+These two months are the most widely spaced months with complete data available for all countries within the period covered by the dataset.
+May 2020 represents the initial phase following the WHO's declaration of COVID-19 as a global pandemic on March 11, 2020, marked by stringent measures.
+March 2022 represents a later phase, roughly six months after most restrictions had been lifted in the countries included in the analysis.
+Figure @Fig-impact presents boxplots depicting the distribution of percentage change in movement counts across areas grouped by population density and deprivation index.
+The boxplots show mobility changes for each category in May 2020 and March 2022, with baseline levels marked by the dotted line at $y = 0$.
+Values above this line indicate increased mobility from pre-pandemic levels, while values below show a reduction.
 These changes show remarkably consistent patterns across countries for both types of classification.
 
-Focusing first on the early pandemic days, we observe a decrease in the number of inflows for all countries and most categories of analysis.
-This decrease displays a systematic gradient across categories of analysis whereby the more densely populated or the more affluent an area is, the higher the losses in the number of inflows.
-Differences in the levels of mobility across population density or RDI categories tend to diminish with time, with a less pronounced gradient across categories of analysis approximately two years later.
+Focusing first on the patterns by population density category, we generally observe a decrease in the number of inflows for all countries and most categories of analysis in the early pandemic days.
+The extent of this decrease tends to be larger for high-density areas.
+Although the majority of areas saw a decline in mobility during May 2020, some locations actually recorded higher movement levels compared to the pre-pandemic period, particularly in regions classified as low-density.
+In fact, some of these areas saw their mobility levels double, as indicated by outliers with percentage changes exceeding 100%.
+However, this does not imply that these areas correspond to large population numbers, as low-density regions typically contain very few residents, meaning that minor fluctuations in population can lead to significant percentage changes.
+Nonetheless, these percentage changes illustrate the extent of the impacts experienced by these sparsely populated regions.
+In March 2022, nearly two years later, mobility levels have generally rebounded closer to pre-pandemic baseline levels across different population density categories.
+Notably, mobility levels in areas characterised by higher population density still remain below the baseline.
+Low-density areas show even greater variability than during the early days of the pandemic, with mobility levels often exceeding the baseline.
+These high variability in sparsely populated regions may reflect the ongoing effects of the pandemic on mobility, but they could also be influenced by seasonal variation (e.g.
+March compared to May).
+Overall, the observed patterns by population density category are consistent with those reported in a previous work CITE, where the same data was analysed without applying the data processing methodology proposed in this study.
+
+Moving to the patterns by relative deprivation index category, we observe tendencies that mirror the patterns by population density category.
+Firstly, in March 2020, there is an overall decrease in the number of inflows for all countries and most categories of analysis.
+Exceptions to this general tendency notably occur in the most deprived (less affluent) areas, especially in Colombia, where mobility levels in some regions are actually above the baseline.
+This decrease displays a gradient across categories of analysis whereby the more affluent an area is, the higher the losses in the number of inflows.
+After almost two years, we find evidence for a general recovery of mobility closer to baseline levels across groups regardless of how deprived they are.
+However, disparities between groups are still visible, indicating that the pandemic has exacerbated behavioral differences among groups with varying socioeconomic statuses.
 
 These patterns can be attributed to...
-remote work, movement to rural areas, affluent people being more able to abide to lockdowns, etc...
-and the patterns are consistent with what has been found in ...
+THIS GOES TO THE DISCUSSION
 
-![Changes in the intensity of inflows as the percentage of change with respect to the pre-pandemic baseline levels, for each country and category of analysis.](figures/Fig2.png)
+![Changes in the intensity of inflows as the percentage of change with respect to the pre-pandemic baseline levels, for each country and category of analysis.](figures/Fig-impact.png)
 
-## Recovery trajectories towards baseline mobility levels are heterogeneous across population categories in Latin America
+## Initial drop in mobility drives current levels, as recovery pace remains consistent across population groups
+
+Next we focus on
 
@@
 The extent of the initial drop and the speed of recovery varied by country and by analysis category.
-These general patterns can be visually explored in Figure @Fig3 and @Fig4 for a classification by population density category and by RDI respectively.
+These general patterns can be visually explored in Figure @Fig-recovery-density and @Fig-recovery-rdi for a classification by population density category and by RDI respectively.
 Below we further explore these patterns by following a three-step method consisting in i) time series decomposition into general trend, seasonality and noise, ii) modelling the general trend , iii) analysing the factors that influence the values of the parameters estimated in ii) to characterise the general trend.
 
-![](figures/Fig3.png)
+![](figures/Fig-recovery-density.png)
 
-![](figures/Fig4.png)
+![](figures/Fig-recovery-rdi.png)
+![](figures/Fig-re-density.pdf)
+
+![](figures/Fig-re-rdi.pdf)
 
@@
 
-# Data and methods {.unnumbered}
+# Data {.unnumbered}
 
@@
 
+# Methods
+
 ## Classification of tiles according to level of urbanisation and socioeconomic deprivation
@@
 <!-- Decide whether I'm going for natural breaks or my own splitting. The disadvantage of own splitting is that there is no common logic for all four countries !-->
+
+The geographic distribution of categories for each of the countries is displayed in Figures @Fig1-ARG a), @Fig1-CHL a) and @Fig1-COL a) for Argentina, Chile and Colombia respectively.
+
+A key challenge in the in the analysis of population counts and movements with Facebook user data is the absence of small count records, which is a result of privacy-protection techniques applied to ensure that the location of individuals or small groups cannot be identified.
+As a consequence, the countries of the analysis have locations with null values for Facebook population counts and flows.
+These missing data are not distributed randomly.
+For example, the missing values for FB population counts display high spatial autocorrelation as shown in Supplementary Figure SF1.
+Therefore, simply removing the missing records from the analysis could lead to geographically biased results (Afghari et al., 2019).
+To address this, we designed a data processing method to estimate the missing values, both for Facebook population counts and flows.
+Furthermore, this method also applies a correction factor to eliminate the fluctuations in the daily number of observations, assuming that the representativeness of Facebook data across spatial units remains stable during the study period.
+The data processing method is described in section X.
+
+Even after applying the imputation method to estimate missing values, we acknowledge that Facebook data may still overrepresent certain population groups while underrepresenting others.
+For example, Figure @Fig1 shows, for each country, the proportion of the population across the various analysis categories.
+These comparisons use WorldPop population estimates and Facebook population counts, with the latter reflecting active users on an average day during a pre-pandemic baseline period (see Section X of the methodology for details).
+In all three countries, notable discrepancies appear between the population distributions according to WorldPop and Facebook data.
+For instance, in Argentina and Chile, WorldPop data indicates a higher proportion of people living in low-density areas, suggesting that Facebook data underrepresents populations in these regions.
+Similarly, in these countries, Facebook data shows an overrepresentation of the most affluent socioeconomic group.
+While addressing this kind of representativity bias is beyond the scope of this paper, we recognise its significance and the importance of addressing it in future analyses.
+This issue is revisited in the Discussion section.
+
+![Classification of spatial units into categories by population density and by relative deprivation index.
+Spatial distribution of categories and population share in each category, by country.](figures/Fig-classification.png){#Fig1}
 

### 2024-10-30 manuscript/manuscript-nature/manuscript-nature.qmd
res-part-ii

@@
 
-We first focus the analysis on the variability in COVID-19's impact on daily mobility levels across different population groups.
+We first focus on analysing the variability in COVID-19's impact on daily mobility levels across different population groups.
 Figure @Fig-impact shows changes in the intensity of movement measured as the percentage change in the number of inflows relative to pre-pandemic baseline levels, at two key points during the period of analysis, May 2020 and March 2022.
@@
 March compared to May).
-Overall, the observed patterns by population density category are consistent with those reported in a previous work CITE, where the same data was analysed without applying the data processing methodology proposed in this study.
+Overall, the observed patterns by population density category are consistent with those reported in CITE for Argentina and Chile, where the data from the same source corresponding to those two countries was analysed without applying the data processing methodology proposed in this study.
 
@@
 
-## Initial drop in mobility drives current levels, as recovery pace remains consistent across population groups
+## Initial drop in mobility drives intergroup variation in mobility levels, as recovery pace is uniform across population groups
 
-Next we focus on
+Next we focus on modelling intergroup differences in mobility trends over time towards baseline levels of mobility, highlighting the variability in the pace of recovery toward baseline mobility levels across population groups.
+Figure @Fig-recovery-density and @Fig-recovery-rdi show, for each country, the evolution over time in the intensity of movement, grouped into population density and relative deprivation index categories respectively.
+Time is represented on the $x$-axis, spanning from mid-April 2020 to mid-April 2022, while the intensity of movement is depicted on the $y$-axis, measured as the percentage change in the number of flows relative to pre-pandemic baseline levels.
+The background of each plot in Figures @Fig-recovery-density and @Fig-recovery-rdi is colored according to the stringency index, which quantifies the level of non-pharmaceutical interventions implemented at each point in time to mitigate the spread of COVID-19, including measures such as social distancing and lockdowns.
+This data provides context for the observed trends, reflecting the mobility restrictions imposed by each country during the specified periods.
 
-In this section we explore trends in the intensity of movements over time across population groups.
-Generally, all categories of analysis experienced an initial drop in the number of movements followed by an upwards general trend of recovery towards baseline levels of mobility.
-The extent of the initial drop and the speed of recovery varied by country and by analysis category.
-These general patterns can be visually explored in Figure @Fig-recovery-density and @Fig-recovery-rdi for a classification by population density category and by RDI respectively.
-Below we further explore these patterns by following a three-step method consisting in i) time series decomposition into general trend, seasonality and noise, ii) modelling the general trend , iii) analysing the factors that influence the values of the parameters estimated in ii) to characterise the general trend.
+A visual inspection of the evolution patterns by population density category, as shown in @Fig-recovery-density, reveals two main commonalities across the three countries.
+First, more densely populated areas consistently record lower levels of mobility throughout the study period.
+Second, there is a general upward trend in mobility across all population density categories for all countries.
+However, variations exist in the patterns for each country.
+While Chile experiences greater overall losses in mobility compared to Argentina, certain areas in Argentina reach levels comparable to the pre-pandemic baseline.
+In contrast, Colombia displays more varied trends across population density categories; low-density areas exhibit above-baseline levels of mobility, whereas high-density areas experience significantly larger losses.
 
@@
 
+Based on a visual inspection of @Fig-recovery-rdi, the evolution patterns by relative deprivation index category mirror those observed by population density category.
+First, less deprived areas consistently record lower levels of mobility throughout the study period.
+In contrast, more deprived areas exhibit higher levels of mobility, particularly in Colombia, where these levels are well above the baseline.
+However, these significant relative changes do not necessarily correspond to large population numbers.
+In Colombia, for example, more deprived regions are often low-density areas that typically contain small populations.
+As a result, minor fluctuations in mobility can lead to substantial percentage changes.
+Nevertheless, the positive percentage changes observed in more deprived areas highlight the variability of the impact across different population groups based on their socioeconomic status.
+
 ![](figures/Fig-recovery-rdi.png)
+Next, we quantitatively model the patterns observed in @Fig-recovery-density and @Fig-recovery-rdi
+
 ![](figures/Fig-re-density.pdf)

### 2024-10-30 manuscript/manuscript-nature/manuscript-nature.qmd
res-part-ii

@@
 We focus on capturing how mobility patterns for different population groups have been impacted by COVID-19 in Latin America.
-We consider populations groups by classifying the spatial units of analysis into categories according to their population density and their socioeconomic level.
+We consider populations groups by classifying the spatial units of analysis into categories according to their population density and their relative level of socioeconomic deprivation.
 More details for the classification method are provided in section X.
 The two criteria for classification are chosen due to their relevance for ...
-We use five population density categories and three socioeconomic categories defined according to the relative deprivation index.
-
 The analysis of mobility is done using Facebook movement data that has been pre-processed according to our methodology.
@@
 We first focus on analysing the variability in COVID-19's impact on daily mobility levels across different population groups.
-Figure @Fig-impact shows changes in the intensity of movement measured as the percentage change in the number of inflows relative to pre-pandemic baseline levels, at two key points during the period of analysis, May 2020 and March 2022.
+@fig-impact shows changes in the intensity of movement measured as the percentage change in the number of inflows relative to pre-pandemic baseline levels, at two key points during the period of analysis, May 2020 and March 2022.
 These two months are the most widely spaced months with complete data available for all countries within the period covered by the dataset.
@@
 March 2022 represents a later phase, roughly six months after most restrictions had been lifted in the countries included in the analysis.
-Figure @Fig-impact presents boxplots depicting the distribution of percentage change in movement counts across areas grouped by population density and deprivation index.
+@fig-impact presents boxplots depicting the distribution of percentage change in movement counts across areas grouped by population density and deprivation index.
 The boxplots show mobility changes for each category in May 2020 and March 2022, with baseline levels marked by the dotted line at $y = 0$.
@@
 Low-density areas show even greater variability than during the early days of the pandemic, with mobility levels often exceeding the baseline.
-These high variability in sparsely populated regions may reflect the ongoing effects of the pandemic on mobility, but they could also be influenced by seasonal variation (e.g.
-March compared to May).
+These high variability in sparsely populated regions may reflect the ongoing effects of the pandemic on mobility, but they could also be influenced by seasonal variation (e.g. March compared to May).
 Overall, the observed patterns by population density category are consistent with those reported in CITE for Argentina and Chile, where the data from the same source corresponding to those two countries was analysed without applying the data processing methodology proposed in this study.
@@
 
-![Changes in the intensity of inflows as the percentage of change with respect to the pre-pandemic baseline levels, for each country and category of analysis.](figures/Fig-impact.png)
+![Changes in the intensity of flows as the percentage of change with respect to the pre-pandemic baseline levels, for each country and category of analysis at the origin location of each flow.](figures/Fig-impact.pdf){#fig-impact}
 
@@
 
-Next we focus on modelling intergroup differences in mobility trends over time towards baseline levels of mobility, highlighting the variability in the pace of recovery toward baseline mobility levels across population groups.
-Figure @Fig-recovery-density and @Fig-recovery-rdi show, for each country, the evolution over time in the intensity of movement, grouped into population density and relative deprivation index categories respectively.
+Next we focus on modelling intergroup differences in mobility trends over time, highlighting the variability in the pace of recovery toward baseline mobility levels across population groups.
+@fig-recovery-density and @fig-recovery-rdi show, for each country, the evolution over time in the intensity of movement, grouped into population density and relative deprivation index categories respectively.
 Time is represented on the $x$-axis, spanning from mid-April 2020 to mid-April 2022, while the intensity of movement is depicted on the $y$-axis, measured as the percentage change in the number of flows relative to pre-pandemic baseline levels.
-The background of each plot in Figures @Fig-recovery-density and @Fig-recovery-rdi is colored according to the stringency index, which quantifies the level of non-pharmaceutical interventions implemented at each point in time to mitigate the spread of COVID-19, including measures such as social distancing and lockdowns.
+The background of each plot in Figures @fig-recovery-density and @fig-recovery-rdi is colored according to the stringency index, which quantifies the level of non-pharmaceutical interventions implemented at each point in time to mitigate the spread of COVID-19, including measures such as social distancing and lockdowns.
 This data provides context for the observed trends, reflecting the mobility restrictions imposed by each country during the specified periods.
 
-A visual inspection of the evolution patterns by population density category, as shown in @Fig-recovery-density, reveals two main commonalities across the three countries.
-First, more densely populated areas consistently record lower levels of mobility throughout the study period.
-Second, there is a general upward trend in mobility across all population density categories for all countries.
-However, variations exist in the patterns for each country.
-While Chile experiences greater overall losses in mobility compared to Argentina, certain areas in Argentina reach levels comparable to the pre-pandemic baseline.
+A visual inspection of the evolution patterns by population density category, as shown in @fig-recovery-density, reveals two main commonalities across the three countries.
+First, there is a general upward trend in mobility across all population density categories for all countries.
+Second, more densely populated areas consistently record lower levels of mobility throughout the study period.
+However, there exist variations in the patterns by country.
+Chile experiences greater overall losses in mobility compared to Argentina, while certain areas in Argentina reach levels comparable to the pre-pandemic baseline.
 In contrast, Colombia displays more varied trends across population density categories; low-density areas exhibit above-baseline levels of mobility, whereas high-density areas experience significantly larger losses.
+Generally, the termporal patterns display fluctuations from the general trend which are likely driven by seasonality effects and by the changing levels of stringency of COVID-19 interventions.
 
-![](figures/Fig-recovery-density.png)
+![](figures/Fig-recovery-density.png){#fig-recovery-density}
 
-Based on a visual inspection of @Fig-recovery-rdi, the evolution patterns by relative deprivation index category mirror those observed by population density category.
-First, less deprived areas consistently record lower levels of mobility throughout the study period.
+The evolution patterns by relative deprivation index category mirror those observed by population density category, as it can be deduced from @Fig-recovery-rdi based on visual inspection. 
+First, we observe a general upward trend in mobility across all relative deprivation index categories for all countries.
+Second, less deprived areas consistently record lower levels of mobility throughout the study period.
 In contrast, more deprived areas exhibit higher levels of mobility, particularly in Colombia, where these levels are well above the baseline.
@@
 
-![](figures/Fig-recovery-rdi.png)
-Next, we quantitatively model the patterns observed in @Fig-recovery-density and @Fig-recovery-rdi
+![](figures/Fig-recovery-rdi.png){#fig-recovery-rdi}
 
-![](figures/Fig-re-density.pdf)
+Next, we quantitatively model the patterns observed in @fig-recovery-density and @fig-recovery-rdi
 
-![](figures/Fig-re-rdi.pdf)
+![](figures/Fig-re-density.pdf){#fig-re-density}
+
+![](figures/Fig-re-rdi.pdf){#fig-re-rdi}
+
+![](figures/Fig-re-rdi.pdf){#Fig-re-rdi}
 

### 2024-10-31 manuscript/manuscript-nature/manuscript-nature.qmd
update-manuscript

@@
 
-A visual inspection of the evolution patterns by population density category, as shown in @fig-recovery-density, reveals two main commonalities across the three countries.
-First, there is a general upward trend in mobility across all population density categories for all countries.
+We highlight four key observations from a visual inspection of the evolution patterns by population density category, as shown in @fig-recovery-density.
+First, there is a general upward trend in mobility across all population density categories for all countries, reflecting a recovery of mobility levels following the early days of COVID-19.
 Second, more densely populated areas consistently record lower levels of mobility throughout the study period.
-However, there exist variations in the patterns by country.
+Third, there exist some variations in the patterns by country.
 Chile experiences greater overall losses in mobility compared to Argentina, while certain areas in Argentina reach levels comparable to the pre-pandemic baseline.
-In contrast, Colombia displays more varied trends across population density categories; low-density areas exhibit above-baseline levels of mobility, whereas high-density areas experience significantly larger losses.
-Generally, the termporal patterns display fluctuations from the general trend which are likely driven by seasonality effects and by the changing levels of stringency of COVID-19 interventions.
+In contrast, Colombia displays more varied trends across population density categories, with above-baseline levels of mobility in low-density areas, and significantly larger losses in high-density areas.
+Fourth, the temporal patterns display fluctuations from the general trend.
+These fluctuations are likely driven by seasonality effects and by the changing levels of stringency of COVID-19 interventions.
 
@@
 
-The evolution patterns by relative deprivation index category mirror those observed by population density category, as it can be deduced from @Fig-recovery-rdi based on visual inspection. 
+Visual inspection of @Fig-recovery-rdi reveals the following four patterns by relative deprivation index category.
 First, we observe a general upward trend in mobility across all relative deprivation index categories for all countries.
 Second, less deprived areas consistently record lower levels of mobility throughout the study period.
-In contrast, more deprived areas exhibit higher levels of mobility, particularly in Colombia, where these levels are well above the baseline.
-However, these significant relative changes do not necessarily correspond to large population numbers.
-In Colombia, for example, more deprived regions are often low-density areas that typically contain small populations.
+In contrast, more deprived areas exhibit higher levels of mobility.
+Third, there is variability by country.
+Colombia stands out for recording mobility levels in the most deprived areas that are well above the baseline.
+These big relative changes for the most deprived category does not necessarily correspond to large population numbers since for example, in Colombia, more deprived regions are often low-density areas that typically contain small populations.
 As a result, minor fluctuations in mobility can lead to substantial percentage changes.
 Nevertheless, the positive percentage changes observed in more deprived areas highlight the variability of the impact across different population groups based on their socioeconomic status.
+Fourth
 
@@
 
-Next, we quantitatively model the patterns observed in @fig-recovery-density and @fig-recovery-rdi
+Next, we quantitatively model intergroup differences in mobility trends over @fig-recovery-density and @fig-recovery-rdi.
+For this purpose we apply a time series decomposition into three components: trend, seasonal and residual.
+We then focus on the trend, which we model according to five model specifications.
 

### 2024-11-27 manuscript/manuscript-nature/manuscript-nature.qmd
changes-manuscript-2711

@@
 Next, we quantitatively model intergroup differences in mobility trends over @fig-recovery-density and @fig-recovery-rdi.
-For this purpose we apply a time series decomposition into three components: trend, seasonal and residual.
-We then focus on the trend, which we model according to five model specifications.
+To achieve this, we perform a time series decomposition into three components: trend, seasonal, and residual.
+Our analysis focuses on the trend component, which we model using five specifications within a linear mixed-effects modeling framework, all yielding consistent results.
+All models include time as a fixed effect and estimate an intercept and a slope associated with this variable.
+The intercept is interpreted as the magnitude of the initial drop in mobility, while the slope represents the rate at which mobility levels recover over time.
+
+Models 1 and 2 consider only fixed effects.
+Models 3, 4, and 5 incorporate time as a fixed effect while also accounting for random effects to capture variation in the intercept, slope, or both, based on the category of the origin of the flows.
+Further details on the model specifications are provided in Methodology Section X.
+Figures @fig-re-density and @fig-re-rdi present the random variations in intercept and slope estimated by Models 3, 4, and 5, which provide insights into the drivers of intergroup variation in the recovery towards baseline mobility levels.
 
@@
 
-![](figures/Fig-re-rdi.pdf){#Fig-re-rdi}
+INTERPRET
 
@@
 
-<!-- Decide whether I'm going for natural breaks or my own splitting. The disadvantage of own splitting is that there is no common logic for all four countries !-->
-
 The geographic distribution of categories for each of the countries is displayed in Figures @Fig1-ARG a), @Fig1-CHL a) and @Fig1-COL a) for Argentina, Chile and Colombia respectively.
@@
 
-We then use the complete baseline Facebook population values to compute missing Facebook population counts during the crisis period.
+We then use the complete baseline of Facebook population values to compute missing Facebook population counts during the crisis period.
 This is possible because, as mentioned above, Meta reports the percentage change in the number of counts with respect to the baseline, even if the counts are not reported due to low value.
@@
 Mathematically, this model can be expressed as \begin{equation}
-\mu_{ijw} = \beta_0 + \beta_1pop_{iw} +\beta_2pop_{jw} + \beta_3d_{ij} + \beta_4W + \beta_5C + \varepsilon
+\mu_{ijw} = \beta_0 + \beta_1pop_{iw} +\beta_2pop_{jw} + \beta_3d_{ij} + \beta_4W + \varepsilon
 \end{equation}
 
-where $\mu_{ijw}$ is the expectation of the flow of people from tile $i$ to tile $j$ on the weekday $w$; $\beta_0$ is an intercept $pop_{iw}$ and $pop_{jw}$ are the Facebook population counts at the origin and destination on weekday $w$ during the baseline period, $d_{ij}$ is the distance between origin and destination, $W$ is a series of indicator variables capturing the day of the week and similarly $C$ is a series of indicator variables reflecting the pair of population density classes for the origin and destination tiles, resulting in X pairs (10 classes × 10 classes minus one so it is not collinear with $\beta_0$), $\beta_{0,1,2,3,4,5}$ are model parameters to be estimated from the observed data.
+where $\mu_{ijw}$ is the expectation of the flow of people from tile $i$ to tile $j$ on the weekday $w$; $\beta_0$ is an intercept $pop_{iw}$ and $pop_{jw}$ are the Facebook population counts at the origin and destination on weekday $w$ during the baseline period, $d_{ij}$ is the distance between origin and destination, $W$ is a series of indicator variables capturing the day of the week, $\beta_{0,1,2,3,4}$ are model parameters to be estimated from the observed data.
 The error term is denoted by \varepsilon.
-
-To estimate the model parameters, we used a count data regression model.
-Specifically, we used a negative binomial regression which is a generalised linear model (GLM) where overdispersion of the error term is assumed, i.e. the variance exceeds the mean.
+To estimate the model parameters, we used a Gaussian regression model, taking the log of the population at origin and at destination, and the log of the distance.
+Residual plots for the spatial interaction model are provided in the Supplementary Figure \ref{fig:residuals-sim}.
 
@@
 
-### Correction factors??? For representativeness
+### Correction factors
 
-## Data analysis
+The total number of active users within a given time window exhibits daily fluctuations due to limitations in Internet connectivity and user data access options [@Maas19].
+Following [@yabe2020] and [@duan2024], we applied a correction factor to account for fluctuations in the daily number of observed users.
+This approach assumes that the representativeness of Facebook data across spatial units remains consistent throughout the study period.
 
-```{=html}
-<!-- EXPLAIN - might include the things I had already written below
+This was done to mitigate the potential impact of variations in the total number of collected users on the results, which could mask the mobility trends within population density and relative deprivation categories.
+The adjusted number of movements between tile $i$ and tile $j$ on day $d$, $y'_{ijd}$ was obtained as:
 
-## Spatiotemporal data aggregation
+$$
+y'_{ijd} = k_d \times y_{ijd}
+$$
 
-Since the focus of this work is on urban mobility, we focus the analysis on Functional Urban Areas (FUAs), defined by the Global Human Settlement Layer (GHSL).
-The spatial extent of the FUAs is often large and may include several towns and neighbourhoods displaying a variety of socioeconomic characteristics, hence using FUAs as the spatial units of aggregation would considerably mask the heterogeneity in the mobility patterns.
-While the original unit of aggregation for the movement data, i.e.
-the tiles from the Bing Maps Tile System, offers the highest degree of spatial granularity available, the interpretation of findings from an analysis based on administrative units is often more valuable.
-For this reason, we perform a spatial join to aggregate the movement data into administrative units at the GADM level 2 or 3.
-Only flows of people starting or ending within the boundaries of a FUA are considered.
+where $y_{ijd}$ is the original number of movements between tiels $i$ and $j$ on day $d$ and $k_d$ is a correction factor obtained as the median of the sum of user counts on day $d$ divided by the sum of user counts across all spatial units on day $d$.
+Mathematically,
 
-While the original movement data is originally aggregated into 8-hour windows, this resolution is too fine for our analysis.
-Since the analysis is focused on the longer-term evolution of patterns of urban mobility, we aggregate the movement data by month.
+$$k_d = \dfrac{\mbox{Median}\big\{\sum_i{P_{id}}\big\}_d}{\sum_i{P_{id}}}$$.
 
-In our analysis, we use the Relative Deprivation Index (RDI) as a measure of socioeocnomic deprivation.
-The RDI data is made available via NASA's Socioeconomic Data and Applications Centre (SEDAC), with a spatial resolution of 1km pixels.
-We perform a spatial join of the gridded data and the administrative units and compute the average RDI within each of these administrative units.
+### Time series smoothing
 
-## Time series analysis
+A time series for each origin-destination tile pair was generated using data processed as described in Sections X to Y.
+However, these time series still contained missing values due to days when no data was reported for specific location pairs.
+To ensure continuity, we addressed this issue by imputing missing values, replacing them with the average of the nearest 15 observations within the time series.
+This number was chosen to provide an optimal balance between maintaining temporal proximity and ensuring sufficient data coverage.
 
-For the countries included in the analysis, the movement data is available until March 2022.
-In order to gain a better understanding of the recovery trends after the pandemic, we generate a 12-month forecast up to March 2023.
-This is done using Prophet [@prophet17], a procedure developed by Meta's Core Data Science team to forecast time series data based on an additive model.
-Prophet stands out from other forecasting methods due to its ability to capture recurring seasonal trends inherent in mobility data, such as fluctuations during holidays, or changes in the general trend due to the impact of particular circumstances, such as the COVID-19 pandemic.
-It is particularly intuitive to use compared to other models due to its automated trend-detection capabilities, making it accessible to users with no expertise on the underlying model, while being robust to missing data and outliers.
-For the analysis in this paper, yearly seasonality effects are included as they yield the most realistic predictions.
+Additionally, to reduce noise and highlight underlying trends that might be obscured by short-term fluctuations, we applied a rolling-average smoothing technique to the resulting time series.
+This approach was used in the time series presented in Figures X, Y, Z.
 
-Figures 2 and 3 have been created by removing statistical outliers.
-We define these as the administrative units which, at any given month, have a z-score greater than 4 for the percentage change in the number of movements.
-Furthermore, a Savitzky-Golay filter is applied to smooth the time series data, using a length of 4 units for the filter window and order 2 polynomials to fit the samples.
+## Trend analysis
 
-!-->
-```
+Model 1 includes time as the sole explanatory variable, while Model 2 incorporates spatial heterogeneity by adding an indicator variable for the population density or RDI category of the origin.
+Models 3, 4, and 5 include time as a fixed effect but also incorporate random effects.
+Model 3 accounts for a random intercept based on the origin category, Model 4 includes a random slope for the origin category, and Model 5 combines both a random intercept and a random slope by origin category.
+Figures @fig-re-density and @fig-re-rdi illustrate the random variation in intercept and slope estimated by Models 3, 4, and 5.
+Full details on the parameter estimates are provided in Supplementary Tables SX for each country.
+
 # Code availability {.unnumbered}

### 2024-11-27 manuscript/manuscript-nature/manuscript-nature.qmd
manuscript-and-fig-cities

@@
 
-# Main {.unnumbered}
+# Main {#sec-introduction}
 
-<!--# CCA: Fr to rewrite section -->
+*`@Francisco to rewrite this section`*
 
@@
 
-# Results
-
-## Analysing the variability in the impact of COVID-19 on mobility across population groups
+# Results {#sec-results}
 
@@
 
-## Affluent and densely populated areas saw steepest decline in daily mobility, but all population groups progress towards baseline levels
+## Affluent and densely populated areas saw steepest decline in daily mobility, but all population groups progress towards baseline levels {#sec-resultsevooverview}
 
@@
 These high variability in sparsely populated regions may reflect the ongoing effects of the pandemic on mobility, but they could also be influenced by seasonal variation (e.g. March compared to May).
-Overall, the observed patterns by population density category are consistent with those reported in CITE for Argentina and Chile, where the data from the same source corresponding to those two countries was analysed without applying the data processing methodology proposed in this study.
+Overall, the observed patterns by population density category are consistent with those reported in [@RoweCCA24] for Argentina and Chile, where data from the same source corresponding to those two countries was analysed without applying the data processing methodology proposed in this study.
 
@@
 
-These patterns can be attributed to...
-THIS GOES TO THE DISCUSSION
+*`@Miguel, can you elaborate the following in the discussion? - The pattern that we observe in this section can be attributed to... THIS GOES TO THE DISCUSSION`*
 
@@
 
-## Initial drop in mobility drives intergroup variation in mobility levels, as recovery pace is uniform across population groups
+## Initial drop in mobility drives intergroup variation in mobility levels, as recovery pace is uniform across population groups {#sec-resultsintergrouprecovery}
 
 Next we focus on modelling intergroup differences in mobility trends over time, highlighting the variability in the pace of recovery toward baseline mobility levels across population groups.
-@fig-recovery-density and @fig-recovery-rdi show, for each country, the evolution over time in the intensity of movement, grouped into population density and relative deprivation index categories respectively.
+@fig-recoverydensity and @fig-recoveryrdi show, for each country, the evolution over time in the intensity of movement, grouped into population density and relative deprivation index categories respectively.
 Time is represented on the $x$-axis, spanning from mid-April 2020 to mid-April 2022, while the intensity of movement is depicted on the $y$-axis, measured as the percentage change in the number of flows relative to pre-pandemic baseline levels.
-The background of each plot in Figures @fig-recovery-density and @fig-recovery-rdi is colored according to the stringency index, which quantifies the level of non-pharmaceutical interventions implemented at each point in time to mitigate the spread of COVID-19, including measures such as social distancing and lockdowns.
+The background of each plot in Figures @fig-recoverydensity and @fig-recoveryrdi is colored according to the stringency index, which quantifies the level of non-pharmaceutical interventions implemented at each point in time to mitigate the spread of COVID-19, including measures such as social distancing and lockdowns.
 This data provides context for the observed trends, reflecting the mobility restrictions imposed by each country during the specified periods.
 
-We highlight four key observations from a visual inspection of the evolution patterns by population density category, as shown in @fig-recovery-density.
+*`@Francisco and @Miguel to check that what I wrote below makes sense`*
+
+We highlight four key observations from a visual inspection of the evolution patterns by population density category, as shown in @fig-recoverydensity.
 First, there is a general upward trend in mobility across all population density categories for all countries, reflecting a recovery of mobility levels following the early days of COVID-19.
@@
 
-![](figures/Fig-recovery-density.png){#fig-recovery-density}
+![Figures](figures/Fig-recovery-density.png){#fig-recoverydensity}
 
-Visual inspection of @Fig-recovery-rdi reveals the following four patterns by relative deprivation index category.
+Visual inspection of @fig-recoveryrdi reveals the following four patterns by relative deprivation index category.
 First, we observe a general upward trend in mobility across all relative deprivation index categories for all countries.
@@
 Colombia stands out for recording mobility levels in the most deprived areas that are well above the baseline.
-These big relative changes for the most deprived category does not necessarily correspond to large population numbers since for example, in Colombia, more deprived regions are often low-density areas that typically contain small populations.
+These big relative changes for the most deprived category do not necessarily correspond to large population numbers since for example, more deprived regions in Colombia are often low-density areas that typically contain small populations.
 As a result, minor fluctuations in mobility can lead to substantial percentage changes.
-Nevertheless, the positive percentage changes observed in more deprived areas highlight the variability of the impact across different population groups based on their socioeconomic status.
-Fourth
+Nevertheless, the positive percentage change observed in more deprived areas highlight the variability of the impact across different population groups based on their socioeconomic status.
+Fourth, like in @fig-recoverydensity, the temporal patterns display fluctuations from the general trend which are likely driven by seasonality effects and by the changing levels of COVID-19 stringency index.
 
-![](figures/Fig-recovery-rdi.png){#fig-recovery-rdi}
+![Figures RDI](figures/Fig-recovery-rdi.png){#fig-recoveryrdi}
 
-Next, we quantitatively model intergroup differences in mobility trends over @fig-recovery-density and @fig-recovery-rdi.
+Next, we quantitatively model intergroup differences in mobility trends over @fig-recoverydensity and @fig-recoveryrdi.
 To achieve this, we perform a time series decomposition into three components: trend, seasonal, and residual.
 Our analysis focuses on the trend component, which we model using five specifications within a linear mixed-effects modeling framework, all yielding consistent results.
-All models include time as a fixed effect and estimate an intercept and a slope associated with this variable.
+All five models include time as a fixed effect and estimate an intercept and a slope associated with this variable.
 The intercept is interpreted as the magnitude of the initial drop in mobility, while the slope represents the rate at which mobility levels recover over time.
+Models 3, 4, and 5 incorporate random effects to capture variation in the intercept, slope, or both, based on the category of the origin of the flows.
+Detailed descriptions of all model specifications are provided in Methodology Section X.
 
-Models 1 and 2 consider only fixed effects.
-Models 3, 4, and 5 incorporate time as a fixed effect while also accounting for random effects to capture variation in the intercept, slope, or both, based on the category of the origin of the flows.
-Further details on the model specifications are provided in Methodology Section X.
-Figures @fig-re-density and @fig-re-rdi present the random variations in intercept and slope estimated by Models 3, 4, and 5, which provide insights into the drivers of intergroup variation in the recovery towards baseline mobility levels.
+Figures @fig-redensity and @fig-rerdi present, for each country, the random variations in intercept and slope estimated by Models 3, 4, and 5, providing insights into the drivers of intergroup variation in the recovery towards baseline mobility levels.
+Focusing first on @fig-redensity, we observe that
 
-![](figures/Fig-re-density.pdf){#fig-re-density}
+![Figure random effects density](figures/Fig-re-density.pdf){#fig-redensity}
 
-![](figures/Fig-re-rdi.pdf){#fig-re-rdi}
+![Figure random effects RDI](figures/Fig-re-rdi.pdf){#fig-rerdi}
 
-INTERPRET
+## Focus on cities {#sec-cities}
+
+![FUAS netflows](figures/Fig-fuas-netflows.pdf){#fig-cities}
 

### 2024-11-29 manuscript/manuscript-nature/manuscript-nature.qmd
manuscript-updates

@@
 In most countries, affluent individuals tended to record the greatest drops in mobility levels as they are predominantly employed in knowledge-intensive jobs which can be done fully or partly remotely [@fraiberger2020; @bonaccorsi2020; @weill2020; @dueñas2021; @santana2023].
-During the COVID-19 pandemic, the adoption of remote work reduced the need of commuting for knowledge-intensive, non-public facing jobs [@florida2021].At the same time, individuals from less privileged socioeconomic backgrounds displayed less pronounced declines mirroring the nature of their work requiring public-facing, face-to-face interaction, and thus a requirement for daily work commutes [@dueñas2021; @santana2023].
+During the COVID-19 pandemic, the adoption of remote work reduced the need of commuting for knowledge-intensive, non-public facing jobs [@florida2021].
+At the same time, individuals from less privileged socioeconomic backgrounds displayed less pronounced declines mirroring the nature of their work requiring public-facing, face-to-face interaction, and thus a requirement for daily work commutes [@dueñas2021; @santana2023].
 
@@
 We consider populations groups by classifying the spatial units of analysis into categories according to their population density and their relative level of socioeconomic deprivation.
-More details for the classification method are provided in section X.
-The two criteria for classification are chosen due to their relevance for ...
-The analysis of mobility is done using Facebook movement data that has been pre-processed according to our methodology.
+More details for the classification method are provided in the Methods section.
+The two criteria for classification are chosen due to their relevance for *`@Francisco, can you help here?`*.
+The analysis of mobility is done using Facebook movement data that has been processed according to our methodology.
 
@@
 Although the majority of areas saw a decline in mobility during May 2020, some locations actually recorded higher movement levels compared to the pre-pandemic period, particularly in regions classified as low-density.
-In fact, some of these areas saw their mobility levels double, as indicated by outliers with percentage changes exceeding 100%.
-However, this does not imply that these areas correspond to large population numbers, as low-density regions typically contain very few residents, meaning that minor fluctuations in population can lead to significant percentage changes.
-Nonetheless, these percentage changes illustrate the extent of the impacts experienced by these sparsely populated regions.
+In fact, some of these areas saw their mobility levels double, as indicated by outliers with percentage change exceeding 100%.
+However, this does not imply that these areas correspond to large population numbers, as low-density regions typically contain very few residents, meaning that minor fluctuations in population can lead to significant percentage change.
+Nonetheless, these relative changes illustrate the extent of the impacts experienced by these sparsely populated regions.
 In March 2022, nearly two years later, mobility levels have generally rebounded closer to pre-pandemic baseline levels across different population density categories.
@@
 
-![Figures](figures/Fig-recovery-density.png){#fig-recoverydensity}
+![Evolution of number of flows by population density category of origin for Argentina, Chile and Colombia, as percentage change with respect to pre-pandemic baseline levels.](figures/Fig-recovery-density.png){#fig-recoverydensity}
 
@@
 
-![Figures RDI](figures/Fig-recovery-rdi.png){#fig-recoveryrdi}
+![Evolution of number of flows by population density category of origin, as percentage change with respect to pre-pandemic baseline levels.](figures/Fig-recovery-density.png)
 
@@
 Figures @fig-redensity and @fig-rerdi present, for each country, the random variations in intercept and slope estimated by Models 3, 4, and 5, providing insights into the drivers of intergroup variation in the recovery towards baseline mobility levels.
-Focusing first on @fig-redensity, we observe that
+Focusing first on @fig-redensity, we observe that...
+*`@Francisco to write this interpretation`*
 
@@
 
-## Focus on cities {#sec-cities}
+## Focus on cities - *`@Francisco @Miguel, we need a better title here - any ideas?`* {#sec-cities}
 
-![FUAS netflows](figures/Fig-fuas-netflows.pdf){#fig-cities}
+Turning our focus to cities, we present a city-centric analysis of the evolution of movements.
+Specifically, we examine movements between urban centers and the rest of spatial units within each country, categorising these spatial units by population density to account for variability across the urban hierarchy.
+Urban centers are defined as the most densely populated spatial unit within each Functional Urban Area (FUA).
+The boundaries of these Functional Urban Areas are derived from the European Commission's Human Settlement Layer (GHSL) project [@ceu.jrc.2019] .
+
+For each country, we analyze inflows to all FUA centers, considering two types of movements: those originating from any spatial unit (first column of @fig-fuasflows) and those from spatial units within a 100 km radius of the FUA centers (second column of @fig-fuasflows).
+Additionally, focusing only on urban centres corresponding to the capital FUAs -Buenos Aires, Santiago and Bogotá- we also look at inflows from any spatial unit and from those situated within a 100 km radius (columns third and fourth columns in @fig-fuasflows).
+The analysis is similarly repeated for outflows from urban centers to other spatial units.
+
+*`@Miguel to write this section. What can we learn about the figure? The target journal is nature cities, so we need to talk about cities...`*
+
+![Evolution of number of inflows (and outflows) to (from) functional urban area centres, by population desnity category of origin (destiantion) for Argentina, Chile and Colombia, as percentage change with respect to pre-pandemic baseline levels.](figures/Fig-fuas-flows.png){#fig-fuasflows}
 
 # Discussion {#sec-discussion}
+
+*`@Miguel to revisit this section - the text here was written a long time ago, before the analysis had taken the direction it's taken.
+Needs to be rewritten.
+Focus on trajectories of recovery.
+Focus on differences across population groups.
+Talk about remaining un-addressed biases in the data.
+With our methodology what we achieve is 1) data imputation for missing counts in regions that have low counts due to a variety of reasons -discuss- 2) data adjustment for fluctuations in daily number of users.
+However, our method does not explicitely adjust underrepresentation of specific population groups.
+This remains as future work and we should acknowledge the importance of this to make research based on digital traces more relevant for policy.`*
 
@@
 
-## Meta-Facebook data
+## Meta-Facebook data {#sec-datameta}
 
@@
 
-## WorldPpop population data
+## WorldPpop population data {#sec-dataworldpop}
 
@@
 
-## Socioeconomic deprivation data
+## Socioeconomic deprivation data {#sec-datardi}
 
@@
 
-# Methods
+# Methods {#sec-methods}
 
-## Classification of tiles according to level of urbanisation and socioeconomic deprivation
+## Classification of tiles according to level of urbanisation and socioeconomic deprivation {#sec-methodsclassify}
 
-The geographic distribution of categories for each of the countries is displayed in Figures @Fig1-ARG a), @Fig1-CHL a) and @Fig1-COL a) for Argentina, Chile and Colombia respectively.
+The geographic distribution of categories for each of the countries is displayed in @fig-classification for Argentina, Chile and Colombia respectively.
 
@@
 Even after applying the imputation method to estimate missing values, we acknowledge that Facebook data may still overrepresent certain population groups while underrepresenting others.
-For example, Figure @Fig1 shows, for each country, the proportion of the population across the various analysis categories.
+For example, Figure @fig-classification shows, for each country, the proportion of the population across the various analysis categories.
 These comparisons use WorldPop population estimates and Facebook population counts, with the latter reflecting active users on an average day during a pre-pandemic baseline period (see Section X of the methodology for details).
@@
 ![Classification of spatial units into categories by population density and by relative deprivation index.
-Spatial distribution of categories and population share in each category, by country.](figures/Fig-classification.png){#Fig1}
+Spatial distribution of categories and population share in each category, by country.](figures/Fig-classification.png){#fig-classification}
 
-## Processing Facebook data
+## Processing Facebook data {#sec-methodsprocess}
 
@@
 
-## Trend analysis
+## Trend analysis {#sec-methodstrend}
 
@@
 
-# Code availability {.unnumbered}
+# Code availability
 
@@
 # References
-
-::: {#refs}
-:::
 

### 2024-12-01 manuscript/manuscript-nature/manuscript-nature.qmd
manuscript-work

@@
     - note: "2"
-      text: Centre for Demographic Urban and Environmental Studies, El Colegio de México, Ciudad de México, México
+      text: Centre for Demographic Urban and Environmental Studies, El Colegio de Mexico, Ciudad de México, México
     - note: "*"
@@
 
-![Evolution of number of flows by population density category of origin, as percentage change with respect to pre-pandemic baseline levels.](figures/Fig-recovery-density.png)
+![Evolution of number of flows by population density category of origin, as percentage change with respect to pre-pandemic baseline levels.](figures/Fig-recovery-density.png){#fig-recoveryrdi}
 
@@
 
-<!-- Include here population data too !-->
+### Facebook Movements
 
-To capture population movements, we used anonymised aggregate mobile phone location data from Meta users for Argentina, Colombia, Chile and Mexico, covering a 24-month period from April 2020 to March 2022.
-We used the dataset Facebook Movements created by Meta and accessed through their Data for Good Initiative (<https://dataforgood.facebook.com>).
+To capture population movements, we used anonymised aggregate mobile phone location data from Meta users for Argentina, Colombia and Chile, covering a 24-month period from April 2020 to March 2022.
+We used the Facebook Movements datasets created by Meta and accessed through their Data for Good Initiative (<https://dataforgood.facebook.com>).
 The data are built from Facebook app users who have the location services setting turned on on their smartphone.
 Prior to releasing the datasets, Meta ensures privacy and anonymity by removing personal information and applying privacy-preserving techniques [@Maas19].
-Small-count dropping is one of these techniques.
-A data entry is removed if the population or movement count for an area is lower than 10.
-The removal of small counts may mean that population counts in small sparsely populated areas are not captured.
-A second technique consists in adding a small undisclosed amount of random noise to ensure that it is not possible to ascertain precise, true counts for sparsely populated locations.
-Third, spatial smoothing using inverse distance-weighted averaging is also applied is applied to produce a smooth population count surface.
-The Facebook Movements dataset offers information on the total number of Facebook users moving between and within spatial units in the form of origin-destination matrices.
-The data is temporally aggregated into three daily 8-hour time windows (i.e. 00:00-08:00, 08:00-16:00 and 16:00-00:00).
-The dataset includes a baseline capturing the number of movements before COVID-19 based on a 45-day period ending on March 10th 2020.
-The baseline is computed using an average for the same time of the day and day of the week in the period preceding March 10th.
-For instance, the baseline for Monday 00:00-08:00 time window is obtained by averaging over data collected on Mondays from 00:00 to 8:00 for the 45-day period.
-Details about the baseline can be found in [@Maas19].
-The Bing Maps Tile System developed by Microsoft (Microsoft) is used a spatial framework to organise the data.
-The Tile System is a geospatial indexing system that partitions the world into tile cells in a hierarchical way, comprising 23 different levels of detail (Microsoft).
-At the lowest level of detail (Level 1), the world is divided into four tiles with a coarse spatial resolution.
-At each successive level, the resolution increases by a factor of two.
-The data that we used are spatially aggregated into Bing tile levels 13.
+One such technique involves adding a small, undisclosed amount of random noise to prevent the determination of precise counts for these areas.
+Another technique consists in applying spatial smoothing, using inverse distance-weighted averaging, to create a smoother data surface.
+Finally, small-count dropping is applied to exclude records where the data counts are below 10.
+
+The Facebook Movements datasets provide information on the aggregated number of Facebook app users moving between pairs of locations, during the 24-month crisis period during COVID-19.
+The data is spatially aggregated into tiles according to the Bing Maps Tile System developed by Microsoft [@microsoftBingMaps].
+The Facebook Movement data for Argentina, Chile and Colombia are spatially aggregated into Bing tile levels 13...
+respectively *`@Carmen to check Bing Tile levels for movement data`*.
 That is about 4.9 x 4.9km at the Equator [@Maas19].
+The data is temporally aggregated into three daily 8-hour time windows which are 00:00-08:00, 08:00-16:00 and 16:00-00:00, Pacific Time (PT).
+These time windows translate to 05:00-13:00, 13:00-21:00, 21:00-05:00 for Argentina and Chile; and 03:00-11:00, 11:00-19:00, 19:00-03:00 for Colombia.
+For each time window, the origin location of a user is defined as the most frequently visited location in the previous time window, while the destination location is the most frequently visited location in the current time window.
+
+In addition, each dataset includes baseline movement counts, which represent the estimated number of people moving prior to COVID-19.
+These baseline counts are calculated based on a 45-day period ending on March 10, 2020, using the average for the same time of day and day of the week within this period.
+For example, the baseline for the 00:00–08:00 (PT) window on Mondays is calculated by averaging all Monday records for that time window across the 45-day period.
+Further details about the baseline can be found in [@Maas19].
+
+Due to the small-count dropping privacy measure, movement counts during both the crisis and baseline periods are excluded if they fall below 10.
+However, even when counts are not reported, the datasets consistently provide the percentage change in movement counts relative to the baseline.
+This percentage change, denoted as $y^{\%}_{ijdt}$, is computed as follows [@Maas19]:
+
+$$
+y^{\%}_{ijdt} = \dfrac{y^{c}_{ijdt} - y^b_{ijdt}}{y^b_{ijdt} - \varepsilon}\times 100.
+$$
+
+Here, $y^c_{ijdt}$ and $y^b_{ijdt}$ represent the crisis and baseline movement counts, respectively, between origin tile $i$ to destination tile $j$ on day $d$ and during time window $t$.
+For the baseline values $y^b_{ijwt}$, the subindex $w$ denotes the day of the week corresponding to day $d$, as the baseline values are recorded by weekday, not by specific date.
+A small constant $\varepsilon$, usually set to 1, is added to the denominator to avoid division by zero [@Maas19].
+
+### Facebook Populations
+
+As part of our proposed data processing methodology, we also utilize the Facebook Population datasets.
+Like the Facebook Movements data, these datasets cover a 24-month period from April 2020 to March 2022 and provide information on the number of active Facebook users at specific locations during the crisis period.
+
+The Facebook Population data is aggregated spatially into Bing tiles and temporally into 8-hour time windows.
+A user's location is defined as the most frequently visited location within each 8-hour time window.
+Similar to the Facebook Movements datasets, the Population datasets include baseline counts, calculated by averaging values for the same day of the week and time of day over the 45-day pre-pandemic baseline period.
+
+The same privacy-preserving techniques are applied to the Facebook Population datasets.
+As a result, population counts during both the crisis and baseline periods are excluded if they fall below 10.
+However, the datasets consistently report the percentage change in active user counts relative to the baseline.
+For tile $i$, day $d$ and time window $t$, the relative change in active user counts $p^{\%}_{idt}$ is computed as:
+
+$$
+p^{\%}_{idt} = \dfrac{p^{c}_{idt} - p^b_{iwt}}{p^b_{iwt} - \varepsilon}\times 100,
+$$ where $p^c_{idt}$ and $p^b_{iwt}$ represent the crisis and baseline active user counts at tile $i$ on day $d$ and during time window $t$.
+Once again, the subindex $w$ in baseline values $p^b_{iwt}$ denotes the day of the week corresponding to day $d$, and $\varepsilon$ is a small constant usually set to 1, added to the denominator to avoid division by zero [@Maas19].
+
+In the remainder of the paper, we sometimes use expressions such as the "Facebook population" or the "population of Facebook users" to refer to the number of active Facebook app users.
+
+### Meta-Facebook data pre-processing
+
+For our analysis, we filter the data to include only one time window per day: 08:00–16:00 Pacific Time (PT), which corresponds to 13:00–21:00 in Argentina and Chile, and 11:00–19:00 in Colombia.
+This window is likely to capture the bulk of socioeconomic daytime activity, as well as a significant portion of morning commutes.
+The choice of considering only one time window per day is driven by the objective of analysing the evolution of movement patterns throughout the pandemic period, without focusing on variations through the day.
+By limiting the analysis to a single time window each day, we reduce potential noise and variability that could arise from considering multiple time windows per day.
+Furthermore, including all time windows could result in opposing movement trends that may cancel each other out when the data is aggregated by day.
+As a result of filtering the data to include only one time window per day, we can drop the subindex $t$ from the variables defined above $y^{c,b,\%}_{ijd}$ and $p^{c,b,\%}_{ijd}$, as the time window is no longer a distinguishing factor in the analysis.
+
+As a pre-processing step, we ensure that both the Facebook movement and population data are aligned in terms of spatial resolution for each country.
+Since the raw Facebook Population datasets are aggregated at a finer spatial resolution, we re-aggregate them to match the spatial resolution of the Facebook Movements data.
 
@@
 
-We used data from WorldPop [@tatem2017] to classify the spatial units of analysis according to their level of urbanisation, and to estimate missing baseline values in the Facebook population data.
+We used data from WorldPop [@tatem2017] to classify the spatial units of analysis according to their population density, and to estimate missing baseline values in the Facebook population data.
 WorldPop offers open access gridded population estimates at a resolution as small as 3 arc-seconds approximately 100m and 1km at the Equator, respectively.
@@
 
-We use the Global Gridded Relative Deprivation Index (GRDI), Version 1 (GRDIv1) data set as a measure of socioeocnomic deprivation.
+We use the Global Gridded Relative Deprivation Index (GRDI), Version 1 (GRDIv1) [@RDI2022] data set as a measure of socioeconomic deprivation.
 The GRDI data is made available via NASA's Socioeconomic Data and Applications Centre (SEDAC), at a spatial resolution of 30 arc-seconds, or 1 km$^2$ approximately.
@@
 
+`@Francisco to revise the whole section`
+
+## Processing Facebook data {#sec-methodsprocess}
+
+A significant challenge in analysing population counts and movements using Facebook user data is the absence of records for small counts.
+This limitation stems from privacy-protection measures designed to prevent the identification of individuals or small groups based on their location.
+The missing data in the Facebook Movements and Facebook Population datasets are not distributed randomly.
+Spatially, these missing values display high spatial autocorrelation as shown in Supplementary Figure [@fig-autocorrelation] for the Facebook Population datasets.
+Furthermore, due to the non-random nature of these missing values, spatial units that are sparsely populated or that have a low number of Facebook app users could potentially be underrepresented in the analysis.
+Therefore, simply removing the missing records from the analysis could lead to geographically biased results [@afghari2019].
+To address this, we designed a data processing method for missing data imputation.
+An overview of this data imputation method is provided in [@fig-dataimputation].
+
+![Workflow diagram for data imputation method.](figures/workflow-diagram.png){#fig-dataimputation}
+
+Additionally, we apply correction factors and time series smoothing to eliminate the fluctuations in the daily number of observations, assuming that the representativeness of Facebook data across spatial units remains stable during the study period.
+The steps of our data-processing workflow are further described in the next four subsections.
+
+### Imputing Facebook Population data
+
+We begin by identifying all the baseline values reported in the Facebook Population datasets, which form the basis for creating a new dataset that includes all available baseline information.
+Next, following [@duan2024], we use linear models to estimate the missing Facebook population baseline counts based on Worldpop population data.
+These models are fitted using ordinary least squares regression, with Worldpop data as the explanatory variable and the available baseline counts from the Facebook Population dataset as the dependent variable.
+The estimation process and its accuracy are illustrated in Supplementary Figure [@fig-imputbaselinepop].
+
+We then use the complete baseline of Facebook population counts to compute missing Facebook population counts during the crisis period.
+This is possible because, as mentioned above, Meta reports the percentage change in the number of counts with respect to the baseline $p^{\%}_{id}$, even if the counts are not reported due to low value.
+
+### Imputing Facebook movement data
+
+The imputation of Facebook movement baseline values is done according to a spatial interaction model (see e.g. [@rowe2022a]).
+We consider baseline movement counts $y^b_{ijw}$ between pairs of origin $i$ and a destination $j$ tiles on weekday $w$, and model this variable as a function of the Facebook population count at the origin tile on the same weekday, the Facebook population count at the destination on the same weekday and the distance between origin and destination.
+We also include indicator variables to capture the day of the week.
+Mathematically, this model can be expressed as \begin{equation}
+\mu^b_{ijw} = \beta_0 + \beta_1p^b_{iw} +\beta_2p^b_{jw} + \beta_3d_{ij} + \beta_4w + \varepsilon
+\end{equation}
+
+where $\mu^b_{ijw} = E[y^b_{ijw}]$ is the expectation of the flow of people from tile $i$ to tile $j$ on the weekday $w$ during the baseline period; $\beta_0$ is an intercept, $p^b_{iw}$ and $p^b_{jw}$ are the Facebook population counts at the origin and destination on weekday $w$ during the baseline period, $d_{ij}$ is the distance between origin and destination, $w$ is a series of indicator variables capturing the day of the week, and $\beta_{0,1,2,3,4}$ are model parameters to be estimated from the observed data.
+The error term is denoted by \varepsilon.
+To estimate the model parameters, we used a Gaussian regression model, taking the log of the population at origin and at destination, and the log of the distance.
+Residual plots for the spatial interaction model are provided in the Supplementary Figure @fig-residualsim.
+
+We compute missing Facebook movement counts during the crisis period by considering the complete Facebook movements baseline and the percentage change in the number of counts with respect to the baseline, which is reported in the Facebook Movements datasets even when the count is not reported due to its low value.
+
+### Applying correction factors
+
+The total number of active users within a given time window exhibits daily fluctuations due to limitations in Internet connectivity and user data access options [@Maas19].
+Following [@yabe2020] and [@duan2024], we applied a correction factor to mitigate the potential impact of these daily fluctuations on the results, since they could mask the mobility trends.
+This approach assumes that the representativeness of Facebook data across spatial units remains consistent throughout the study period.
+
+The adjusted number of movements between tile $i$ and tile $j$ on day $d$, $y'_{ijd}$ was obtained as:
+
+$$
+y'_{ijd} = k_d \times y^c_{ijd}
+$$
+
+where $y^c_{ijd}$ is the original number of movements between tiles $i$ and $j$ on day $d$ during the crises period and $k_d$ is a correction factor computed as the median of the sum of active user counts across all days $d$ divided by the sum of active user counts across all spatial units on day $d$.
+Mathematically,
+
+$$k_d = \dfrac{\mbox{med}_d\big(\sum_i{p^c_{id}}\big)}{\sum_i{p^c_{id}}}.$$
+
+### Time series smoothing
+
+A time series for each pair of origin-destination tiles was generated using data processed as described above.
+However, the resulting time series still contained missing values due to days when no data was reported for specific location pairs.
+To ensure continuity, we addressed this issue by imputing missing values within the time series, replacing them with the average of the nearest 15 observations within the time series.
+This number was chosen to provide an optimal balance between maintaining temporal proximity and ensuring sufficient data coverage.
+
+Additionally, to reduce noise and highlight underlying trends that might be obscured by short-term fluctuations, we applied a rolling-average smoothing technique to the resulting time series.
+This approach was used in the time series presented in @fig-recoverydensity, @fig-recoveryrdi and @fig-fuasflows.
+
 ## Classification of tiles according to level of urbanisation and socioeconomic deprivation {#sec-methodsclassify}
@@
 The geographic distribution of categories for each of the countries is displayed in @fig-classification for Argentina, Chile and Colombia respectively.
+@fig-classification also shows, for each country, the proportion of the population across the various analysis categories is presented, based on both Facebook population counts and Worldpop population estimates.
+The Facebook population counts reflect the average number of active users across all weekdays in the pre-pandemic baseline period.
 
-A key challenge in the in the analysis of population counts and movements with Facebook user data is the absence of small count records, which is a result of privacy-protection techniques applied to ensure that the location of individuals or small groups cannot be identified.
-As a consequence, the countries of the analysis have locations with null values for Facebook population counts and flows.
-These missing data are not distributed randomly.
-For example, the missing values for FB population counts display high spatial autocorrelation as shown in Supplementary Figure SF1.
-Therefore, simply removing the missing records from the analysis could lead to geographically biased results (Afghari et al., 2019).
-To address this, we designed a data processing method to estimate the missing values, both for Facebook population counts and flows.
-Furthermore, this method also applies a correction factor to eliminate the fluctuations in the daily number of observations, assuming that the representativeness of Facebook data across spatial units remains stable during the study period.
-The data processing method is described in section X.
-
-Even after applying the imputation method to estimate missing values, we acknowledge that Facebook data may still overrepresent certain population groups while underrepresenting others.
-For example, Figure @fig-classification shows, for each country, the proportion of the population across the various analysis categories.
-These comparisons use WorldPop population estimates and Facebook population counts, with the latter reflecting active users on an average day during a pre-pandemic baseline period (see Section X of the methodology for details).
 In all three countries, notable discrepancies appear between the population distributions according to WorldPop and Facebook data.
@@
 
-## Processing Facebook data {#sec-methodsprocess}
+## Trend analysis {#sec-methodstrend}
 
-To ensure the privacy and anonymity of the users' data, Meta removes information corresponding to data entries where the population or movement count is less than 10 for a specific time or day, retaining only information about the percentage change in the number of counts with respect to the baseline period is retained [@Maas19].
-Consequently, we observe many tiles where the population or movement count for either the baseline or crisis period are blank.
-<!-- Justify why we do the imputation.
-Spatial bias (maybe show Moran's I for missing values on a selected day?
-Or everyday?).
-Small number bias, underrespresentation of dataentries with low counts that we know for sure exist !-->
-
-### Facebook population data
-
-To imput Facebook population baseline values, we first identify all the baseline values that are available for each spatial unit and weekday (of the three available time windows, we only consider one per day).
-We then estimate the missing baseline values for each weekday, based on a linear model for the relationship between the Worldpop population and the Facebook population counts, which are fitted using ordinary least squares.
-This is illustrated in Supplementary Figure \ref{fig:imput-baseline-population}.
-
-We then use the complete baseline of Facebook population values to compute missing Facebook population counts during the crisis period.
-This is possible because, as mentioned above, Meta reports the percentage change in the number of counts with respect to the baseline, even if the counts are not reported due to low value.
-
-### Facebook movement data
-
-The imputation of Facebook movement baseline values is done according to a spatial interaction model (see e.g. [@rowe2022a]).
-We considered the population flow between an origin and a destination tile, and model this variable as a function of the Facebook population count at the origin tile on the same weekday, the Facebook population count at the destination on the same weekday and the distance between origin and destination; we also included indicator variables to capture the day of the week and the pair of population density classes corresponding to the origin and destination tiles.
-Mathematically, this model can be expressed as \begin{equation}
-\mu_{ijw} = \beta_0 + \beta_1pop_{iw} +\beta_2pop_{jw} + \beta_3d_{ij} + \beta_4W + \varepsilon
-\end{equation}
-
-where $\mu_{ijw}$ is the expectation of the flow of people from tile $i$ to tile $j$ on the weekday $w$; $\beta_0$ is an intercept $pop_{iw}$ and $pop_{jw}$ are the Facebook population counts at the origin and destination on weekday $w$ during the baseline period, $d_{ij}$ is the distance between origin and destination, $W$ is a series of indicator variables capturing the day of the week, $\beta_{0,1,2,3,4}$ are model parameters to be estimated from the observed data.
-The error term is denoted by \varepsilon.
-To estimate the model parameters, we used a Gaussian regression model, taking the log of the population at origin and at destination, and the log of the distance.
-Residual plots for the spatial interaction model are provided in the Supplementary Figure \ref{fig:residuals-sim}.
-
-We compute missing Facebook movement counts during the crisis period by considering the complete Facebook movements baseline and the percentage change in the number of counts with respect to the baseline, which is reported in the Facebook Movements datasets even when the count is not reported due to its low value.
-
-### Correction factors
-
-The total number of active users within a given time window exhibits daily fluctuations due to limitations in Internet connectivity and user data access options [@Maas19].
-Following [@yabe2020] and [@duan2024], we applied a correction factor to account for fluctuations in the daily number of observed users.
-This approach assumes that the representativeness of Facebook data across spatial units remains consistent throughout the study period.
-
-This was done to mitigate the potential impact of variations in the total number of collected users on the results, which could mask the mobility trends within population density and relative deprivation categories.
-The adjusted number of movements between tile $i$ and tile $j$ on day $d$, $y'_{ijd}$ was obtained as:
-
-$$
-y'_{ijd} = k_d \times y_{ijd}
-$$
-
-where $y_{ijd}$ is the original number of movements between tiels $i$ and $j$ on day $d$ and $k_d$ is a correction factor obtained as the median of the sum of user counts on day $d$ divided by the sum of user counts across all spatial units on day $d$.
-Mathematically,
-
-$$k_d = \dfrac{\mbox{Median}\big\{\sum_i{P_{id}}\big\}_d}{\sum_i{P_{id}}}$$.
-
-### Time series smoothing
-
-A time series for each origin-destination tile pair was generated using data processed as described in Sections X to Y.
-However, these time series still contained missing values due to days when no data was reported for specific location pairs.
-To ensure continuity, we addressed this issue by imputing missing values, replacing them with the average of the nearest 15 observations within the time series.
-This number was chosen to provide an optimal balance between maintaining temporal proximity and ensuring sufficient data coverage.
-
-Additionally, to reduce noise and highlight underlying trends that might be obscured by short-term fluctuations, we applied a rolling-average smoothing technique to the resulting time series.
-This approach was used in the time series presented in Figures X, Y, Z.
-
-## Trend analysis {#sec-methodstrend}
+In order to quantify intergroup differences in the evolution towards pre-pandemic mobility patterns, we model the time series displayed in @fig-recoverydensity and @fig-recoveryrdi.
+To this end, we extract the trend, seasonal, and noise components of each time series using the `seasonal_decompose()` method from the time-series models and methods API in Python's `statsmodels` package (version 0.14.4).
+We then model the trend component according to five linear mixed-effects model specifications, using R's `glmmTMB` library (version 1.1.10).
 
@@
 Model 3 accounts for a random intercept based on the origin category, Model 4 includes a random slope for the origin category, and Model 5 combines both a random intercept and a random slope by origin category.
-Figures @fig-re-density and @fig-re-rdi illustrate the random variation in intercept and slope estimated by Models 3, 4, and 5.
-Full details on the parameter estimates are provided in Supplementary Tables SX for each country.
+Figures @fig-redensity and @fig-rerdi illustrate the random variation in intercept and slope estimated by Models 3, 4, and 5.
+Full details on the parameter estimates are provided in Supplementary Tables @tab-trendanalysis for each country.
 

### 2025-05-16 code/analysis/09-lme-with-R-cities.qmd
updates-code

---
title: "lme-r"
format: html
editor: visual
---

## Dependencies

```{r}
# data wrangling
library(tidyverse)
# data utilities
library(timetk)
# estimating mixed effects models
library(lme4)
library(merTools)
library(glmmTMB)
library(nlme)
# correlograms
library(ggcorrplot)
library(corrplot)
# data visualisation
library(viridis)
library(ggthemes)
library(ggpubr)
library(zoo)
library(showtext)
library(patchwork)
library(ggrepel)
# display regression equation
library(equatiomatic)
# standardise input variables
  #library(arm)
  # reporting regression results
library(broom.mixed)
library(gtsummary)
library(sjPlot)
```

## Data import

```{r}
# Set working directory
wd <- '/Users/carmen/Library/CloudStorage/OneDrive-TheUniversityofLiverpool/research/RECAST/latin-mobility-covid-local-files'

# Set the country
country <- 'Chile'

# Initialize variables for country_short and country_code
country_short <- ""
country_code <- ""

# Assign country_short and country_code based on the country
if (country == 'Argentina') {
    country_short <- 'ARG'
    country_code <- 'AR'
} else if (country == 'Chile') {
    country_short <- 'CHL'
    country_code <- 'CL'
} else if (country == 'Colombia') {
    country_short <- 'COL'
    country_code <- 'CO'
} 
```

```{r}
category <- "density"
capital <- TRUE

if (capital == TRUE) {
  place <- 'capital'
} else {
  place <- 'fuas'
}
```

```{r}
capital
# Read the CSV file
file_path <- paste0(wd, '/data/outputs/', country_short, '/mov-analysis/by-', category, '/trend_', place, '.csv')
df <- read.csv(file_path)

df$time2 <- df$time^2
```

```{r}
dates <- read.csv(paste0(wd, '/data/outputs/', country_short, '/mov-analysis/dates.csv'))

df_stringency <- read.csv(paste0(wd, '/data/inputs/covid-stringency/owid-covid-data.csv')) %>%
  filter(location == "Argentina") %>% 
  filter(date %in% dates$date) %>% 
  slice(seq(1, n(), by = 7)) %>% 
  mutate(time = seq(0, n() - 1)) %>%
  select(time, stringency_index)

df <- df %>% left_join(df_stringency, by = "time")

df <- df[, !(names(df) == "X")]

```

## Define models

```{r}

eq1 <- y ~ time
m1 <- glmmTMB(eq1, data = df)

###

eq1s <- y ~ time + stringency_index
m1s <- glmmTMB(eq1s, data = df)

###

eq2cat <- y ~ time + cat
m2cat <- glmmTMB(eq2cat, data = df)

eq2cats <- y ~ time + cat + stringency_index
m2cats <- glmmTMB(eq2cats, data = df)

###

eq3cat <- y ~ time + (1|cat)
m3cat <- glmmTMB(eq3cat, data = df)
m3cat_re <- plot_model(m3cat, type ="re")

eq3cats <- y ~ time + (1|cat) + stringency_index
m3cats <- glmmTMB(eq3cats, data = df)
m3cats_re <- plot_model(m3cats, type ="re")

###

df$catf <- factor(df$cat)
eq4cat <- y ~ time + (0 + time|catf) 
m4cat <- glmmTMB(eq4cat, data = df)
m4cat_re <- plot_model(m4cat, type ="re")

df$catf <- factor(df$cat)
eq4cats <- y ~ time + (0 + time|catf) + stringency_index
m4cats <- glmmTMB(eq4cats, data = df)
m4cats_re <- plot_model(m4cats, type ="re")

###

df$catf <- factor(df$cat)
eq5cat <- y ~ time + (time|catf) 
m5cat <- glmmTMB(eq5cat, data = df)
m5cat_re <- plot_model(m5cat, type ="re")

df$catf <- factor(df$cat)
eq5cats <- y ~ time + (time|catf) + stringency_index
m5cats <- glmmTMB(eq5cats, data = df)
m5cats_re <- plot_model(m5cats, type ="re")

###

eq6cat <- y ~ time + time2 + (time|cat) 
m6cat <- glmmTMB(eq6cat, data = df)
m6cat_re <- plot_model(m6cat, type ="re")

eq6cats <- y ~ time + time2 + (time|cat) + stringency_index
m6cats <- glmmTMB(eq6cats, data = df)
m6cats_re <- plot_model(m6cats, type ="re")

###

eq7cat <- y ~ time + time2 + (time + time2 | cat)
m7cat <- glmmTMB(eq7cat, data = df)
m7cat_re <- plot_model(m7cat, type ="re")

eq7cats <- y ~ time + time2 + (time + time2|cat) + stringency_index
m7cats <- glmmTMB(eq7cats, data = df)
m7cats_re <- plot_model(m7cats, type ="re")

```

## Save model data

```{r}
models_cat <- list(
  m1,
  m2cat,
  m3cat,
  m4cat,
  m5cat,
  m6cat,
  m7cat
  )
tidy_results_cat <- map_df(models_cat, tidy, .id = "model")
file_tidy_results_cat <- paste0(wd, '/data/outputs/', country_short, '/mov-analysis/by-', category, '/data_trend_netflows_', place, '_param_tidy_results_cat.csv')
write.csv(tidy_results_cat, file_tidy_results_cat, row.names = FALSE)

glance_results_cat <- map_df(models_cat, glance, .id = "model")
file_glance_results_cat <- paste0(wd, '/data/outputs/', country_short, '/mov-analysis/by-', category, '/data_trend_netflows_', place, '_param_glance_results_cat.csv')
write.csv(glance_results_cat, file_glance_results_cat, row.names = FALSE)

models_cats <- list(
  m1s,
  m2cats,
  m3cats,
  m4cats,
  m5cats,
  m6cats,
  m7cats
  )
tidy_results_cats <- map_df(models_cats, tidy, .id = "model")
file_tidy_results_cats <- paste0(wd, '/data/outputs/', country_short, '/mov-analysis/by-', category, '/data_trend_netflows_', place, '_param_tidy_results_cats.csv')
write.csv(tidy_results_cats, file_tidy_results_cats, row.names = FALSE)

glance_results_cats <- map_df(models_cats, glance, .id = "model")
file_glance_results_cats <- paste0(wd, '/data/outputs/', country_short, '/mov-analysis/by-', category, '/data_trend_netflows_', place, '_param_glance_results_cats.csv')
write.csv(glance_results_cats, file_glance_results_cats, row.names = FALSE)
```

## Save random effects data

```{r}

file_m3cat_re <- paste0(wd, '/data/outputs/', country_short, '/mov-analysis/by-', category, '/data_trend_param_', place, '_m3cat_re.csv')
write.csv(m3cat_re[[1]], file_m3cat_re, row.names = FALSE)

file_m3cats_re <- paste0(wd, '/data/outputs/', country_short, '/mov-analysis/by-', category, '/data_trend_param_', place, '_m3cats_re.csv')
write.csv(m3cats_re[[1]], file_m3cats_re, row.names = FALSE)

###

file_m4cat_re <- paste0(wd, '/data/outputs/', country_short, '/mov-analysis/by-', category, '/data_trend_param_', place, '_m4cat_re.csv')
write.csv(m4cat_re[[1]], file_m4cat_re, row.names = FALSE)

file_m4cats_re <- paste0(wd, '/data/outputs/', country_short, '/mov-analysis/by-', category, '/data_trend_param_', place, '_m4cats_re.csv')
write.csv(m4cats_re[[1]], file_m4cats_re, row.names = FALSE)

###

file_m5cat_re <- paste0(wd, '/data/outputs/', country_short, '/mov-analysis/by-', category, '/data_trend_param_', place, '_m5cat_re.csv')
write.csv(m5cat_re[[1]], file_m5cat_re, row.names = FALSE)

file_m5cats_re <- paste0(wd, '/data/outputs/', country_short, '/mov-analysis/by-', category, '/data_trend_param_', place, '_m5cats_re.csv')
write.csv(m5cats_re[[1]], file_m5cats_re, row.names = FALSE)

###

file_m6cat_re <- paste0(wd, '/data/outputs/', country_short, '/mov-analysis/by-', category, '/data_trend_param_', place, '_m6cat_re.csv')
write.csv(m6cat_re[[1]], file_m6cat_re, row.names = FALSE)

file_m6cats_re <- paste0(wd, '/data/outputs/', country_short, '/mov-analysis/by-', category, '/data_trend_param_', place, '_m6cats_re.csv')
write.csv(m6cats_re[[1]], file_m6cats_re, row.names = FALSE)

###

file_m7cat_re <- paste0(wd, '/data/outputs/', country_short, '/mov-analysis/by-', category, '/data_trend_param_', place, '_m7cat_re.csv')
write.csv(m7cat_re[[1]], file_m7cat_re, row.names = FALSE)

file_m7cats_re <- paste0(wd, '/data/outputs/', country_short, '/mov-analysis/by-', category, '/data_trend_param_', place, '_m7cats_re.csv')
write.csv(m7cats_re[[1]], file_m7cats_re, row.names = FALSE)

```

## Other

```{r}
# p1_m1me <- plot_model(m3o_re, terms = c("1", "time")) %>% 
#   .$data %>% ggplot(aes(x=estimate, y=term, color=group)) +
#   geom_point(size = 3.5,  colour = "black") +
#   geom_errorbar(aes(xmin = conf.low, xmax = conf.high), 
#                 width=.05, color= "black") +
#   geom_pointrange(aes(xmin = conf.low, xmax = conf.high)) +
#   scale_color_manual(values=c("#88ccee")) +
#   theme(legend.position = "none",
#         axis.title.y = element_blank(),
#         axis.text = element_text(size = 10.5)
#   ) +
#   scale_y_discrete(labels=c("time" = "Week", 
#                             "1" = "Intercept")) +
#   geom_vline(xintercept=0, 
#              linetype="solid", 
#              color = "grey20",
#              linewidth=1,
#              alpha =.2) +
#   labs(title = "Main fixed effects")
# 
# p1_m1me
```

```{r}
# m2_re <- plot_model(m,
#            type ="re", terms = "O")
# 
# m2_re <- m2_re[[1]]$data
# 
# p2_m2re <- m2_re %>% filter(facet == "O (Intercept)") %>% 
#   ggplot(aes(x=estimate, y=term, color=group)) +
#   geom_point(size = 3.5,  colour = "black") +
#   geom_errorbar(aes(xmin = conf.low, xmax = conf.high), 
#                 width=.8, color= "black") +
#   geom_pointrange(aes(xmin = conf.low, xmax = conf.high)) +
#   scale_color_manual(values=c("#ddcc77", "#88ccee")) +
#   theme(legend.position = "none",
#         axis.title.y = element_blank(),
#         axis.text = element_text(size = 10.5)
#         ) +
#   geom_vline(xintercept=0, 
#              linetype="solid", 
#              color = "grey20",
#              size=1,
#              alpha =.2) +
#   labs(title = "d")
# 
# p2_m2re
```

```{r}
# eq <- y ~  time + (time | O) + (time | D)
# model <- glmmTMB(eq, data = df)
```

```{r}
# pm3o <- plot_model(m3o, terms = c("time", "O")) %>% 
#   .$df %>% ggplot(aes(x=estimate, y=term, color=group)) +
#   geom_point(size = 3.5,  colour = "black") +
#   geom_errorbar(aes(xmin = conf.low, xmax = conf.high), 
#                 width=.05, color= "black") +
#   geom_pointrange(aes(xmin = conf.low, xmax = conf.high)) +
#   scale_color_manual(values=c("#88ccee")) +
#   theme(legend.position = "none",
#         axis.title.y = element_blank(),
#         axis.text = element_text(size = 10.5)
#   ) +
#   scale_y_discrete(labels=c("z_stringency_index" = "Stringency t", 
#                             "z_cases" = "New cases t",
#                             "lag(z_stringency_index, 1)" = "Stringency t-1")) +
#   geom_vline(xintercept=0, 
#              linetype="solid", 
#              color = "grey20",
#              linewidth=1,
#              alpha =.2) +
#   labs(title = "c",
#        subtitle = " Main fixed effects")
# 
# p1_m
```

## Plots of RE

```{r}
m3cat_re
m3cats_re
m4cat_re
m4cats_re
m5cat_re
m5cats_re
m6cat_re
m6cats_re
m7cat_re
m7cats_re
```

```{r}
m6cat_re[[1]]$estimate[[6]]

m6cat$fit$par
```

```{r}
# Create predicted data frame for time 1 to 1000
pred_df_1 <- tibble(time = 1:300, time2 = time^2,
  fitted = (m6cat$fit$par[[1]] + m6cat_re[[1]]$estimate[[1]]) +
    (m6cat$fit$par[[2]] + m6cat_re[[1]]$estimate[[6]]) * time + 
    m6cat$fit$par[[3]] * time2)

# Create predicted data frame for time 1 to 1000
pred_df_2 <- tibble(time = 1:300, time2 = time^2,
  fitted = (m6cat$fit$par[[1]] + m6cat_re[[1]]$estimate[[2]]) +
    (m6cat$fit$par[[2]] + m6cat_re[[1]]$estimate[[7]]) * time + 
    m6cat$fit$par[[3]] * time2)

# Create predicted data frame for time 1 to 1000
pred_df_3 <- tibble(time = 1:300, time2 = time^2,
  fitted = (m6cat$fit$par[[1]] + m6cat_re[[1]]$estimate[[3]]) +
    (m6cat$fit$par[[2]] + m6cat_re[[1]]$estimate[[8]]) * time + 
    m6cat$fit$par[[3]] * time2)

# Create predicted data frame for time 1 to 1000
pred_df_4 <- tibble(time = 1:300, time2 = time^2,
  fitted = (m6cat$fit$par[[1]] + m6cat_re[[1]]$estimate[[4]]) +
    (m6cat$fit$par[[2]] + m6cat_re[[1]]$estimate[[9]]) * time + 
    m6cat$fit$par[[3]] * time2)

# Create predicted data frame for time 1 to 1000
pred_df_5 <- tibble(time = 1:300, time2 = time^2,
  fitted = (m6cat$fit$par[[1]] + m6cat_re[[1]]$estimate[[5]]) +
    (m6cat$fit$par[[2]] + m6cat_re[[1]]$estimate[[10]]) * time + 
    m6cat$fit$par[[3]] * time2)

# Plot using ggplot
ggplot() +
  geom_point(data = df, aes(x = time, y = y), color = "blue", alpha = 0.6, size = 2) +
  geom_point(data = pred_df_1, aes(x = time, y = fitted), color = "red", alpha = 0.6, size = 1.5) +
  geom_point(data = pred_df_2, aes(x = time, y = fitted), color = "red3", alpha = 0.6, size = 1.5) +
  geom_point(data = pred_df_3, aes(x = time, y = fitted), color = "darkred", alpha = 0.6, size = 1.5) +
  geom_point(data = pred_df_4, aes(x = time, y = fitted), color = "purple4", alpha = 0.6, size = 1.5) +
  geom_point(data = pred_df_5, aes(x = time, y = fitted), color = "midnightblue", alpha = 0.6, size = 1.5) +
  labs(title = "Observed vs Predicted (Quadratic Fit)",
       x = "Time", y = "Y") +
  xlim(0, 300) +
  ylim(min(c(df$y, pred_df$fitted), na.rm = TRUE),
       max(c(df$y, pred_df$fitted, 20), na.rm = TRUE)) +
  theme_minimal()

```

### 2025-05-16 code/analysis/09b-gamm-with-R.qmd
updates-code

---
title: "lme-r"
format: html
editor: visual
---

```{r}
# data wrangling
library(tidyverse)
# data utilities
library(timetk)
# estimating mixed effects models
library(lme4)
library(merTools)
library(glmmTMB)
library(nlme)
# correlograms
library(ggcorrplot)
library(corrplot)
# data visualisation
library(viridis)
library(ggthemes)
library(ggpubr)
library(zoo)
library(showtext)
library(patchwork)
library(ggrepel)
# display regression equation
library(equatiomatic)
# standardise input variables
  #library(arm)
  # reporting regression results
library(broom.mixed)
library(gtsummary)
library(sjPlot)
library(mgcv)
library(car)    # For VIF (optional, if you want to check multicollinearity)
library(lmtest) # For Durbin-Watson test
library(MASS)   # For stepwise AIC
```

```{r}
# Set working directory
wd <- '/Users/carmen/Library/CloudStorage/OneDrive-TheUniversityofLiverpool/research/RECAST/latin-mobility-covid-local-files'

# Set the country
country <- 'Chile'

# Initialize variables for country_short and country_code
country_short <- ""
country_code <- ""

# Assign country_short and country_code based on the country
if (country == 'Argentina') {
    country_short <- 'ARG'
    country_code <- 'AR'
} else if (country == 'Chile') {
    country_short <- 'CHL'
    country_code <- 'CL'
} else if (country == 'Colombia') {
    country_short <- 'COL'
    country_code <- 'CO'
} 
```

```{r}
category <- "density"
```

```{r}
# Read the CSV file
file_path <- paste0(wd, '/data/outputs/', country_short, '/mov-analysis/by-', category, '/time_series_capital.csv')
df <- read.csv(file_path)
df <- df[, !(names(df) == "X")]

df$time2 <- df$time^2

```

```{r}
dates <- read.csv(paste0(wd, '/data/outputs/', country_short, '/mov-analysis/dates.csv'))

df_stringency <- read.csv(paste0(wd, '/data/inputs/covid-stringency/owid-covid-data.csv')) %>%
  filter(location == "Argentina") %>% 
  filter(date %in% dates$date) %>% 
  slice(seq(1, n(), by = 7)) %>% 
  mutate(time = seq(0, n() - 1)) %>%
  select(time, stringency_index)

df <- df %>% left_join(df_stringency, by = "time")
```

```{r}
# Fit a base model (simple linear with random intercept for cat)
model_1 <- gamm(y ~ time , 
                random = list(cat = ~ 1), 
                data = df)

# Durbin-Watson test to check for autocorrelation in residuals
dw_test <- dwtest(model_1$gam)
print(dw_test)  # p-value < 0.05 indicates autocorrelation

# Check residuals for autocorrelation (ACF plot)
acf(model_1$gam$residuals)  # Look for any spikes at lag 1 (positive autocorrelation)

# Model with nonlinear effect of time (still no autocorrelation)
model_2 <- gamm(y ~ s(time) , 
                random = list(cat = ~ 1), 
                data = df)

# Model with random slopes for time (still no autocorrelation)
model_3 <- gamm(y ~ s(time), 
                random = list(cat = ~ time), 
                data = df)

# Model with AR(1) autocorrelation structure (to address autocorrelation)
model_4 <- gamm(y ~ s(time) , 
                random = list(cat = ~ time), 
                correlation = corAR1(form = ~ time | cat),
                data = df)

# Model with AR(2) autocorrelation structure (to address autocorrelation)
model_5 <- gamm(y ~ s(time) , 
                random = list(cat = ~ time), 
                correlation = corARMA(p = 2, form = ~ time | cat),
                data = df)

# Model with AR(2) autocorrelation structure (to address autocorrelation)
model_6 <- gamm(y ~ s(time) , 
                random = list(cat = ~ time), 
                correlation = corARMA(p = 4, form = ~ time | cat),
                data = df)

```

```{r}
summary(model_1$lme)
```

```{r}
print(summary(model_1$lme)$AIC)
print(summary(model_2$lme)$AIC)
print(summary(model_3$lme)$AIC)
print(summary(model_4$lme)$AIC)
print(summary(model_5$lme)$AIC)
print(summary(model_6$lme)$AIC)
```

```{r}
df_new <- data.frame(
  time = rep(0:199, times = 5),
  cat = rep(0:4, each = 200)
)
```

```{r}
# Predict on the original data (or new data if you have it)
predictions <- predict(model_6$gam, 
                       newdata = df_new, 
                       re.form=NULL,
                       type = "response")

df_new$y_pred <- predictions
```

```{r}

# Extract random effects for each level of 'cat' from the original model
random_effects <- ranef(model_3$lme)$cat

# Make predictions for the fixed effects (smooth function of time)
predictions_fixed <- predict(model_3$gam, 
                             newdata = df_new, 
                             type = "response")

# Add random effects for the corresponding 'cat' levels in new_data
df_new$predicted_y_with_random <- predictions_fixed + 
  random_effects[match(df_new$cat, names(random_effects))]

# View the updated data with predictions
print(df_new)
```

### 2025-06-04 code/analysis/05-mixed-models-R.qmd
updates-code-analysis

---
title: "lme-r"
format: html
editor: visual
---

## Dependencies

```{r}
# data wrangling
library(tidyverse)
# data utilities
library(timetk)
# estimating mixed effects models
library(lme4)
library(merTools)
library(glmmTMB)
library(nlme)
# correlograms
library(ggcorrplot)
library(corrplot)
# data visualisation
library(viridis)
library(ggthemes)
library(ggpubr)
library(zoo)
library(showtext)
library(patchwork)
library(ggrepel)
# display regression equation
library(equatiomatic)
# standardise input variables
  #library(arm)
  # reporting regression results
library(broom.mixed)
library(gtsummary)
library(sjPlot)
library(modelsummary)
```

## Data import

```{r}
# Set working directory
wd <- '/Users/carmen/Library/CloudStorage/OneDrive-TheUniversityofLiverpool/research/RECAST/latin-mobility-covid-local-files'

# Set the country
country <- 'Argentina'

# Initialize variables for country_short and country_code
country_short <- ""
country_code <- ""

# Assign country_short and country_code based on the country
if (country == 'Argentina') {
    country_short <- 'ARG'
    country_code <- 'AR'
} else if (country == 'Chile') {
    country_short <- 'CHL'
    country_code <- 'CL'
} else if (country == 'Colombia') {
    country_short <- 'COL'
    country_code <- 'CO'
} 
```

```{r}
category <- "rdi"
```

```{r}
# Read the CSV file
file_path <- paste0(wd, '/data/outputs/', country_short, '/mov-analysis/by-', category, '/data_trend_new.csv')
df <- read.csv(file_path)

df$time2 <- df$time^2
```

```{r}
dates <- read.csv(paste0(wd, '/data/outputs/', country_short, '/mov-analysis/dates.csv'))

df_stringency <- read.csv(paste0(wd, '/data/inputs/covid-stringency/owid-covid-data.csv')) %>%
  filter(location == "Chile") %>% 
  filter(date %in% dates$date) %>% 
  slice(seq(1, n(), by = 7)) %>% 
  mutate(time = seq(0, n() - 1)) %>%
  select(time, stringency_index)

df <- df %>% left_join(df_stringency, by = "time") %>% rename("O" = "cat")
```

```{r}
df_stringency$stringency_index
```

## Define models

```{r}

eq1 <- y ~ time
m1 <- glmmTMB(eq1, data = df)

###

eq1s <- y ~ time + stringency_index
m1s <- glmmTMB(eq1s, data = df)

###

# eq2 <- y ~ time + O + D
# m2 <- glmmTMB(eq2, data = df)
# 
# eq2s <- y ~ time + O + D + stringency_index
# m2s <- glmmTMB(eq2s, data = df)

eq2o <- y ~ time + O
m2o <- glmmTMB(eq2o, data = df)

eq2os <- y ~ time + O + stringency_index
m2os <- glmmTMB(eq2os, data = df)

###

# eq3 <- y ~ time + (1|O) + (1|D)
# m3 <- glmmTMB(eq3, data = df)
# m3_re <- plot_model(m3, type ="re")
# 
# eq3s <- y ~ time + (1|O) + (1|D) + stringency_index
# m3s <- glmmTMB(eq3s, data = df)
# m3s_re <- plot_model(m3s, type ="re")

eq3o <- y ~ time + (1|O)
m3o <- glmmTMB(eq3o, data = df)
m3o_re <- plot_model(m3o, type ="re")

eq3os <- y ~ time + (1|O) + stringency_index
m3os <- glmmTMB(eq3os, data = df)
m3os_re <- plot_model(m3os, type ="re")

###

# df$Of <- factor(df$O)
# df$Df <- factor(df$D)
# eq4 <- y ~ time + (0 + time|Of) + (0 + time|Df)
# m4 <- glmmTMB(eq4, data = df)
# m4_re <- plot_model(m4, type ="re")
# 
# df$Of <- factor(df$O)
# df$Df <- factor(df$D)
# eq4s <- y ~ time + (0 + time|Of) + (0 + time|Df) + stringency_index
# m4s <- glmmTMB(eq4s, data = df)
# m4s_re <- plot_model(m4s, type ="re")

df$Of <- factor(df$O)
eq4o <- y ~ time + (0 + time|Of) 
m4o <- glmmTMB(eq4o, data = df)
m4o_re <- plot_model(m4o, type ="re")

df$Of <- factor(df$O)
eq4os <- y ~ time + (0 + time|Of) + stringency_index
m4os <- glmmTMB(eq4os, data = df)
m4os_re <- plot_model(m4os, type ="re")

###

# df$Of <- factor(df$O)
# df$Df <- factor(df$D)
# eq5 <- y ~ time + (time|Of) + (time|Df)
# m5 <- glmmTMB(eq5, data = df)
# m5_re <- plot_model(m5, type ="re")
# 
# df$Of <- factor(df$O)
# df$Df <- factor(df$D)
# eq5s <- y ~ time + (time|Of) + (time|Df) + stringency_index
# m5s <- glmmTMB(eq5s, data = df)
# m5s_re <- plot_model(m5s, type ="re")

df$Of <- factor(df$O)
eq5o <- y ~ time + (time|Of) 
m5o <- glmmTMB(eq5o, data = df)
m5o_re <- plot_model(m5o, type ="re")

df$Of <- factor(df$O)
eq5os <- y ~ time + (time|Of) + stringency_index
m5os <- glmmTMB(eq5os, data = df)
m5os_re <- plot_model(m5os, type ="re")

###

# eq6 <- y ~ time + time2 + (time|O) + (time|D)
# m6 <- glmmTMB(eq6, data = df)
# m6_re <- plot_model(m6, type ="re")
# 
# eq6s <- y ~ time + time2 + (time|O) + (time|D) + stringency_index
# m6s <- glmmTMB(eq6s, data = df)
# m6s_re <- plot_model(m6s, type ="re")

eq6o <- y ~ time + time2 + (time|O) 
m6o <- glmmTMB(eq6o, data = df)
m6o_re <- plot_model(m6o, type ="re")

eq6os <- y ~ time + time2 + (time|O) + stringency_index
m6os <- glmmTMB(eq6os, data = df)
m6os_re <- plot_model(m6os, type ="re")

###

# eq7 <- y ~ time + time2 + (time + time2|O) + (time + time2|D)
# m7 <- glmmTMB(eq7, data = df)
# m7_re <- plot_model(m7, type ="re")
# 
# eq7s <- y ~ time + time2 + (time + time2|O) + (time + time2|D) + stringency_index
# m7s <- glmmTMB(eq7s, data = df)
# m7s_re <- plot_model(m7s, type ="re")

eq7o <- y ~ time + time2 + (time + time2|O) 
m7o <- glmmTMB(eq7o, data = df)
m7o_re <- plot_model(m7o, type ="re")

eq7os <- y ~ time + time2 + (time + time2|O) + stringency_index
m7os <- glmmTMB(eq7os, data = df)
m7os_re <- plot_model(m7os, type ="re")

```

## Save model data

```{r}
models_o <- list(
  m1,
  m2o,
  m3o,
  m4o,
  m5o,
  m6o,
  m7o
  )
tidy_results_o <- map_df(models_o, tidy, .id = "model")
tidy_results_o <- map_df(models_o, tidy, .id = "model")
file_tidy_results_o <- paste0(wd, '/data/outputs/', country_short, '/mov-analysis/by-', category, '/data_trend_outflows_param_tidy_results_o_new.csv')
write.csv(tidy_results_o, file_tidy_results_o, row.names = FALSE)

glance_results_o <- map_df(models_o, glance, .id = "model")
file_glance_results_o <- paste0(wd, '/data/outputs/', country_short, '/mov-analysis/by-', category, '/data_trend_outflows_param_glance_results_o_new.csv')
write.csv(glance_results_o, file_glance_results_o, row.names = FALSE)

models_os <- list(
  m1s,
  m2os,
  m3os,
  m4os,
  m5os,
  m6os,
  m7os
  )
tidy_results_os <- map_df(models_os, tidy, .id = "model")
file_tidy_results_os <- paste0(wd, '/data/outputs/', country_short, '/mov-analysis/by-', category, '/data_trend_outflows_param_tidy_results_os_new.csv')
write.csv(tidy_results_os, file_tidy_results_os, row.names = FALSE)

glance_results_os <- map_df(models_os, glance, .id = "model")
file_glance_results_os <- paste0(wd, '/data/outputs/', country_short, '/mov-analysis/by-', category, '/data_trend_outflows_param_glance_results_os_new.csv')
write.csv(glance_results_os, file_glance_results_os, row.names = FALSE)
```

## Save random effects data

```{r}

file_m3o_re <- paste0(wd, '/data/outputs/', country_short, '/mov-analysis/by-', category, '/data_trend_param_quad_m3o_re_new.csv')
write.csv(m3o_re[[1]], file_m3o_re, row.names = FALSE)

file_m3os_re <- paste0(wd, '/data/outputs/', country_short, '/mov-analysis/by-', category, '/data_trend_param_quad_m3os_re_new.csv')
write.csv(m3os_re[[1]], file_m3os_re, row.names = FALSE)

###

file_m4o_re <- paste0(wd, '/data/outputs/', country_short, '/mov-analysis/by-', category, '/data_trend_param_quad_m4o_re_new.csv')
write.csv(m4o_re[[1]], file_m4o_re, row.names = FALSE)

file_m4os_re <- paste0(wd, '/data/outputs/', country_short, '/mov-analysis/by-', category, '/data_trend_param_quad_m4os_re_new.csv')
write.csv(m4os_re[[1]], file_m4os_re, row.names = FALSE)

###

file_m5o_re <- paste0(wd, '/data/outputs/', country_short, '/mov-analysis/by-', category, '/data_trend_param_quad_m5o_re_new.csv')
write.csv(m5o_re[[1]], file_m5o_re, row.names = FALSE)

file_m5os_re <- paste0(wd, '/data/outputs/', country_short, '/mov-analysis/by-', category, '/data_trend_param_quad_m5os_re_new.csv')
write.csv(m5os_re[[1]], file_m5os_re, row.names = FALSE)

###

file_m6o_re <- paste0(wd, '/data/outputs/', country_short, '/mov-analysis/by-', category, '/data_trend_param_quad_m6o_re_new.csv')
write.csv(m6o_re[[1]], file_m6o_re, row.names = FALSE)

file_m6os_re <- paste0(wd, '/data/outputs/', country_short, '/mov-analysis/by-', category, '/data_trend_param_quad_m6os_re_new.csv')
write.csv(m6os_re[[1]], file_m6os_re, row.names = FALSE)

###

file_m7o_re <- paste0(wd, '/data/outputs/', country_short, '/mov-analysis/by-', category, '/data_trend_param_quad_m7o_re_new.csv')
write.csv(m7o_re[[1]], file_m7o_re, row.names = FALSE)

file_m7os_re <- paste0(wd, '/data/outputs/', country_short, '/mov-analysis/by-', category, '/data_trend_param_quad_m7os_re_new.csv')
write.csv(m7os_re[[1]], file_m7os_re, row.names = FALSE)

```

## Plots of RE

```{r}
m3o_re
m3os_re
m4o_re
m4os_re
m5o_re
m5os_re
m6o_re
m6os_re
m7o_re
m7os_re
```

### 2025-12-01 notebooks/analysis/05b-mixed-models-R-r1.qmd
revision

--- 2025-06-04 code/analysis/05-mixed-models-R.qmd
@@
 library(modelsummary)
+library(brms)
 ```
@@
 # Set working directory
-wd <- '/Users/carmen/Library/CloudStorage/OneDrive-TheUniversityofLiverpool/research/RECAST/latin-mobility-covid-local-files'
+wd <- '/your/path/to/working/directory/'
+wd <- '/Users/carmen/Library/CloudStorage/OneDrive-TheUniversityofLiverpool/research/latin-mobility-covid'
 
@@
 ```{r}
-category <- "rdi"
+category <- "density"
 ```
@@
 df_stringency <- read.csv(paste0(wd, '/data/inputs/covid-stringency/owid-covid-data.csv')) %>%
-  filter(location == "Chile") %>% 
+  filter(location == country) %>% 
   filter(date %in% dates$date) %>% 
@@
 df <- df %>% left_join(df_stringency, by = "time") %>% rename("O" = "cat")
-```
 
-```{r}
-df_stringency$stringency_index
 ```
@@
 # df$Df <- factor(df$D)
-# eq5 <- y ~ time + (time|Of) + (time|Df)
+# eq5 <- y ~ time + (time|Of) + (time|Df) + ratio
 # m5 <- glmmTMB(eq5, data = df)
@@
 # df$Df <- factor(df$D)
-# eq5s <- y ~ time + (time|Of) + (time|Df) + stringency_index
+# eq5s <- y ~ time + (time|Of) + (time|Df) + stringency_index + ratio
 # m5s <- glmmTMB(eq5s, data = df)
@@
 
-# eq6 <- y ~ time + time2 + (time|O) + (time|D)
+# eq6 <- y ~ time + time2 + (time|O) + (time|D) + ratio
 # m6 <- glmmTMB(eq6, data = df)
@@
 # 
-# eq6s <- y ~ time + time2 + (time|O) + (time|D) + stringency_index
+# eq6s <- y ~ time + time2 + (time|O) + (time|D) + stringency_index + ratio
 # m6s <- glmmTMB(eq6s, data = df)
@@
 
-# eq7 <- y ~ time + time2 + (time + time2|O) + (time + time2|D)
+# eq7 <- y ~ time + time2 + (time + time2|O) + (time + time2|D) + ratio
 # m7 <- glmmTMB(eq7, data = df)
@@
 # 
-# eq7s <- y ~ time + time2 + (time + time2|O) + (time + time2|D) + stringency_index
+# eq7s <- y ~ time + time2 + (time + time2|O) + (time + time2|D) + stringency_index + ratio
 # m7s <- glmmTMB(eq7s, data = df)
@@
 
+df$O  <- factor(df$O)
+eq8o <- y ~ 
+    s(time, k = 8) + 
+    s(time, O, bs = "fs", k = 7)
+m_gam <- gam(eq8o, data = df, method = "REML")
+
+df$O  <- factor(df$O)
+eq8os <- y ~ 
+    s(time, k = 8) + 
+    stringency_index + 
+    s(time, O, bs = "fs", k = 7)
+m_gams <- gam(eq8os, data   = df, method = "REML")
+
+```
+
+```{r}
+summary(m_gam)$s.table
+```
+
+```{r}
+plot(m_gam, pages = 1, rug = FALSE, scheme = 1, shade = TRUE)
 ```
@@
   m6o,
-  m7o
+  m7o, 
+  m_gam
   )
-tidy_results_o <- map_df(models_o, tidy, .id = "model")
-tidy_results_o <- map_df(models_o, tidy, .id = "model")
+tidy_fixed <- map_df(models_o, ~ tidy(.x, parametric = TRUE), .id = "model")
+tidy_smooth <- map_df(models_o, ~ tidy(.x, smooth = TRUE), .id = "model")
+
+tidy_results_o <- bind_rows(tidy_fixed, tidy_smooth) %>%
+  distinct()
+
 file_tidy_results_o <- paste0(wd, '/data/outputs/', country_short, '/mov-analysis/by-', category, '/data_trend_outflows_param_tidy_results_o_new.csv')
@@
   m6os,
-  m7os
+  m7os,
+  m_gams
   )
-tidy_results_os <- map_df(models_os, tidy, .id = "model")
+tidy_fixed <- map_df(models_os, ~ tidy(.x, parametric = TRUE), .id = "model")
+tidy_smooth <- map_df(models_os, ~ tidy(.x, smooth = TRUE), .id = "model")
+tidy_results_os <- bind_rows(tidy_fixed, tidy_smooth) %>%
+  distinct()
 file_tidy_results_os <- paste0(wd, '/data/outputs/', country_short, '/mov-analysis/by-', category, '/data_trend_outflows_param_tidy_results_os_new.csv')
@@
 write.csv(m7os_re[[1]], file_m7os_re, row.names = FALSE)
-
 ```
@@
 ```{r}
-m3o_re
-m3os_re
-m4o_re
-m4os_re
-m5o_re
-m5os_re
-m6o_re
-m6os_re
-m7o_re
-m7os_re
+# m3o_re
+# m3os_re
+# m4o_re
+# m4os_re
+# m5o_re
+# m5os_re
+# m6o_re
+# m6os_re
+# m7o_re
+# m7os_re
 ```
+
+## Compute recovery times for spline model. The 95% confidence intervals are computed via bootstrap.
+
+```{r}
+stringency <- FALSE
+```
+
+```{r}
+# Create new time sequence covering the range of your original data
+new_time <- expand.grid(
+  time = seq(min(df$time), max(df$time)*100, length.out = max(df$time)*100-min(df$time)+1),
+  O = levels(df$O)   # one dataset per group
+)
+
+new_time <- new_time %>%
+  left_join(df[, c("time", "O", "stringency_index")], by = c("time", "O"))
+
+new_time$stringency_index[is.na(new_time$stringency_index)] <- 0
+
+if (stringency == FALSE) {
+  new_time <- expand.grid(
+  time = seq(min(df$time), max(df$time)*100, length.out = max(df$time)*100-min(df$time)+1),
+  O = levels(df$O)   # one dataset per group
+)
+  pred_base <- predict(m_gam, newdata = new_time, se.fit = TRUE)
+} else {
+  pred_base <- predict(m_gams, newdata = new_time, se.fit = TRUE)
+}
+
+new_time$fit <- pred_base$fit
+new_time$se  <- pred_base$se.fit
+
+# Find FIRST root where preds(t) == target
+get_root_threshold <- function(time, preds, target) {
+  diff_sign <- diff(sign(preds - target))
+  crossings <- which(diff_sign != 0)
+  if (length(crossings) == 0) return(NA)
+
+  i <- crossings[1]  # first crossing
+  out <- tryCatch(
+    uniroot(function(t) {
+      approx(time, preds - target, xout = t)$y
+    },
+    lower = time[i],
+    upper = time[i + 1])$root,
+    error = function(e) NA
+  )
+  return(out)
+}
+
+alphas <- seq(0, 1, by = 0.05)   # adjust resolution as needed
+
+# Bootstrap confidence intervals
+set.seed(123)
+n_boot <- 200   # increase to 500–1000 for publication
+
+results_alpha <- map_df(levels(df$O), function(g) {
+
+  group_data <- new_time %>% filter(O == g)
+
+  # group-specific y(t=0): use earliest time in grid for that group
+  idx0 <- which.min(group_data$time)
+  y0_group <- group_data$fit[idx0]
+
+  # For each alpha / target, bootstrap
+  map_df(seq_along(alphas), function(j) {
+    alpha   <- alphas[j]
+    target <- (1 - alpha) * y0_group
+
+    boot_roots <- replicate(n_boot, {
+      # simulate one curve from uncertainty
+      y_sim <- rnorm(nrow(group_data), mean = group_data$fit, sd = group_data$se)
+      get_root_threshold(group_data$time, y_sim, target)
+    })
+
+    data.frame(
+      O        = g,
+      alpha    = alpha,
+      y_target = target,
+      t_alpha  = median(boot_roots, na.rm = TRUE),
+      CI_lower = quantile(boot_roots, 0.025, na.rm = TRUE),
+      CI_upper = quantile(boot_roots, 0.975, na.rm = TRUE),
+      SD_t_alpha = sd(boot_roots, na.rm = TRUE)
+    )
+  })
+})
+
+results_alpha <- as_tibble(results_alpha)
+
+```
+
+## Save recovery times for splines model
+
+```{r}
+
+if (stringency == FALSE) {
+  file_results_alpha <- paste0(wd, '/data/outputs/', country_short, '/mov-analysis/by-', category, '/recovery-times-gam-r1.csv')
+} else {
+  file_results_alpha <- paste0(wd, '/data/outputs/', country_short, '/mov-analysis/by-', category, '/recovery-times-gams-r1.csv')
+}
+write.csv(results_alpha, file_results_alpha, row.names = FALSE)
+
+```
+
+## Plot recovery times splines model for various alpha
+
+```{r}
+ggplot(results_alpha, aes(x = alpha, y = t_alpha, color = O, group = O)) +
+  geom_line(size = 1) +
+  geom_ribbon(aes(ymin = CI_lower, ymax = CI_upper, fill = O),
+              alpha = 0.15, color = NA) +
+  labs(
+    title = "Time to reach threshold level (1 - α)",
+    x = expression(alpha),
+    y = expression(t(alpha)),
+    color = "Group (O)",
+    fill = "Group (O)"
+  ) +
+  theme_minimal(base_size = 14) +
+  theme(
+    legend.position = "right",
+    legend.title = element_text(size = 12),
+    legend.text = element_text(size = 11),
+    plot.title = element_text(face = "bold", size = 16)
+  )
+
+```

### 2025-12-01 notebooks/analysis/08b-mixed-models-R-cities-r1.qmd
revision

--- 2025-05-16 code/analysis/09-lme-with-R-cities.qmd
@@
 category <- "density"
-capital <- TRUE
+capital <- FALSE
 
@@
 
+###
+
+df$cat  <- factor(df$cat)
+eq8cat <- y ~ 
+    s(time, k = 8) + 
+    s(time, cat, bs = "fs", k = 7)
+m_gam <- gam(eq8cat, data = df, method = "REML")
+
+df$cat  <- factor(df$cat)
+eq8cats <- y ~ 
+    s(time, k = 8) + 
+    stringency_index + 
+    s(time, cat, bs = "fs", k = 7)
+m_gams <- gam(eq8cats, data   = df, method = "REML")
+
 ```
@@
   m6cat,
-  m7cat
+  m7cat, 
+  m_gam
   )
-tidy_results_cat <- map_df(models_cat, tidy, .id = "model")
-file_tidy_results_cat <- paste0(wd, '/data/outputs/', country_short, '/mov-analysis/by-', category, '/data_trend_netflows_', place, '_param_tidy_results_cat.csv')
-write.csv(tidy_results_cat, file_tidy_results_cat, row.names = FALSE)
+tidy_fixed <- map_df(models_cat, ~ tidy(.x, parametric = TRUE), .id = "model")
+tidy_smooth <- map_df(models_cat, ~ tidy(.x, smooth = TRUE), .id = "model")
+tidy_results_cat <- bind_rows(tidy_fixed, tidy_smooth) %>%
+  distinct()
+file_tidy_results_cat <- paste0(wd, '/data/outputs/', country_short, '/mov-analysis/by-', category, '/data_trend_netflows_', place, '_param_tidy_results_cat_new.csv')
+# write.csv(tidy_results_cat, file_tidy_results_cat, row.names = FALSE)
 
 glance_results_cat <- map_df(models_cat, glance, .id = "model")
-file_glance_results_cat <- paste0(wd, '/data/outputs/', country_short, '/mov-analysis/by-', category, '/data_trend_netflows_', place, '_param_glance_results_cat.csv')
-write.csv(glance_results_cat, file_glance_results_cat, row.names = FALSE)
+file_glance_results_cat <- paste0(wd, '/data/outputs/', country_short, '/mov-analysis/by-', category, '/data_trend_netflows_', place, '_param_glance_results_cat_new.csv')
+# write.csv(glance_results_cat, file_glance_results_cat, row.names = FALSE)
 
@@
   m6cats,
-  m7cats
+  m7cats,
+  m_gams
   )
-tidy_results_cats <- map_df(models_cats, tidy, .id = "model")
-file_tidy_results_cats <- paste0(wd, '/data/outputs/', country_short, '/mov-analysis/by-', category, '/data_trend_netflows_', place, '_param_tidy_results_cats.csv')
-write.csv(tidy_results_cats, file_tidy_results_cats, row.names = FALSE)
+tidy_fixed <- map_df(models_cats, ~ tidy(.x, parametric = TRUE), .id = "model")
+tidy_smooth <- map_df(models_cats, ~ tidy(.x, smooth = TRUE), .id = "model")
+tidy_results_cats <- bind_rows(tidy_fixed, tidy_smooth) %>%
+  distinct()
+
+file_tidy_results_cats <- paste0(wd, '/data/outputs/', country_short, '/mov-analysis/by-', category, '/data_trend_netflows_', place, '_param_tidy_results_cats_new.csv')
+# write.csv(tidy_results_cats, file_tidy_results_cats, row.names = FALSE)
 
 glance_results_cats <- map_df(models_cats, glance, .id = "model")
-file_glance_results_cats <- paste0(wd, '/data/outputs/', country_short, '/mov-analysis/by-', category, '/data_trend_netflows_', place, '_param_glance_results_cats.csv')
-write.csv(glance_results_cats, file_glance_results_cats, row.names = FALSE)
+file_glance_results_cats <- paste0(wd, '/data/outputs/', country_short, '/mov-analysis/by-', category, '/data_trend_netflows_', place, '_param_glance_results_cats_new.csv')
+# write.csv(glance_results_cats, file_glance_results_cats, row.names = FALSE)
 ```
@@
 
-file_m3cat_re <- paste0(wd, '/data/outputs/', country_short, '/mov-analysis/by-', category, '/data_trend_param_', place, '_m3cat_re.csv')
-write.csv(m3cat_re[[1]], file_m3cat_re, row.names = FALSE)
+file_m3cat_re <- paste0(wd, '/data/outputs/', country_short, '/mov-analysis/by-', category, '/data_trend_param_', place, '_m3cat_re_new.csv')
+# write.csv(m3cat_re[[1]], file_m3cat_re, row.names = FALSE)
 
-file_m3cats_re <- paste0(wd, '/data/outputs/', country_short, '/mov-analysis/by-', category, '/data_trend_param_', place, '_m3cats_re.csv')
-write.csv(m3cats_re[[1]], file_m3cats_re, row.names = FALSE)
+file_m3cats_re <- paste0(wd, '/data/outputs/', country_short, '/mov-analysis/by-', category, '/data_trend_param_', place, '_m3cats_re_new.csv')
+# write.csv(m3cats_re[[1]], file_m3cats_re, row.names = FALSE)
 
@@
 
-file_m4cat_re <- paste0(wd, '/data/outputs/', country_short, '/mov-analysis/by-', category, '/data_trend_param_', place, '_m4cat_re.csv')
-write.csv(m4cat_re[[1]], file_m4cat_re, row.names = FALSE)
+file_m4cat_re <- paste0(wd, '/data/outputs/', country_short, '/mov-analysis/by-', category, '/data_trend_param_', place, '_m4cat_re_new.csv')
+# write.csv(m4cat_re[[1]], file_m4cat_re, row.names = FALSE)
 
-file_m4cats_re <- paste0(wd, '/data/outputs/', country_short, '/mov-analysis/by-', category, '/data_trend_param_', place, '_m4cats_re.csv')
-write.csv(m4cats_re[[1]], file_m4cats_re, row.names = FALSE)
+file_m4cats_re <- paste0(wd, '/data/outputs/', country_short, '/mov-analysis/by-', category, '/data_trend_param_', place, '_m4cats_re_new.csv')
+# write.csv(m4cats_re[[1]], file_m4cats_re, row.names = FALSE)
 
@@
 
-file_m5cat_re <- paste0(wd, '/data/outputs/', country_short, '/mov-analysis/by-', category, '/data_trend_param_', place, '_m5cat_re.csv')
-write.csv(m5cat_re[[1]], file_m5cat_re, row.names = FALSE)
+file_m5cat_re <- paste0(wd, '/data/outputs/', country_short, '/mov-analysis/by-', category, '/data_trend_param_', place, '_m5cat_re_new.csv')
+# write.csv(m5cat_re[[1]], file_m5cat_re, row.names = FALSE)
 
-file_m5cats_re <- paste0(wd, '/data/outputs/', country_short, '/mov-analysis/by-', category, '/data_trend_param_', place, '_m5cats_re.csv')
-write.csv(m5cats_re[[1]], file_m5cats_re, row.names = FALSE)
+file_m5cats_re <- paste0(wd, '/data/outputs/', country_short, '/mov-analysis/by-', category, '/data_trend_param_', place, '_m5cats_re_new.csv')
+# write.csv(m5cats_re[[1]], file_m5cats_re, row.names = FALSE)
 
@@
 
-file_m6cat_re <- paste0(wd, '/data/outputs/', country_short, '/mov-analysis/by-', category, '/data_trend_param_', place, '_m6cat_re.csv')
-write.csv(m6cat_re[[1]], file_m6cat_re, row.names = FALSE)
+file_m6cat_re <- paste0(wd, '/data/outputs/', country_short, '/mov-analysis/by-', category, '/data_trend_param_', place, '_m6cat_re_new.csv')
+# write.csv(m6cat_re[[1]], file_m6cat_re, row.names = FALSE)
 
-file_m6cats_re <- paste0(wd, '/data/outputs/', country_short, '/mov-analysis/by-', category, '/data_trend_param_', place, '_m6cats_re.csv')
-write.csv(m6cats_re[[1]], file_m6cats_re, row.names = FALSE)
+file_m6cats_re <- paste0(wd, '/data/outputs/', country_short, '/mov-analysis/by-', category, '/data_trend_param_', place, '_m6cats_re_new.csv')
+# write.csv(m6cats_re[[1]], file_m6cats_re, row.names = FALSE)
 
@@
 
-file_m7cat_re <- paste0(wd, '/data/outputs/', country_short, '/mov-analysis/by-', category, '/data_trend_param_', place, '_m7cat_re.csv')
-write.csv(m7cat_re[[1]], file_m7cat_re, row.names = FALSE)
+file_m7cat_re <- paste0(wd, '/data/outputs/', country_short, '/mov-analysis/by-', category, '/data_trend_param_', place, '_m7cat_re_new.csv')
+# write.csv(m7cat_re[[1]], file_m7cat_re, row.names = FALSE)
 
-file_m7cats_re <- paste0(wd, '/data/outputs/', country_short, '/mov-analysis/by-', category, '/data_trend_param_', place, '_m7cats_re.csv')
-write.csv(m7cats_re[[1]], file_m7cats_re, row.names = FALSE)
+file_m7cats_re <- paste0(wd, '/data/outputs/', country_short, '/mov-analysis/by-', category, '/data_trend_param_', place, '_m7cats_re_new.csv')
+# write.csv(m7cats_re[[1]], file_m7cats_re, row.names = FALSE)
 
@@
 
-# Plot using ggplot
-ggplot() +
-  geom_point(data = df, aes(x = time, y = y), color = "blue", alpha = 0.6, size = 2) +
-  geom_point(data = pred_df_1, aes(x = time, y = fitted), color = "red", alpha = 0.6, size = 1.5) +
-  geom_point(data = pred_df_2, aes(x = time, y = fitted), color = "red3", alpha = 0.6, size = 1.5) +
-  geom_point(data = pred_df_3, aes(x = time, y = fitted), color = "darkred", alpha = 0.6, size = 1.5) +
-  geom_point(data = pred_df_4, aes(x = time, y = fitted), color = "purple4", alpha = 0.6, size = 1.5) +
-  geom_point(data = pred_df_5, aes(x = time, y = fitted), color = "midnightblue", alpha = 0.6, size = 1.5) +
-  labs(title = "Observed vs Predicted (Quadratic Fit)",
-       x = "Time", y = "Y") +
-  xlim(0, 300) +
-  ylim(min(c(df$y, pred_df$fitted), na.rm = TRUE),
-       max(c(df$y, pred_df$fitted, 20), na.rm = TRUE)) +
-  theme_minimal()
+```
+
+## Compute recovery times for spline model. The 95% confidence intervals are computed via bootstrap.
+
+```{r}
+stringency <- FALSE
+```
+
+```{r}
+# Create new time sequence covering the range of your original data
+new_time <- expand.grid(
+  time = seq(min(df$time), max(df$time)*100, length.out = max(df$time)*100-min(df$time)+1),
+  cat = levels(df$cat)   # one dataset per group
+)
+
+new_time <- new_time %>%
+  left_join(df[, c("time", "cat", "stringency_index")], by = c("time", "cat"))
+
+new_time$stringency_index[is.na(new_time$stringency_index)] <- 0
+
+if (stringency == FALSE) {
+  new_time <- expand.grid(
+  time = seq(min(df$time), max(df$time)*100, length.out = max(df$time)*100-min(df$time)+1),
+  cat = levels(df$cat)   # one dataset per group
+)
+  pred_base <- predict(m_gam, newdata = new_time, se.fit = TRUE)
+} else {
+  pred_base <- predict(m_gams, newdata = new_time, se.fit = TRUE)
+}
+
+new_time$fit <- pred_base$fit
+new_time$se  <- pred_base$se.fit
+
+# Find FIRST root where preds(t) == target
+get_root_threshold <- function(time, preds, target) {
+  diff_sign <- diff(sign(preds - target))
+  crossings <- which(diff_sign != 0)
+  if (length(crossings) == 0) return(NA)
+
+  i <- crossings[1]  # first crossing
+  out <- tryCatch(
+    uniroot(function(t) {
+      approx(time, preds - target, xout = t)$y
+    },
+    lower = time[i],
+    upper = time[i + 1])$root,
+    error = function(e) NA
+  )
+  return(out)
+}
+
+alphas <- seq(0, 1, by = 0.05)   # adjust resolution as needed
+
+# Bootstrap confidence intervals
+set.seed(123)
+n_boot <- 200   # increase to 500–1000 for publication
+
+results_alpha <- map_df(levels(df$cat), function(g) {
+
+  group_data <- new_time %>% filter(cat == g)
+
+  # group-specific y(t=0): use earliest time in grid for that group
+  idx0 <- which.min(group_data$time)
+  y0_group <- group_data$fit[idx0]
+
+  # For each alpha / target, bootstrap
+  map_df(seq_along(alphas), function(j) {
+    alpha   <- alphas[j]
+    target <- (1 - alpha) * y0_group
+
+    boot_roots <- replicate(n_boot, {
+      # simulate one curve from uncertainty
+      y_sim <- rnorm(nrow(group_data), mean = group_data$fit, sd = group_data$se)
+      get_root_threshold(group_data$time, y_sim, target)
+    })
+
+    data.frame(
+      cat        = g,
+      alpha    = alpha,
+      y_target = target,
+      t_alpha  = median(boot_roots, na.rm = TRUE),
+      CI_lower = quantile(boot_roots, 0.025, na.rm = TRUE),
+      CI_upper = quantile(boot_roots, 0.975, na.rm = TRUE),
+      SD_t_alpha = sd(boot_roots, na.rm = TRUE)
+    )
+  })
+})
+
+results_alpha <- as_tibble(results_alpha)
 
 ```
+
+## Save recovery times for splines model
+
+```{r}
+
+if (stringency == FALSE) {
+  if (capital == TRUE) {
+    file_results_alpha <- paste0(wd, '/data/outputs/', country_short, '/mov-analysis/by-', category, '/recovery-times-capital-gam-r1.csv')
+  } else {
+    file_results_alpha <- paste0(wd, '/data/outputs/', country_short, '/mov-analysis/by-', category, '/recovery-times-fuas-gam-r1.csv')
+  }
+} else {
+  if (capital == TRUE) {
+    file_results_alpha <- paste0(wd, '/data/outputs/', country_short, '/mov-analysis/by-', category, '/recovery-times-capital-gams-r1.csv')
+  } else {
+    file_results_alpha <- paste0(wd, '/data/outputs/', country_short, '/mov-analysis/by-', category, '/recovery-times-fuas-gams-r1.csv')
+  }
+}
+write.csv(results_alpha, file_results_alpha, row.names = FALSE)
+
+```
+
+## Plot recovery times splines model for various alpha
+
+```{r}
+ggplot(results_alpha, aes(x = alpha, y = t_alpha, color = cat, group = cat)) +
+  geom_line(size = 1) +
+  geom_ribbon(aes(ymin = CI_lower, ymax = CI_upper, fill = cat),
+              alpha = 0.15, color = NA) +
+  labs(
+    title = "Time to reach threshold level (1 - α)",
+    x = expression(alpha),
+    y = expression(t(alpha)),
+    color = "Group (cat)",
+    fill = "Group (cat)"
+  ) +
+  theme_minimal(base_size = 14) +
+  theme(
+    legend.position = "right",
+    legend.title = element_text(size = 12),
+    legend.text = element_text(size = 11),
+    plot.title = element_text(face = "bold", size = 16)
+  )
+
+```

### 2025-12-04 notebooks/analysis/05b-mixed-models-R-r1.qmd
updates-gam

@@
 library(modelsummary)
-library(brms)
+library(mgcv)
+
 ```
